### 2025-01-06 paper.qmd
Initial commit

---
title: "Extracting sunbeams from cucumbers: Reproducibility as a necessity for the future of archaeological science"
format: html
bibliography: references.bib
---

```{=html}
<!--

authoritative, agenda-setting or hypothesis-setting papers, rather than descriptive compilations of the literature.

have the vision and authority to guide the next generation of archaeological science research

cross-ref Vaiglova 2025 this issue

-->
```

## Introduction

In their paper celebrating the 40th anniversary of this journal Torrence et al [-@torrenceFortyYearsStill2015] noted that reproducibility was an issue important to the reputation and sustainability of the discipline, and necessary for archaeological science to behave like a science. As part of the celebration of the 50th anniversary, and of Torrence's leadership of the journal, my contribution revisits these topics of archaeology's status as a science, this journal's place in the landscape of archaeological science, and how the journal has responded to a growing recognition of the importance of reproducibility. I first present bibliometric evidence of the position of archaeology as a whole, and this journal in particular, in the hierarchy of the sciences. Next I report on the journal's progress in supporting reproducible research, and my work doing a new kind of peer review for JAS, one that evaluates the computational reproducibilty of the research submitted for publication.

The question of archaeology's status as a science usually comes up in the context of what the discipline should or should not be. One of the first landmarks in tackling this question is the debate published by *Antiquity* between classical archaeologist Jacquetta Hawkes and palaeoanthropologist Glynn Isaac. @hawkesProperStudyMankind1968, advocating a humanistic archaeology, was concerned that scientific approaches to archaeology were causing researchers to be "swamped by a vast accumulation of insignificant, disparate facts, like a terrible tide of mud, quite beyond the capacity of any man to contain and mould into historical form". More optimistic about the integration of science and archaeology, @isaacWhitherArchaeology1971 counters that "New levels of precision in presenting data and in interpreting them can surely lead to briefer and more interesting technical reports as well as providing the basis for more lively literary portrayals of what happened in prehistory. Expanding on Isaac's perspective, @binford1962archaeology argued that archaeology should operate as a science after the model proposed by philosopher Carl Hempel, which prescribed hypothesis-driven approaches, leading to generalizable laws of human behavior. Counter-arguments came from numerous directions, notably @hodder1985postprocessual who rejected the quest for generalisations and instead argued that archaeology should be subjective and reflective, focussed on symbolic and relational meanings of material culture and the historical particularity of past human cultures. These debates, and the many more similar ones summarised by @martinon-torresArchaeologicalTheoriesArchaeological2013, have become a genre in archaeological writing that can be characterised as mostly based on personal observations, microscopic dissections of a handful of cherry-picked case studies of good or bad practice, and discussion of various philosophers and sociologists.

What has been missing from these debates is a macroscopic observation of what the majority of archaeologists are actually doing, and an objective comparison to a broad spectrum of relatively harder and softer disciplines.

Computational reproducibilty has become central to archaeology's status a science, especially as the definition and practice of science shifts over time. Computers have become an essential field and laboratory instrument for almost all of us, so we should document how we change our data as it flows through silicon just as carefully as we document the operating parameters of a mass spectrometer or other laboratory instrument. In making the case for reproducibilty as a core value in archaeological science

In recent years the Journal of Archaeological Science has been one of a number of archaeology journals that introduced an unusual new kind of peer review, one that evaluates the computational reproducibilty of the research submitted for publication. This journal was among the first of these and is currently by far the most active in conducting these reviews. In this article I reflect on how this new kind of review came about, why it is important for the future of archaeological science, and offer suggestions for prospective authors based on my experience as a reproducibilty reviewer for this journal on how we can ensure archaeological science remains worthy of the name.

My motivation is the view that It is customary in significant round-number anniversary

### 2025-01-06 code/001-authors.R

@@ -1,0 +1,75 @@
+library(tidyverse)
+
+items_df <- read_rds("data/wos-data-df.rds")
+
+# number of authors + greater scope and need for collaboration [9,44]
+
+# all on one
+
+items_df %>% 
+  filter(!is.na(year)) %>% 
+  ggplot(aes(year,
+             log(authors_n))) +
+  geom_boxplot(colour = "red", 
+               size = 1)  +
+  scale_y_continuous(limits = c(0, 5)) +
+  scale_x_continuous(labels = NULL,
+                     name = NULL) +
+  theme_minimal() +
+  theme(panel.grid  = element_blank()) +
+  coord_equal(ratio = 6)
+
+ggsave("figures/box_num_authors.svg")
+
+# over time
+items_df %>% 
+  filter(!is.na(year)) %>% 
+  ggplot(aes(year,
+             log(authors_n),
+             group = year)) +
+  geom_jitter(alpha = 0.2) +
+  theme_minimal() +
+  ylab("Number of authors (ln)")
+
+ggsave("figures/box_num_authors_over_time.png")
+
+# by journal, wow yes!
+library(ggridges)
+items_df %>% 
+  mutate(journal = str_wrap(journal, 20)) %>% 
+  ggplot(aes(y = reorder(journal,
+                     authors_n), 
+             x = log(authors_n),
+             fill = ..x..,
+             height = ..density..)) +
+  geom_density_ridges_gradient(stat = "density",
+                               colour = "white") +
+  scale_fill_viridis_c() +
+  guides(fill = 'none') +
+  theme_minimal() +
+  theme(axis.text.y = element_text(size = 6)) +
+  ylab("") + 
+  xlab("Number of authors (ln)")
+
+
+ggsave("figures/box_num_authors_by_journal.png")
+
+# histogram
+
+tally_authors_n <- 
+  items_df %>% 
+  group_by(authors_n) %>% 
+  tally() %>% 
+  mutate(perc = n / sum(n) * 100) %>% 
+  mutate(cum_per = cumsum(perc)) %>% 
+  arrange(cum_per) # 90% of papers have 4 or fewer authors
+
+items_df %>% 
+  ggplot(aes(authors_n)) +
+  geom_histogram()  +
+  theme_minimal() +
+  scale_x_log10() +
+  xlab("Number of authors") + 
+  ylab("Count") 
+
+ggsave("figures/box_num_histogram.png")

### 2025-01-06 code/002-article-length.R

@@ -1,0 +1,53 @@
+#  ------------------------------------------------------------------
+# length of article - less need to introduce, justify and explain study [47,52]
+
+# all on one
+library(tidyverse)
+
+items_df %>% 
+  ggplot(aes(year,
+             log(pages_n))) +
+  geom_boxplot(colour = "red", 
+               size = 1)  +
+  scale_y_continuous(limits = c(0, 5)) +
+  scale_x_continuous(labels = NULL,
+                     name = NULL) +
+  theme_minimal() +
+  theme(panel.grid  = element_blank()) +
+  coord_equal(ratio = 6)
+
+ggsave("figures/box_num_pages.svg")
+
+
+# over time
+items_df %>% 
+  ggplot(aes(year,
+             log(pages_n),
+             group = year)) +
+  geom_boxplot() +
+  geom_jitter(alpha = 0.01) +
+  theme_minimal() +
+  ylab("Number of pages (ln)")
+
+ggsave("figures/box_num_pages_over_time.png")
+
+# by journal, wow yes!
+library(ggforce)
+items_df %>% 
+  mutate(journal = str_wrap(journal, 20)) %>% 
+  ggplot(aes(y = reorder(journal,
+                     pages_n), 
+             x = log(pages_n),
+             fill = ..x..,
+             height = ..density..)) +
+  geom_density_ridges_gradient(stat = "density",
+                               colour = "white") +
+  scale_fill_viridis_c() +
+  guides(fill = 'none') +
+  theme_minimal() +
+  theme(axis.text.y = element_text(size = 6)) +
+  xlab("Number of pages (ln)") +
+  ylab("")
+
+ggsave("figures/box_num_pages_by_journal.png")
+

### 2025-01-06 code/005-first-person.R

@@ -1,0 +1,61 @@
+
+# use of first person (singular vs. plural) in abstracts, - universal validity of claims; less scope for argumentation; fewer appeals to opinion and authority [59], the proportion of first person pronouns, both singular and plural (i.e. ‘‘I’’, ‘‘me’’, ‘‘mine’’, ‘‘we’’, ‘‘our’’ etc.) among all words in the abstract.
+
+i <- c(" i'd ", " i'll ", " i'm ", " i've ", " i ")
+me <- c(" me ", " my ", " mine ")
+us <- c(" us ", " our ", " ours ")
+first_persons <- c(i, me, us)
+
+library(stringi)
+
+# function to compute proportion of first person pronouns
+first_person_prop <- function(x){
+  x <- tolower(x)
+  firsts <- sum(str_count(x, first_persons))
+  abstract_n_words <- stri_count_words(x)
+  firsts / abstract_n_words
+}
+
+
+# compute proportions, single author papers only
+items_df_abstract <- 
+  items_df %>% 
+  mutate(abstract_first_person = first_person_prop(abstract)) 
+
+# plots
+# all on one
+items_df_abstract %>% 
+  ggplot(aes(year,
+             abstract_first_person)) +
+  geom_boxplot()  +
+  scale_x_continuous(labels = NULL,
+                     name = NULL) +
+  theme_minimal()
+
+# over time
+items_df_abstract %>% 
+  ggplot(aes(year,
+             abstract_first_person,
+             group = year)) +
+  geom_boxplot() +
+  geom_jitter(alpha = 0.01) +
+ # scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
+  theme_minimal()
+
+# by journal
+items_df_abstract %>% 
+  mutate(journal = str_wrap(journal, 20)) %>% 
+  ggplot(aes(y = reorder(journal,
+                     abstract_first_person), 
+             x = abstract_first_person,
+             fill = ..x..,
+             height = ..density..)) +
+  geom_density_ridges_gradient(stat = "density",
+                               colour = "white") +
+  scale_fill_viridis_c() +
+  guides(fill = 'none') +
+  theme_minimal() +
+  theme(axis.text.y = element_text(size = 6))  +
+  ylab("") +
+  xlab("Proportion of first person pronouns (singular vs. plural) in abstracts")
+

### 2025-01-06 code/006-bibmetrics-summary.R

@@ -1,0 +1,79 @@
+
+#--------------------------------
+# Six variables over time
+
+# https://github.com/benmarwick/March-2019-Cambridge-Big-Data-Archaeology/blob/master/code/002-fanelli-hierarchy-of-sciences.R
+
+over_time <- 
+  items_df %>% 
+  left_join(items_df_title) %>% 
+  left_join(shannon_per_item) %>% 
+  mutate(log_authors_n = log(authors_n),
+         log_pages_n = log(pages_n),
+         sqrt_refs_n = sqrt(refs_n),
+         journal_wrp = str_wrap(journal, 30)) %>% 
+  select(year,
+         log_authors_n,
+         log_pages_n,
+         prices_index, 
+         shannon,
+         sqrt_refs_n, 
+         relative_title_length)
+
+over_time_long <- 
+  over_time %>% 
+  ungroup() %>% 
+  select(-journal) %>% 
+  gather(variable, 
+         value,
+         -year) %>% 
+  filter(value != -Inf,
+         value !=  Inf) %>% 
+  mutate(variable = case_when(
+    variable == "log_authors_n" ~ "N. of authors (ln)",
+    variable == "log_pages_n"   ~ "N. of pages (ln)",
+    variable == "prices_index"  ~ "Price's index",
+    variable == "shannon"  ~ "Shannon div. of sources",
+    variable == "sqrt_refs_n"  ~ "N. of refs (sqrt)",
+    variable == "relative_title_length"  ~ "Relative title length (ln)"
+  ))
+
+# compute beta estimates so we can colour lines to indicate more or
+# less scientific
+library(broom)
+over_time_long_models <- 
+  over_time_long %>% 
+  group_nest(variable) %>% 
+  mutate(model = map(data, ~tidy(lm(value ~ year, data = .)))) %>% 
+  unnest(model) %>% 
+  filter(term == 'year') %>% 
+  mutate(becoming_more_scientific = case_when(
+    variable == "N. of authors (ln)" & estimate > 0 ~ "TRUE",
+    variable == "N. of pages (ln)"   & estimate < 0 ~ "TRUE",
+    variable == "N. of refs (sqrt)"  & estimate < 0 ~ "TRUE",
+    variable == "Price's index"     & estimate > 0 ~ "TRUE",
+    variable == " Relative title length (ln)"     & estimate > 0 ~ "TRUE",
+    variable == "Shannon div. of sources"     & estimate < 0 ~ "TRUE",
[…]

### 2025-01-06 code/007-disruptive-articles.R

@@ -1,0 +1,257 @@
+
+# https://github.com/benmarwick/March-2019-Cambridge-Big-Data-Archaeology/blob/master/code/003-disruptiveness-index.R
+# https://pubsonline.informs.org/doi/10.1287/mnsc.2015.2366
+# https://www.nature.com/articles/s41586-019-0941-9
+
+library(tidyverse)
+library(here)
+
+# get articles with DOIs and reference lists
+items_wth_dois_and_refs <- 
+  items_df %>% 
+  filter(str_detect(doi, "^10.")) %>% 
+  mutate(doi = str_remove_all(doi, "\\[|\\]")) %>% 
+  mutate(doi_article = doi) %>% 
+  select(-doi)
+
+# get items in the reference lists with DOIs
+items_df_one_row_per_ref_with_doi <- 
+  items_wth_dois_and_refs %>% 
+  separate_rows(refs, sep = "\n  ") %>% 
+  ungroup %>% 
+  mutate(refs = tolower(str_trim(refs))) %>% 
+  mutate(refs = str_extract(refs, "doi.*")) %>% 
+  mutate(refs = str_remove_all(refs, "\\[|\\]")) %>% 
+  mutate(doi = str_remove_all(refs, "doi"))  %>% 
+  mutate(doi_cited = str_trim(doi)) %>%  
+  filter(!is.na(doi_cited)) %>% 
+  select(-doi)
+
+# ------- compute Disruption  ----
+# for a focal article:
+# - find the articles that cite the focal article
+# - find what articles that citing article cites 
+# - find the ratio of citing articles that cite focal articles refs vs those that don't
+
+disruption_output <- list(length = length(items_wth_dois_and_refs$doi_article))
+
+for(i in seq_along(items_wth_dois_and_refs$doi_article)){
+  
+  this_article_doi <- items_wth_dois_and_refs$doi_article[i] # 1000
+  
+  # what articles are cited by the focal paper
+  cited_by_focal_paper <- 
+    items_df_one_row_per_ref_with_doi %>% 
+    filter(doi_article %in% this_article_doi) 
+  
+  # what articles cite the focal paper, that are in our list
+  citing_the_focal_paper <- 
+    items_df_one_row_per_ref_with_doi %>% 
+    filter(doi_cited %in% this_article_doi)
+  
+  # what articles are cited by the citing papers
+  cited_by_citing_papers <- 
+    items_df_one_row_per_ref_with_doi %>% 
+    filter(doi_article %in% citing_the_focal_paper$doi_article)
+  
[…]

### 2025-01-08 paper.qmd
intro

@@
 
-What has been missing from these debates is a macroscopic observation of what the majority of archaeologists are actually doing, and an objective comparison to a broad spectrum of relatively harder and softer disciplines.
+What has been missing from these debates is a macroscopic observation of what the majority of archaeologists are actually doing, and an empirical comparison to a broad spectrum of relatively harder and softer disciplines.
+At the 'hard' end of the spectrum, scholars more typically share an established set of theories, facts, and methods, facilitating fairly rapid agreement on the validity and significance of new results.
+At the 'soft' end of the spectrum, the theories, facts, and methods on which there is widespread consensus is smaller, and agreement is slower and less frequently reached about the significance of new finding and the continuing relevance of previous work.
+Hardness and softness is a controversial distinction, in part because it is often used to imply a rank order of disciplines that encodes legitimacy, productivity, perceived value to society, and worthiness of funding [@DifferentAgenda2012; @coleHierarchySciences1983].
+Independent of these value judgments, empirical analysis of scholarly articles does indicate a spectrum of variation in practice linked to differing degrees of consensus in a discipline [@clevelandGraphsScientificPublications1984; @smithScientificGraphsHierarchy2000].
+Furthermore, the hard-soft spectrum is evident in surveys of how researchers view their own work relative to others [@biglanCharacteristicsSubjectMatter1973].
+
+n an ideal science, scholars share a common background of established theories, facts and methods.
+This allows them to agree (usually after debate and further evidence) on the validity and significance of a new research finding, making it the basis for further theorizing and research.
+Harder sciences are hypothesised to come closer to this ideal.
+Moving towards “softer” fields, this consensus becomes less likely to be reached, the common background shrinks and fractures, and so data become less able to “speak for themselves” [\[](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0066938#pone.0066938-Fanelli1)
 

### 2025-01-09 paper.qmd
intro and methods

@@
 Counter-arguments came from numerous directions, notably @hodder1985postprocessual who rejected the quest for generalisations and instead argued that archaeology should be subjective and reflective, focussed on symbolic and relational meanings of material culture and the historical particularity of past human cultures.
-These debates, and the many more similar ones summarised by @martinon-torresArchaeologicalTheoriesArchaeological2013, have become a genre in archaeological writing that can be characterised as mostly based on personal observations, microscopic dissections of a handful of cherry-picked case studies of good or bad practice, and discussion of various philosophers and sociologists.
+These debates, and the many more similar ones summarised by @martinon-torresArchaeologicalTheoriesArchaeological2013, have become a genre in archaeological writing that can be characterized as mostly based on personal observations, microscopic dissections of a handful of cherry-picked case studies of good or bad practice, and discussion of various philosophers and sociologists.
 
 What has been missing from these debates is a macroscopic observation of what the majority of archaeologists are actually doing, and an empirical comparison to a broad spectrum of relatively harder and softer disciplines.
-At the 'hard' end of the spectrum, scholars more typically share an established set of theories, facts, and methods, facilitating fairly rapid agreement on the validity and significance of new results.
-At the 'soft' end of the spectrum, the theories, facts, and methods on which there is widespread consensus is smaller, and agreement is slower and less frequently reached about the significance of new finding and the continuing relevance of previous work.
+At the 'hard' end of the spectrum (e.g.
+physics and chemistry), scholars more typically share an established set of theories, facts, and methods, facilitating fairly rapid agreement on the validity and significance of new results.
+At the 'soft' end of the spectrum (e.g.
+economics and psychology), the theories, facts, and methods on which there is widespread consensus is smaller, and agreement is slower and less frequently reached about the significance of new findings and the continuing relevance of previous work.
 Hardness and softness is a controversial distinction, in part because it is often used to imply a rank order of disciplines that encodes legitimacy, productivity, perceived value to society, and worthiness of funding [@DifferentAgenda2012; @coleHierarchySciences1983].
-Independent of these value judgments, empirical analysis of scholarly articles does indicate a spectrum of variation in practice linked to differing degrees of consensus in a discipline [@clevelandGraphsScientificPublications1984; @smithScientificGraphsHierarchy2000].
-Furthermore, the hard-soft spectrum is evident in surveys of how researchers view their own work relative to others [@biglanCharacteristicsSubjectMatter1973].
+However, independent of these value judgments, empirical analysis of scholarly articles does indicate a spectrum of variation in practice linked to differing degrees of consensus in a discipline, for example in approaches to data visualisation [@clevelandGraphsScientificPublications1984; @smithScientificGraphsHierarchy2000].
+Similarly, quantitative analysis of the frequency of positive results (ie.
+full or partial support for a research hypothesis) in publications is significantly correlated with hardness, consistent with a model where researchers in harder fields more readily accept any result their research produces, while those in softer fields have more freedom to choose which theories and hypotheses to test and how to interpret results [@PositiveResultsIncrease].
+The hard-soft spectrum is also evident in surveys of how researchers view their own work relative to those in other fields [@biglanCharacteristicsSubjectMatter1973].
+
+Here I take a bibliometric approach to objectively quantify the relative hardness or softness of archaeology to evaluate its status as a science, and the place of this journal in the discipline.
+For a large sample of peer-reviewed journal articles, I quantify the number of authors, length of article, relative title length, number of cited references, age of references, and diversity of references.
+These parameters are useful because of how they signify consensus in a research community.
+A larger number of authors on a paper reflects collaboration of people with specialized roles, each of whom has the ability to study a part of the problem with high accuracy and detail [@zuckerman1972age].
+Harder disciplines tend to have higher average numbers of authors on papers.
+In low-consensus, or softer, fields, papers must be longer to present justification, nuance and contextualization of results.
+I used page numbers to measure article length.
+Similarly, the number of cited references will typically be higher in low-consensus disciplines where relatively fewer claims are taken for granted and thus need citations for support, relative to harder fields [@skilton2006comparative].
+The age of works cited has long been used as a measure of a field's hardness [@moed1998new; @bornerAtlasScience2010], based on the assumption that harder fields assimilate new results more rapidly that softer fields [@price1970citation].
+I calculated the Price index for each article which is the proportion of citated worked published in the five years preceding the paper.
 

### 2025-01-12 paper.qmd
methods
methods and materials

@@
 
-```{=html}
 <!--
@@
 -->
-```
 
@@
 At the 'hard' end of the spectrum (e.g.
-physics and chemistry), scholars more typically share an established set of theories, facts, and methods, facilitating fairly rapid agreement on the validity and significance of new results.
+physics and chemistry), scholars more typically share a large set of established set of theories, facts, and methods, facilitating fairly rapid agreement on the validity and significance of new results.
 At the 'soft' end of the spectrum (e.g.
-economics and psychology), the theories, facts, and methods on which there is widespread consensus is smaller, and agreement is slower and less frequently reached about the significance of new findings and the continuing relevance of previous work.
+economics and psychology), the set of theories, facts, and methods on which there is widespread consensus is smaller, and agreement is slower and less frequently reached about the significance of new findings and the continuing relevance of previous work.
 Hardness and softness is a controversial distinction, in part because it is often used to imply a rank order of disciplines that encodes legitimacy, productivity, perceived value to society, and worthiness of funding [@DifferentAgenda2012; @coleHierarchySciences1983].
@@
 
-Here I take a bibliometric approach to objectively quantify the relative hardness or softness of archaeology to evaluate its status as a science, and the place of this journal in the discipline.
-For a large sample of peer-reviewed journal articles, I quantify the number of authors, length of article, relative title length, number of cited references, age of references, and diversity of references.
+## Methods and materials
+
+To objectively quantify the relative hardness or softness of archaeology, as an evaluation of its status as a science, and the place of this journal in context of other archaeology journals, I take a bibliometric approach.
+This approach is based on @fanelliBibliometricEvidenceHierarchy2013, who examined the hardness and softness of 12 disciplines using parameters of scholarly publications.
+Following @fanelliBibliometricEvidenceHierarchy2013, I quantify the number of authors, length of article, relative title length, number of cited references, age of references, and diversity of references for a large sample of peer-reviewed journal articles.
 These parameters are useful because of how they signify consensus in a research community.
-A larger number of authors on a paper reflects collaboration of people with specialized roles, each of whom has the ability to study a part of the problem with high accuracy and detail [@zuckerman1972age].
-Harder disciplines tend to have higher average numbers of authors on papers.
+A larger number of authors on a paper reflects collaboration of people working together on a common goal.
+Collaborators have specialized roles, each of whom has the ability to study a part of the problem with high accuracy and detail, with harder fields having larger groups of collaborators[@zuckerman1972age].
+Reflecting this collaboration group size, harder disciplines tend to have higher average numbers of authors on papers.
+Article length has an inverse correlation with field hardness.
 In low-consensus, or softer, fields, papers must be longer to present justification, nuance and contextualization of results.
 I used page numbers to measure article length.
-Similarly, the number of cited references will typically be higher in low-consensus disciplines where relatively fewer claims are taken for granted and thus need citations for support, relative to harder fields [@skilton2006comparative].
+Similarly, the number of cited references will typically be higher in low-consensus disciplines where relatively fewer claims are taken for granted and thus many claims need citations for support, relative to harder fields [@skilton2006comparative].
 The age of works cited has long been used as a measure of a field's hardness [@moed1998new; @bornerAtlasScience2010], based on the assumption that harder fields assimilate new results more rapidly that softer fields [@price1970citation].
-I calculated the Price index for each article which is the proportion of citated worked published in the five years preceding the paper.
+I calculated the Price index for each article, which is the proportion of all cited works that were published in the five years preceding the paper.
 
-n an ideal science, scholars share a common background of established theories, facts and methods.
+```{r}
+library(tidyverse)
+
+items_df <- read_rds("data/wos-data-df.rds")
+
+n_articles <- nrow(items_df)
+year_max <- max(items_df$year)
+year_min <- min(items_df$year)
+```
+
+I sampled `r n_articles` papers published during `r year_min`-`r year_max`.
+This sample was collected from Clarivate's Web of Science database by first selecting the Web of Science category 'Archaeology' and the Document type 'article' (n = 28,871).
+To focus on journals of broad relevance to most archaeologists, and that are representative of substantial communities of practice, I then filtered the results to keep only articles published in the top-ranking 25 journals according to their h-indices as reported by Clarivate's Journal Citation Indicator.
+Finally I excluded journals with less than 100 articles in the database.
+
+The entire R code [@rcoreteamLanguageEnvironmentStatistical2024] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at https://doi.org/xxx/xxx/xxx to enable re-use of materials and improve reproducibility and transparency [@marwickComputationalReproducibilityArchaeological2017].
+All the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this compendium [@doi:10.1080/00031305.2017.1375986].
+The R code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.
+
+## Results
+
+### How does archaeology compare to other fields?
+
+```{r}
+library(ggrepel)
+
+boxlplot_n_authors <- 
+items_df %>% 
+  filter(!is.na(year)) %>% 
+  ggplot(aes(1,
+             log(authors_n))) +
+  geom_boxplot(colour = "red", 
+               size = 1)  +
+  # median for physics, from Fanelli & Glänzel Fig 2
+  geom_hline(yintercept = 1.4,
+             colour = "grey70") +
+  geom_text_repel(data = tibble(x = 1, 
+                                y = 1.4,
+                                label = "p"),
+                  aes(x, y, label = label),
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0) +
+  # median for soc sci, from Fanelli & Glänzel Fig 2
+  geom_hline(yintercept = 0.8,
+             colour = "grey70") +
+  geom_text_repel(data = tibble(x = 1, 
+                                y = 0.8,
+                                label = "s"),
+                  aes(x, y, label = label),
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0) +
+    # median for humanities, from Fanelli & Glänzel Fig 2
+  geom_hline(yintercept = 0.0,
+             colour = "grey70") +
+  geom_text_repel(data = tibble(x = 1, 
+                                y = 0.0,
+                                label = "h"),
+                  aes(x, y, label = label),
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0) +
+  scale_y_continuous(limits = c(0, 5)) +
+  scale_x_continuous(labels = NULL) +
+  theme_minimal() + 
+  theme(panel.grid  = element_blank()) +
+  ylab("N. of authors (ln)") +
+  xlab("Collaborator group size")
+
+boxlplot_n_pages <- 
+items_df %>% 
+  ggplot(aes(1,
+             log(pages_n))) +
+  geom_boxplot(colour = "red", 
+               size = 1)  +
+   # median for physics, from Fanelli & Glänzel Fig 2
+  geom_hline(yintercept = 1.8,
+             colour = "grey70") +
+  geom_text_repel(data = tibble(x = 1, 
+                                y = 1.8,
+                                label = "p"),
+                  aes(x, y, label = label),
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0) +
+  # median for soc sci, from Fanelli & Glänzel Fig 2
+  geom_hline(yintercept = 2.5,
+             colour = "grey70") +
+  geom_text_repel(data = tibble(x = 1, 
+                                y = 2.5,
+                                label = "s"),
+                  aes(x, y, label = label),
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0) +
+    # median for humanities, from Fanelli & Glänzel Fig 2
+  geom_hline(yintercept = 2.8,
+             colour = "grey70") +
+  geom_text_repel(data = tibble(x = 1, 
+                                y = 2.8,
+                                label = "h"),
+                  aes(x, y, label = label),
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0) +
+  scale_y_continuous(limits = c(0, 5)) +
+  scale_x_continuous(labels = NULL) +
+  theme_minimal() + 
+  theme(panel.grid  = element_blank()) +
+  ylab("N. of pages (ln)") +
+  xlab("Article length")
+
+library(cowplot)
+
+plot_grid(boxlplot_n_authors, 
+          boxlplot_n_pages)
+```
+
+### How has the hardness of archaeology varied over time?
+
+### How do archaeology journals vary in hardness?
+
+```{=html}
+<!-- n an ideal science, scholars share a common background of established theories, facts and methods.
 This allows them to agree (usually after debate and further evidence) on the validity and significance of a new research finding, making it the basis for further theorizing and research.
@@
 
-My motivation is the view that It is customary in significant round-number anniversary
+My motivation is the view that It is customary in significant round-number anniversary -->
+```

### 2025-01-12 code/001-authors.R

@@ -8,16 +8,15 @@ items_df <- read_rds("data/wos-data-df.rds")
 
 items_df %>% 
   filter(!is.na(year)) %>% 
-  ggplot(aes(year,
+  ggplot(aes(1,
              log(authors_n))) +
   geom_boxplot(colour = "red", 
                size = 1)  +
   scale_y_continuous(limits = c(0, 5)) +
-  scale_x_continuous(labels = NULL,
-                     name = NULL) +
+  scale_x_continuous(labels = NULL) +
   theme_minimal() +
-  theme(panel.grid  = element_blank()) +
-  coord_equal(ratio = 6)
+  ylab("N. of authors (ln)") +
+  xlab("Collaborator group size")
 
 ggsave("figures/box_num_authors.svg")
 

### 2025-01-12 code/002-article-length.R

@@ -27,7 +27,7 @@ items_df %>%
   geom_boxplot() +
   geom_jitter(alpha = 0.01) +
   theme_minimal() +
-  ylab("Number of pages (ln)")
+  ylab("Number of pages (ln)") 
 
 ggsave("figures/box_num_pages_over_time.png")
 

### 2025-01-12 paper.qmd
writing results

@@
 bibliography: references.bib
+execute: 
+  echo: false
+  warning: false
 ---
 
+```{=html}
 <!--
@@
 -->
+```
 
@@
 To objectively quantify the relative hardness or softness of archaeology, as an evaluation of its status as a science, and the place of this journal in context of other archaeology journals, I take a bibliometric approach.
-This approach is based on @fanelliBibliometricEvidenceHierarchy2013, who examined the hardness and softness of 12 disciplines using parameters of scholarly publications.
-Following @fanelliBibliometricEvidenceHierarchy2013, I quantify the number of authors, length of article, relative title length, number of cited references, age of references, and diversity of references for a large sample of peer-reviewed journal articles.
+This approach is based on @fanelliBibliometricEvidenceHierarchy2013, who examined the hardness and softness of 12 disciplines using scholarly publication parameters.
+@fanelliBibliometricEvidenceHierarchy2013 found a spectrum of statistically significant variation in bibliometric variables from the physical to the social sciences, with papers tending at the softer end of the spectrum tending to have fewer co-authors, use less substantive titles, have longer texts, cite older literature, and have a higher diversity of sources.
+In @fanelliBibliometricEvidenceHierarchy2013's analysis harder sciences include Space Science, Physics, Chemistry, less hard sciences include social sciences include Psychiatry/Psychology, Economics, Business and General Social Sciences, and the Humanities define the soft end of the spectrum.
+Following @fanelliBibliometricEvidenceHierarchy2013, I quantify the number of authors, length of article, relative title length, age of references, and diversity of references for a large sample of peer-reviewed journal articles.
+
 These parameters are useful because of how they signify consensus in a research community.
@@
 In low-consensus, or softer, fields, papers must be longer to present justification, nuance and contextualization of results.
-I used page numbers to measure article length.
-Similarly, the number of cited references will typically be higher in low-consensus disciplines where relatively fewer claims are taken for granted and thus many claims need citations for support, relative to harder fields [@skilton2006comparative].
+The number of substantive and informative words in an article's title tends to be positively correlated with article length in harder disciplines [@yitzhakiRelationTitleLength2002; @yitzhakiVariationInformativityTitles1997], reflecting a focus on empiricism and efficiency that is characteristic of high-consensus disciplines.
+While @yitzhakiRelationTitleLength2002 removed stop-words (e.g.
+prepositions, articles, conjunctions, etc.) to calculate article length, in order to generate results for comparison with @fanelliBibliometricEvidenceHierarchy2013 I follow his method of dividing the total word count of the article title by the total number of pages of the article to compute relative title length.
+
 The age of works cited has long been used as a measure of a field's hardness [@moed1998new; @bornerAtlasScience2010], based on the assumption that harder fields assimilate new results more rapidly that softer fields [@price1970citation].
-I calculated the Price index for each article, which is the proportion of all cited works that were published in the five years preceding the paper.
+I calculated a recency of references index for each article (also known as the Price index), which is the proportion of all cited works that were published in the five years preceding the paper.
+The diversity of references
 
@@
 
+# Number of authors ------------------
 boxlplot_n_authors <- 
@@
              log(authors_n))) +
-  geom_boxplot(colour = "red", 
+  geom_boxplot(
                size = 1)  +
@@
 
+# Relative title length  ----------------
+
+items_df_title <- 
+  items_df %>% 
+  filter(!is.na(pages_n)) %>%  
+  filter(!is.na(title_n)) %>% 
+  mutate(relative_title_length = log(title_n / pages_n))
+
+boxlplot_rel_title_length <- 
+items_df_title %>% 
+  filter(!is.na(year)) %>% 
+  ggplot(aes(1,
+             relative_title_length)) +
+  geom_boxplot(
+               size = 1)  +
+  # median for physics, from Fanelli & Glänzel Fig 2
+  geom_hline(yintercept = 0.8,
+             colour = "grey70") +
+  geom_text_repel(data = tibble(x = 1, 
+                                y = 0.8,
+                                label = "p"),
+                  aes(x, y, label = label),
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0) +
+  # median for soc sci, from Fanelli & Glänzel Fig 2
+  geom_hline(yintercept = 0.0,
+             colour = "grey70") +
+  geom_text_repel(data = tibble(x = 1, 
+                                y = 0.0,
+                                label = "s"),
+                  aes(x, y, label = label),
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0) +
+    # median for humanities, from Fanelli & Glänzel Fig 2
+  geom_hline(yintercept = -0.2,
+             colour = "grey70") +
+  geom_text_repel(data = tibble(x = 1, 
+                                y = -0.2,
+                                label = "h"),
+                  aes(x, y, label = label),
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0) +
+
+  scale_y_continuous(limits = c(-4.5, 3),
+                     breaks =  seq(-5, 5, 1),
+                     labels = seq(-5, 5, 1)) +
+  scale_x_continuous(labels = NULL) +
+  theme_minimal() +
+  theme(panel.grid  = element_blank()) +
+  ylab("Ratio of title length to article length (ln)") +
+  xlab("Relative title length")
+
+# Number of pages ------------------
 boxlplot_n_pages <- 
@@
              log(pages_n))) +
-  geom_boxplot(colour = "red", 
+  geom_boxplot(
                size = 1)  +
@@
                   force = 0) +
-  scale_y_continuous(limits = c(0, 5)) +
+  scale_y_reverse(limits = c(5, 0)) +
   scale_x_continuous(labels = NULL) +
@@
 
+# Price's index - recency of references ------------------
+library(stringr)
+
+# output storage
+prices_index <- vector("list", length = nrow(items_df))
+
+# loop, this takes a moment
+for(i in seq_len(nrow(items_df))){
+
+  refs <-  items_df$refs[i]
+  year <-  items_df$year[i]
+
+  ref_years <- 
+    as.numeric(str_match(str_extract_all(refs, ", [0-9]{4}, ")[[1]], "\\d{4}"))
+
+  preceeding_five_years <-  
+    seq(year - 5, year, 1)
+
+  refs_n_in_preceeding_five_years <- 
+    ref_years[ref_years %in% preceeding_five_years]
+
+  prices_index[[i]] <- 
+    length(refs_n_in_preceeding_five_years) / length(ref_years)
+
+  # for debugging
+  # print(i)
+
+}
+
+prices_index <- flatten_dbl(prices_index)
+
+# add to data frame
+items_df$prices_index <-  prices_index
+
+# plot
+boxlplot_price_index <- 
+items_df %>% 
+  ggplot(aes(1,
+             prices_index)) +
+  geom_boxplot(
+               size = 1)  +
+  # median for physics, from Fanelli & Glänzel Fig 2
+  geom_hline(yintercept = 0.38,
+             colour = "grey70") +
+  geom_text_repel(data = tibble(x = 1, 
+                                y = 0.38,
+                                label = "p"),
+                  aes(x, y, label = label),
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0) +
+  # median for soc sci, from Fanelli & Glänzel Fig 2
+  geom_hline(yintercept = 0.29,
+             colour = "grey70") +
+  geom_text_repel(data = tibble(x = 1, 
+                                y = 0.29,
+                                label = "s"),
+                  aes(x, y, label = label),
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0) +
+    # median for humanities, from Fanelli & Glänzel Fig 2
+  geom_hline(yintercept = 0.19,
+             colour = "grey70") +
+  geom_text_repel(data = tibble(x = 1, 
+                                y = 0.19,
+                                label = "h"),
+                  aes(x, y, label = label),
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0) +
+  scale_y_continuous(limits = c(0, 1)) +
+  scale_x_continuous(labels = NULL) +
+  theme_minimal() +
+  theme(panel.grid  = element_blank()) +
+  ylab("Prop. refs in last 5 years") +
+  xlab("Recency of references")
+
+# Shannon index - diversity of references ------------------
+# journal name as species, article as habitat
+
+# simplify the refs, since they are a bit inconsistent, some of
+# these steps take a few seconds
+ref_list1 <- map(items_df$refs, ~tolower(.x))
+ref_list2 <- map(ref_list1, ~str_replace_all(.x, "\\.|,| ", ""))
+ref_list3 <- map(ref_list2, ~str_split(.x, "\n"))
+ref_list4 <- map(ref_list3, ~data_frame(x = .x))
+ref_list5 <- bind_rows(ref_list4, .id = "id") 
+ref_list6 <- unnest(ref_list5)
+
+# get the journal names out of the refs
+ref_list7 <- 
+  ref_list6 %>% 
+  mutate(journal_name = gsub("\\-", "", x)) %>% 
+  mutate(journal_name = gsub("\\:", "", journal_name)) %>% 
+  mutate(journal_name = gsub("^[a-z'\\(\\)\\:]+[0-9]{4}", "", journal_name)) %>% 
+  mutate(journal_name = gsub("v[0-9]+.*", "", journal_name)) %>% 
+  mutate(journal_name = gsub("p[0-9]+$", "", journal_name))
+
+# prepare to compute shannon and join with other variables
+items_df$id <- 1:nrow(items_df)
+
+# tally of all referenced items
+all_cited_items <- 
+  ref_list7 %>% 
+  select(x) %>% 
+  group_by(x) %>% 
+  tally() %>% 
+  arrange(desc(n))
+
+# get a list of the top journals
+top_journals <- 
+  ref_list7 %>% 
+  select(journal_name) %>% 
+  group_by(journal_name) %>% 
+  tally() %>% 
+  filter(n > 50) %>% 
+  arrange(desc(n))
+
+# In the Shannon index, p_i is the proportion (n/N) of individuals of one particular species (journal) found (n) divided by the total number of individuals found (N), ln is the natural log, Σ is the sum of the calculations, and s is the number of species.
+
+# compute diversity of all citations for each article (habitat)
+shannon_per_item <- 
+  ref_list7 %>% 
+  group_by(id, x) %>% 
+  tally() %>% 
+  mutate(n_in_article = n) %>% 
+  select(-n) %>% 
+  left_join(all_cited_items) %>% 
+  mutate(p_i = n / sum(n, na.rm = TRUE)) %>% 
+  mutate(p_i_ln = log(p_i)) %>% 
+  group_by(id) %>% 
+  summarise(shannon = -sum(p_i * p_i_ln, na.rm = TRUE)) %>% 
+  mutate(id = as.numeric(id)) %>% 
+  arrange(id)  %>% 
+  left_join(items_df)
+
+# plot
+boxlplot_shannon_index <- 
+shannon_per_item %>% 
+  filter(!is.na(year)) %>% 
+  ggplot(aes(1,
+             shannon)) +
+  geom_boxplot(
+               size = 1)  +
+  # median for physics, from Fanelli & Glänzel Fig 2
+  geom_hline(yintercept = 2.6,
+             colour = "grey70") +
+  geom_text_repel(data = tibble(x = 1, 
+                                y = 2.6,
+                                label = "p"),
+                  aes(x, y, label = label),
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0) +
+  # median for soc sci, from Fanelli & Glänzel Fig 2
+  geom_hline(yintercept = 3.2,
+             colour = "grey70") +
+  geom_text_repel(data = tibble(x = 1, 
+                                y = 3.2,
+                                label = "s"),
+                  aes(x, y, label = label),
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0) +
+    # median for humanities, from Fanelli & Glänzel Fig 2
+  geom_hline(yintercept = 3.3,
+             colour = "grey70") +
+  geom_text_repel(data = tibble(x = 1, 
+                                y = 3.3,
+                                label = "h"),
+                  aes(x, y, 
+                      label = label,
+                      segment.colour = NA),
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0,
+                  nudge_x = 0.05) +
+  scale_y_reverse(limits = c(6, 0)) +
+  scale_x_continuous(labels = NULL) +
+  theme_minimal()  +
+  theme(panel.grid  = element_blank()) +
+  ylab("Shannon Index") +
+  xlab("Diversity of references")
+```
+
+```{r}
+#| label: fig-compare-other-fields
+#| fig-cap: Distributions of article characteristics hypothesised to reflect the level of consensus.
+The boxplot shows the distribution of values of archaeology articles.
+The thick black line in the middle of the boxplot is the median value, the box represents the inter-quartile range (the range between the 25th and 75th percentiles, where 50% of the data are located), and individual points represent outliers.
+The thin grey horizontal lines in each boxplot indicate the median values computed by Fanelli and Glanzel (2013), where p = physics, s = social sciences, h = humanities.
+
 library(cowplot)
@@
 plot_grid(boxlplot_n_authors, 
-          boxlplot_n_pages)
+          boxlplot_rel_title_length,
+          boxlplot_n_pages,
+          boxlplot_price_index, 
+          boxlplot_shannon_index,
+          nrow = 2)
 ```
+
+@fig-compare-other-fields shows the distribution of bibliometric variables for archaeology in context of data from other fields presented by @fanelliBibliometricEvidenceHierarchy2013.
+The most striking indicator of archaeology as a hard science is the number of authors, where it is between the social sciences and physics.
+Archaeology is a close fit with the social sciences in relative title length.
+It is between the the social sciences and humanities in recency of references and diversity of references.
+The clearest indicator of archaeology as a soft science is article length where it is like the humanities.
+Overall archaeology does not sit squarely at either end of the hard-soft spectrum.
+It is generally not a harder science than the social sciences, with the exception of collaborator group sizes.
 
@@
 
+```{r}
+
+over_time <- 
+  items_df %>% 
+  left_join(items_df_title) %>% 
+  left_join(shannon_per_item) %>% 
+  mutate(log_authors_n = log(authors_n),
+         log_pages_n = log(pages_n),
+         sqrt_refs_n = sqrt(refs_n),
+         journal_wrp = str_wrap(journal, 30)) %>% 
+  select(year,
+         log_authors_n,
+         log_pages_n,
+         prices_index, 
+         shannon,
+         sqrt_refs_n, 
+         relative_title_length)
+
+over_time_long <- 
+  over_time %>% 
+  ungroup() %>% 
+  select(-journal) %>% 
+  gather(variable, 
+         value,
+         -year) %>% 
+  filter(value != -Inf,
+         value !=  Inf) %>% 
+  mutate(variable = case_when(
+    variable == "log_authors_n" ~ "N. of authors (ln)",
+    variable == "log_pages_n"   ~ "N. of pages (ln)",
+    variable == "prices_index"  ~ "Recency of references",
+    variable == "shannon"  ~ "Diversity of references",
+    variable == "sqrt_refs_n"  ~ "N. of refs (sqrt)",
+    variable == "relative_title_length"  ~ "Relative title length (ln)"
+  )) %>% 
+  filter(!is.na(variable)) %>% 
+  filter(!is.nan(value)) %>% 
+  filter(!is.na(value)) %>% 
+  filter(value != "NaN") %>% 
+  mutate(value = parse_number(value))
+
+# compute beta estimates so we can colour lines to indicate more or
+# less scientific
+library(broom)
+over_time_long_models <- 
+  over_time_long %>% 
+  group_nest(variable) %>% 
+  mutate(model = map(data, ~tidy(lm(value ~ year, data = .)))) %>% 
+  unnest(model) %>% 
+  filter(term == 'year') %>% 
+  mutate(becoming_more_scientific = case_when(
+    variable == "N. of authors (ln)"         & estimate > 0 ~ "TRUE",
+    variable == "N. of pages (ln)"           & estimate < 0 ~ "TRUE",
+    variable == "N. of refs (sqrt)"          & estimate < 0 ~ "TRUE",
+    variable == "Recency of references"      & estimate > 0 ~ "TRUE",
+    variable == "Relative title length (ln)" & estimate > 0 ~ "TRUE",
+    variable == "Diversity of references"    & estimate < 0 ~ "TRUE",
+    TRUE ~ "FALSE"
+  ))
+
+# join with data
+over_time_long_colour <- 
+  over_time_long %>% 
+  left_join(over_time_long_models)
+```
+
+```{r}
+#| label: fig-change-over-time
+#| fig-cap: Distribution of article characteristics for archaeology articles over time.
+Individual articles are shown by data points.
+The Ordinary Least Squares method was used to fit the straight lines summarising the relationships between the variables and the time series.
+ggplot(over_time_long_colour,
+       aes(year, 
+           value,
+           colour = becoming_more_scientific)) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(method = "lm", 
+              size = 3,
+              colour = "#7570b3") +
+  facet_wrap( ~ variable,
+              scales = "free_y") +
+  theme_bw(base_size = 12) +
+  scale_color_manual(values = c("#d95f02", "#1b9e77" )) +
+  guides(colour = "none") +
+  ylab("")
+```
+
+@fig-change-over-time shows how the bibliometric indicators of field hardness have changed of time for archaeology articles.
+By two measures, the number of authors and relative title length, archaeology has become increasingly harder over time.
+On the other hand, three metrics indicate that archaeology has become softer (diversity of references, article length and recently of references).
+
 ### How do archaeology journals vary in hardness?
+
+## References
+
+## Acknowledgements
 

### 2025-01-12 code/006-bibmetrics-summary.R

@@ -32,11 +32,16 @@ over_time_long <-
   mutate(variable = case_when(
     variable == "log_authors_n" ~ "N. of authors (ln)",
     variable == "log_pages_n"   ~ "N. of pages (ln)",
-    variable == "prices_index"  ~ "Price's index",
-    variable == "shannon"  ~ "Shannon div. of sources",
+    variable == "prices_index"  ~ "Recency of references",
+    variable == "shannon"  ~ "Diversity of references",
     variable == "sqrt_refs_n"  ~ "N. of refs (sqrt)",
     variable == "relative_title_length"  ~ "Relative title length (ln)"
-  ))
+  )) %>% 
+  filter(!is.na(variable)) %>% 
+  filter(!is.nan(value)) %>% 
+  filter(!is.na(value)) %>% 
+  filter(value != "NaN") %>% 
+  mutate(value = parse_number(value))
 
 # compute beta estimates so we can colour lines to indicate more or
 # less scientific
@@ -48,12 +53,12 @@ over_time_long_models <-
   unnest(model) %>% 
   filter(term == 'year') %>% 
   mutate(becoming_more_scientific = case_when(
-    variable == "N. of authors (ln)" & estimate > 0 ~ "TRUE",
-    variable == "N. of pages (ln)"   & estimate < 0 ~ "TRUE",
-    variable == "N. of refs (sqrt)"  & estimate < 0 ~ "TRUE",
-    variable == "Price's index"     & estimate > 0 ~ "TRUE",
-    variable == " Relative title length (ln)"     & estimate > 0 ~ "TRUE",
-    variable == "Shannon div. of sources"     & estimate < 0 ~ "TRUE",
+    variable == "N. of authors (ln)"         & estimate > 0 ~ "TRUE",
+    variable == "N. of pages (ln)"           & estimate < 0 ~ "TRUE",
+    variable == "N. of refs (sqrt)"          & estimate < 0 ~ "TRUE",
+    variable == "Recency of references"      & estimate > 0 ~ "TRUE",
+    variable == "Relative title length (ln)" & estimate > 0 ~ "TRUE",
+    variable == "Diversity of references"    & estimate < 0 ~ "TRUE",
     TRUE ~ "FALSE"
   )) 
 
@@ -66,13 +71,14 @@ ggplot(over_time_long_colour,
        aes(year, 
            value,
            colour = becoming_more_scientific)) +
-  geom_point(alpha = 0.005) +
[…]

### 2025-01-13 paper.qmd
where did my text go?

@@
 @fanelliBibliometricEvidenceHierarchy2013 found a spectrum of statistically significant variation in bibliometric variables from the physical to the social sciences, with papers tending at the softer end of the spectrum tending to have fewer co-authors, use less substantive titles, have longer texts, cite older literature, and have a higher diversity of sources.
-In @fanelliBibliometricEvidenceHierarchy2013's analysis harder sciences include Space Science, Physics, Chemistry, less hard sciences include social sciences include Psychiatry/Psychology, Economics, Business and General Social Sciences, and the Humanities define the soft end of the spectrum.
+In @fanelliBibliometricEvidenceHierarchy2013's analysis harder sciences include Space Science, Physics, Chemistry, less hard sciences include social sciences include Psychiatry, Psychology, Economics, Business and General Social Sciences, and the Humanities define the soft end of the spectrum.
 Following @fanelliBibliometricEvidenceHierarchy2013, I quantify the number of authors, length of article, relative title length, age of references, and diversity of references for a large sample of peer-reviewed journal articles.
@@
 A larger number of authors on a paper reflects collaboration of people working together on a common goal.
-Collaborators have specialized roles, each of whom has the ability to study a part of the problem with high accuracy and detail, with harder fields having larger groups of collaborators[@zuckerman1972age].
+Collaborators have specialized roles, each of whom has the ability to study a part of the problem with high accuracy and detail, with harder fields having larger groups of collaborators [@zuckerman1972age].
 Reflecting this collaboration group size, harder disciplines tend to have higher average numbers of authors on papers.
@@
 
-I sampled `r n_articles` papers published during `r year_min`-`r year_max`.
-This sample was collected from Clarivate's Web of Science database by first selecting the Web of Science category 'Archaeology' and the Document type 'article' (n = 28,871).
-To focus on journals of broad relevance to most archaeologists, and that are representative of substantial communities of practice, I then filtered the results to keep only articles published in the top-ranking 25 journals according to their h-indices as reported by Clarivate's Journal Citation Indicator.
+While Fanelli and Glänzel (2013) analysed papers published in a single year (2012), I found only 303 papers for that same year, and 70.2691554% of papers in the sample published after that date.
+To make efficient use of the available data and ensure robust representation from different areas of archaeology, including those with lower frequencies of journal article publication, I analysed 9697 papers published during 1975-2025.
+This sample was collected from Clarivate’s Web of Science database by first selecting the Web of Science category ‘Archaeology’ and the Document type ‘article’ (n = 28,871).
+To focus on journals of broad relevance to most archaeologists, and that are representative of substantial communities of practice, I then filtered the results to keep only articles published in the top-ranking 25 journals according to their h-indices as reported by Clarivate’s Journal Citation Indicator.
 Finally I excluded journals with less than 100 articles in the database.
@@
 
-# Price's index - recency of references ------------------
+# Number of references ------------------
+boxlplot_n_refs <- 
+items_df %>% 
+  ggplot(aes(1,
+             sqrt(refs_n))) +
+  geom_boxplot(
+               size = 1)  +
+  # median for physics, from Fanelli & Glänzel Fig 2
+  geom_hline(yintercept = 6,
+             colour = "grey70") +
+  geom_text_repel(data = tibble(x = 1, 
+                                y = 6,
+                                label = "p"),
+                  aes(x, y, label = label),
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0) +
+  # median for soc sci, from Fanelli & Glänzel Fig 2
+  geom_hline(yintercept = 7,
+             colour = "grey70") +
+  geom_text_repel(data = tibble(x = 1, 
+                                y = 7,
+                                label = "s"),
+                  aes(x, y, label = label),
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0) +
+    # median for humanities, from Fanelli & Glänzel Fig 2
+  geom_hline(yintercept = 5,
+             colour = "grey70") +
+  geom_text_repel(data = tibble(x = 1, 
+                                y = 5,
+                                label = "h"),
+                  aes(x, y, label = label),
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0) +
+  scale_y_reverse(limits = c(20, 0)) +
+  scale_x_continuous(labels = NULL) +
+  theme_minimal() +
+  theme(panel.grid  = element_blank()) +
+  ylab("N. of refs (sqrt)") +
+  xlab("Number of references")
+
+# Price's index - age of references ------------------
 library(stringr)
@@
           boxlplot_n_pages,
+          boxlplot_n_refs,
           boxlplot_price_index, 
@@
 
-@fig-compare-other-fields shows the distribution of bibliometric variables for archaeology in context of data from other fields presented by @fanelliBibliometricEvidenceHierarchy2013.
+@fig-mpare-other-fields shows the distribution of bibliometric variables for archaeology in context of data from other fields presented by @fanelliBibliometricEvidenceHierarchy2013.
 The most striking indicator of archaeology as a hard science is the number of authors, where it is between the social sciences and physics.
@@
 It is between the the social sciences and humanities in recency of references and diversity of references.
-The clearest indicator of archaeology as a soft science is article length where it is like the humanities.
+The clearest indicator of archaeology as a soft science is article length where it is similar to the humanities.
 Overall archaeology does not sit squarely at either end of the hard-soft spectrum.
@@
 # compute beta estimates so we can colour lines to indicate more or
-# less scientific
+# less hard
+
 library(broom)
+
 over_time_long_models <- 
@@
   left_join(over_time_long_models)
+
 ```
@@
 #| fig-cap: Distribution of article characteristics for archaeology articles over time.
-Individual articles are shown by data points.
+D ata points represent individual articles.
 The Ordinary Least Squares method was used to fit the straight lines summarising the relationships between the variables and the time series.
+
+library(ggpmisc)
+formula <-  y ~ x
+
 ggplot(over_time_long_colour, 
@@
               colour = "#7570b3") +
+  stat_poly_eq(use_label("R2", "R2.CI", "P"), 
+               formula = formula) +
   facet_wrap( ~ variable,
@@
   guides(colour = "none") +
-  ylab("")
+  ylab("") +
+  stat_poly_eq(use_label("R2", "R2.CI", "P"), 
+               formula = formula)
+
 ```
@@
 ### How do archaeology journals vary in hardness?
-
-## References
-
-## Acknowledgements
 

### 2025-01-13 code/006-bibmetrics-summary.R

@@ -1,85 +1,0 @@
-
-#--------------------------------
-# Six variables over time
-
-# https://github.com/benmarwick/March-2019-Cambridge-Big-Data-Archaeology/blob/master/code/002-fanelli-hierarchy-of-sciences.R
-
-over_time <- 
-  items_df %>% 
-  left_join(items_df_title) %>% 
-  left_join(shannon_per_item) %>% 
-  mutate(log_authors_n = log(authors_n),
-         log_pages_n = log(pages_n),
-         sqrt_refs_n = sqrt(refs_n),
-         journal_wrp = str_wrap(journal, 30)) %>% 
-  select(year,
-         log_authors_n,
-         log_pages_n,
-         prices_index, 
-         shannon,
-         sqrt_refs_n, 
-         relative_title_length)
-
-over_time_long <- 
-  over_time %>% 
-  ungroup() %>% 
-  select(-journal) %>% 
-  gather(variable, 
-         value,
-         -year) %>% 
-  filter(value != -Inf,
-         value !=  Inf) %>% 
-  mutate(variable = case_when(
-    variable == "log_authors_n" ~ "N. of authors (ln)",
-    variable == "log_pages_n"   ~ "N. of pages (ln)",
-    variable == "prices_index"  ~ "Recency of references",
-    variable == "shannon"  ~ "Diversity of references",
-    variable == "sqrt_refs_n"  ~ "N. of refs (sqrt)",
-    variable == "relative_title_length"  ~ "Relative title length (ln)"
-  )) %>% 
-  filter(!is.na(variable)) %>% 
-  filter(!is.nan(value)) %>% 
-  filter(!is.na(value)) %>% 
-  filter(value != "NaN") %>% 
-  mutate(value = parse_number(value))
-
-# compute beta estimates so we can colour lines to indicate more or
-# less scientific
-library(broom)
-over_time_long_models <- 
-  over_time_long %>% 
-  group_nest(variable) %>% 
-  mutate(model = map(data, ~tidy(lm(value ~ year, data = .)))) %>% 
-  unnest(model) %>% 
-  filter(term == 'year') %>% 
-  mutate(becoming_more_scientific = case_when(
-    variable == "N. of authors (ln)"         & estimate > 0 ~ "TRUE",
-    variable == "N. of pages (ln)"           & estimate < 0 ~ "TRUE",
-    variable == "N. of refs (sqrt)"          & estimate < 0 ~ "TRUE",
[…]

### 2025-01-13 paper.qmd
reinsert some inline R code

@@
 year_min <- min(items_df$year)
+
+items_df_2012 <- 
+  items_df %>% 
+  filter(year == 2012)
+
+# how many archaeology articles in 2012
+n_items_df_2012 <- nrow(items_df_2012)
+
+# how many after 2012?
+n_items_df_after_2012 <- 
+  items_df %>% 
+  filter(year %in% 2013:year_max) %>% 
+  nrow()
+
+# what proportion of archaeology articles published after 2012
+prop_pub_after2012 <-  n_items_df_after_2012  / n_articles
+
 ```
 
-While Fanelli and Glänzel (2013) analysed papers published in a single year (2012), I found only 303 papers for that same year, and 70.2691554% of papers in the sample published after that date.
+While Fanelli and Glänzel (2013) analysed papers published in a single year (2012), I found only `r n_items_df_2012` papers for that same year, and `r round(prop_pub_after2012*100, 0)`% of papers in the sample published after that date.
 To make efficient use of the available data and ensure robust representation from different areas of archaeology, including those with lower frequencies of journal article publication, I analysed 9697 papers published during 1975-2025.

### 2025-01-13 paper.qmd
finish results section: change over time

@@
 
-# Number of references ------------------
-boxlplot_n_refs <- 
-items_df %>% 
-  ggplot(aes(1,
-             sqrt(refs_n))) +
-  geom_boxplot(
-               size = 1)  +
-  # median for physics, from Fanelli & Glänzel Fig 2
-  geom_hline(yintercept = 6,
-             colour = "grey70") +
-  geom_text_repel(data = tibble(x = 1, 
-                                y = 6,
-                                label = "p"),
-                  aes(x, y, label = label),
-                  bg.colour = "white", 
-                  bg.r = .2, 
-                  force = 0) +
-  # median for soc sci, from Fanelli & Glänzel Fig 2
-  geom_hline(yintercept = 7,
-             colour = "grey70") +
-  geom_text_repel(data = tibble(x = 1, 
-                                y = 7,
-                                label = "s"),
-                  aes(x, y, label = label),
-                  bg.colour = "white", 
-                  bg.r = .2, 
-                  force = 0) +
-    # median for humanities, from Fanelli & Glänzel Fig 2
-  geom_hline(yintercept = 5,
-             colour = "grey70") +
-  geom_text_repel(data = tibble(x = 1, 
-                                y = 5,
-                                label = "h"),
-                  aes(x, y, label = label),
-                  bg.colour = "white", 
-                  bg.r = .2, 
-                  force = 0) +
-  scale_y_reverse(limits = c(20, 0)) +
-  scale_x_continuous(labels = NULL) +
-  theme_minimal() +
-  theme(panel.grid  = element_blank()) +
-  ylab("N. of refs (sqrt)") +
-  xlab("Number of references")
-
 # Price's index - age of references ------------------
@@
           boxlplot_n_pages,
-          boxlplot_n_refs,
           boxlplot_price_index, 
@@
          log_pages_n = log(pages_n), 
-         sqrt_refs_n = sqrt(refs_n), 
          journal_wrp = str_wrap(journal, 30)) %>%  
@@
     variable == "shannon"  ~ "Diversity of references", 
-    variable == "sqrt_refs_n"  ~ "N. of refs (sqrt)", 
     variable == "relative_title_length"  ~ "Relative title length (ln)" 
@@
               colour = "#7570b3") +
-  stat_poly_eq(use_label("R2", "R2.CI", "P"), 
-               formula = formula) +
   facet_wrap( ~ variable,
@@
   theme_bw(base_size = 12) +
-  scale_color_manual(values = c("#d95f02", "#1b9e77" )) +
+  scale_color_manual(values = c("#d95f02",
+                                "#1b9e77" )) +
   guides(colour = "none") +
   ylab("") +
-  stat_poly_eq(use_label("R2", "R2.CI", "P"), 
-               formula = formula)
+  stat_poly_eq(use_label("eq", "P"),
+               label.y = 0.95) +
+  stat_poly_eq(use_label("adj.R2", "R2.CI"),
+               label.y = 0.9)
 
@@
 On the other hand, three metrics indicate that archaeology has become softer (diversity of references, article length and recently of references).
+Although all the relationships are statistically significant, generally these temporal trends are very weak with low slope values, indicating very slow change over time.
+Similarly the r-squared values are very low, demonstrating that much of the variability in these metrics is independent of time.
+The  most striking change over time is in the increase in the number of authors, which has the highest r-squared value of these metrics.
+One interesting detail evident in @fig-change-over-time is the increase in the range of diversity of references after about 2010.
+This may be due to some broader changes in academic publishing around this time, such as moves to digital-first continuous publishing, new journals appearing (e.g.
+Archaeological and Anthropological Sciences in 2009 and Journal of Island & Coastal Archaeology in 2010), and non-archaeology journals becoming more relevant to archaeologists.
+For example, _PLOS ONE_ received its first impact factor in 2010 and in 2011 Nature's _Scientific Reports_ began publishing [@malashichev2017open].
 

### 2025-01-13 paper.qmd
results: variation among journals

@@
                   bg.r = .2, 
-                  force = 0) +
+                  force = 0,
+                  nudge_x = 0.05) +
   # median for soc sci, from Fanelli & Glänzel Fig 2
@@
 The thin grey horizontal lines in each boxplot indicate the median values computed by Fanelli and Glanzel (2013), where p = physics, s = social sciences, h = humanities.
+#| fig-height: 10
 
@@
          shannon, 
-         sqrt_refs_n,  
          relative_title_length)
@@
 The Ordinary Least Squares method was used to fit the straight lines summarising the relationships between the variables and the time series.
+#| fig-height: 10
 
@@
 
+```{r}
+library(ggridges)
+
+journal_title_size <- 4
+
+plot_journals_authors <- 
+items_df %>% 
+  mutate(journal = str_wrap(journal, 20)) %>% 
+  ggplot(aes(y = reorder(journal,
+                     authors_n), 
+             x = log(authors_n),
+             fill =   after_stat(x),
+             height = after_stat(density))) +
+  geom_density_ridges_gradient(stat = "density",
+                               colour = "white") +
+  scale_fill_viridis_c() +
+  guides(fill = 'none') +
+  theme_minimal() +
+  theme(axis.text.y = element_text(size = journal_title_size)) +
+  ylab("") + 
+  xlab("Number of authors (ln)")
+
+plot_journals_article_length <- 
+items_df %>% 
+  mutate(journal = str_wrap(journal, 20)) %>% 
+  ggplot(aes(y = reorder(journal,
+                     -pages_n), 
+             x = log(pages_n),
+             fill =   after_stat(x),
+             height = after_stat(density))) +
+  geom_density_ridges_gradient(stat = "density",
+                               colour = "white") +
+  scale_fill_viridis_c() +
+  guides(fill = 'none') +
+  theme_minimal() +
+  theme(axis.text.y = element_text(size = journal_title_size)) +
+  xlab("Number of pages (ln)") +
+  ylab("")
+
+plot_journals_title_length <- 
+items_df_title %>% 
+  mutate(journal = str_wrap(journal, 20)) %>% 
+  ggplot(aes(y = reorder(journal,
+                     relative_title_length), 
+             x = relative_title_length,
+             fill =   after_stat(x),
+             height = after_stat(density))) +
+  geom_density_ridges_gradient(stat = "density",
+                               colour = "white") +
+  scale_fill_viridis_c() +
+  guides(fill = 'none') +
+  theme_minimal() +
+  theme(axis.text.y = element_text(size = journal_title_size)) +
+  ylab("") +
+  xlab("Relative title length (ln)")
+
+plot_journals_ref_recency <- 
+items_df %>% 
+  mutate(journal = str_wrap(journal, 20)) %>% 
+  group_by(journal) %>% 
+  ggplot(aes(y = reorder(journal,
+                     prices_index), 
+             x = prices_index,            
+             fill =   after_stat(x),
+             height = after_stat(density))) +
+  geom_density_ridges_gradient(stat = "density",
+                               colour = "white") +
+  scale_fill_viridis_c() +
+  guides(fill = 'none') +
+  theme_minimal() +
+  theme(axis.text.y = element_text(size = journal_title_size)) +
+  ylab("") +
+  xlab("Recency of references")
+
+plot_journals_ref_diversity <- 
+shannon_per_item %>%  
+  mutate(journal = str_wrap(journal, 20)) %>% 
+  group_by(journal) %>% 
+  ggplot(aes(y = reorder(journal,
+                     -shannon), 
+             x = shannon,
+             fill =   after_stat(x),
+             height = after_stat(density))) +
+  geom_density_ridges_gradient(stat = "density",
+                               colour = "white") +
+  scale_fill_viridis_c() +
+  guides(fill = 'none') +
+  theme_minimal() +
+  theme(axis.text.y = element_text(size = journal_title_size)) +
+  ylab("") +
+  xlab("Diversity of references")
+```
+
+```{r}
+#| label: fig-variation-by-journal
+#| fig-cap: Variation by journal
+#| fig-height: 10
+
+library(cowplot)
+
+plot_grid(plot_journals_authors, 
+          plot_journals_article_length,
+          plot_journals_title_length,
+          plot_journals_ref_recency, 
+          plot_journals_ref_diversity,
+          nrow = 2)
+```
+
+@fig-variation-by-journal shows
+
 ```{=html}

### 2025-01-22 paper.qmd
change lm to gam for change over time plots

@@
                   bg.r = .2, 
-                  force = 0,
-                  nudge_x = 0.05) +
+                  force = 0) +
   # median for soc sci, from Fanelli & Glänzel Fig 2
@@
                   bg.r = .2, 
-                  force = 0) +
+                  force = 0,
+                  nudge_x = 0.05) +
 
@@
 
-@fig-mpare-other-fields shows the distribution of bibliometric variables for archaeology in context of data from other fields presented by @fanelliBibliometricEvidenceHierarchy2013.
+@fig-compare-other-fields shows the distribution of bibliometric variables for archaeology in context of data from other fields presented by @fanelliBibliometricEvidenceHierarchy2013.
 The most striking indicator of archaeology as a hard science is the number of authors, where it is between the social sciences and physics.
@@
 #| fig-cap: Distribution of article characteristics for archaeology articles over time.
-D ata points represent individual articles.
-The Ordinary Least Squares method was used to fit the straight lines summarising the relationships between the variables and the time series.
+Data points represent individual articles.
+The colour of the points indicates if the overall trend is toward softer (orange) or harder (green) Generalized Additive Models were computed to fit the lines summarising the relationships between the variables and the time series.
 #| fig-height: 10
@@
 library(ggpmisc)
+library(mgcv)
 formula <-  y ~ x
 
-ggplot(over_time_long_colour, 
+over_time_long_colour_gams <- 
+over_time_long_colour %>% 
+  nest(.by = variable) %>% 
+  mutate(mod_gam = lapply(data, 
+                          function(df) gam(year ~ s(value, bs = "cr"), 
+                                           data = df)))
+
+over_time_long_colour_gams_summary <- 
+over_time_long_colour %>% 
+  nest(.by = variable) %>% 
+  mutate(fit = map(data, ~mgcv::gam(year ~ s(value, bs = "cs"), data = .)),
+         results = map(fit, glance),
+         R.square = map_dbl(fit, ~ summary(.)$r.sq)) %>%
+  unnest(results) %>%
+  select(-data, -fit) %>% 
+  select(variable, adj.r.squared)
+
+over_time_long_colour_gams_summary_df <- 
+over_time_long_colour %>% 
+  left_join(over_time_long_colour_gams_summary)
+
+ggplot() + 
+  geom_point(data = over_time_long_colour_gams_summary_df, 
        aes(year,  
            value, 
-           colour = becoming_more_scientific)) + 
-  geom_point(alpha = 0.5) + 
-  geom_smooth(method = "lm",  
+           colour = becoming_more_scientific),
+       alpha = 0.5) + 
+  geom_smooth(data = over_time_long_colour_gams_summary_df,
+              aes(year, value),
+              method="gam", 
+              formula = y ~ s(x, bs = "cs"),
+              se = FALSE,  
               size = 3, 
@@
   ylab("") +
-  stat_poly_eq(use_label("eq", "P"),
-               label.y = 0.95) +
-  stat_poly_eq(use_label("adj.R2", "R2.CI"),
-               label.y = 0.9)
+  geom_text(data = over_time_long_colour_gams_summary_df %>% 
+              group_by(variable) %>% 
+              summarise(max_value = max(value),
+                        adj.r.squared = unique(adj.r.squared)),
+           aes(
+           x = 1980, 
+           y = max_value, 
+           label = paste("Pseudo R² = ", 
+                         signif(adj.r.squared, 
+                                digits = 3))),
+           hjust = 0, vjust = 1.5)
 

### 2025-01-23 paper.qmd
hardness by journal

@@
 bibliography: references.bib
+fig-dpi: 600
 execute: 
@@
 Similarly, quantitative analysis of the frequency of positive results (ie.
-full or partial support for a research hypothesis) in publications is significantly correlated with hardness, consistent with a model where researchers in harder fields more readily accept any result their research produces, while those in softer fields have more freedom to choose which theories and hypotheses to test and how to interpret results [@PositiveResultsIncrease].
+full or partial support for a research hypothesis) in publications is significantly correlated with hardness, consistent with a model where researchers in harder fields more readily accept any result their research produces, while those in softer fields have more freedom to choose which theories and hypotheses to test and how to interpret results [@fanelliPositiveResultsIncrease2010].
 The hard-soft spectrum is also evident in surveys of how researchers view their own work relative to those in other fields [@biglanCharacteristicsSubjectMatter1973].
@@
 
+# how many distinct journals?
+n_journals <- n_distinct(items_df$journal)
 ```
@@
 To focus on journals of broad relevance to most archaeologists, and that are representative of substantial communities of practice, I then filtered the results to keep only articles published in the top-ranking 25 journals according to their h-indices as reported by Clarivate’s Journal Citation Indicator.
-Finally I excluded journals with less than 100 articles in the database.
+Finally I excluded journals with less than 100 articles in the database, resulting in `r n_journals` journals.
 
@@
 The thin grey horizontal lines in each boxplot indicate the median values computed by Fanelli and Glanzel (2013), where p = physics, s = social sciences, h = humanities.
-#| fig-height: 10
+ln denotes the natural logarithm, or logarithm to the base e.
+#| fig-height: 8
 
@@
   left_join(shannon_per_item) %>%  
+      filter(relative_title_length != -Inf, 
+            relative_title_length !=  Inf,
+            prices_index != "NaN" 
+           ) %>% 
   mutate(log_authors_n = log(authors_n), 
@@
 The colour of the points indicates if the overall trend is toward softer (orange) or harder (green) Generalized Additive Models were computed to fit the lines summarising the relationships between the variables and the time series.
-#| fig-height: 10
+#| fig-height: 6
 
@@
 Similarly the r-squared values are very low, demonstrating that much of the variability in these metrics is independent of time.
+
 The most striking change over time is in the increase in the number of authors, which has the highest r-squared value of these metrics.
@@
 This may be due to some broader changes in academic publishing around this time, such as moves to digital-first continuous publishing, new journals appearing (e.g.
-Archaeological and Anthropological Sciences in 2009 and Journal of Island & Coastal Archaeology in 2010), and non-archaeology journals becoming more relevant to archaeologists.
-For example, _PLOS ONE_ received its first impact factor in 2010 and in 2011 Nature's _Scientific Reports_ began publishing [@malashichev2017open].
+_Archaeological and Anthropological Sciences_ in 2009 and _Journal of Island & Coastal Archaeology_ in 2010), and non-archaeology journals becoming more relevant to archaeologists.
+For example, *PLOS ONE* received its first impact factor in 2010 and in 2011 Nature's *Scientific Reports* began publishing [@malashichev2017open].
 
@@
 ```{r}
-library(ggridges)
 
@@
 
+# get rank order of journals by these bibliometic variables
+
+journal_metrics_for_plotting <- 
+items_df %>% 
+  left_join(items_df_title) %>% 
+  left_join(shannon_per_item) %>% 
+  ungroup() %>% 
+  select(journal,
+         authors_n,  # log
+         pages_n,    # log
+         relative_title_length,
+         prices_index,
+         shannon
+         ) %>% 
+    filter(relative_title_length != -Inf, 
+           relative_title_length !=  Inf,
+           prices_index != "NaN" 
+           ) %>% 
+  mutate(
+    log_authors = log(authors_n),
+    log_pages = log(pages_n)
+  )
+
+journal_metrics_for_plotting_summary <- 
+  journal_metrics_for_plotting %>% 
+  group_by(journal) %>% 
+  summarise(mean_log_authors = mean(log_authors),
+            mean_log_pages =   mean(log_pages),
+            mean_relative_title_length = mean(relative_title_length),
+            mean_prices_index = mean(prices_index),
+            mean_shannon =      mean(shannon))
+
+library(FactoMineR)
+library(factoextra)
+
+journal_metrics_for_plotting_summary_pca <- 
+journal_metrics_for_plotting_summary %>% 
+  column_to_rownames("journal") %>% 
+  PCA(graph = FALSE)
+
+  fviz_pca_biplot(journal_metrics_for_plotting_summary_pca,
+                  label = "var", repel = TRUE) + 
+  geom_text_repel(aes(label = str_wrap(journal_metrics_for_plotting_summary$journal, 20)), 
+            size = 2) +
+    ggtitle("")
+
+```
+
+```{r}
+# looking into rankings of the journals
+
+journal_summary_metrics_ranks <- 
+journal_metrics_for_plotting_summary %>% 
+  mutate(across(starts_with("mean"), 
+                ~ rank(-.), 
+                .names = "rank_{.col}")) %>% 
+  select(journal, starts_with("rank")) %>% 
+  # reorder by hardness
+  mutate(rank_mean_log_pages = 21 - rank_mean_log_pages,
+         rank_mean_shannon =   21 - rank_mean_shannon)
+
+library(irr)
+journal_summary_metrics_ranks_test <- 
+journal_summary_metrics_ranks %>% 
+  select(-journal) %>% 
+  kendall(correct = TRUE)
+
+# Convert to scientific text
+pretty_print_sci <- function(num){
+  scientific_text <- paste0(gsub("e", " x 10^",  # Replace 'e' with ' x 10^'
+                          sprintf("%.2e", num)), "^") # round to 2 sf
+  return(scientific_text)
+  }
+
+borda_count_tbl <- function(votes_tbl) {
+  # Number of voters
+  num_voters <- ncol(votes_tbl) - 1
+
+  # Calculate scores for each option
+  scores <- votes_tbl %>%
+    rowwise() %>%
+    mutate(Score = sum(num_voters - c_across(starts_with("rank_")))) %>%
+    ungroup() %>%
+    select(1, Score)
+
+  # Return scores
+  return(scores)
+}
+
+# Calculate Borda Count scores
+borda_scores <- 
+journal_summary_metrics_ranks %>% 
+  borda_count_tbl() %>% 
+  rename("Journal" = "journal") %>% 
+  arrange(desc(Score))
+
+plot_borda_scores <- 
+  borda_scores %>% 
+  mutate(Journal = str_wrap(Journal, 20)) %>% 
+ggplot() +
+  aes(reorder(Journal, Score),
+      Score) +
+  geom_col() +
+  coord_flip() +
+  ylab("Borda Count scores") + 
+  xlab("") +
+  theme_minimal() +
+  theme(axis.text.y = element_text(size = journal_title_size))
+
+```
+
+```{r}
+library(ggridges)
+
 plot_journals_authors <- 
-items_df %>% 
+journal_metrics_for_plotting %>% 
   mutate(journal = str_wrap(journal, 20)) %>% 
   ggplot(aes(y = reorder(journal, 
-                     authors_n), 
-             x = log(authors_n),
+                         log_authors,
+                     FUN = mean),
+             x = log_authors,
              fill =   after_stat(x),
@@
 plot_journals_article_length <- 
-items_df %>% 
+journal_metrics_for_plotting %>% 
   mutate(journal = str_wrap(journal, 20)) %>% 
   ggplot(aes(y = reorder(journal,
-                     -pages_n), 
-             x = log(pages_n),
+                     -log_pages,
+                     FUN = mean), 
+             x = log_pages,
              fill =   after_stat(x),
@@
 plot_journals_title_length <- 
-items_df_title %>% 
+journal_metrics_for_plotting %>% 
   mutate(journal = str_wrap(journal, 20)) %>% 
   ggplot(aes(y = reorder(journal,
-                     relative_title_length), 
+                         relative_title_length,
+                         FUN = mean), 
              x = relative_title_length,
@@
 plot_journals_ref_recency <- 
-items_df %>% 
+journal_metrics_for_plotting %>% 
   mutate(journal = str_wrap(journal, 20)) %>% 
@@
   ggplot(aes(y = reorder(journal,
-                     prices_index), 
+                     prices_index,
+                     FUN = mean), 
              x = prices_index,            
@@
 plot_journals_ref_diversity <- 
-shannon_per_item %>%  
+journal_metrics_for_plotting %>%  
   mutate(journal = str_wrap(journal, 20)) %>% 
@@
   ggplot(aes(y = reorder(journal,
-                     -shannon), 
+                     -shannon,
+                     FUN = mean), 
              x = shannon,
@@
 #| label: fig-variation-by-journal
-#| fig-cap: Variation by journal
-#| fig-height: 10
+#| fig-cap: Variation in bibliometric indicators of hardness for 20 archaeological journals.
+The journals are ordered for each indicator so that within each plot, the harder journals are at the top of the plot and the softer journals are at the base.
+The lower right panel shows a bar plot that is the single consensus ranking computed from all five variables, using the Borda Count ranking algorithm.
+#| fig-height: 12
 
@@
           plot_journals_ref_diversity,
-          nrow = 2)
+          plot_borda_scores,
+          nrow = 2,
+          labels = "AUTO")
+
 ```
 
-@fig-variation-by-journal shows
+@fig-variation-by-journal shows the distribution of our bibliometric variables of hardness for each of the `r n_journals` journals in the sample.
+Overall agreement between these bibliometric variables in ranking these journals on a hard-soft spectrum is moderate to strong, with a Kendall's coefficient of concordance (Wt) value of `r round(journal_summary_metrics_ranks_test$value, 2)` (in a 0-1 range, where 1 is perfect agreement) and a p-value of `r pretty_print_sci(journal_summary_metrics_ranks_test$p.value)`.
+Panel F of @fig-variation-by-journal shows an overall consensus ranking of all journals in the sample.
+In this consensus ranking the *Journal of Archaeological Science* among the top five archaeology journals for hardness.
+It is placed at the harder end of the hard-soft spectrum especially by the number of pages and relative title length, and to lesser degrees by the number of authors and recency of references.
+However, according to the diversity of references, the *Journal of Archaeological Science* is at the middle of the spectrum.
+
+The *Journal of Cultural Heritage* is the only journal that consistently ranks as hard across all variables, occurring in the top five journals for all five metrics.
+This journal primarily publishes materials science and computational analyses related to conservation and preservation of historic objects in museums and other collections.
+Authors of papers in recent issues have affiliations with museums, cultural heritage programs, and chemistry, engineering, and physics departments at European and Chinese universities.
+Notably, papers in this journal typically do not engage in questions or debates about past human behaviour or culture.
+The absence of these questions in research published in this journal makes it an outlier here, since these questions are central to a common definition of archaeology as 'cultural anthropology of the past', a phrase first found in @leroigourhan1946 and repeated in widely-used contemporary undergraduate textbooks such as @renfrew2024.
+Most archaeologists would likely be surprised at the decision by Clarivate to include the *Journal of Cultural Heritage* in their category of archaeology journals, leading to this result in @fig-variation-by-journal where the hardest archaeology journal publishes papers that are not very archaeological at all because they do not engage with anthropological topics.
+
+The *Journal of Archaeological Research* is notable because it consistently ranks as soft, ranking as the softest journal for four of our five bibliometric variables.
+This is a predicable result for a review journal, which is a distinct type of journal dedicated to summarizing, analyzing, and synthesizing existing research in a particular field.
+The stated aim of the *Journal of Archaeological Research* is to 'bring together the most recent international research summaries on a broad range of topics and geographical areas' [@feinmanAimsScopeJournal2024].
+A typical article is a long single-authored synthesis of archaeology in a region or on a topic.
+As the only review journal in this sample, this is a stark contrast to the other journals here that present original research findings, and like the *Journal of Cultural Heritage*, may be considered an outlier in this sample.
+
+## Dicsussion
+
+This macroscopic perspective on the way thousands of archaeologists communicate their research has produced a complex picture of archaeology as a science.
+In the context of a broad spectrum of other research areas, archaeologists behave like social scientists.
+We are harder than typical social scientists in tending to form larger groups of collaborators more often, and softer in sometimes writing longer articles than more resemble humanities scholarship.
+The outlook for the future of archaeology is also complex, with three out of five of the bibliometric variables trending towards more humanistic styles of working, but the discipline showing more extreme values in some metrics towards both hard and soft sciences after about 2010.
+Among archaeology journals, we see distinct communities of practice
 

### 2025-01-23 paper.qmd
pivoting to repro

@@
   journal_metrics_for_plotting %>% 
+  mutate(journal = str_wrap(journal, 20)) %>% 
   group_by(journal) %>% 
@@
 
-library(FactoMineR)
-library(factoextra)
-
+# PCA of journal means
 journal_metrics_for_plotting_summary_pca <- 
@@
   column_to_rownames("journal") %>% 
-  PCA(graph = FALSE)
+  prcomp(scale = TRUE)
 
-  fviz_pca_biplot(journal_metrics_for_plotting_summary_pca,
-                  label = "var", repel = TRUE) + 
-  geom_text_repel(aes(label = str_wrap(journal_metrics_for_plotting_summary$journal, 20)), 
+# Tidy the PCA results
+pca_means_tidy <- journal_metrics_for_plotting_summary_pca %>% tidy(matrix = "pcs")
+# first two PCs explain how much?
+# Get the summary of the PCA
+pca_summary <- summary(journal_metrics_for_plotting_summary_pca)
+# Extract the proportion of variance explained by PC1 and PC2
+variance_explained <- round(pca_summary$importance[2, 1:2] * 100, 0)
+
+# Get the PCA scores
+pca_scores_means <- journal_metrics_for_plotting_summary_pca %>%  augment(journal_metrics_for_plotting_summary)
+
+# Get the PCA loadings
+pca_loadings_means <- 
+  journal_metrics_for_plotting_summary_pca %>% 
+  tidy(matrix = "rotation") %>% 
+    pivot_wider(names_from = "PC", 
+              values_from = "value",
+              names_prefix = "PC") %>% 
+  mutate(column = case_when(
+    column == "mean_log_authors" ~ "Number of\nauthors",
+    column == "mean_log_pages" ~ "Number of\npages",
+    column == "mean_relative_title_length" ~ "Relative\ntitle\nlength",
+    column == "mean_prices_index" ~ "Recency of\nreferences",
+    column == "mean_shannon" ~ "Diversity of\nreferences",
+  ))
+
+# Plot the PCA results
+plot_pca_means <- 
+ggplot() + 
+  labs(x = paste0("PC1 (", variance_explained[1], "%)"),
+       y = paste0("PC2 (", variance_explained[2], "%)")) +
+  geom_point(data = pca_scores_means,
+               aes(.fittedPC1,
+                   .fittedPC2),
              size = 2) +
-    ggtitle("")
+  geom_text_repel(data = pca_scores_means,
+            aes(.fittedPC1,
+                .fittedPC2,
+                label = journal),
+            lineheight = 0.8,
+            size = 2,
+            bg.color = "white",  # Color of the halo
+            bg.r = 0.2,) +
+     geom_segment(data = pca_loadings_means, 
+               aes(x = 0, 
+                   y = 0, 
+                   xend = PC1, 
+                   yend = PC2),
+               arrow = arrow(length = unit(0.2, "cm")), 
+               color = "grey70") +
+  geom_text_repel(data = pca_loadings_means,
+            aes(PC1,
+                PC2,
+                label = column),
+            size = 2,
+            lineheight = 0.7,
+            force = 0,
+            segment.color = NA,
+            color = "grey40",
+           bg.color = "white",  # Color of the halo
+          bg.r = 0.2) +
+  theme_minimal()
 
@@
 #| label: fig-variation-by-journal
-#| fig-cap: Variation in bibliometric indicators of hardness for 20 archaeological journals.
+#| fig-cap: "Panels A-E: Variation in bibliometric indicators of hardness for 20 archaeological journals.
 The journals are ordered for each indicator so that within each plot, the harder journals are at the top of the plot and the softer journals are at the base.
-The lower right panel shows a bar plot that is the single consensus ranking computed from all five variables, using the Borda Count ranking algorithm.
-#| fig-height: 12
+Panel F shows a bar plot that is the single consensus ranking computed from all five variables, using the Borda Count ranking algorithm.
+Panel G is a biplot of the first and second principal components of a PCA computed on the journal means of the five bibliometric variables."
+#| fig-height: 14
 
@@
 
+plot_grid(
   plot_grid(plot_journals_authors, 
@@
            nrow = 2,
-          labels = "AUTO")
+           labels = LETTERS[1:6]),
+  plot_pca_means,
+  nrow = 2,
+  labels = c("", LETTERS[7]),
+  rel_heights = c(1, 0.5))
 
@@
 
+The PCA results in panel G show of @fig-variation-by-journal that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left.
+The distribution of PC1 values is skewed left, with most of the journals concentrated at the harder end of the spectrum.
+The distribution of PC2 values reveals additional structure to the data and can be roughly separated into generalist journals in the negative range (e.g.
+_American Antiquity_, _Antiquity_, _Advances in Archaeological Practice_), and more specialised journals in the positive range (e.g.
+_Environmental Archaeology_, _Geoarchaeology_, _Archaeologial Research in Asia_, _Journal of Island and Coastal Archaeology_).
+The _Journal of Archaeological Science_ sits about midway between these two groups, reflecting its relevance to specialised communities of practice in archaeology, as well as general
+
 ## Dicsussion
 
-This macroscopic perspective on the way thousands of archaeologists communicate their research has produced a complex picture of archaeology as a science.
+This macroscopic perspective derived from an analysis of the ways thousands of archaeologists communicate their research has produced a complex picture of archaeology as a science.
 In the context of a broad spectrum of other research areas, archaeologists behave like social scientists.
@@
 The outlook for the future of archaeology is also complex, with three out of five of the bibliometric variables trending towards more humanistic styles of working, but the discipline showing more extreme values in some metrics towards both hard and soft sciences after about 2010.
-Among archaeology journals, we see distinct communities of practice
+Among archaeology journals, we see distinct communities of practice reflected in the PCA results that are very close together on the hard-soft spectrum, but have minor differences in their communication styles, perhaps due to differences in writing traditions inherited from parent disciplines such as geology and biology.
+
+While these bibliometric variables provide several interesting insights into the status of archaeology as a science, via measurement of consensus, and are important for moving the debate beyond discussions of a small number of case studies, they miss a crucial factor that separates scientific practice from non-science.
+This is reproducibility, which, according to a report by the US National Science Foundation [@cacioppo2015social], "refers to the ability of a researcher to duplicate the results of a prior study using the same materials as were used by the original investigator.
+That is, a second researcher might use the same raw data to build the same analysis files and implement the same statistical analysis in an attempt to yield the same results...
+Reproducibility is a minimum necessary condition for a finding to be believable and informative." The importance of this factor can be traced to Irish chemist Robert Boyle (1627-1691), best known for his experiments with vacuum pumps [@shapin2011leviathan].
+Boyle was concerned about the secrecy common among experimentalists in the 17th century and aimed to shift the culture from valuing direct in-person witnessing of scientific demonstrations towards meticulous written communications that were detailed enough to enable a reader to successfully undertake the experiment themselves, independent of the original author.
 

### 2025-01-28 paper.qmd
typing

@@
 This may be due to some broader changes in academic publishing around this time, such as moves to digital-first continuous publishing, new journals appearing (e.g.
-_Archaeological and Anthropological Sciences_ in 2009 and _Journal of Island & Coastal Archaeology_ in 2010), and non-archaeology journals becoming more relevant to archaeologists.
+*Archaeological and Anthropological Sciences* in 2009 and *Journal of Island & Coastal Archaeology* in 2010), and non-archaeology journals becoming more relevant to archaeologists.
 For example, *PLOS ONE* received its first impact factor in 2010 and in 2011 Nature's *Scientific Reports* began publishing [@malashichev2017open].
@@
 The distribution of PC2 values reveals additional structure to the data and can be roughly separated into generalist journals in the negative range (e.g.
-_American Antiquity_, _Antiquity_, _Advances in Archaeological Practice_), and more specialised journals in the positive range (e.g.
-_Environmental Archaeology_, _Geoarchaeology_, _Archaeologial Research in Asia_, _Journal of Island and Coastal Archaeology_).
-The _Journal of Archaeological Science_ sits about midway between these two groups, reflecting its relevance to specialised communities of practice in archaeology, as well as general
+*American Antiquity*, *Antiquity*, *Advances in Archaeological Practice*), and more specialised journals in the positive range (e.g.
+*Environmental Archaeology*, *Geoarchaeology*, *Archaeologial Research in Asia*, *Journal of Island and Coastal Archaeology*).
+The *Journal of Archaeological Science* sits about midway between these two groups, reflecting its relevance to specialised communities of practice in archaeology, as well as general
 
-## Dicsussion
+## Discussion
 
@@
 Boyle was concerned about the secrecy common among experimentalists in the 17th century and aimed to shift the culture from valuing direct in-person witnessing of scientific demonstrations towards meticulous written communications that were detailed enough to enable a reader to successfully undertake the experiment themselves, independent of the original author.
+
+With many disciplines making increasing use of computationally intensive analyses in recent years there has been renewed interest in reproducibility [@levequeReproducibleResearchScientific2012].
+In part, this is because computationally intensive research is difficult to communicate within the constraints of the methods section of a traditional journal article — the reader needs also needs the computer code written by the original authors, not just the article text.
+There is also the broader context of rising pressure to publish in prestigious journals and intense competition for funds that create strong incentives for malpractice in research [@edwards2017academic].
+These two factors have led to widespread concerns of a reproducibility crisis in many fields [@Baker2016].
+Estimates of the extent of this problem come from both empirical and computational analyses.
+Empirical replications of 100 studies published in three psychology journals found that 36% of replications had statistically significant results, compared to 96% of the original studies [@OpenScienceCollaboration2015].
+Similar empirical replications of social science studies and experiemental economics studies successfully replicated 61% and 62% of their target studies respectively [@camererEvaluatingReplicabilitySocial2018; @camererEvaluatingReplicabilityLaboratory2016].
+An attempt at reproducing the computational results of 204 papers in *Science* resulted in success for reproduce the findings for 26% [@stoddenEmpiricalAnalysisJournal2018].
 

### 2025-01-29 paper.qmd
writing AER report section

@@
 These two factors have led to widespread concerns of a reproducibility crisis in many fields [@Baker2016].
-Estimates of the extent of this problem come from both empirical and computational analyses.
+Estimates of scientific reproducibility in several fields confirm the extent of this problem.
 Empirical replications of 100 studies published in three psychology journals found that 36% of replications had statistically significant results, compared to 96% of the original studies [@OpenScienceCollaboration2015].
-Similar empirical replications of social science studies and experiemental economics studies successfully replicated 61% and 62% of their target studies respectively [@camererEvaluatingReplicabilitySocial2018; @camererEvaluatingReplicabilityLaboratory2016].
-An attempt at reproducing the computational results of 204 papers in *Science* resulted in success for reproduce the findings for 26% [@stoddenEmpiricalAnalysisJournal2018].
+Similar empirical replications of large numbers of social science studies and experiemental economics studies successfully replicated 61% and 62% of their target studies respectively [@camererEvaluatingReplicabilitySocial2018; @camererEvaluatingReplicabilityLaboratory2016].
+
+Similarly bleak results come from measurements specifically of the reproducibility of computational analyses of scientific studies.
+An attempt at reproducing the computational results of 204 papers in *Science* resulted in success in reproducing the findings for 26% [@stoddenEmpiricalAnalysisJournal2018].
+The computational results of two out of 41 geoscience papers could be fully reproduced on the first attempt [@konkolComputationalReproducibilityGeoscientific2019].
+In the biomedical field, code in 1,203 out of 27,271 (4%) notebooks associated with 3,467 publications could be run without errors [@samuelComputationalReproducibilityJupyter2024].
+Economists have been especially active in researching computational reproducibility, with studies indicating successful reproduction of results using code and data provided by authors for 30% of 67 papers [@ChangLi2015], 14% of 203 papers [@gertler2018make], 44% of 152 papers [@herbert2021reproducibility], 30% of 419 articles [@fivsar2024reproducibility], and 28% of 168 papers [@perignonComputationalReproducibilityFinance2024].
+These efforts confirm that the reproducibility of published research is widely recognised as a cornerstone of rigorous science, and work on evaluating how successful a research community is at generating reproducible results has become a distinctive and important meta-research activity in many fields.
+
+How does archaeology compare to these other fields in terms of reproducibility?
+Empirical reproducibility has long been valued in field archaeology.
+Throughout the history of archaeology, well-known sites have been repeatedly revisited to test old hypotheses with new evidence or methods, for example, Olduvai Gorge (Tanzania), Cahokia (USA), Çatalhöyük (Turkey), and Madjedbebe (Australia).
+Similarly among many experimental archaeologists, empirical reproducibility is a key concern, for example in lithic use-wear identification [@hayesLearningBlindTests2017] and the measurement of lithics [@pargeterReplicabilityLithicAnalysis2023].
+To what extent does this concern extend to computational reproducibility among archaeologists?
+
+## Investigating computational reproducibility in archaeology
+
+In 2024 the *Journal of Archaeological Science* introduced a new kind of peer review that has provided an opportunity to investigate computational reproducibility in archaeology.
+In January 2024 I accepted the position of 'Associate Editor for Reproducibility' and conducted peer reviews of submissions that mentioned programming languages such as R or Python in the methods sections.
+
+it is hard to assess whether reported findings are credible.
+
+While not yet the norm in archaeology, the systemic verification of the computational reproducibility of the results of conditionally accepted papers is now required in many leading academic journals including the *American Economic Review*, *American Economic Journal*, *Econometrica*, *Review of Economic Studies*, *Economic Journal*, *Journal of the American Statistical Association*, *American Journal of Political Science*, and the *Journal of Politics*.
+
+Computationally reproducible research can still be plagued by various types of errors, ranging from honest mistakes (e.g., typos in the code) to outright fraud (e.g., data fabrication).
+A practical hurdle in ascertaining computational reproducibility for the published literature is that the verification is resource-intensive
 

### 2025-01-30 paper.qmd
reporting on AER work

@@
 
-In 2024 the *Journal of Archaeological Science* introduced a new kind of peer review that has provided an opportunity to investigate computational reproducibility in archaeology.
-In January 2024 I accepted the position of 'Associate Editor for Reproducibility' and conducted peer reviews of submissions that mentioned programming languages such as R or Python in the methods sections.
+In 2024 the *Journal of Archaeological Science* introduced a new kind of peer review that has provided an opportunity to tackle this question about computational reproducibility in archaeology.
+In January 2024 I accepted the position of 'Associate Editor for Reproducibility' (AER) for JAS and conducted reproducibility reviews of submissions that mentioned programming languages such as R or Python in the methods sections.
+A reproducibility review examines the code and data used to generate the results presented in the paper, and attempts to run the authors' code to reproduce the their results.
+This new AER role is based on similar positions (i.e.
+'data editor' or 'reproducibility editor') that journals in economics [@vilhuber2019report], statistics [@wrobel2024partnering], astronomy [@muench2023roles], ecology [@bolnick2022ensuring], and environmental studies [@rosenbergReproducibleResultsPolicy2021] have had, in some cases for over a decade.
+In 2024, three archaeology journals, in addition to JAS, added AERs to their editorial communities: *Advances in Archaeological Practice* [@marwick2024introducing, one paper reviewed], *Journal of Field Archaeology* [@farahani2024reproducibility, two papers reviewed], and *American Antiquity* [@martin2024editors, no papers reviewed].
+
+```{r}
+# reporting on my AER work
+jas_rr_data <- read_csv("data/JAS AER data analysis.csv")
+base_size <- 10
+
+# what software did the authors use?
+aer_plot_software <- 
+jas_rr_data %>% 
+  separate_longer_delim(Language, ",") %>% 
+  mutate(Language = str_squish(Language)) %>% 
+  count(Language) %>% 
+  drop_na() %>% 
+  mutate(Language = fct_reorder(Language, n, .desc = FALSE)) %>% 
+  ggplot() +
+  aes(Language, n) +
+  geom_bar(stat = 'identity') +
+  coord_flip() +
+  xlab("") +
+  ylab("Number of papers reviewed") +
+  theme_minimal(base_size = base_size)
+
+# what methods did the authors use? ML includes CNN, VAR, DL
+aer_plot_methods <- 
+jas_rr_data %>% 
+  separate_longer_delim(Methods2, ",") %>% 
+  mutate(Methods2 = str_squish(Methods2)) %>% 
+  count(Methods2) %>% 
+  drop_na() %>% 
+  mutate(Methods = fct_reorder(Methods2, n, .desc = FALSE)) %>% 
+ggplot() +
+  aes(Methods, n) +
+  geom_bar(stat = 'identity') +
+  coord_flip() +
+  xlab("") +
+  ylab("Number of papers reviewed") +
+  theme_minimal(base_size = base_size)
+
+# where did they put their materials?
+aer_plot_repo <- 
+jas_rr_data %>% 
+  separate_longer_delim(Repo, ",") %>% 
+  mutate(Repo = str_squish(Repo)) %>% 
+  count(Repo) %>% 
+  drop_na() %>% 
+  mutate(Repo = fct_reorder(Repo, n, .desc = FALSE)) %>% 
+  ggplot() +
+  aes(Repo, n) +
+  geom_bar(stat = 'identity') +
+  coord_flip() +
+  xlab("") +
+  ylab("Number of papers reviewed") +
+  theme_minimal(base_size = base_size) +
+  scale_y_continuous(breaks = scales::breaks_pretty())
+
+# could reproduce the results on the first try?
+aer_first_try <- 
+jas_rr_data %>% 
+  count(`Run on first try?`) %>% 
+  drop_na() %>% 
+  mutate(`Run on first try?` = fct_reorder(`Run on first try?`, n, .desc = FALSE))
+
+# why not? issues with code
+aer_plot_issues <- 
+jas_rr_data %>% 
+  separate_longer_delim(`Code issue`, ",") %>% 
+  mutate(`Code issue` = str_squish(`Code issue`)) %>% 
+    mutate(`Code issue` = ifelse(`Code issue` == "Proprietary software", 
+                               "Proprietary\nsoftware",
+                               `Code issue`)) %>% 
+  count(`Code issue`) %>% 
+  drop_na() %>% 
+  mutate(`Code issue` = fct_reorder(`Code issue`, n, .desc = FALSE)) %>% 
+  ggplot() +
+  aes(`Code issue`, n) +
+  geom_bar(stat = 'identity') +
+  coord_flip() +
+  xlab("") +
+  ylab("Number of papers reviewed") +
+  theme_minimal(base_size = base_size)  +
+  scale_y_continuous(breaks = scales::breaks_pretty())
+
+# relationship between code issues and language?
+aer_plot_issues_lang <- 
+jas_rr_data %>% 
+  separate_longer_delim(`Code issue`, ",") %>% 
+  separate_longer_delim(Language, ",") %>% 
+  mutate(Language = str_squish(Language)) %>% 
+  mutate(`Code issue` = str_squish(`Code issue`)) %>% 
+  group_by(`Code issue`, 
+           Language) %>% 
+  tally() %>% 
+  drop_na() %>% 
+  ggplot() +
+  aes(`Code issue`, 
+      Language,
+      size = n) +
+  geom_point() +
+  xlab("Problem with the code") +
+  ylab("Software") +
+  scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
+  theme_minimal(base_size = base_size) +
+  scale_size_area(breaks = c(1, 2, 3, 5)) +
+  theme(legend.position = c(.9,.7),
+  legend.box.background=element_rect(fill="white", 
+                                     color="black")) 
+```
+
+```{r}
+#| label: fig-aer-summary
+#| fig-cap: "Summary of reproducibility reviews for JAS.
+A: Primary software used for the computational analysis reported in a manuscript.
+B: Computational or statistical method used by the authors (GMM = geometric morphometrics, Frequentist = hypothesis tests such as chi-square and ANOVA, AI/ML = artifical intellegence and machine learning, including neural networks and deep learning, Network = statitical analysis of social networks, 3D = analysis of 3D data such as artefact models, Composition = compositional analysis of artefacts).
+C: Locations where authors deposited their code and data files.
+D: Issues that prevented the reproducible review from succeeding on the first attempt.
+E: Relationship between software used and issues that make research irreproducible."
+
+library(patchwork)
+(aer_plot_software + aer_plot_methods + aer_plot_repo) / (aer_plot_issues + aer_plot_issues_lang) + plot_annotation(tag_levels = 'A')
+
+```
+
+At the time of writing (Feburary 2025) I have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS.
+Of these, 11 have been published in JAS so far.
+Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate so of 63%.
+Four of the seven papers could be fully reproduced on my first attempt, the others required additional input from the authors.
 

### 2025-01-30 paper.qmd
recommendations

@@
 
-## Methods and materials
+## How to measure the hardness or softness of a science?
 
@@
 
-## Discussion
+## Reproducibility: A key measure of how scientific a field is
 
@@
 In the biomedical field, code in 1,203 out of 27,271 (4%) notebooks associated with 3,467 publications could be run without errors [@samuelComputationalReproducibilityJupyter2024].
+Statisticians could reproduce 15% of 93 papers [@xiongStatePlayReproducibility2023].
 Economists have been especially active in researching computational reproducibility, with studies indicating successful reproduction of results using code and data provided by authors for 30% of 67 papers [@ChangLi2015], 14% of 203 papers [@gertler2018make], 44% of 152 papers [@herbert2021reproducibility], 30% of 419 articles [@fivsar2024reproducibility], and 28% of 168 papers [@perignonComputationalReproducibilityFinance2024].
@@
 
-## Investigating computational reproducibility in archaeology
+## How reproducible is archaeology? Investigating computational reproducibility
 
@@
 In January 2024 I accepted the position of 'Associate Editor for Reproducibility' (AER) for JAS and conducted reproducibility reviews of submissions that mentioned programming languages such as R or Python in the methods sections.
-A reproducibility review examines the code and data used to generate the results presented in the paper, and attempts to run the authors' code to reproduce the their results.
+A reproducibility review examines the code and data used to generate the results presented in the paper, and attempts to run the authors' code to reproduce the their results (see @editorsReproducibilityJournalArchaeological for more details about this process).
 This new AER role is based on similar positions (i.e.
@@
 At the time of writing (Feburary 2025) I have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS.
-Of these, 11 have been published in JAS so far.
-Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate so of 63%.
+Of these, 11 have been published in JAS to date.
+Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate of 63%.
 Four of the seven papers could be fully reproduced on my first attempt, the others required additional input from the authors.
-
+For comparison with reproducibility studies in other fields reported above, the seven fully reproducible papers should be divided by the 25 reviewed for reproducibility, resulting in a 28% success rate.
+Expanding the denominator to include the total number of research articles published in JAS from May 2024 (when the first article to pass the reproducibility review, @herskindComputationalLinguisticMethodology2024, was published) to January 2025 (n = 97), which are articles that could have been eligible for reproducibility review had the authors used an open source programming language (e.g.
+instead of commercial software such as Microsoft Excel or SPSS, etc.) the success rate is 7%.
+In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values obtained from a variety of harder and soft sciences.
+
+@fig-aer-summary shows a summary of basic characteristics of the 25 articles that have been through to the reproducibility review so far.
+The most commonly used software is R, followed by Python.
+Results generated with proprietary or closed-source software are out of scope for reproducibility reviews.
+Several distinct types of analyses are well-represented in this sample, especially geometric morphometry, network statistics, and analyses using artificial intelligence or machine learning algorithms (this includes deep learning and neural networks).
+Most authors are sharing their code and data files via Zenodo, a non-profit generic research data repository hosted by CERN that accepts any file format and freely assigns all publicly available uploads a DOI to make the files easily and uniquely citable [@petersZenodoSpotlightTraditional2017].
+In this same category of DOI-issuing, research-grade repositories is OSF (the Open Science Foundation), Figshare, and university repositories.
+GitHub, a commercial service owned by Microsoft is a code hosting platform that is convenient for collaboration, is also popular among JAS authors, but is a problematic choice because does not offer DOIs or the same commitments to long-term availability of Zenodo.
+Some authors attached their code and data as journal article supplementary files, but this is a poor choice for long-term availability because these files are typically renamed and converted to different formats during the article production process, making it difficult or impossible for a reader to combine the code and data to reproduce the results.
+
+Panels D and E of @fig-aer-summary summarise the common issues that resulted in irreproducible results.
+The most common issue was an incomplete compendium.
+This ranges from missing data files down to missing lines of code, and in most cases seems to be due to carelessness.
+Unspecified or under-specified dependencies is another common issue that prevents code from running.
+This refers to the software packages in addition to R or Python that an author used to do specialized analyses and visualisations, (e.g.
+dplyr for R or numpy for Python).
+If an author does not clearly specify the name and version number of the packages that they used for their analysis, it can be very time-consuming or impossible to correctly identify these because many packages have functions with similar names, and functions in any one package can change they way they behave as the developers update their package.
+Other reasons why papers failed the reproducibility review is that the paths to data files were incorrectly specified (likely a result of the author reorganising their compendium after completing their analysis, or omitting data files from the materials submitted for review), and errors returned by functions, which have diverse causes.
+
+Poor communication about dependencies is widespread in code written by researchers in many disciplines, and a common source of irreproducible results [@trisovicLargescaleStudyResearch2022; @samuelComputationalReproducibilityJupyter2024].
+In our sample issues relating to dependencies are strongly associated with the use of Python.
+One possible reason for this is that relative to R, Python uses more package managers, more environments, and deeper dependency chains with more complex inter-dependencies that change more rapidly [@decan2016topology; @decanEmpiricalComparisonDependency2019; @korkmazModelingImpactPython2020].
+Another reason may be that there is a bigger and more established community of R users in archaeology [@schmidtToolDrivenRevolutionsArchaeological2020; @batistOpenArchaeologyOpen2024] that highly values code that is easy for others to reuse [e.g.
+@bilottiPointPatternAnalysis2024; @willExploringUtilityUnretouched2025].
+
+## How to improve the computational reproducibility of archaeology?
+
+Despite the small number of reproducibility reviews reported on here, there are patterns of common issues that point to efforts that have high potential to increase reproducibility.
+The problem of incomplete materials can be tackled in several ways.
+First, authors should use a clear and logical folder structure for their code and data.
+There are many simple templates to guide authors on this, e.g.
+@doi:10.1080/00031305.2017.1375986, @figueiredoSimpleKitUse2022, \@Cookiecutter, @cooperGuideReproducibleCode2017 and @wilsonGoodEnoughPractices2017 Second, for analyses that are not highly time-consuming, authors should re-run their code more than once, and ideally on more than one machine, before submission to confirm everything works as expected.
+Complex and time-consuming analyses should use pipeline or workflow management tools, e.g.
+GNU Make, Luigi, Snakemake, or Targets, to document the relationship of the files and folders in a machine-readable format and simplify running and re-running code [@wrattenReproducibleScalableShareable2021; @landau2021targets].
+Third, authors should include in their compendium a README document that describes to a reader the folders and files contained in the project.
+The README file is typically the first file that a reader will look at in a compendium so it should include brief instructions to guide the user to a successful reproduction of the original results (e.g.
+what order to run the code files in).
+Fourth, authors should ensure clear, direct and obvious connections between their code and the results they present in their paper.
+A simple way to do this is to have one code file for each figure and table, and name the code files with the figure or table number and some key words in the caption.
+Another way some authors are accomplishing this is by using literate programming tools, such as Quarto and Jupyter notebooks [@quarto; @kluyver2016jupyter] that enable the research narrative and code for data analysis to be woven together in one document.
+Archaeologists are using these tools write complete journal articles, e.g.
+@vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024.
+
+Dependencies
+
+```{=html}
+<!--
+
+so efforts aimed at systemic improvements of dependency handling—for example, as per [134]—have great potential to increase reproducibility....
 it is hard to assess whether reported findings are credible.
 
-While not yet the norm in archaeology, the systemic verification of the computational reproducibility of the results of conditionally accepted papers is now required in many leading academic journals including the *American Economic Review*, *American Economic Journal*, *Econometrica*, *Review of Economic Studies*, *Economic Journal*, *Journal of the American Statistical Association*, *American Journal of Political Science*, and the *Journal of Politics*.
+While not yet the norm in archaeology, the systemic verification of the computational reproducibility of the results of conditionally accepted papers is now required in many leading academic journals including the American Economic Review, American Economic Journal, Econometrica, Review of Economic Studies, Economic Journal, Journal of the American Statistical Association, American Journal of Political Science, and the Journal of Politics.
 
@@
 
-```{=html}
-<!-- n an ideal science, scholars share a common background of established theories, facts and methods.
+n an ideal science, scholars share a common background of established theories, facts and methods.
 This allows them to agree (usually after debate and further evidence) on the validity and significance of a new research finding, making it the basis for further theorizing and research.

### 2025-01-30 paper.qmd
up to the conclusion

@@
 Of these, 11 have been published in JAS to date.
-Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate of 63%.
+Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate, by a generous measure, of 63%.
 Four of the seven papers could be fully reproduced on my first attempt, the others required additional input from the authors.
@@
 
-Poor communication about dependencies is widespread in code written by researchers in many disciplines, and a common source of irreproducible results [@trisovicLargescaleStudyResearch2022; @samuelComputationalReproducibilityJupyter2024].
-In our sample issues relating to dependencies are strongly associated with the use of Python.
-One possible reason for this is that relative to R, Python uses more package managers, more environments, and deeper dependency chains with more complex inter-dependencies that change more rapidly [@decan2016topology; @decanEmpiricalComparisonDependency2019; @korkmazModelingImpactPython2020].
-Another reason may be that there is a bigger and more established community of R users in archaeology [@schmidtToolDrivenRevolutionsArchaeological2020; @batistOpenArchaeologyOpen2024] that highly values code that is easy for others to reuse [e.g.
-@bilottiPointPatternAnalysis2024; @willExploringUtilityUnretouched2025].
-
 ## How to improve the computational reproducibility of archaeology?
 
-Despite the small number of reproducibility reviews reported on here, there are patterns of common issues that point to efforts that have high potential to increase reproducibility.
-The problem of incomplete materials can be tackled in several ways.
+Despite the small number of reproducibility reviews reported on here, there are patterns of common issues that point to a small set of simple tasks that have high potential to increase reproducibility.
+The problem of incomplete materials can be tackled in several basic ways.
 First, authors should use a clear and logical folder structure for their code and data.
 There are many simple templates to guide authors on this, e.g.
-@doi:10.1080/00031305.2017.1375986, @figueiredoSimpleKitUse2022, \@Cookiecutter, @cooperGuideReproducibleCode2017 and @wilsonGoodEnoughPractices2017 Second, for analyses that are not highly time-consuming, authors should re-run their code more than once, and ideally on more than one machine, before submission to confirm everything works as expected.
+@doi:10.1080/00031305.2017.1375986, @figueiredoSimpleKitUse2022, @Cookiecutter, @cooperGuideReproducibleCode2017 and @wilsonGoodEnoughPractices2017 Second, for analyses that are not highly time-consuming, authors should re-run their code more than once, and ideally on more than one machine (e.g.
+by another co-author of the paper), before submission to confirm everything works as expected.
+This would likely also prevent issues relating to path and function errors.
 Complex and time-consuming analyses should use pipeline or workflow management tools, e.g.
@@
 Another way some authors are accomplishing this is by using literate programming tools, such as Quarto and Jupyter notebooks [@quarto; @kluyver2016jupyter] that enable the research narrative and code for data analysis to be woven together in one document.
-Archaeologists are using these tools write complete journal articles, e.g.
-@vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024.
-
-Dependencies
+Quarto was a popular tool among the JAS papers in the reproducibility review sample, for example, @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024 wrote their entire manuscript using Quarto.
+These four simple steps have high potential to substantially increase the computational reproducibility of archaeological science.
+
+Our finding that dependencies are a common cause of irreproducible results is consistent with previous studies that have identified this as a widespread weakness in communicating computationally intensive research [@trisovicLargescaleStudyResearch2022; @samuelComputationalReproducibilityJupyter2024].
+In our sample, issues relating to dependencies are strongly associated with the use of Python.
+One possible reason for this is that relative to R, Python uses more package managers, more environments, and deeper dependency chains with more complex inter-dependencies that change more rapidly [@decan2016topology; @decanEmpiricalComparisonDependency2019; @korkmazModelingImpactPython2020].
+Another reason may be that there is a bigger and more established community of R users in archaeology [@schmidtToolDrivenRevolutionsArchaeological2020; @batistOpenArchaeologyOpen2024] that highly values code that is easy for others to reuse and has evolved practices to effectively communicate dependencies [e.g.
+@bilottiPointPatternAnalysis2024; @willExploringUtilityUnretouched2025].
+The simplest way for archaeologists, to improve here is to write the names and version numbers of the software and packages they used in their README file, as we see in @herskindComputationalLinguisticMethodology2024 and @monnaStudyingSerialityMaterial2024.
+For more complex research projects, i.e.
+those using five or more packages or machine learning algorithms, authors should use dependency management tools to keep track of the packages and version numbers needed to reproduce their results.
+This is an active area of development, and while there are many tools currently available, the most robust and widely used include renv for R [@renv, see examples in @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024] and conda and poetry for Python [@conda2023; @poetry2023].
+A more comprehensive solution, and the leading best practice for managing dependencies in many computationally intensive fields using Python in particular, is to include a Dockerfile in the compendium [@moreauContainersComputationalReproducibility2023].
+This is a set of a machine- and human-readable instructions that enables a user to recreate the author's computational environment (including those requirements beyond the R or Python packages) on another computer [@nustTenSimpleRules2020].
+Dockerfiles are gradually being adopted by archaeologists, see @cremaModellingDiffusionInnovation2024 and @liaoApplyingMaskRCNN2024 for examples, and most of my reproducibility reviews include a recommendation that the author include a Dockerfile to manage dependencies efficiently.
+
+## Conclusion
 
@@
 
-so efforts aimed at systemic improvements of dependency handling—for example, as per [134]—have great potential to increase reproducibility....
-it is hard to assess whether reported findings are credible.
+limitations: not looked at FAIR or CARE attributes of data, not looked at code quality (efficiency, readability, style)
 

### 2025-01-31 paper.qmd
minor edits

@@
 Of these, 11 have been published in JAS to date.
-Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate, by a generous measure, of 63%.
+Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate, by one measure, of 63%.
 Four of the seven papers could be fully reproduced on my first attempt, the others required additional input from the authors.
@@
 
-@fig-aer-summary shows a summary of basic characteristics of the 25 articles that have been through to the reproducibility review so far.
+@fig-aer-summary shows a summary of basic characteristics of the 25 articles that have been through the reproducibility review process so far.
 The most commonly used software is R, followed by Python.
@@
 In this same category of DOI-issuing, research-grade repositories is OSF (the Open Science Foundation), Figshare, and university repositories.
-GitHub, a commercial service owned by Microsoft is a code hosting platform that is convenient for collaboration, is also popular among JAS authors, but is a problematic choice because does not offer DOIs or the same commitments to long-term availability of Zenodo.
+GitHub, a commercial service owned by Microsoft is a code hosting platform that is convenient for collaboration, is also popular among JAS authors, but is a problematic choice because does not offer DOIs or the same commitments to long-term availability as Zenodo.
 Some authors attached their code and data as journal article supplementary files, but this is a poor choice for long-term availability because these files are typically renamed and converted to different formats during the article production process, making it difficult or impossible for a reader to combine the code and data to reproduce the results.
@@
 The most common issue was an incomplete compendium.
-This ranges from missing data files down to missing lines of code, and in most cases seems to be due to carelessness.
+This ranges from missing data files down to missing lines of code.
+In most cases this can attributed to accidental carelessness, with the exception of two cases where data was unavailable due to licensing restrictions.
 Unspecified or under-specified dependencies is another common issue that prevents code from running.

### 2025-02-03 paper.qmd
edits

--- 2025-01-30 paper.qmd
@@
 ---
-title: "Extracting sunbeams from cucumbers: Reproducibility as a necessity for the future of archaeological science"
+title: "Extracting sunbeams from cucumbers: Computational reproducibility is essential for archaeological science"
 format: html
@@
 As part of the celebration of the 50th anniversary, and of Torrence's leadership of the journal, my contribution revisits these topics of archaeology's status as a science, this journal's place in the landscape of archaeological science, and how the journal has responded to a growing recognition of the importance of reproducibility.
-I first present bibliometric evidence of the position of archaeology as a whole, and this journal in particular, in the hierarchy of the sciences.
+I first present bibliometric evidence of the position of archaeology as a whole, and this journal in particular, in the sciences.
 Next I report on the journal's progress in supporting reproducible research, and my work doing a new kind of peer review for JAS, one that evaluates the computational reproducibilty of the research submitted for publication.
+Finally, I analyse twelve months of reproducibility reviews to identify common weaknesses in the ways archaeologists are working currently, and provide simple recommendations for researchers to overcome these and contribute to the improvement computational reproducibility in archaeological science.
 
@@
 For comparison with reproducibility studies in other fields reported above, the seven fully reproducible papers should be divided by the 25 reviewed for reproducibility, resulting in a 28% success rate.
-Expanding the denominator to include the total number of research articles published in JAS from May 2024 (when the first article to pass the reproducibility review, @herskindComputationalLinguisticMethodology2024, was published) to January 2025 (n = 97), which are articles that could have been eligible for reproducibility review had the authors used an open source programming language (e.g.
-instead of commercial software such as Microsoft Excel or SPSS, etc.) the success rate is 7%.
+Expanding the denominator to include the total number of research articles published in JAS from May 2024 (when the first article to pass the reproducibility review, @herskindComputationalLinguisticMethodology2024, was published) to January 2025 (n = 97), the success rate is 7%.
+In my opinion this is the proper denominator because these are articles that report quantitative analyses, and all quantitative analyses should be done using an open source programming language to provide transparency (those authors used commercial software such as Microsoft Excel or SPSS, etc.).
 In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values obtained from a variety of harder and soft sciences.
@@
 
+While consensus is one measure of the hardness of a science, the computability
+
 While not yet the norm in archaeology, the systemic verification of the computational reproducibility of the results of conditionally accepted papers is now required in many leading academic journals including the American Economic Review, American Economic Journal, Econometrica, Review of Economic Studies, Economic Journal, Journal of the American Statistical Association, American Journal of Political Science, and the Journal of Politics.

### 2025-02-03 paper.qmd

--- 2025-01-31 paper.qmd
@@
 
-At the time of writing (Feburary 2025) I have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS.
+At the time of writing (January 2025) I have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS.
 Of these, 11 have been published in JAS to date.
@@
 instead of commercial software such as Microsoft Excel or SPSS, etc.) the success rate is 7%.
-In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values obtained from a variety of harder and soft sciences.
+In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values obtained from a variety of hard and soft sciences.
 

### 2025-02-05 paper.qmd
writing conclusion

@@
 
-At the time of writing (January 2025) I have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS.
+At the time of writing (January 2025) we have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS.
 Of these, 11 have been published in JAS to date.
@@
 
-Despite the small number of reproducibility reviews reported on here, there are patterns of common issues that point to a small set of simple tasks that have high potential to increase reproducibility.
+Despite the relatively small number of reproducibility reviews reported on here, there are patterns of common issues that point to a small set of simple tasks authors can do that have high potential to increase reproducibility.
 The problem of incomplete materials can be tackled in several basic ways.
 First, authors should use a clear and logical folder structure for their code and data.
-There are many simple templates to guide authors on this, e.g.
-@doi:10.1080/00031305.2017.1375986, @figueiredoSimpleKitUse2022, @Cookiecutter, @cooperGuideReproducibleCode2017 and @wilsonGoodEnoughPractices2017 Second, for analyses that are not highly time-consuming, authors should re-run their code more than once, and ideally on more than one machine (e.g.
+There are many excellent, simple, and widely-used templates that authors can choose from, e.g.
+@doi:10.1080/00031305.2017.1375986, @figueiredoSimpleKitUse2022, @Cookiecutter, @cooperGuideReproducibleCode2017 and @wilsonGoodEnoughPractices2017
+
+Second, for analyses that are not highly time-consuming (which was over 90% of the sample), authors should re-run their code more than once, and ideally not on the same computer (e.g.
 by another co-author of the paper), before submission to confirm everything works as expected.
-This would likely also prevent issues relating to path and function errors.
+This would ensure portability and likely also prevent issues relating to path and function errors.
 Complex and time-consuming analyses should use pipeline or workflow management tools, e.g.
 GNU Make, Luigi, Snakemake, or Targets, to document the relationship of the files and folders in a machine-readable format and simplify running and re-running code [@wrattenReproducibleScalableShareable2021; @landau2021targets].
-Third, authors should include in their compendium a README document that describes to a reader the folders and files contained in the project.
+
+Third, authors should include in their compendium a README document that describes to readers the folders and files contained in the project.
 The README file is typically the first file that a reader will look at in a compendium so it should include brief instructions to guide the user to a successful reproduction of the original results (e.g.
 what order to run the code files in).
+A README should also briefly describe the contents of the compendium, where other necessary files can be obtained (e.g.
+data files not included in the compendium because they require registration), and the key software packages needed and the version numbers that the authors used.
+
 Fourth, authors should ensure clear, direct and obvious connections between their code and the results they present in their paper.
-A simple way to do this is to have one code file for each figure and table, and name the code files with the figure or table number and some key words in the caption.
+One simple way to do this is to have one code file for each figure and table, and name the code files with the figure or table number and some key words in the caption.
 Another way some authors are accomplishing this is by using literate programming tools, such as Quarto and Jupyter notebooks [@quarto; @kluyver2016jupyter] that enable the research narrative and code for data analysis to be woven together in one document.
-Quarto was a popular tool among the JAS papers in the reproducibility review sample, for example, @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024 wrote their entire manuscript using Quarto.
+Quarto was a popular tool among the JAS papers in the reproducibility review sample, for example, @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024 wrote their entire manuscripts using Quarto.
 These four simple steps have high potential to substantially increase the computational reproducibility of archaeological science.
@@
 @bilottiPointPatternAnalysis2024; @willExploringUtilityUnretouched2025].
+
 The simplest way for archaeologists, to improve here is to write the names and version numbers of the software and packages they used in their README file, as we see in @herskindComputationalLinguisticMethodology2024 and @monnaStudyingSerialityMaterial2024.
@@
 This is an active area of development, and while there are many tools currently available, the most robust and widely used include renv for R [@renv, see examples in @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024] and conda and poetry for Python [@conda2023; @poetry2023].
+
 A more comprehensive solution, and the leading best practice for managing dependencies in many computationally intensive fields using Python in particular, is to include a Dockerfile in the compendium [@moreauContainersComputationalReproducibility2023].
 This is a set of a machine- and human-readable instructions that enables a user to recreate the author's computational environment (including those requirements beyond the R or Python packages) on another computer [@nustTenSimpleRules2020].
-Dockerfiles are gradually being adopted by archaeologists, see @cremaModellingDiffusionInnovation2024 and @liaoApplyingMaskRCNN2024 for examples, and most of my reproducibility reviews include a recommendation that the author include a Dockerfile to manage dependencies efficiently.
+Dockerfiles are gradually being adopted by archaeologists, see @cremaModellingDiffusionInnovation2024 and @liaoApplyingMaskRCNN2024 for examples, and most of our reproducibility reviews include a recommendation that the author include a Dockerfile to manage complex dependencies efficiently.
 
 ## Conclusion
+
+I have presented a bibliometric analysis on the status of archaeology as a science, showing that it is If archaeological science is to continue to progress through increased consensus through the accumulation of reliable facts and methods, it is essential for researchers to take computational reproducibility seriously.
+Computers have become a central field and laboratory instrument for so much of our work, so we have an ethical duty to document how we change our data as it flows through silicon just as carefully as we document the operating parameters of a mass spectrometer or any other field or laboratory instrument.
+However, the current state of quantitative archaeology, with most researchers not using open source code, is comparable to the secrecy of alchemy prior to the emergence of chemistry.
+Abandoning this habit of secrecy in favour of reproducibility is vital if we are to avoid a future where our journals are filled with pretty pictures that the reader has no hope of repeating or adapting in their own work.
+It is time for computational reproducibility to be considered a minimum requirement for evaluating the integrity and usefulness of empirical results.
+
+Computational reproducibility is not a panacea, results that are fully reproducible can contain errors and fraud.
+It is no guarantee of code quality, or that statistics have been used appropriately (cf.
+Vaiglova this issue), or that data management is consistent with FAIR and CARE principles [cf.
+@carrollOperationalizingCAREFAIR2021].
+It is also time-consuming for authors to ensure their computational work can be reproduced, and for reviewers to evaluate.
+In an professional environment where job security and career progression is often associated with pressure to publish many high-impact papers, demands for authors to spend time on reproducibility, resulting in less time for publishing more papers, may seem frustrating.
+This may seem especially unfair to early career researchers on short-term contracts, who may feel the goalposts for career success are being moved and that they are being asked to do work their graduate training has not prepared them for.
+This highlights the need for a culture shift among senior archaeological scientists to value reproducibility in hiring and promotion decision-making.
+Professors should contribute to this culture shift by nurturing a positive culture of reanalysis and reproducibility in their graduate teaching and evaluation [e.g.
+@marwickHowUseReplication2020].
 
@@
 While consensus is one measure of the hardness of a science, the computability
-
-While not yet the norm in archaeology, the systemic verification of the computational reproducibility of the results of conditionally accepted papers is now required in many leading academic journals including the American Economic Review, American Economic Journal, Econometrica, Review of Economic Studies, Economic Journal, Journal of the American Statistical Association, American Journal of Political Science, and the Journal of Politics.
-
-Computationally reproducible research can still be plagued by various types of errors, ranging from honest mistakes (e.g., typos in the code) to outright fraud (e.g., data fabrication).
-A practical hurdle in ascertaining computational reproducibility for the published literature is that the verification is resource-intensive
 
@@
 
-Computational reproducibilty has become central to archaeology's status a science, especially as the definition and practice of science shifts over time.
-Computers have become an essential field and laboratory instrument for almost all of us, so we should document how we change our data as it flows through silicon just as carefully as we document the operating parameters of a mass spectrometer or other laboratory instrument.
-In making the case for reproducibilty as a core value in archaeological science
-
-In recent years the Journal of Archaeological Science has been one of a number of archaeology journals that introduced an unusual new kind of peer review, one that evaluates the computational reproducibilty of the research submitted for publication.
-This journal was among the first of these and is currently by far the most active in conducting these reviews.
-In this article I reflect on how this new kind of review came about, why it is important for the future of archaeological science, and offer suggestions for prospective authors based on my experience as a reproducibilty reviewer for this journal on how we can ensure archaeological science remains worthy of the name.
-
-My motivation is the view that It is customary in significant round-number anniversary -->
+ -->
 ```

### 2025-02-10 paper.qmd
writing conclusion

@@
 
-I have presented a bibliometric analysis on the status of archaeology as a science, showing that it is If archaeological science is to continue to progress through increased consensus through the accumulation of reliable facts and methods, it is essential for researchers to take computational reproducibility seriously.
+In Jonathon Swift's classic satirical novel *Gulliver's Travels* (1726) Gulliver visits the fictional Grand Academy of Lagado in Balnibarbi, a caricature of the Royal Society of London, and meets severals researchers working on wildly impractical projects, including one attempting to extracting sunbeams out of cucumbers.
+
+I have presented a bibliometric analysis on the status of archaeology as a science, showing that we generally behave as social scientists If archaeological science is to continue to progress through increased consensus through the accumulation of reliable facts and methods, it is essential for researchers to take computational reproducibility seriously.
 Computers have become a central field and laboratory instrument for so much of our work, so we have an ethical duty to document how we change our data as it flows through silicon just as carefully as we document the operating parameters of a mass spectrometer or any other field or laboratory instrument.

### 2025-02-11 README.Rmd
simplify structure and add readme

---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)

```

# Extracting sunbeams from cucumbers: Computational reproducibility is essential for archaeological science

This repository contains the data and code for my paper to be submitted
to the Journal of Archaeological Science. Earlier versions were
presented at the University of Cambridge
[(2019)](https://github.com/benmarwick/March-2019-Cambridge-Big-Data-Archaeology)
and New York University
[(2025)](https://github.com/benmarwick/nyu-2025-workshop)

## Contents

The **analysis** directory contains:

-   [:file_folder: paper](/analysis/paper): Quarto source document for
    manuscript. Includes code to reproduce the figures and tables
    generated by the analysis.
-   [:file_folder: data](/analysis/data): Data used in the analysis.
-   [:file_folder: figures](/analysis/figures): Plots and other
    illustrations

## How to run in your browser or download and run locally

This research compendium has been developed using the statistical
programming language R. To work with the compendium, you will need
installed on your computer the [R
software](https://cloud.r-project.org/) itself and optionally [RStudio
Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:

-   open the `.Rproj` file in RStudio\
-   install the packages used in the Quarto document\
-   finally, render the `analysis/paper/paper.qmd` document to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console

### Licenses

**Text and figures :**
[CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [LICENSE](LICENSE.md) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/)
attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please
see our [contributor guidelines](CONTRIBUTING.md). Please note that this
project is released with a [Contributor Code of Conduct](CONDUCT.md). By
participating in this project you agree to abide by its terms.

### 2025-02-11 analysis/paper/paper.qmd
simplify structure and add readme

--- 2025-02-10 paper.qmd
@@
 
-items_df <- read_rds("data/wos-data-df.rds")
+items_df <- 
+  read_rds(here::here("analysis/data/wos-data-df.rds"))
 
@@
 # reporting on my AER work
-jas_rr_data <- read_csv("data/JAS AER data analysis.csv")
+jas_rr_data <- 
+  read_csv(here::here("analysis/data/JAS AER data analysis.csv"))
+
 base_size <- 10

### 2025-02-11 analysis/code/000-import-raw-data.R

@@ -1,0 +1,113 @@
+# read in the data, Jan 2012 data for comparison with Fanelli?
+wos_files <- dir("data",  full.names = TRUE)
+
+library(tidyverse)
+text <- map(wos_files, 
+            ~read_file(.x, 
+                       locale = locale(encoding = "latin1")))
+
+# split on article delimiter provided by WOS
+items <- unlist(str_split(text, 
+                          pattern = "\nPT J\n"))
+
+# get rid of the advertising
+items <- str_replace_all(items, 
+                         "FN Clarivate Analytics Web of Science\nVR 1.0", "")
+items <- items[items != ""] 
+
+#  length(items) # each item is one article
+
+# function to automate getting the variables out of each item
+extractor <- function(i){
+  
+  # debug with 
+  # i <- items[[3]]
+  
+  authors =     gsub(".*AU *(.*?) *\nAF .*", "\\1", i)
+  authors_n =   str_count(authors, "\n") + 1
+  title =       gsub(".*\nTI *(.*?) *\nSO .*", "\\1", i)
+  title_n =     str_count(title, "\\w+") 
+  journal =     gsub(".*\nSO *(.*?) *\nLA .*", "\\1", i)
+  abstract =    gsub(".*\nAB *(.*?) *\nC1 .*", "\\1", i)
+  refs =        gsub(".*\nCR *(.*?) *\nNR .*", "\\1", i)
+  refs_n =      as.numeric(gsub(".*\nNR *(.*?) *\nTC .*", "\\1", i))
+  pages_n =     as.numeric(gsub(".*\nPG *(.*?) *\nWC .*", "\\1", i))
+  year =        as.numeric(gsub(".*\nPY *(.*?) *\nVL .*", "\\1", i))
+  doi =         gsub(".*\nDI *(.*?) *\nPG .*", "\\1", i)
+  
+  dplyr::data_frame(
+    authors =         authors,
+    authors_n =       authors_n,
+    title =           title ,
+    title_n =         title_n,
+    journal =         journal,
+    abstract  =       abstract,
[…]

### 2025-02-11 analysis/code/001-authors.R

@@ -1,0 +1,74 @@
+library(tidyverse)
+
+items_df <- read_rds("data/wos-data-df.rds")
+
+# number of authors + greater scope and need for collaboration [9,44]
+
+# all on one
+
+items_df %>% 
+  filter(!is.na(year)) %>% 
+  ggplot(aes(1,
+             log(authors_n))) +
+  geom_boxplot(colour = "red", 
+               size = 1)  +
+  scale_y_continuous(limits = c(0, 5)) +
+  scale_x_continuous(labels = NULL) +
+  theme_minimal() +
+  ylab("N. of authors (ln)") +
+  xlab("Collaborator group size")
+
+ggsave("figures/box_num_authors.svg")
+
+# over time
+items_df %>% 
+  filter(!is.na(year)) %>% 
+  ggplot(aes(year,
+             log(authors_n),
+             group = year)) +
+  geom_jitter(alpha = 0.2) +
+  theme_minimal() +
+  ylab("Number of authors (ln)")
+
+ggsave("figures/box_num_authors_over_time.png")
+
+# by journal, wow yes!
+library(ggridges)
+items_df %>% 
+  mutate(journal = str_wrap(journal, 20)) %>% 
+  ggplot(aes(y = reorder(journal,
+                     authors_n), 
+             x = log(authors_n),
+             fill = ..x..,
+             height = ..density..)) +
+  geom_density_ridges_gradient(stat = "density",
+                               colour = "white") +
+  scale_fill_viridis_c() +
+  guides(fill = 'none') +
+  theme_minimal() +
+  theme(axis.text.y = element_text(size = 6)) +
+  ylab("") + 
+  xlab("Number of authors (ln)")
+
+
+ggsave("figures/box_num_authors_by_journal.png")
+
+# histogram
+
+tally_authors_n <- 
+  items_df %>% 
+  group_by(authors_n) %>% 
+  tally() %>% 
+  mutate(perc = n / sum(n) * 100) %>% 
+  mutate(cum_per = cumsum(perc)) %>% 
+  arrange(cum_per) # 90% of papers have 4 or fewer authors
+
+items_df %>% 
+  ggplot(aes(authors_n)) +
+  geom_histogram()  +
+  theme_minimal() +
+  scale_x_log10() +
+  xlab("Number of authors") + 
+  ylab("Count") 
+
+ggsave("figures/box_num_histogram.png")

### 2025-02-11 analysis/code/002-article-length.R

@@ -1,0 +1,53 @@
+#  ------------------------------------------------------------------
+# length of article - less need to introduce, justify and explain study [47,52]
+
+# all on one
+library(tidyverse)
+
+items_df %>% 
+  ggplot(aes(year,
+             log(pages_n))) +
+  geom_boxplot(colour = "red", 
+               size = 1)  +
+  scale_y_continuous(limits = c(0, 5)) +
+  scale_x_continuous(labels = NULL,
+                     name = NULL) +
+  theme_minimal() +
+  theme(panel.grid  = element_blank()) +
+  coord_equal(ratio = 6)
+
+ggsave("figures/box_num_pages.svg")
+
+
+# over time
+items_df %>% 
+  ggplot(aes(year,
+             log(pages_n),
+             group = year)) +
+  geom_boxplot() +
+  geom_jitter(alpha = 0.01) +
+  theme_minimal() +
+  ylab("Number of pages (ln)") 
+
+ggsave("figures/box_num_pages_over_time.png")
+
+# by journal, wow yes!
+library(ggforce)
+items_df %>% 
+  mutate(journal = str_wrap(journal, 20)) %>% 
+  ggplot(aes(y = reorder(journal,
+                     pages_n), 
+             x = log(pages_n),
+             fill = ..x..,
+             height = ..density..)) +
+  geom_density_ridges_gradient(stat = "density",
+                               colour = "white") +
+  scale_fill_viridis_c() +
+  guides(fill = 'none') +
+  theme_minimal() +
+  theme(axis.text.y = element_text(size = 6)) +
+  xlab("Number of pages (ln)") +
+  ylab("")
+
+ggsave("figures/box_num_pages_by_journal.png")
+

### 2025-02-11 analysis/code/003-title-length.R

@@ -1,0 +1,59 @@
+
+# relative title length + clearly defined, substantive research questions [52,58], total number of words, divided by total number of pages.
+
+items_df_title <- 
+  items_df %>% 
+  filter(!is.na(pages_n)) %>%  
+  filter(!is.na(title_n)) %>% 
+  mutate(relative_title_length = log(title_n / pages_n))
+
+# all on one
+items_df_title %>% 
+  filter(!is.na(year)) %>% 
+  ggplot(aes(year,
+             relative_title_length)) +
+  geom_boxplot(colour = "red", 
+               size = 1)  +
+  scale_y_continuous(limits = c(-5, 5),
+                     breaks =  seq(-5, 5, 1),
+                     labels = seq(-5, 5, 1)) +
+  scale_x_continuous(labels = NULL,
+                     name = NULL) +
+  theme_minimal() +
+  theme(panel.grid  = element_blank()) +
+  coord_equal(ratio = 5)
+
+ggsave("figures/box_title_length.svg")
+
+# over time
+items_df_title %>% 
+  filter(!is.na(year)) %>% 
+  ggplot(aes(year,
+             relative_title_length,
+             group = year)) +
+  geom_boxplot() +
+  geom_jitter(alpha = 0.01)+
+  theme_minimal() +
+  ylab("Relative title length")
+
+ggsave("figures/box_title_length_over_time.png")
+
+# by journal
+items_df_title %>% 
+  mutate(journal = str_wrap(journal, 20)) %>% 
+  ggplot(aes(y = reorder(journal,
+                     relative_title_length), 
+             x = relative_title_length,
+             fill = ..x..,
+             height = ..density..)) +
+  geom_density_ridges_gradient(stat = "density",
+                               colour = "white") +
+  scale_fill_viridis_c() +
+  guides(fill = 'none') +
+  theme_minimal() +
+  theme(axis.text.y = element_text(size = 6)) +
+  xlab("") +
+  ylab("Relative title length")
+
+ggsave("figures/box_title_length_over_time.png")
+

### 2025-02-11 analysis/code/004-references.R

@@ -1,0 +1,257 @@
+#  ------------------------------------------------------------------
+# number of references - less need to justify, explain and support study [47]
+
+
+# all on one
+
+items_df %>% 
+  ggplot(aes(year,
+             sqrt(refs_n))) +
+  geom_boxplot(colour = "red", 
+               size = 1)  +
+  scale_y_continuous(limits = c(0, 20)) +
+  scale_x_continuous(labels = NULL,
+                     name = NULL) +
+  theme_minimal() +
+  theme(panel.grid  = element_blank()) +
+  coord_equal(ratio = 6)
+
+ggsave("figures/box_num_refs.svg")
+
+
+# over time
+items_df %>% 
+  ggplot(aes(year,
+             sqrt(refs_n),
+             group = year)) +
+  geom_boxplot() +
+  geom_jitter(alpha = 0.01)+
+  theme_minimal() +
+  ylab("Number of references (ln)")
+
+ggsave("figures/box_num_refs_over_time.png")
+
+# by journal, wow yes!
+items_df %>% 
+  mutate(journal = str_wrap(journal, 20)) %>% 
+  ggplot(aes(y = reorder(journal,
+                     refs_n), 
+             x = sqrt(refs_n),
+             fill = ..x..,
+             height = ..density..)) +
+  geom_density_ridges_gradient(stat = "density",
+                               colour = "white") +
+  scale_fill_viridis_c() +
+  guides(fill = 'none') +
+  theme_minimal()  +
+  theme(axis.text.y = element_text(size = 6)) +
+  xlab("Number of references (ln)") +
+  ylab("")
+
+ggsave("figures/box_num_refs_by_journal.png")
+
+#  ------------------------------------------------------------------
+#  references to monographs - focus on simpler questions; less need to justify, explain and support study [53,54]
+
+# not easy...
+#  ------------------------------------------------------------------
+# age of references - faster settling of disagreements; greater potential to build research upon previous findings [44,56]
+
+# Derek de Solla Price proposed an index, which measures the proportion of cited references published in the five years preceding the citing paper
+
+
+# years are ", YYYY, "
+library(stringr)
+
[…]

### 2025-02-11 code/000-import-raw-data.R

@@ -1,113 +1,0 @@
-# read in the data, Jan 2012 data for comparison with Fanelli?
-wos_files <- dir("data",  full.names = TRUE)
-
-library(tidyverse)
-text <- map(wos_files, 
-            ~read_file(.x, 
-                       locale = locale(encoding = "latin1")))
-
-# split on article delimiter provided by WOS
-items <- unlist(str_split(text, 
-                          pattern = "\nPT J\n"))
-
-# get rid of the advertising
-items <- str_replace_all(items, 
-                         "FN Clarivate Analytics Web of Science\nVR 1.0", "")
-items <- items[items != ""] 
-
-#  length(items) # each item is one article
-
-# function to automate getting the variables out of each item
-extractor <- function(i){
-  
-  # debug with 
-  # i <- items[[3]]
-  
-  authors =     gsub(".*AU *(.*?) *\nAF .*", "\\1", i)
-  authors_n =   str_count(authors, "\n") + 1
-  title =       gsub(".*\nTI *(.*?) *\nSO .*", "\\1", i)
-  title_n =     str_count(title, "\\w+") 
-  journal =     gsub(".*\nSO *(.*?) *\nLA .*", "\\1", i)
-  abstract =    gsub(".*\nAB *(.*?) *\nC1 .*", "\\1", i)
-  refs =        gsub(".*\nCR *(.*?) *\nNR .*", "\\1", i)
-  refs_n =      as.numeric(gsub(".*\nNR *(.*?) *\nTC .*", "\\1", i))
-  pages_n =     as.numeric(gsub(".*\nPG *(.*?) *\nWC .*", "\\1", i))
-  year =        as.numeric(gsub(".*\nPY *(.*?) *\nVL .*", "\\1", i))
-  doi =         gsub(".*\nDI *(.*?) *\nPG .*", "\\1", i)
-  
-  dplyr::data_frame(
-    authors =         authors,
-    authors_n =       authors_n,
-    title =           title ,
-    title_n =         title_n,
-    journal =         journal,
-    abstract  =       abstract,
[…]

### 2025-02-11 code/001-authors.R

@@ -1,74 +1,0 @@
-library(tidyverse)
-
-items_df <- read_rds("data/wos-data-df.rds")
-
-# number of authors + greater scope and need for collaboration [9,44]
-
-# all on one
-
-items_df %>% 
-  filter(!is.na(year)) %>% 
-  ggplot(aes(1,
-             log(authors_n))) +
-  geom_boxplot(colour = "red", 
-               size = 1)  +
-  scale_y_continuous(limits = c(0, 5)) +
-  scale_x_continuous(labels = NULL) +
-  theme_minimal() +
-  ylab("N. of authors (ln)") +
-  xlab("Collaborator group size")
-
-ggsave("figures/box_num_authors.svg")
-
-# over time
-items_df %>% 
-  filter(!is.na(year)) %>% 
-  ggplot(aes(year,
-             log(authors_n),
-             group = year)) +
-  geom_jitter(alpha = 0.2) +
-  theme_minimal() +
-  ylab("Number of authors (ln)")
-
-ggsave("figures/box_num_authors_over_time.png")
-
-# by journal, wow yes!
-library(ggridges)
-items_df %>% 
-  mutate(journal = str_wrap(journal, 20)) %>% 
-  ggplot(aes(y = reorder(journal,
-                     authors_n), 
-             x = log(authors_n),
-             fill = ..x..,
-             height = ..density..)) +
-  geom_density_ridges_gradient(stat = "density",
-                               colour = "white") +
-  scale_fill_viridis_c() +
-  guides(fill = 'none') +
-  theme_minimal() +
-  theme(axis.text.y = element_text(size = 6)) +
-  ylab("") + 
-  xlab("Number of authors (ln)")
-
-
-ggsave("figures/box_num_authors_by_journal.png")
-
-# histogram
-
-tally_authors_n <- 
-  items_df %>% 
-  group_by(authors_n) %>% 
-  tally() %>% 
-  mutate(perc = n / sum(n) * 100) %>% 
-  mutate(cum_per = cumsum(perc)) %>% 
-  arrange(cum_per) # 90% of papers have 4 or fewer authors
-
-items_df %>% 
-  ggplot(aes(authors_n)) +
-  geom_histogram()  +
-  theme_minimal() +
-  scale_x_log10() +
-  xlab("Number of authors") + 
-  ylab("Count") 
-
-ggsave("figures/box_num_histogram.png")

### 2025-02-11 code/003-title-length.R

@@ -1,59 +1,0 @@
-
-# relative title length + clearly defined, substantive research questions [52,58], total number of words, divided by total number of pages.
-
-items_df_title <- 
-  items_df %>% 
-  filter(!is.na(pages_n)) %>%  
-  filter(!is.na(title_n)) %>% 
-  mutate(relative_title_length = log(title_n / pages_n))
-
-# all on one
-items_df_title %>% 
-  filter(!is.na(year)) %>% 
-  ggplot(aes(year,
-             relative_title_length)) +
-  geom_boxplot(colour = "red", 
-               size = 1)  +
-  scale_y_continuous(limits = c(-5, 5),
-                     breaks =  seq(-5, 5, 1),
-                     labels = seq(-5, 5, 1)) +
-  scale_x_continuous(labels = NULL,
-                     name = NULL) +
-  theme_minimal() +
-  theme(panel.grid  = element_blank()) +
-  coord_equal(ratio = 5)
-
-ggsave("figures/box_title_length.svg")
-
-# over time
-items_df_title %>% 
-  filter(!is.na(year)) %>% 
-  ggplot(aes(year,
-             relative_title_length,
-             group = year)) +
-  geom_boxplot() +
-  geom_jitter(alpha = 0.01)+
-  theme_minimal() +
-  ylab("Relative title length")
-
-ggsave("figures/box_title_length_over_time.png")
-
-# by journal
-items_df_title %>% 
-  mutate(journal = str_wrap(journal, 20)) %>% 
-  ggplot(aes(y = reorder(journal,
-                     relative_title_length), 
-             x = relative_title_length,
-             fill = ..x..,
-             height = ..density..)) +
-  geom_density_ridges_gradient(stat = "density",
-                               colour = "white") +
-  scale_fill_viridis_c() +
-  guides(fill = 'none') +
-  theme_minimal() +
-  theme(axis.text.y = element_text(size = 6)) +
-  xlab("") +
-  ylab("Relative title length")
-
-ggsave("figures/box_title_length_over_time.png")
-

### 2025-02-11 code/004-references.R

@@ -1,257 +1,0 @@
-#  ------------------------------------------------------------------
-# number of references - less need to justify, explain and support study [47]
-
-
-# all on one
-
-items_df %>% 
-  ggplot(aes(year,
-             sqrt(refs_n))) +
-  geom_boxplot(colour = "red", 
-               size = 1)  +
-  scale_y_continuous(limits = c(0, 20)) +
-  scale_x_continuous(labels = NULL,
-                     name = NULL) +
-  theme_minimal() +
-  theme(panel.grid  = element_blank()) +
-  coord_equal(ratio = 6)
-
-ggsave("figures/box_num_refs.svg")
-
-
-# over time
-items_df %>% 
-  ggplot(aes(year,
-             sqrt(refs_n),
-             group = year)) +
-  geom_boxplot() +
-  geom_jitter(alpha = 0.01)+
-  theme_minimal() +
-  ylab("Number of references (ln)")
-
-ggsave("figures/box_num_refs_over_time.png")
-
-# by journal, wow yes!
-items_df %>% 
-  mutate(journal = str_wrap(journal, 20)) %>% 
-  ggplot(aes(y = reorder(journal,
-                     refs_n), 
-             x = sqrt(refs_n),
-             fill = ..x..,
-             height = ..density..)) +
-  geom_density_ridges_gradient(stat = "density",
-                               colour = "white") +
-  scale_fill_viridis_c() +
-  guides(fill = 'none') +
-  theme_minimal()  +
-  theme(axis.text.y = element_text(size = 6)) +
-  xlab("Number of references (ln)") +
-  ylab("")
-
-ggsave("figures/box_num_refs_by_journal.png")
-
-#  ------------------------------------------------------------------
-#  references to monographs - focus on simpler questions; less need to justify, explain and support study [53,54]
-
-# not easy...
-#  ------------------------------------------------------------------
-# age of references - faster settling of disagreements; greater potential to build research upon previous findings [44,56]
-
-# Derek de Solla Price proposed an index, which measures the proportion of cited references published in the five years preceding the citing paper
-
-
-# years are ", YYYY, "
-library(stringr)
-
[…]

### 2025-02-11 code/007-disruptive-articles.R

@@ -1,257 +1,0 @@
-
-# https://github.com/benmarwick/March-2019-Cambridge-Big-Data-Archaeology/blob/master/code/003-disruptiveness-index.R
-# https://pubsonline.informs.org/doi/10.1287/mnsc.2015.2366
-# https://www.nature.com/articles/s41586-019-0941-9
-
-library(tidyverse)
-library(here)
-
-# get articles with DOIs and reference lists
-items_wth_dois_and_refs <- 
-  items_df %>% 
-  filter(str_detect(doi, "^10.")) %>% 
-  mutate(doi = str_remove_all(doi, "\\[|\\]")) %>% 
-  mutate(doi_article = doi) %>% 
-  select(-doi)
-
-# get items in the reference lists with DOIs
-items_df_one_row_per_ref_with_doi <- 
-  items_wth_dois_and_refs %>% 
-  separate_rows(refs, sep = "\n  ") %>% 
-  ungroup %>% 
-  mutate(refs = tolower(str_trim(refs))) %>% 
-  mutate(refs = str_extract(refs, "doi.*")) %>% 
-  mutate(refs = str_remove_all(refs, "\\[|\\]")) %>% 
-  mutate(doi = str_remove_all(refs, "doi"))  %>% 
-  mutate(doi_cited = str_trim(doi)) %>%  
-  filter(!is.na(doi_cited)) %>% 
-  select(-doi)
-
-# ------- compute Disruption  ----
-# for a focal article:
-# - find the articles that cite the focal article
-# - find what articles that citing article cites 
-# - find the ratio of citing articles that cite focal articles refs vs those that don't
-
-disruption_output <- list(length = length(items_wth_dois_and_refs$doi_article))
-
-for(i in seq_along(items_wth_dois_and_refs$doi_article)){
-  
-  this_article_doi <- items_wth_dois_and_refs$doi_article[i] # 1000
-  
-  # what articles are cited by the focal paper
-  cited_by_focal_paper <- 
-    items_df_one_row_per_ref_with_doi %>% 
-    filter(doi_article %in% this_article_doi) 
-  
-  # what articles cite the focal paper, that are in our list
-  citing_the_focal_paper <- 
-    items_df_one_row_per_ref_with_doi %>% 
-    filter(doi_cited %in% this_article_doi)
-  
-  # what articles are cited by the citing papers
-  cited_by_citing_papers <- 
-    items_df_one_row_per_ref_with_doi %>% 
-    filter(doi_article %in% citing_the_focal_paper$doi_article)
-  
[…]

### 2025-02-14 paper.qmd
update data paths

@@
 
-items_df <- read_rds("data/wos-data-df.rds")
+items_df <- read_rds(here::here("analysis/data/wos-data-df.rds"))
 
@@
 # reporting on my AER work
-jas_rr_data <- read_csv("data/JAS AER data analysis.csv")
+jas_rr_data <-
+  read_csv(here::here("analysis/data/JAS AER data analysis.csv"))
 base_size <- 10

### 2025-02-15 README.Rmd
update dockerfile

@@
 
+[![.github/workflows/knit-in-docker.yml](https://github.com/benmarwick/web-of-science-archaeology/actions/workflows/knit-in-docker.yml/badge.svg)](https://github.com/benmarwick/web-of-science-archaeology/actions/workflows/knit-in-docker.yml)
+
 This repository contains the data and code for my paper to be submitted
@@
 
-## How to run in your browser or download and run locally
+## How to download and run locally
 
-This research compendium has been developed using the statistical
-programming language R. To work with the compendium, you will need
-installed on your computer the [R
-software](https://cloud.r-project.org/) itself and optionally [RStudio
-Desktop](https://rstudio.com/products/rstudio/download/).
+This research compendium has been developed using the statistical programming language R.
+To work with the compendium, you will need installed on your computer the [R software](https://cloud.r-project.org/) itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).
 
@@
 -   finally, render the `analysis/paper/paper.qmd` document to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console
+
+## How to download and run locally in a Docker container
+
+This research compendium includes a [Dockerfile](Dockerfile) that specifies the computational environment used to generate the results presented in the paper.
+You can recreate this environment on your computer by following these steps:
+
+1. Download the compendium as a zip from from this URL:
+[master.zip](/archive/master.zip) and unzip    
+2. Install [Docker](https://www.docker.com/get-started/)    
+3. Open the compendium in the terminal and run `docker build -t wos .` to build the container on your computer, it will take a few minutes and require an internet connection.
+4. Run in your terminal `docker run --rm -it -e ROOT=TRUE -e PASSWORD=rstudio -dp 8787:8787 wos` to run the container
+5. Go to http://localhost:8787/ with your browser and enter  rstudio in both fields to start RStudio in the browser, running in the Docker container
+6. In RStudio, open the project and the Quarto document, and render it
 

### 2025-02-15 analysis/paper/paper.qmd
Docker has renv pkgs available in RStudio when proj opens

@@
 instead of commercial software such as Microsoft Excel or SPSS, etc.) the success rate is 7%.
-In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values obtained from a variety of hard and soft sciences.
+In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values available from a variety of hard and soft sciences.
 
@@
 Complex and time-consuming analyses should use pipeline or workflow management tools, e.g.
-GNU Make, Luigi, Snakemake, or Targets, to document the relationship of the files and folders in a machine-readable format and simplify running and re-running code [@wrattenReproducibleScalableShareable2021; @landau2021targets].
+GNU Make, Luigi, Snakemake, or Targets, to document the relationship of the files and folders in a machine-readable format and simplify running and re-running code by others [@wrattenReproducibleScalableShareable2021; @landau2021targets].
 
@@
 A README should also briefly describe the contents of the compendium, where other necessary files can be obtained (e.g.
-data files not included in the compendium because they require registration), and the key software packages needed and the version numbers that the authors used.
+data files not included in the compendium), and the key software packages needed and the version numbers that the authors used.
 

### 2025-02-15 README.Rmd
dockerfile and instructions

--- 2025-02-11 README.Rmd
@@
 
-This repository contains the data and code for my paper to be submitted
-to the Journal of Archaeological Science. Earlier versions were
-presented at the University of Cambridge
-[(2019)](https://github.com/benmarwick/March-2019-Cambridge-Big-Data-Archaeology)
-and New York University
-[(2025)](https://github.com/benmarwick/nyu-2025-workshop)
+This repository contains the data and code for my paper to be submitted to the Journal of Archaeological Science.
+Earlier versions were presented at the University of Cambridge [(2019)](https://github.com/benmarwick/March-2019-Cambridge-Big-Data-Archaeology) and New York University [(2025)](https://github.com/benmarwick/nyu-2025-workshop)
 
@@
 
-## How to run in your browser or download and run locally
+## How to download and run locally in a Docker container
 
-This research compendium has been developed using the statistical
-programming language R. To work with the compendium, you will need
-installed on your computer the [R
-software](https://cloud.r-project.org/) itself and optionally [RStudio
-Desktop](https://rstudio.com/products/rstudio/download/).
+This research compendium includes a [Dockerfile](Dockerfile) that specifies the computational environment used to generate the results presented in the paper.
+You can recreate this environment, which includes R, RStudio and all necessary R packages, on your computer by following these steps:
 
-You can download the compendium as a zip from from this URL:
-[master.zip](/archive/master.zip). After unzipping:
-
--   open the `.Rproj` file in RStudio\
--   install the packages used in the Quarto document\
--   finally, render the `analysis/paper/paper.qmd` document to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console
+1. Install [Docker](https://www.docker.com/get-started/) on your computer 
+2. Download the compendium as a zip from from this URL:
+[master.zip](/archive/master.zip) and unzip    
+3. Set your terminal working directory to the compendium, and run `docker build -t wos .` to build the container on your computer, it will take a few minutes and require an internet connection.
+4. Run in your terminal `docker run --rm -it -e ROOT=TRUE -e PASSWORD=rstudio -dp 8787:8787 wos` to run the container
+5. Go to http://localhost:8787/ with your browser and enter rstudio in both fields to start RStudio in the browser, running in the Docker container
+6. In RStudio, run `rstudioapi::openProject("/home/rstudio/web-of-science-archaeology")` to open this project
+7. Run `rmarkdown::render("analysis/paper/paper.qmd")` to render the Quarto document
+8. When finished, to remove the Docker container and files, run this in the terminal `docker ps -aq | xargs docker stop | xargs docker rm`
 

### 2025-02-16 README.Rmd
update zip link
update readme

@@
 1. Install [Docker](https://www.docker.com/get-started/) on your computer 
-2. Download the compendium as a zip from from this URL:
-[master.zip](/archive/master.zip) and unzip    
+2. Download the compendium from GitHub by clicking on the green `< > Code` button above, then click `Download ZIP`, and unzip on your computer
 3. Set your terminal working directory to the compendium, and run `docker build -t wos .` to build the container on your computer, it will take a few minutes and require an internet connection.
-4. Run in your terminal `docker run --rm -it -e ROOT=TRUE -e PASSWORD=rstudio -dp 8787:8787 wos` to run the container
-5. Go to http://localhost:8787/ with your browser and enter rstudio in both fields to start RStudio in the browser, running in the Docker container
-6. In RStudio, run `rstudioapi::openProject("/home/rstudio/web-of-science-archaeology")` to open this project
-7. Run `rmarkdown::render("analysis/paper/paper.qmd")` to render the Quarto document
-8. When finished, to remove the Docker container and files, run this in the terminal `docker ps -aq | xargs docker stop | xargs docker rm`
+4. Run in your terminal `docker run --rm -it -e ROOT=TRUE -e PASSWORD=rstudio -dp 8787:8787 wos` to start the container on your computer
+5. Go to http://localhost:8787/ with your browser and enter `rstudio` in both fields to start RStudio in your browser, running in the Docker container with all the required depedencies installed
+6. In RStudio, run `rstudioapi::openProject("/home/rstudio/web-of-science-archaeology")` to open the compendium's project
+7. Run `rmarkdown::render("analysis/paper/paper.qmd")` to render the Quarto document that is the manuscript submitted for publication
+8. When finished, to delete the Docker container from your computer, run this in the terminal `docker ps -aq | xargs docker stop | xargs docker rm`
 

### 2025-02-16 analysis/paper/paper.qmd
update readme

@@
 
-In Jonathon Swift's classic satirical novel *Gulliver's Travels* (1726) Gulliver visits the fictional Grand Academy of Lagado in Balnibarbi, a caricature of the Royal Society of London, and meets severals researchers working on wildly impractical projects, including one attempting to extracting sunbeams out of cucumbers.
-
-I have presented a bibliometric analysis on the status of archaeology as a science, showing that we generally behave as social scientists If archaeological science is to continue to progress through increased consensus through the accumulation of reliable facts and methods, it is essential for researchers to take computational reproducibility seriously.
+I have presented a bibliometric analysis on the status of archaeology as a science, showing distinct disunity, but that we generally behave as social scientists.
+If archaeological science is to continue to progress through increased consensus through the accumulation of reliable facts and methods, it is essential for researchers to take computational reproducibility seriously.
 Computers have become a central field and laboratory instrument for so much of our work, so we have an ethical duty to document how we change our data as it flows through silicon just as carefully as we document the operating parameters of a mass spectrometer or any other field or laboratory instrument.
@@
 
+In Jonathon Swift's classic satirical novel *Gulliver's Travels* (1726) Gulliver visits the fictional Grand Academy of Lagado in Balnibarbi, a caricature of the Royal Society of London, and meets several researchers working on wildly impractical projects, including one attempting to extracting sunbeams out of cucumbers.
+
 ```{=html}

### 2025-02-18 analysis/paper/paper.qmd

@@
 
+In the classic satirical novel *Gulliver's Travels* (1726) by Irish writer Jonathon Swift Gulliver visits the fictional Grand Academy of Lagado in Balnibarbi, a caricature of the Royal Society of London, and meets several researchers working on wildly impractical projects, including one attempting to extracting sunbeams out of cucumbers.
+This is usually interpreted as a subversive and anti-colonial parody depicting institutionalized research as an absurd fund-raising activity with no practical benefits to society [@alff2014swift; @nicolson1937scientific].
+@feinbergExtractingSunbeamsCucumbers2011
+
 I have presented a bibliometric analysis on the status of archaeology as a science, showing distinct disunity, but that we generally behave as social scientists.
@@
 
-In Jonathon Swift's classic satirical novel *Gulliver's Travels* (1726) Gulliver visits the fictional Grand Academy of Lagado in Balnibarbi, a caricature of the Royal Society of London, and meets several researchers working on wildly impractical projects, including one attempting to extracting sunbeams out of cucumbers.
-
 ```{=html}

### 2025-02-18 analysis/paper/paper.qmd
writing conclusion

@@
 This is usually interpreted as a subversive and anti-colonial parody depicting institutionalized research as an absurd fund-raising activity with no practical benefits to society [@alff2014swift; @nicolson1937scientific].
-@feinbergExtractingSunbeamsCucumbers2011
-
-I have presented a bibliometric analysis on the status of archaeology as a science, showing distinct disunity, but that we generally behave as social scientists.
+It has also been used as a metaphor for the difficulty of extracting useful data out of published  @feinbergExtractingSunbeamsCucumbers2011
+
+I have presented a bibliometric analysis on the status of archaeology as a science, showing distinct disunity that is increasing over time,.
+On average we generally behave as social scientists, with some elements in common with harder sciences.
 If archaeological science is to continue to progress through increased consensus through the accumulation of reliable facts and methods, it is essential for researchers to take computational reproducibility seriously.

### 2025-02-19 analysis/paper/paper.qmd
final polishing before submission

@@
 title: "Extracting sunbeams from cucumbers: Computational reproducibility is essential for archaeological science"
-format: html
-bibliography: references.bib
+author:
+  - Ben Marwick:
+      correspondence: "yes"
+      email: bmarwick@uw.edu
+      orcid: 0000-0003-1689-0557
+      institute: uw
+institute:
+  - uw:
+      name: Department of Anthropology, University of Washington
+      address: 4218 Memorial Way Northeast, Seattle, WA 98105
 fig-dpi: 600
+title-block-published: "Last updated"  
+date: now
+date-format: long
+format: 
+  docx:
+    reference-doc: "templates/template.docx" # Insert path for the DOCX file
 execute:
@@
   warning: false
+  message: false
+  comment: "#>"
+  fig-path: "../figures/"
+  fig-dpi: 600
+filters:
+  - templates/scholarly-metadata.lua
+  - templates/author-info-blocks.lua
+  - templates/pagebreak.lua
+bibliography: references.bib
+csl: "templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
+abstract: |
+  Computational reproducibility is increasingly essential for advancing archaeological science, yet its implementation remains inconsistent across the discipline.
+This study presents a bibliometric assessment of archaeology’s status relative to other fields alongside a systematic review of computational reproducibility in published archaeological research.
+Analyzing nearly 10,000 articles, I identify trends in authorship, citation practices, and related metrics that position archaeology between the natural and social sciences.
+A review of reproducibility assessments for the Journal of Archaeological Science reveals persistent challenges, including missing data, unspecified dependencies, and inadequate documentation.
+To address these issues, I recommend to authors basic practical steps such as standardized project organization and explicit dependency documentation.
+Strengthening reproducibility will enhance archaeology’s scientific rigor and ensure the verifiability of research findings.
+This study underscores the urgent need for cultural and technical shifts to establish reproducibility as a cornerstone of rigorous, accountable, and impactful archaeological science.
+keywords: |
+  Computational reproducibility; Archaeological science; Bibliometric analysis; Open science
+highlights: |
+  Archaeology occupies an intermediate position between the natural and social sciences.
+  Computational reproducibility in archaeology is low, with key barriers in incomplete compendiums, unspecified dependencies, and errors in code.
+  Practical steps like logical folder structures, README files, and dependency management tools can substantially improve computational reproducibility.
+  A cultural shift toward valuing reproducibility in hiring, promotion, and training is essential for advancing archaeological science.
 ---
-
-```{=html}
-<!--
-
-authoritative, agenda-setting or hypothesis-setting papers, rather than descriptive compilations of the literature.
-
-have the vision and authority to guide the next generation of archaeological science research
-
-cross-ref Vaiglova 2025 this issue
-
--->
-```
 
@@
 
+base_size <- 6
+
 # Number of authors ------------------
@@
   scale_x_continuous(labels = NULL) +
-  theme_minimal() + 
+  theme_minimal(base_size = base_size) + 
   theme(panel.grid  = element_blank()) +
@@
                   nudge_x = 0.05) +
-
   scale_y_continuous(limits = c(-4.5, 3),
@@
   scale_x_continuous(labels = NULL) +
-  theme_minimal() +
+  theme_minimal(base_size = base_size) +
   theme(panel.grid  = element_blank()) +
@@
   scale_x_continuous(labels = NULL) +
-  theme_minimal()  +
+  theme_minimal(base_size = base_size)  +
   theme(panel.grid  = element_blank()) +
@@
 ln denotes the natural logarithm, or logarithm to the base e.
-#| fig-height: 8
+#| fig-height: 4
 
@@
 The colour of the points indicates if the overall trend is toward softer (orange) or harder (green) Generalized Additive Models were computed to fit the lines summarising the relationships between the variables and the time series.
-#| fig-height: 6
+#| fig-height: 4
 
@@
               se = FALSE,  
-              size = 3, 
+              size = 2, 
               colour = "#7570b3") +
@@
               scales = "free_y") + 
-  theme_bw(base_size = 12) +
+  theme_bw(base_size = base_size) +
   scale_color_manual(values = c("#d95f02",
@@
                                 digits = 3))),
-           hjust = 0, vjust = 1.5)
+           hjust = 0, 
+           vjust = 1.5,
+           size = 2)
 
@@
 
-journal_title_size <- 4
+journal_title_size <- 2
 
@@
                    .fittedPC2),
-             size = 2) +
+             size = 1) +
   geom_text_repel(data = pca_scores_means,
@@
             lineheight = 0.8,
-            size = 2,
+            segment.color = NA,
+            size = 1,
             bg.color = "white",  # Color of the halo
-            bg.r = 0.2,) +
+            bg.r = 0.2) +
      geom_segment(data = pca_loadings_means, 
@@
                 label = column),
-            size = 2,
+            size = 1,
             lineheight = 0.7,
@@
           bg.r = 0.2) +
-  theme_minimal()
+  theme_minimal(base_size = base_size) +
+  coord_equal()
 
@@
   xlab("") +
-  theme_minimal() +
+  theme_minimal(base_size = base_size) +
   theme(axis.text.y = element_text(size = journal_title_size))
@@
   guides(fill = 'none') +
-  theme_minimal() +
+  theme_minimal(base_size = base_size) +
   theme(axis.text.y = element_text(size = journal_title_size)) +
@@
 Panel G is a biplot of the first and second principal components of a PCA computed on the journal means of the five bibliometric variables."
-#| fig-height: 14
+#| fig-height: 6
 
@@
            nrow = 2,
-           labels = LETTERS[1:6]),
+           labels = LETTERS[1:6],
+           label_size = 6),
   plot_pca_means,
@@
   labels = c("", LETTERS[7]),
-  rel_heights = c(1, 0.5))
+  rel_heights = c(1, 0.5),
+  label_size = 6)
 
@@
 
-base_size <- 10
+base_size <- 4
 
@@
   theme(legend.position = c(.9,.7),
+        legend.title = element_text(hjust = 0.5),
+        legend.spacing = unit(0.1, "cm"),  # Reduce space between legend items
+        legend.margin = margin(1, 5, 1, 1),  # Reduce margin around the legend
+        legend.key.size = unit(0.2, "cm"),  # Reduce size of legend keys
         legend.box.background=element_rect(fill="white", 
@@
 E: Relationship between software used and issues that make research irreproducible."
+#| fig-height: 3
 
@@
 The problem of incomplete materials can be tackled in several basic ways.
-First, authors should use a clear and logical folder structure for their code and data.
+First, authors should use a simple and logical folder structure to organise their code and data to be as self-contained as possible (ie.
+not requiring files outside of the project folder).
 There are many excellent, simple, and widely-used templates that authors can choose from, e.g.
@@
 In the classic satirical novel *Gulliver's Travels* (1726) by Irish writer Jonathon Swift Gulliver visits the fictional Grand Academy of Lagado in Balnibarbi, a caricature of the Royal Society of London, and meets several researchers working on wildly impractical projects, including one attempting to extracting sunbeams out of cucumbers.
-This is usually interpreted as a subversive and anti-colonial parody depicting institutionalized research as an absurd fund-raising activity with no practical benefits to society [@alff2014swift; @nicolson1937scientific].
-It has also been used as a metaphor for the difficulty of extracting useful data out of published  @feinbergExtractingSunbeamsCucumbers2011
-
-I have presented a bibliometric analysis on the status of archaeology as a science, showing distinct disunity that is increasing over time,.
+This is usually interpreted as a subversive anti-colonial parody depicting institutionalized research as an absurd fund-raising activity with no practical benefits to society [@alff2014swift; @nicolson1937scientific].
+It has also been used as a metaphor for the difficulty of getting insights from data tables in scholarly publications [@feinbergExtractingSunbeamsCucumbers2011].
+This metaphor is relevant in our current age of computationally intensive research, where extracting useful code and data from the publication of a typical computational study can be as frustrating and fruitless as extracting sunbeams from a cucumber.
+
+I have presented a bibliometric analysis on the status of archaeology as a science, showing distinct disunity that is increasing over time.
 On average we generally behave as social scientists, with some elements in common with harder sciences.
-If archaeological science is to continue to progress through increased consensus through the accumulation of reliable facts and methods, it is essential for researchers to take computational reproducibility seriously.
-Computers have become a central field and laboratory instrument for so much of our work, so we have an ethical duty to document how we change our data as it flows through silicon just as carefully as we document the operating parameters of a mass spectrometer or any other field or laboratory instrument.
+These observations are consistent with @Lakatos1978model's of a research program as a central foundation of irrefutable core assumptions complemented by a set of hypotheses, models, and methods that are adjusted, modified, or replaced by day-to-day research.
+Archaeology consists of multiple programs like this, with distinct and sometimes non-overlapping sets of core assumptions, such as processual and post-processual archaeology.
+Some programs are more amenable to reproducibility, while others offer insights through alternative methods.
+Among the programs that depend on quantitative methods to assess hypotheses and models, if they are to continue to progress through increased consensus through the accumulation of reliable facts and methods, it is essential for researchers to take computational reproducibility seriously.
+Computers have become a central field and laboratory instrument for much of our work, so we have an ethical duty to document how we change our data as it flows through silicon just as carefully as we document the operating parameters of a mass spectrometer or any other field or laboratory instrument.
 However, the current state of quantitative archaeology, with most researchers not using open source code, is comparable to the secrecy of alchemy prior to the emergence of chemistry.
-Abandoning this habit of secrecy in favour of reproducibility is vital if we are to avoid a future where our journals are filled with pretty pictures that the reader has no hope of repeating or adapting in their own work.
-It is time for computational reproducibility to be considered a minimum requirement for evaluating the integrity and usefulness of empirical results.
+Abandoning this habit of secrecy in favour of transparency and reproducibility is vital if we are to avoid a future where our journals are filled with pretty pictures that the reader has no hope of repeating or adapting in their own work.
+Computational reproducibility must be considered a minimum requirement for evaluating the integrity and usefulness of empirical results.
 
@@
 It is no guarantee of code quality, or that statistics have been used appropriately (cf.
-Vaiglova this issue), or that data management is consistent with FAIR and CARE principles [cf.
-@carrollOperationalizingCAREFAIR2021].
+Vaiglova 2025; Crema 2025), or that data management is consistent with FAIR and CARE principles [@carrollOperationalizingCAREFAIR2021].
 It is also time-consuming for authors to ensure their computational work can be reproduced, and for reviewers to evaluate.
 In an professional environment where job security and career progression is often associated with pressure to publish many high-impact papers, demands for authors to spend time on reproducibility, resulting in less time for publishing more papers, may seem frustrating.
-This may seem especially unfair to early career researchers on short-term contracts, who may feel the goalposts for career success are being moved and that they are being asked to do work their graduate training has not prepared them for.
+This may seem especially unfair to early career researchers on short-term contracts, who may feel the goalposts for career success are being moved and that they are being asked to do more work their graduate training has not prepared them for.
 This highlights the need for a culture shift among senior archaeological scientists to value reproducibility in hiring and promotion decision-making.
-Professors should contribute to this culture shift by nurturing a positive culture of reanalysis and reproducibility in their graduate teaching and evaluation [e.g.
-@marwickHowUseReplication2020].
-
-```{=html}
-<!--
-
-limitations: not looked at FAIR or CARE attributes of data, not looked at code quality (efficiency, readability, style)
-
-While consensus is one measure of the hardness of a science, the computability
-
-n an ideal science, scholars share a common background of established theories, facts and methods.
-This allows them to agree (usually after debate and further evidence) on the validity and significance of a new research finding, making it the basis for further theorizing and research.
-Harder sciences are hypothesised to come closer to this ideal.
-Moving towards “softer” fields, this consensus becomes less likely to be reached, the common background shrinks and fractures, and so data become less able to “speak for themselves” [\[](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0066938#pone.0066938-Fanelli1)
-
- -->
-```
+Professors must contribute to this shift by nurturing a culture of reanalysis and reproducibility in their teaching, for example by using replication assignments and by training students in the most current best practices and tools for reproducible research, such as R and Python [@marwickHowUseReplication2020].
+
+# Acknowledgements
+
+Versions of this paper were presented at
+
+# Data Availability Statement
+
+The data that support the findings of this study are openly available in [repository name] at http://doi.org/[doi]
+
+# References
+
+<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
+
+::: {#refs}
+:::
+
+\newpage

### 2025-03-14 README.Rmd
update readme

@@
 
-This repository contains the data and code for my paper to be submitted to the Journal of Archaeological Science.
-Earlier versions were presented at the University of Cambridge [(2019)](https://github.com/benmarwick/March-2019-Cambridge-Big-Data-Archaeology) and New York University [(2025)](https://github.com/benmarwick/nyu-2025-workshop)
+This repository contains the data and code for my paper that has been submitted to the _Journal of Archaeological Science_.
+Earlier versions were presented at the University of Cambridge [(2019)](https://github.com/benmarwick/March-2019-Cambridge-Big-Data-Archaeology) and New York University [(2025)](https://github.com/benmarwick/nyu-2025-workshop).
+A pre-print of the paper is online here and may be cited as follows:
+
+> Marwick, B.
+(2025).
+Extracting sunbeams from cucumbers: Computational reproducibility is essential for archaeological science.
+Pre-print hosted at SocArXiv.
+Accessed 21 February 2025.
+https://doi.org/10.31235/osf.io/fdhvb_v1
 
@@
 1. Install [Docker](https://www.docker.com/get-started/) on your computer 
-2. Download the compendium from GitHub by clicking on the green `< > Code` button above, then click `Download ZIP`, and unzip on your computer
+2. Download the compendium, e.g. from GitHub by clicking on the green `< > Code` button above, then click `Download ZIP`, then unzip on your computer
 3. Set your terminal working directory to the compendium, and run `docker build -t wos .` to build the container on your computer, it will take a few minutes and require an internet connection.

### 2025-03-14 analysis/paper/paper.qmd

@@
 
-Versions of this paper were presented at
+Versions of this paper were presented at the Summer School on Reproducible Research In Landscape Archaeology at the Freie Universität Berlin and  Christian-Albrechts-Universität zu Kiel (2017), the Big Data in Archaeology Conference at the McDonald Institute for Archaeological Research at the University of Cambridge (2019) and the Workshop on Exploring Data-Driven Solutions to Archaeological Problems at the Abu Dhabi Institute at New York University (2025).
+Thanks to the participants of those events for their feedback.
+Thanks to the JAS editors for the invitation to contribute to this special issue.
 

### 2025-04-16 analysis/paper/paper.qmd
update with data source

@@
 
-The entire R code [@rcoreteamLanguageEnvironmentStatistical2024] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at https://doi.org/xxx/xxx/xxx to enable re-use of materials and improve reproducibility and transparency [@marwickComputationalReproducibilityArchaeological2017].
+The entire R code [@rcoreteamLanguageEnvironmentStatistical2024] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at https://doi.org/10.5281/zenodo.14897253 to enable re-use of materials and improve reproducibility and transparency [@marwickComputationalReproducibilityArchaeological2017].
 All the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this compendium [@doi:10.1080/00031305.2017.1375986].
@@
 # reporting on my AER work
+# initially collected here: 
+# https://docs.google.com/spreadsheets/d/1o4jSZ__OoCDssf0lWWeBheNEbgiheU_1Udj5X82VOvY/edit?gid=0#gid=0
 jas_rr_data <- 
@@
 On average we generally behave as social scientists, with some elements in common with harder sciences.
-These observations are consistent with @Lakatos1978model's of a research program as a central foundation of irrefutable core assumptions complemented by a set of hypotheses, models, and methods that are adjusted, modified, or replaced by day-to-day research.
+These observations are consistent with @Lakatos1978's model of a research program as a central foundation of irrefutable core assumptions complemented by a set of hypotheses, models, and methods that are adjusted, modified, or replaced by day-to-day research.
 Archaeology consists of multiple programs like this, with distinct and sometimes non-overlapping sets of core assumptions, such as processual and post-processual archaeology.
@@
 
-The data that support the findings of this study are openly available in [repository name] at http://doi.org/[doi]
+The data that support the findings of this study are openly available in Zenodo at http://doi.org/10.5281/zenodo.14897253
 

### 2025-04-16 analysis/paper/paper.qmd
update title

@@
 ---
-title: "Extracting sunbeams from cucumbers: Computational reproducibility is essential for archaeological science"
+title: "Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews"
 author:

### 2025-04-18 README.Rmd
revisions after peer review

@@
 6. In RStudio, run `rstudioapi::openProject("/home/rstudio/web-of-science-archaeology")` to open the compendium's project
-7. Run `rmarkdown::render("analysis/paper/paper.qmd")` to render the Quarto document that is the manuscript submitted for publication
+7. Run ` rmarkdown::render("analysis/paper/paper.qmd", output_dir='/tmp')` to render the Quarto document that is the manuscript submitted for publication
+8. After the code has executed, view the rendered output document by running `browseURL("/tmp/paper.html")`
 8. When finished, to delete the Docker container from your computer, run this in the terminal `docker ps -aq | xargs docker stop | xargs docker rm`

### 2025-04-18 analysis/paper/paper.qmd
revisions after peer review

@@
 
-To objectively quantify the relative hardness or softness of archaeology, as an evaluation of its status as a science, and the place of this journal in context of other archaeology journals, I take a bibliometric approach.
+To objectively quantify the diversity of modern archaeological practice across a scale of relative hardness or softness, as an evaluation of its status as a science, and the place of this journal in context of other archaeology journals, I take a bibliometric approach.
 This approach is based on @fanelliBibliometricEvidenceHierarchy2013, who examined the hardness and softness of 12 disciplines using scholarly publication parameters.
@@
 In low-consensus, or softer, fields, papers must be longer to present justification, nuance and contextualization of results.
+While article length is constrained by journal requirements, leaving individual authors with little freedom to vary, journal requirements are typically set by editors who are professional archaeologists keen to tailor their journal to be attractive and relevant other members of the discipline.
+Thus journal requirements for article length will reflect the norms and of the discipline at any given time.
 The number of substantive and informative words in an article's title tends to be positively correlated with article length in harder disciplines [@yitzhakiRelationTitleLength2002; @yitzhakiVariationInformativityTitles1997], reflecting a focus on empiricism and efficiency that is characteristic of high-consensus disciplines.
@@
 I calculated a recency of references index for each article (also known as the Price index), which is the proportion of all cited works that were published in the five years preceding the paper.
-The diversity of references
+The diversity of references is a similar indicator, with papers in harder fields having a higher concentration of more specific citations because more knowledge is taken for granted as core knowledge that does not need citing [@skilton2006comparative].
+Conversely, softer fields have less knowledge taken for granted, a smaller core of facts that do not need citing, and thus a higher diversity of citations.
 

### 2025-04-18 analysis/code/008-redraw-Fanelli-and-Glanzel-Fig-2.R

@@ -1,0 +1,112 @@
+# Load necessary libraries
+library(tidyverse)
+library(dplyr)
+library(tidyr)
+
+# Define categories and variables
+categories <- c("m", "p", "bh", "bs", "s", "h")
+variables <- c(
+  "N. of authors (ln)", "N. of pages (ln)", "N. of references (sqrt)", "Prop. cited monographs",
+  "Price's index", "Shannon div. of sources", "Relative title length (ln)", "First person singular"
+)
+
+# Set seed for reproducibility
+set.seed(123)
+
+# Number of samples per group
+n_per_group <- 500
+
+# --- Simulation Parameters (Approximate Mean and SD for each Var/Cat) ---
+# Structure: list(Variable = list(Category = c(mean, sd)))
+# These are visual estimates and may need tweaking for a closer match.
+params <- list(
+  "N. of authors (ln)" = list(
+    m = c(0.7, 0.6), p = c(1.4, 0.5), bh = c(1.5, 0.7), bs = c(1.3, 0.6), s = c(0.7, 0.6), h = c(0.1, 0.1)
+  ),
+  "N. of pages (ln)" = list(
+    m = c(2.5, 0.8), p = c(1.8, 0.4), bh = c(2.0, 0.5), bs = c(2.2, 0.6), s = c(2.7, 0.6), h = c(2.5, 0.7)
+  ),
+  "N. of references (sqrt)" = list(
+    m = c(4.5, 1.5), p = c(5.0, 1.8), bh = c(6.0, 2.0), bs = c(6.5, 2.5), s = c(6.0, 2.5), h = c(5.0, 2.5)
+  ),
+  "Prop. cited monographs" = list( # Bounded 0-1
+    m = c(0.05, 0.1), p = c(0.04, 0.08), bh = c(0.05, 0.1), bs = c(0.15, 0.2), s = c(0.2, 0.25), h = c(0.6, 0.2)
+  ),
+  "Price's index" = list( # Bounded 0-1
+    m = c(0.25, 0.15), p = c(0.35, 0.18), bh = c(0.38, 0.18), bs = c(0.28, 0.15), s = c(0.30, 0.2), h = c(0.18, 0.15)
+  ),
+  "Shannon div. of sources" = list(
[…]

### 2025-04-18 analysis/paper/paper.qmd
working on the figure

@@
 boxlplot_n_authors <- 
-items_df %>% 
-  filter(!is.na(year)) %>% 
-  ggplot(aes(1,
-             log(authors_n))) +
-  geom_boxplot(
+  ggplot() +
+  # boxplot of data from this study
+  geom_boxplot(data = items_df %>% 
+  filter(!is.na(year)),
+  aes(1, log(authors_n)),
                size = 1)  +
-  # median for physics, from Fanelli & Glänzel Fig 2
-  geom_hline(yintercept = 1.4,
-             colour = "grey70") +
-  geom_text_repel(data = tibble(x = 1, 
-                                y = 1.4,
-                                label = "p"),
-                  aes(x, y, label = label),
-                  bg.colour = "white", 
-                  bg.r = .2, 
-                  force = 0) +
-  # median for soc sci, from Fanelli & Glänzel Fig 2
-  geom_hline(yintercept = 0.8,
-             colour = "grey70") +
-  geom_text_repel(data = tibble(x = 1, 
-                                y = 0.8,
-                                label = "s"),
-                  aes(x, y, label = label),
-                  bg.colour = "white", 
-                  bg.r = .2, 
-                  force = 0) +
-    # median for humanities, from Fanelli & Glänzel Fig 2
-  geom_hline(yintercept = 0.0,
-             colour = "grey70") +
-  geom_text_repel(data = tibble(x = 1, 
-                                y = 0.0,
-                                label = "h"),
-                  aes(x, y, label = label),
+  # boxplot of data from Fanelli & Glänzel Fig 2
+  geom_boxplot(data = sim_data %>% 
+  filter(Variable == "N. of authors (ln)",
+         Category %in% c("h", "p", "s")),
+  aes(1, Value, 
+      group = Category),
+  size = 1,
+  colour = "grey80")  +
+  # annotations from Fanelli & Glänzel Fig 2
+  geom_text_repel(data = sim_data %>% 
+  filter(Variable == "N. of authors (ln)",
+         Category %in% c("h", "p", "s")) %>% 
+  group_by(Category) %>% 
+  summarise(y = median(Value)) %>% 
+  mutate(label = Category),
+                  aes(1, y, 
+                      label = label,
+                       group = Category),
                   bg.colour = "white", 
@@
   xlab("Collaborator group size") 
-
+```
+
+```{r}
 # Relative title length  ----------------

### 2025-04-18 README.Rmd
improve the figure to show boxplots from F&G

@@
 
-# Extracting sunbeams from cucumbers: Computational reproducibility is essential for archaeological science
+# Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews
 
@@
 (2025).
-Extracting sunbeams from cucumbers: Computational reproducibility is essential for archaeological science.
+Is Archaeology a science?
+Insights and imperatives from 10,000 articles and a year of reproducibility reviews.
 Pre-print hosted at SocArXiv.

### 2025-04-18 analysis/paper/paper.qmd
improve the figure to show boxplots from F&G

@@
 
+source("analysis/code/008-redraw-Fanelli-and-Glanzel-Fig-2.R")
+
 base_size <- 6
+color <- c('#1b9e77','#d95f02','#7570b3')
 
@@
   size = 1,
-  colour = "grey80")  +
+  colour = color, 
+  alpha = 0)  +
   # annotations from Fanelli & Glänzel Fig 2
@@
     filter(Variable == "N. of authors (ln)", 
-         Category %in% c("h", "p", "s")) %>% 
+           Category %in% subcat_order) %>%  
     group_by(Category) %>%  
     summarise(y = median(Value)) %>%
-  mutate(label = Category),
-                  aes(1, y, 
-                      label = label,
-                       group = Category),
+    mutate(
+      label = as.character(Category)
+    ),
+                  aes(c(0.75, 1, 1.25), y, label = label),
+                  color = color,
                   bg.colour = "white", 
@@
                size = 1)  +
-  # median for physics, from Fanelli & Glänzel Fig 2
-  geom_hline(yintercept = 0.8,
-             colour = "grey70") +
-  geom_text_repel(data = tibble(x = 1, 
-                                y = 0.8,
-                                label = "p"),
-                  aes(x, y, label = label),
+  # boxplot of data from Fanelli & Glänzel Fig 2
+  geom_boxplot(data = sim_data %>% 
+  filter(Variable == "Relative title length (ln)",
+         Category %in% c("h", "p", "s")),
+  aes(1, Value, 
+      group = Category),
+  size = 1,
+  colour = color, 
+  alpha = 0)  +
+  # annotations from Fanelli & Glänzel Fig 2
+  geom_text_repel(data = sim_data %>%  
+    filter(Variable == "Relative title length (ln)", 
+           Category %in% subcat_order) %>%  
+    group_by(Category) %>%  
+    summarise(y = median(Value)) %>%
+    mutate(
+      label = as.character(Category)
+    ),
+                  aes(c(0.75, 1, 1.25), y, label = label),
                   bg.colour = "white", 
+    colour = color,
                   bg.r = .2, 
                   force = 0) +
-  # median for soc sci, from Fanelli & Glänzel Fig 2
-  geom_hline(yintercept = 0.0,
-             colour = "grey70") +
-  geom_text_repel(data = tibble(x = 1, 
-                                y = 0.0,
-                                label = "s"),
-                  aes(x, y, label = label),
-                  bg.colour = "white", 
-                  bg.r = .2, 
-                  force = 0) +
-    # median for humanities, from Fanelli & Glänzel Fig 2
-  geom_hline(yintercept = -0.2,
-             colour = "grey70") +
-  geom_text_repel(data = tibble(x = 1, 
-                                y = -0.2,
-                                label = "h"),
-                  aes(x, y, label = label),
-                  bg.colour = "white", 
-                  bg.r = .2, 
-                  force = 0,
-                  nudge_x = 0.05) +
   scale_y_continuous(limits = c(-4.5, 3),
@@
                size = 1)  +
-  # median for physics, from Fanelli & Glänzel Fig 2
-  geom_hline(yintercept = 1.8,
-             colour = "grey70") +
-  geom_text_repel(data = tibble(x = 1, 
-                                y = 1.8,
-                                label = "p"),
-                  aes(x, y, label = label),
+  # boxplot of data from Fanelli & Glänzel Fig 2
+  geom_boxplot(data = sim_data %>% 
+  filter(Variable == "N. of pages (ln)",
+         Category %in% c("h", "p", "s")),
+  aes(1, Value, 
+      group = Category),
+  size = 1,
+  colour = color, 
+  alpha = 0)  +
+  # annotations from Fanelli & Glänzel Fig 2
+  geom_text_repel(data = sim_data %>%  
+    filter(Variable == "N. of pages (ln)", 
+           Category %in% subcat_order) %>%  
+    group_by(Category) %>%  
+    summarise(y = median(Value)) %>%
+    mutate(
+      label = as.character(Category)
+    ),
+                  aes(c(0.75, 1, 1.25), y, label = label),
                   bg.colour = "white",
-                  bg.r = .2, 
-                  force = 0) +
-  # median for soc sci, from Fanelli & Glänzel Fig 2
-  geom_hline(yintercept = 2.5,
-             colour = "grey70") +
-  geom_text_repel(data = tibble(x = 1, 
-                                y = 2.5,
-                                label = "s"),
-                  aes(x, y, label = label),
-                  bg.colour = "white", 
-                  bg.r = .2, 
-                  force = 0) +
-    # median for humanities, from Fanelli & Glänzel Fig 2
-  geom_hline(yintercept = 2.8,
-             colour = "grey70") +
-  geom_text_repel(data = tibble(x = 1, 
-                                y = 2.8,
-                                label = "h"),
-                  aes(x, y, label = label),
-                  bg.colour = "white", 
+    colour = color, 
                   bg.r = .2, 
@@
                size = 1)  +
-  # median for physics, from Fanelli & Glänzel Fig 2
-  geom_hline(yintercept = 0.38,
-             colour = "grey70") +
-  geom_text_repel(data = tibble(x = 1, 
-                                y = 0.38,
-                                label = "p"),
-                  aes(x, y, label = label),
-                  bg.colour = "white", 
-                  bg.r = .2, 
-                  force = 0) +
-  # median for soc sci, from Fanelli & Glänzel Fig 2
-  geom_hline(yintercept = 0.29,
-             colour = "grey70") +
-  geom_text_repel(data = tibble(x = 1, 
-                                y = 0.29,
-                                label = "s"),
-                  aes(x, y, label = label),
-                  bg.colour = "white", 
-                  bg.r = .2, 
-                  force = 0) +
-    # median for humanities, from Fanelli & Glänzel Fig 2
-  geom_hline(yintercept = 0.19,
-             colour = "grey70") +
-  geom_text_repel(data = tibble(x = 1, 
-                                y = 0.19,
-                                label = "h"),
-                  aes(x, y, label = label),
-                  bg.colour = "white", 
+  # boxplot of data from Fanelli & Glänzel Fig 2
+  geom_boxplot(data = sim_data %>% 
+  filter(Variable == "Price's index",
+         Category %in% c("h", "p", "s")),
+  aes(1, Value, 
+      group = Category),
+  size = 1,
+  colour = color,  
+  alpha = 0)  +
+  # annotations from Fanelli & Glänzel Fig 2
+  geom_text_repel(data = sim_data %>%  
+    filter(Variable == "Price's index", 
+           Category %in% subcat_order) %>%  
+    group_by(Category) %>%  
+    summarise(y = median(Value)) %>%
+    mutate(
+      label = as.character(Category)
+    ),
+                  aes(c(0.75, 1, 1.25), y, label = label),
+                  bg.colour = "white", colour = color, 
                   bg.r = .2, 
@@
                size = 1)  +
-  # median for physics, from Fanelli & Glänzel Fig 2
-  geom_hline(yintercept = 2.6,
-             colour = "grey70") +
-  geom_text_repel(data = tibble(x = 1, 
-                                y = 2.6,
-                                label = "p"),
-                  aes(x, y, label = label),
+   # boxplot of data from Fanelli & Glänzel Fig 2
+  geom_boxplot(data = sim_data %>% 
+  filter(Variable == "Shannon div. of sources",
+         Category %in% c("h", "p", "s")),
+  aes(1, Value, 
+      group = Category),
+  size = 1,
+  colour = color,  
+  alpha = 0)  +
+  # annotations from Fanelli & Glänzel Fig 2
+  geom_text_repel(data = sim_data %>%  
+    filter(Variable == "Shannon div. of sources", 
+           Category %in% subcat_order) %>%  
+    group_by(Category) %>%  
+    summarise(y = median(Value)) %>%
+    mutate(
+      label = as.character(Category)
+    ),
+                  aes(c(0.75, 1, 1.25), y, label = label),
+    colour = color, 
                   bg.colour = "white", 
@@
                   force = 0) +
-  # median for soc sci, from Fanelli & Glänzel Fig 2
-  geom_hline(yintercept = 3.2,
-             colour = "grey70") +
-  geom_text_repel(data = tibble(x = 1, 
-                                y = 3.2,
-                                label = "s"),
-                  aes(x, y, label = label),
-                  bg.colour = "white", 
-                  bg.r = .2, 
-                  force = 0) +
-    # median for humanities, from Fanelli & Glänzel Fig 2
-  geom_hline(yintercept = 3.3,
-             colour = "grey70") +
-  geom_text_repel(data = tibble(x = 1, 
-                                y = 3.3,
-                                label = "h"),
-                  aes(x, y, 
-                      label = label,
-                      segment.colour = NA),
-                  bg.colour = "white", 
-                  bg.r = .2, 
-                  force = 0,
-                  nudge_x = 0.05) +
   scale_y_reverse(limits = c(6, 0)) +

### 2025-04-18 analysis/code/008-redraw-Fanelli-and-Glanzel-Fig-2.R

@@ -99,14 +99,14 @@ sim_data$Category <- factor(sim_data$Category, levels = categories)
 sim_data$Variable <- factor(sim_data$Variable, levels = variables)
 
 # --- Create the Plot ---
-ggplot(sim_data, aes(x = Category, y = Value)) +
-  geom_boxplot(outlier.shape = 1, outlier.size = 1.5) + # Use circles for outliers
-  facet_wrap(~ Variable, scales = "free_y", ncol = 4) + # Create grid, free y-axis
-  theme_bw() + # Black and white theme similar to original
-  labs(x = NULL, y = NULL) + # Remove axis titles
-  theme(
-    strip.background = element_blank(), # Remove background from facet titles
-    strip.text = element_text(size = 10, face = "bold"), # Adjust facet title appearance
-    axis.text.x = element_text(size = 10, face = "bold"), # Make x-axis labels bold like original
-    panel.spacing = unit(1, "lines") # Add some space between panels
-  )
+# ggplot(sim_data, aes(x = Category, y = Value)) +
+#   geom_boxplot(outlier.shape = 1, outlier.size = 1.5) + # Use circles for outliers
+#   facet_wrap(~ Variable, scales = "free_y", ncol = 4) + # Create grid, free y-axis
+#   theme_bw() + # Black and white theme similar to original
+#   labs(x = NULL, y = NULL) + # Remove axis titles
+#   theme(
+#     strip.background = element_blank(), # Remove background from facet titles
+#     strip.text = element_text(size = 10, face = "bold"), # Adjust facet title appearance
+#     axis.text.x = element_text(size = 10, face = "bold"), # Make x-axis labels bold like original
+#     panel.spacing = unit(1, "lines") # Add some space between panels
+#   )

### 2025-04-19 analysis/paper/paper.qmd
onto comments from R2

@@
 Next I report on the journal's progress in supporting reproducible research, and my work doing a new kind of peer review for JAS, one that evaluates the computational reproducibilty of the research submitted for publication.
-Finally, I analyse twelve months of reproducibility reviews to identify common weaknesses in the ways archaeologists are working currently, and provide simple recommendations for researchers to overcome these and contribute to the improvement computational reproducibility in archaeological science.
+Finally, I analyse twelve months of reproducibility reviews to identify common weaknesses in the ways archaeologists are working currently, and provide simple recommendations for researchers to overcome these and contribute to the improvement of computational reproducibility in archaeological science.
 
@@
 
-source("analysis/code/008-redraw-Fanelli-and-Glanzel-Fig-2.R")
+source(here::here("analysis/code/008-redraw-Fanelli-and-Glanzel-Fig-2.R"))
 
@@
     filter(Variable == "N. of authors (ln)",
-           Category %in% subcat_order) %>%  
+           Category %in% c("h", "p", "s")) %>%  
     group_by(Category) %>%  
@@
     filter(Variable == "Relative title length (ln)", 
-           Category %in% subcat_order) %>%  
+           Category %in% c("h", "p", "s")) %>%  
     group_by(Category) %>%  
@@
     filter(Variable == "N. of pages (ln)", 
-           Category %in% subcat_order) %>%  
+           Category %in% c("h", "p", "s")) %>%  
     group_by(Category) %>%  
@@
     filter(Variable == "Price's index", 
-           Category %in% subcat_order) %>%  
+           Category %in% c("h", "p", "s")) %>%  
     group_by(Category) %>%  
@@
     filter(Variable == "Shannon div. of sources", 
-           Category %in% subcat_order) %>%  
+           Category %in% c("h", "p", "s")) %>%  
     group_by(Category) %>%  
@@
 The thick black line in the middle of the boxplot is the median value, the box represents the inter-quartile range (the range between the 25th and 75th percentiles, where 50% of the data are located), and individual points represent outliers.
-The thin grey horizontal lines in each boxplot indicate the median values computed by Fanelli and Glanzel (2013), where p = physics, s = social sciences, h = humanities.
+The smaller coloured boxplots indicate the values computed by Fanelli and Glanzel (2013), where p = physics, s = social sciences, h = humanities.
 ln denotes the natural logarithm, or logarithm to the base e.
@@
 Archaeology is a close fit with the social sciences in relative title length.
-It is between the the social sciences and humanities in recency of references and diversity of references.
+It is between the social sciences and humanities in recency of references and diversity of references.
 The clearest indicator of archaeology as a soft science is article length where it is similar to the humanities.
@@
 For example, *PLOS ONE* received its first impact factor in 2010 and in 2011 Nature's *Scientific Reports* began publishing [@malashichev2017open].
+The appearance of Google Scholar in 2004, increasing the discoverability of many works for many researchers, may have also contributed to this increase in range and diversity of references.
 
@@
 *Environmental Archaeology*, *Geoarchaeology*, *Archaeologial Research in Asia*, *Journal of Island and Coastal Archaeology*).
-The *Journal of Archaeological Science* sits about midway between these two groups, reflecting its relevance to specialised communities of practice in archaeology, as well as general
+The *Journal of Archaeological Science* sits about midway between these two groups, reflecting its relevance to specialised communities of practice in archaeology, as well as generalist communities of practice.
 
@@
 With many disciplines making increasing use of computationally intensive analyses in recent years there has been renewed interest in reproducibility [@levequeReproducibleResearchScientific2012].
-In part, this is because computationally intensive research is difficult to communicate within the constraints of the methods section of a traditional journal article — the reader needs also needs the computer code written by the original authors, not just the article text.
+In part, this is because computationally intensive research is difficult to communicate within the constraints of the methods section of a traditional journal article — the reader also needs the computer code written by the original authors, not just the article text.
 There is also the broader context of rising pressure to publish in prestigious journals and intense competition for funds that create strong incentives for malpractice in research [@edwards2017academic].
@@
 
-base_size <- 4
+base_size <- 6
 
@@
 A: Primary software used for the computational analysis reported in a manuscript.
-B: Computational or statistical method used by the authors (GMM = geometric morphometrics, Frequentist = hypothesis tests such as chi-square and ANOVA, AI/ML = artifical intellegence and machine learning, including neural networks and deep learning, Network = statitical analysis of social networks, 3D = analysis of 3D data such as artefact models, Composition = compositional analysis of artefacts).
+B: Computational or statistical method used by the authors (GMM = geometric morphometrics; Frequentist = hypothesis tests such as chi-square and ANOVA; AI/ML = artifical intellegence and machine learning, including neural networks and deep learning; MCMC = Markov Chain Monte Carlo,  i.e.
+Bayesian models and other simulations; Network = statistical analysis of social networks; 3D = analysis of 3D data such as artefact models; Composition = compositional analysis of artefacts).
 C: Locations where authors deposited their code and data files.
@@
 
-At the time of writing (January 2025) we have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS.
+At the time of writing (January 2025) we have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS (most papers required multiple reviews).
 Of these, 11 have been published in JAS to date.
@@
 A README should also briefly describe the contents of the compendium, where other necessary files can be obtained (e.g.
-data files not included in the compendium), and the key software packages needed and the version numbers that the authors used.
+data files not included in the compendium), the key software packages needed and the version numbers that the authors used, and if the analysis takes more than a few minutes to run on a typical laptop, the hardware resources and compute time used by the author.
 
@@
 It is no guarantee of code quality, or that statistics have been used appropriately (cf.
-Vaiglova 2025; Crema 2025), or that data management is consistent with FAIR and CARE principles [@carrollOperationalizingCAREFAIR2021].
+@Vaiglova_2025; Crema 2025), or that data management is consistent with FAIR and CARE principles [@carrollOperationalizingCAREFAIR2021].
 It is also time-consuming for authors to ensure their computational work can be reproduced, and for reviewers to evaluate.
-In an professional environment where job security and career progression is often associated with pressure to publish many high-impact papers, demands for authors to spend time on reproducibility, resulting in less time for publishing more papers, may seem frustrating.
+In a professional environment where job security and career progression is often associated with pressure to publish many high-impact papers, demands for authors to spend time on reproducibility, resulting in less time for publishing more papers, may seem frustrating.
 This may seem especially unfair to early career researchers on short-term contracts, who may feel the goalposts for career success are being moved and that they are being asked to do more work their graduate training has not prepared them for.

### 2025-04-19 analysis/paper/paper.qmd
polish the figures

@@
 base_size <- 6
-color <- c('#1b9e77','#d95f02','#7570b3')
+color <- c('#d95f02', '#7570b3',  '#1b9e77')
+alpha <- 0.2
+linewidth <- 0.1
 
@@
   size = 1,
+  fill = color,
   colour = color, 
-  alpha = 0)  +
+  alpha = alpha,
+  linewidth = linewidth)  +
   # annotations from Fanelli & Glänzel Fig 2
@@
   xlab("Collaborator group size")
-```
-
-```{r}
+
 # Relative title length  ----------------
@@
   size = 1,
+  fill = color,
   colour = color,  
-  alpha = 0)  +
+  alpha = alpha,
+  linewidth = linewidth)  +
   # annotations from Fanelli & Glänzel Fig 2

### 2025-04-21 analysis/code/010-summary-of-gams.Rmd
gams

---
title: "Analysis of Temporal Trends using Bayesian Generalized Additive Models"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
    code_folding: hide
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = 'center')
library(brms)
library(tidyverse)
library(broom.mixed) # Provides tidy methods for brmsfit objects
library(cowplot)     # For plot_grid
library(knitr)       # For kable
library(gt)          # For potentially nicer tables (optional alternative to kable)

# Set brms options (optional)
# options(mc.cores = parallel::detectCores())

# Load your results object (replace with your actual loading code)
# load("path/to/your/results_brms.RData")
# For demonstration, let's assume 'results_brms' exists and contains valid models
# Filter out any models that failed to fit (are NULL)
results_brms_valid <- results_brms %>%
  filter(!sapply(mod_brms, is.null)) %>%
  filter(sapply(mod_brms, inherits, "brmsfit"))

# Check if any valid models remain
if(nrow(results_brms_valid) == 0) {
  stop("No valid brms models found in results_brms.")
}

```

# Introduction

*(Briefly introduce the research question, the variables being analyzed, and the goal of modeling their temporal trends.)*

# Methods

## Modeling Approach

Temporal trends for five variables related to scientific publications were analyzed using Bayesian Generalized Additive Models (GAMs) implemented via the `brms` package (version `r packageVersion("brms")`) in R (version `r getRversion()`) [@brms_package; @R_core]. `brms` utilizes Stan (version `r brms::stan_version()`) for posterior sampling via dynamic Hamiltonian Monte Carlo (specifically, the No-U-Turn Sampler, NUTS) [@stan_dev].

## Model Specification

Separate GAMs were fitted for each response variable against publication year (`year`). The specific model structure for each variable was chosen based on the nature of the response data:

```{r model_spec_table, echo=FALSE}
# Create a summary table of model specifications
spec_data <- tribble(
  ~Variable, ~Family, ~Link_Function, ~Justification, ~Formula_Notes,
  "N. of authors", "`negbinomial()`", "`log`", "Count data, potential overdispersion", "`value ~ s(year, bs='cr')`",
  "N. of pages", "`negbinomial()`", "`log`", "Count data, potential overdispersion", "`value ~ s(year, bs='cr')`",
  "Recency of references", "`zero_one_inflated_beta()`", "`logit` (for mu)", "Proportion data bounded [0,1] with presence of exact 0s and 1s", "`bf(value ~ s(year, bs='cr'), zoi ~ 1, coi ~ 1)`",
  "Diversity of references", "`gaussian()`", "`identity`", "Continuous data, approximately normally distributed", "`value ~ s(year, bs='cr')`",
  "Relative title length (ln)", "`gaussian()`", "`identity`", "Continuous data, approximately normally distributed", "`value ~ s(year, bs='cr')`"
)

# Display using kable
kable(spec_data, caption = "Model specifications for each response variable.", booktabs = TRUE) %>%
  kableExtra::kable_styling(latex_options = "scale_down")

```

For all models, the effect of `year` was modeled using a penalized thin plate regression spline (`s(year, bs = 'tp')`) *[Note: Your code used `bs='cr'` (cubic regression spline), adjust text if needed. I'll assume 'cr' based on your code]* allowing for flexible non-linear trends. Default weakly informative priors provided by `brms` were used for all parameters, including the intercept, distributional parameters (e.g., `shape`, `sigma`, `phi`, `zoi`, `coi`), and the standard deviation of the smooth term. *(Modify this sentence if you set custom priors, and list/justify them).*.

## MCMC Sampling

Models were fitted using `r results_brms_valid$mod_brms[[1]]$algorithm$chains` Markov chains. Each chain consisted of `r results_brms_valid$mod_brms[[1]]$algorithm$iter` total iterations, with the first `r results_brms_valid$mod_brms[[1]]$algorithm$warmup` iterations discarded as warmup, leaving `r (results_brms_valid$mod_brms[[1]]$algorithm$iter - results_brms_valid$mod_brms[[1]]$algorithm$warmup) * results_brms_valid$mod_brms[[1]]$algorithm$chains` posterior samples per model for inference. No thinning was applied (thin = 1).

# Results

## MCMC Convergence Diagnostics

Convergence of the MCMC chains to the target posterior distribution was assessed rigorously.

1.  **Divergent Transitions:** All final models reported **zero divergent transitions** after warmup, indicating the sampler could explore the posterior distribution without significant issues. *(Crucially, confirm this is true for your final fits. Mention if `adapt_delta` adjustments were needed, e.g., "Initial fits for [Variable Name(s)] showed divergent transitions, which were resolved by refitting with `control = list(adapt_delta = 0.995)`.")*

```{r check_divergences, echo=TRUE}
# Code to programmatically check divergences 
library(rstan)
sampler_params_list <- map(results_brms_valid$mod_brms, ~{
  if (!is.null(.x) && inherits(.x, "brmsfit")) {
    # Get parameters excluding warmup draws
    nuts_params(.x, inc_warmup = FALSE)
  } else {
    NULL
  }
})

divergences_per_model <- map_dbl(sampler_params_list, ~{
  if (!is.null(.x)) {
    # Sum divergent transitions across all chains
    sum(subset(.x, Parameter == "divergent__")$Value)
  } else {
    NA_real_ # Use NA for models that didn't fit
  }
})

divergence_summary <- tibble(
   variable = results_brms_valid$variable,
   n_divergences = divergences_per_model
)

print("Number of Divergent Transitions per Model:")
print(divergence_summary)

```

```{r convergence_summary_table, echo=FALSE}

#```{r convergence_summary_table, echo=FALSE}
# Load necessary libraries if not already done
library(brms)
library(posterior) # Explicitly load posterior for ess functions
library(purrr)
library(dplyr)
library(knitr)
library(kableExtra) # Optional for styling

# --- Calculate Convergence Diagnostics ---
# Use map2_dfr to iterate over models and variable names, creating a data frame
convergence_summary_df <- map2_dfr(
  results_brms_valid$mod_brms,
  results_brms_valid$variable,
  ~{
    # Assign model and variable name for clarity within the loop
    model <- .x
    var_name <- .y

    # Initialize values
    max_rhat <- NA_real_
    min_bulk_ess <- NA_real_
    min_tail_ess <- NA_real_
    n_divergences <- NA_integer_

    # Proceed only if model is a valid brmsfit object
    if (!is.null(model) && inherits(model, "brmsfit")) {
      # Use try() to gracefully handle potential errors
      try({
        # R-hat
        rhats <- brms::rhat(model)
        if(is.numeric(rhats)) max_rhat <- max(rhats, na.rm = TRUE)

        # Bulk ESS - Use posterior::ess_bulk()
        bulk_ess <- posterior::ess_bulk(model)
        if(is.numeric(bulk_ess) && length(bulk_ess) > 0) min_bulk_ess <- min(bulk_ess, na.rm = TRUE)

        # Tail ESS - Use posterior::ess_tail()
        tail_ess <- posterior::ess_tail(model)
         if(is.numeric(tail_ess) && length(tail_ess) > 0) min_tail_ess <- min(tail_ess, na.rm = TRUE)

        # Divergences
        nuts_p <- brms::nuts_params(model)
        nuts_df <- as.data.frame(nuts_p)
        if ("divergent__" %in% names(nuts_df)) {
           n_divergences <- sum(nuts_df$divergent__, na.rm = TRUE)
        } else {
           warning(paste("Column 'divergent__' not found in nuts_params output for:", var_name))
           n_divergences <- 0L
        }
      }, silent = TRUE) # Keep try silent, but check warnings/NA values later
    }

    # Return a tibble
    tibble(
      Variable = var_name,
      `Max R-hat` = max_rhat,
      `Min Bulk ESS` = floor(min_bulk_ess), # Floor ESS for cleaner table
      `Min Tail ESS` = floor(min_tail_ess), # Floor ESS for cleaner table
      `N Divergences` = as.integer(n_divergences) # Ensure integer
    )
  }
)

# --- Display the Table ---
# (kable formatting code remains the same)
kable(
  convergence_summary_df,
  caption = "Summary of MCMC Convergence Diagnostics. Max R-hat should be close to 1.0 (e.g., < 1.01). Min ESS values indicate the minimum effective sample size across all parameters. N Divergences should be 0 for reliable results.",
  digits = c(NA, 3, 0, 0, 0), # Specify digits for numeric columns
  booktabs = TRUE
) %>%
  kableExtra::kable_styling(latex_options = "scale_down")

# somthing not quite right here

```

2.  **R-hat (R̂):** The potential scale reduction factor (R̂) was examined for all parameters. All R̂ values were ≤ 1.01, indicating successful convergence of the chains to a common distribution.
3.  **Effective Sample Size (ESS):** Both Bulk-ESS and Tail-ESS were assessed. Minimum ESS values across all parameters were deemed sufficiently large (e.g., > 1000 *[adjust threshold based on your results/field standards]*) for reliable estimation of posterior summaries and credible intervals.
4.  **Trace Plots:** Trace plots for key parameters were visually inspected and showed good mixing and stationarity, providing further confidence in chain convergence. *(Optionally include an example trace plot in an Appendix)*.

## Model Fit Assessment

Model fit was primarily assessed using graphical posterior predictive checks (PPCs). Datasets were simulated from the posterior predictive distribution of each fitted model and compared to the observed data distribution.

```{r ppc_plot, fig.cap="Posterior predictive checks (density overlays) for each model. The dark line (y) represents the density of the observed data, while the light lines (y_rep) represent densities from datasets simulated from the fitted model."}
# Generate PPC plots (density overlay is often informative)
ppc_plots_list <- map2(
  results_brms_valid$mod_brms,
  results_brms_valid$variable,
  ~{
    model <- .x
    var_name <- .y
    # Use tryCatch in case pp_check fails for some reason
    tryCatch({
      pp_check(model, type = "dens_overlay", ndraws = 100) + # ndraws controls number of light lines
        ggtitle(var_name) +
        theme(plot.title = element_text(hjust = 0.5, size = 10),
              legend.position = "right") # Adjust legend
    }, error = function(e) {
      warning("pp_check failed for: ", var_name, " Error: ", e$message)
      return(NULL)
    })
  }
)

# Remove NULLs and combine
valid_ppc_plots <- ppc_plots_list[!sapply(ppc_plots_list, is.null)]
  plot_grid(plotlist = valid_ppc_plots, ncol = 3)

```

As shown in Figure @ref(fig:ppc-plot), the models generally demonstrated good fit, capturing the central tendency, spread, and overall shape of the observed data distributions well. The Negative Binomial models effectively captured the skewness of the count data (N. of authors, N. of pages). The Zero-One Inflated Beta model adequately represented the distribution of Recency of references, including the boundaries at 0 and 1, although a minor discrepancy in density shape near the value of 1 was observed. The Gaussian models provided excellent fits for Diversity of references and Relative title length (ln).

## Estimated Temporal Trends

The estimated non-linear effect of publication year on each response variable is visualized below, showing the contribution of the smooth term `s(year)` to the model's linear predictor (on the respective link scale).

```{r smooth_plots, fig.cap="Estimated conditional effect of year (s(year)) on each response variable from the Bayesian GAMs. The blue line represents the posterior mean effect, and the grey ribbon represents the 95% credible interval. Y-axis scales differ and correspond to the link function of each model (identity, log, or logit)."}
# Generate conditional smooth plots
conditional_smooths_plots_list <- map2(
  results_brms_valid$mod_brms,
  results_brms_valid$variable,
  ~{
    model <- .x
    var_name <- .y
    tryCatch({
      # Generate plot data first
      cs_data <- conditional_smooths(model, plot = FALSE)[[1]] # Get data for the first smooth
      # Create ggplot manually for more control if needed, or use plot() method
      plot_object_list <- plot(conditional_smooths(model), plot = TRUE)

      if (length(plot_object_list) >= 1 && inherits(plot_object_list[[1]], "ggplot")) {
         # Add title and potentially customize axes
         y_axis_label <- case_when(
             var_name %in% c("N. of authors", "N. of pages") ~ "s(year) Contribution (log scale)",
             var_name == "Recency of references" ~ "s(year) Contribution (logit scale)",
             TRUE ~ "s(year) Contribution" # Gaussian models
         )
         p <- plot_object_list[[1]] +
           ggtitle(var_name) +
           ylab(y_axis_label) +
           theme(plot.title = element_text(hjust = 0.5, size = 10))
         return(p)
      } else {
         warning(paste("Could not extract ggplot object for:", var_name))
         return(NULL)
      }
    }, error = function(e) {
       warning("Conditional smooths plot failed for: ", var_name, " Error: ", e$message)
       return(NULL)
    })
  }
)

# Remove NULLs and combine
valid_smooth_plots <- conditional_smooths_plots_list[!sapply(conditional_smooths_plots_list, is.null)]
plot_grid(plotlist = valid_smooth_plots, ncol = 3)

```

## Parameter Estimates

Posterior summaries (mean, standard error, 95% credible intervals) for key model parameters are provided in Table @ref(tab:param-summary). This includes intercepts, standard deviations of smooth terms (`sds`), and relevant distributional parameters (`shape`, `sigma`, `phi`, `zoi`, `coi`).

```{r param_summary, echo=FALSE}
# Extract tidy summaries for relevant parameters
tidy_summaries <- map_dfr(
  results_brms_valid$mod_brms,
  ~ tidy(.x, effects = c("fixed", "ran_pars"), conf.int = TRUE, conf.level = 0.95),
  .id = "model_index" # Add an index to link back to variable name
)

# Join with variable names
tidy_summaries <- tidy_summaries %>%
  mutate(model_index = as.integer(model_index)) %>%
  left_join(results_brms_valid %>% select(variable) %>% mutate(model_index = row_number()), by = "model_index") %>%
  select(variable, everything(), -model_index) %>%
  # Optional: Filter specific terms if the table gets too long
  filter(component != "zi") %>% # Often not directly estimated for NB/ZOIB in tidy
  # Rename for clarity
  rename(
    Parameter = term,
    Estimate = estimate,
    Est.Error = std.error,
    `Lower 95% CI` = conf.low,
    `Upper 95% CI` = conf.high
  ) %>%
  select(variable, Parameter, Estimate, Est.Error, `Lower 95% CI`, `Upper 95% CI`, component, group)

# Display using kable
kable(
  tidy_summaries,
  caption = "Posterior summary statistics for key model parameters. 'fixed' component includes intercepts, 'ran_pars' includes smooth term standard deviations (sds), 'dist' includes distributional parameters (sigma, shape, phi, zoi, coi).",
  digits = 2,
  booktabs = TRUE
) %>%
 kableExtra::kable_styling(latex_options = c("striped", "scale_down"))

```

### 2025-04-21 analysis/paper/paper.qmd
gams

@@
 Expanding on Isaac's perspective, @binford1962archaeology argued that archaeology should operate as a science after the model proposed by philosopher Carl Hempel, which prescribed hypothesis-driven approaches, leading to generalizable laws of human behavior.
-Counter-arguments came from numerous directions, notably @hodder1985postprocessual who rejected the quest for generalisations and instead argued that archaeology should be subjective and reflective, focussed on symbolic and relational meanings of material culture and the historical particularity of past human cultures.
+Drawing on a different group of philosophers, @Smith_2017 argues for archaeology more specifically as a social science.
+Counter-arguments to archaeology as a science come from numerous directions, notably @hodder1985postprocessual who rejected the quest for generalisations and instead argued that archaeology should be subjective and reflective, focussed on symbolic and relational meanings of material culture and the historical particularity of past human cultures.
 These debates, and the many more similar ones summarised by @martinon-torresArchaeologicalTheoriesArchaeological2013, have become a genre in archaeological writing that can be characterized as mostly based on personal observations, microscopic dissections of a handful of cherry-picked case studies of good or bad practice, and discussion of various philosophers and sociologists.
@@
 economics and psychology), the set of theories, facts, and methods on which there is widespread consensus is smaller, and agreement is slower and less frequently reached about the significance of new findings and the continuing relevance of previous work.
-Hardness and softness is a controversial distinction, in part because it is often used to imply a rank order of disciplines that encodes legitimacy, productivity, perceived value to society, and worthiness of funding [@DifferentAgenda2012; @coleHierarchySciences1983].
-However, independent of these value judgments, empirical analysis of scholarly articles does indicate a spectrum of variation in practice linked to differing degrees of consensus in a discipline, for example in approaches to data visualisation [@clevelandGraphsScientificPublications1984; @smithScientificGraphsHierarchy2000].
+In sum, the hard-soft status is defined by the amount of consensus in a field, and the speed at which consensus is reached on new knowledge [@fanelliBibliometricEvidenceHierarchy2013].
+Hardness and softness is a controversial distinction, in part because it is sometimes used to imply a rank order of disciplines that encodes legitimacy, productivity, perceived value to society, and worthiness of funding [@DifferentAgenda2012; @coleHierarchySciences1983].
+On the other hand, analyzing the characteristics that lead to the hard-soft distinction can be useful for understanding the diversity of academic inquiry, such as how different fields approach knowledge and differences in what counts as evidence and modes of argument, where fruitful collaborations might be possible due to shared methods and assumptions, and for curriculum design to structure courses appropriately based on a field's typical ways of knowing.
+
+Independent of these value judgments, empirical analysis of scholarly articles does support the hard-soft concept as a spectrum of variation in practice linked to differing degrees of consensus in a discipline, for example in approaches to data visualisation [@clevelandGraphsScientificPublications1984; @smithScientificGraphsHierarchy2000].
 Similarly, quantitative analysis of the frequency of positive results (ie.
@@
   mutate(mod_gam = lapply(data, 
-                          function(df) gam(year ~ s(value, bs = "cr"), 
+                          function(df) gam(value ~ s(year, bs = "cr"), 
                                            data = df)))
@@
   nest(.by = variable) %>% 
-  mutate(fit = map(data, ~mgcv::gam(year ~ s(value, bs = "cs"), data = .)),
+  mutate(fit = map(data, ~mgcv::gam(value ~ s(year, bs = "cs"), data = .)),
          results = map(fit, glance),
@@
 That is, a second researcher might use the same raw data to build the same analysis files and implement the same statistical analysis in an attempt to yield the same results...
-Reproducibility is a minimum necessary condition for a finding to be believable and informative." The importance of this factor can be traced to Irish chemist Robert Boyle (1627-1691), best known for his experiments with vacuum pumps [@shapin2011leviathan].
+Reproducibility is a minimum necessary condition for a finding to be believable and informative." Scientific reproducibility is linked to consensus because if more researchers provide sufficient detail for others to reproduce their results, then consensus on new knowledge can more often and more rapidly established.
+The importance of this factor can be traced to Irish chemist Robert Boyle (1627-1691), best known for his experiments with vacuum pumps [@shapin2011leviathan].
 Boyle was concerned about the secrecy common among experimentalists in the 17th century and aimed to shift the culture from valuing direct in-person witnessing of scientific demonstrations towards meticulous written communications that were detailed enough to enable a reader to successfully undertake the experiment themselves, independent of the original author.

### 2025-04-21 analysis/code/009-exploring-gams.R

@@ -1,0 +1,348 @@
+
+over_time <-  
+  items_df %>%  
+  left_join(items_df_title) %>%  
+  left_join(shannon_per_item) %>%  
+  filter(relative_title_length != -Inf, 
+         relative_title_length !=  Inf,
+         prices_index != "NaN" 
+  ) %>% 
+  mutate(log_authors_n = log(authors_n), 
+         log_pages_n = log(pages_n), 
+         journal_wrp = str_wrap(journal, 30)) %>%  
+  select(year, 
+         log_authors_n, 
+         log_pages_n, 
+         prices_index,  
+         shannon, 
+         relative_title_length,
+         authors_n,
+         pages_n) 
+
+over_time_long <-  
+  over_time %>%  
+  ungroup() %>%  
+  select(-journal) %>%  
+  gather(variable,  
+         value, 
+         -year) %>%  
+  filter(value != -Inf, 
+         value !=  Inf) %>%  
+  mutate(variable = case_when( 
+    variable == "prices_index"  ~ "Recency of references", 
+    variable == "shannon"  ~ "Diversity of references", 
+    variable == "relative_title_length"  ~ "Relative title length (ln)", 
+    variable == "authors_n" ~ "N. of authors",
+    variable == "pages_n" ~ "N. of pages"
+  )) %>%  
+  filter(!is.na(variable)) %>%  
+  filter(!is.nan(value)) %>%  
+  filter(!is.na(value)) %>%  
+  filter(value != "NaN") %>%  
+  mutate(value = parse_number(value))
+
+# compute beta estimates so we can colour lines to indicate more or
+# less hard
+
+library(broom)
+
+over_time_long_models <- 
+  over_time_long %>%  
+  group_nest(variable) %>%  
+  mutate(model = map(data, ~tidy(lm(value ~ year, data = .)))) %>%  
+  unnest(model) %>%  
+  filter(term == 'year') %>%  
+  mutate(becoming_more_scientific = case_when( 
+    variable == "N. of authors"         & estimate > 0 ~ "TRUE", 
+    variable == "N. of pages"           & estimate < 0 ~ "TRUE", 
+    variable == "N. of refs (sqrt)"          & estimate < 0 ~ "TRUE", 
+    variable == "Recency of references"      & estimate > 0 ~ "TRUE", 
+    variable == "Relative title length (ln)" & estimate > 0 ~ "TRUE", 
[…]

### 2025-04-21 analysis/code/010-summary-of-gams.Rmd
gams supplement

@@
 
-*(Briefly introduce the research question, the variables being analyzed, and the goal of modeling their temporal trends.)*
-
-# Methods
+This supplementary document provides further details on the General Additive Models (GAMs) used to model change in the bibliometric variables over time.
+GAMs are useful because they can be fit to complex, nonlinear relationships and make good predictions, while still being able to provide inferential statistics and understand and explain the underlying structure of our models.
 
@@
 
-For all models, the effect of `year` was modeled using a penalized thin plate regression spline (`s(year, bs = 'tp')`) *[Note: Your code used `bs='cr'` (cubic regression spline), adjust text if needed.
-I'll assume 'cr' based on your code]* allowing for flexible non-linear trends.
+For all models, the effect of `year` was modeled using a cubic  regression spline (`s(year, bs = 'cr')`)  allowing for flexible non-linear trends.
 Default weakly informative priors provided by `brms` were used for all parameters, including the intercept, distributional parameters (e.g., `shape`, `sigma`, `phi`, `zoi`, `coi`), and the standard deviation of the smooth term.
-*(Modify this sentence if you set custom priors, and list/justify them).*.
 
@@
 **Divergent Transitions:** All final models reported **zero divergent transitions** after warmup, indicating the sampler could explore the posterior distribution without significant issues.
-*(Crucially, confirm this is true for your final fits.
-Mention if `adapt_delta` adjustments were needed, e.g., "Initial fits for [Variable Name(s)] showed divergent transitions, which were resolved by refitting with `control = list(adapt_delta = 0.995)`.")*
+Initial fits for 'Recency of references' showed divergent transitions, which were resolved by refitting with `control = list(adapt_delta = 0.995)`.")
 

### 2025-04-21 analysis/code/009-exploring-gams.R

@@ -85,113 +85,7 @@ stack(each_variable) %>%
   facet_wrap(~ind, scales = "free")
 
 
-over_time_long_colour_gams_testing_nested_df %>% 
-  mutate(mod_gam = case_when(
-    variable == "N. of authors" ~ lapply(data, 
-                            function(df) gam(value ~ s(year, bs = "cr"), 
-                                           family = nb(),
-                                           method = "REML",
-                                           data = df)),
-    variable == "N. of pages" ~ lapply(data, 
-                             function(df) gam(value ~ s(year, bs = "cr"), 
-                                          family = gaussian(),
-                                           method = "REML",
-                                           data = df)),
-    variable == "Recency of references" ~ lapply(data, 
-                              function(df) gam(value ~ s(year, bs = "cr"), 
-                                           family = betar(link = "logit"),
-                                           method = "REML",
-                                           data = df)),
-    variable == "Diversity of references" ~ lapply(data, 
-                              function(df) gam(value ~ s(year, bs = "cr"), 
-                                           family = gaussian(),
-                                           method = "REML",
-                                           data = df)),
-    variable == "Relative title length (ln)" ~ lapply(data, 
-                              function(df) gam(value ~ s(year, bs = "cr"), 
-                                           family = gaussian(),
-                                           method = "REML",
-                                           data = df)),
-    .default = NULL
-    ))
-
-# inspect diagnostics
-map(over_time_long_colour_gams_testing$mod_gam,
-    gam.check)
-
-library(gratia)
-library(patchwork)
-
-plots_with_titles <- 
-over_time_long_colour_gams_testing %>%
-  mutate(appraise_plot = map2(mod_gam, variable, ~ {
-    p <- appraise(.x)  
-    p + plot_annotation(title = .y)
-  }))
-
-plots_with_titles$appraise_plot[2]
-
-
-# the nested df is 'over_time_long_colour_gams_testing_nested_df'
-# with list-column 'data' and character column 'variable'
-
-# Ensure purrr is loaded
-library(purrr)
-library(dplyr)
-library(mgcv)
-
-# the nested df is 'over_time_long_colour_gams_testing_nested_df'
-# with list-column 'data' and character column 'variable'
-
-over_time_long_colour_gams_testing_nested_df_out <- 
-  over_time_long_colour_gams_testing_nested_df %>%
-  mutate(
[…]

### 2025-04-25 analysis/code/010-summary-of-gams.qmd
working on GAMS explanation

--- 2025-04-21 analysis/code/010-summary-of-gams.Rmd
@@
 library(gt)          # For potentially nicer tables (optional alternative to kable)
-
-# Set brms options (optional)
-# options(mc.cores = parallel::detectCores())
-
-# Load your results object (replace with your actual loading code)
-# load("path/to/your/results_brms.RData")
-# For demonstration, let's assume 'results_brms' exists and contains valid models
-# Filter out any models that failed to fit (are NULL)
-results_brms_valid <- results_brms %>%
-  filter(!sapply(mod_brms, is.null)) %>%
-  filter(sapply(mod_brms, inherits, "brmsfit"))
-
-# Check if any valid models remain
-if(nrow(results_brms_valid) == 0) {
-  stop("No valid brms models found in results_brms.")
-}
-
 ```
@@
 
-Separate GAMs were fitted for each response variable against publication year (`year`). The specific model structure for each variable was chosen based on the nature of the response data:
+Separate GAMs were fitted for each response variable against publication year (`year`).
+The specific model structure for each variable was chosen based on the nature of the response data.
+@fig-brms-distros shows the distributions of the five variables related to scientific publications.
+The shape of these distributions informed our decisions about model specification.
+Recency of references: This variable, bounded between 0 and 1, exhibited a right-skewed distribution with notable occurrences at the boundaries (0 and 1).
+Consequently, a Zero-One Inflated Beta (zero_one_inflated_beta()) distribution family was selected.
+The associated logit link function is standard for modeling the mean (mu) of proportions, transforming the (0, 1) interval to the real line (-∞, +∞) and modeling the log-odds of the mean recency score for values strictly between 0 and 1.
+
+Diversity of references & Relative title length (ln): Both variables displayed approximately symmetric, bell-shaped (normal) distributions.
+For these continuous variables, the Gaussian (gaussian()) family was employed with the canonical identity link function, meaning the linear predictor directly models the expected value (mean) of the response variable.
+
+N.
+of authors & N.
+of pages: These count variables both exhibited highly right-skewed distributions.
+Therefore, the Negative Binomial (negbinomial()) family was chosen.
+The standard log link function was applied, ensuring that the predicted mean remains positive (μ = exp(η)) and implying that predictor effects are multiplicative on the original count scale, which is conventional for count data.
+
+```{r}
+#| label: fig-brms-distros
+#| fig-cap: "Distributions of the response variables"
+
+# prepare data for GAMS
+
+# create data objects needed for this document
+qmd_file <- here::here("analysis/paper/paper.qmd")
+
+# A one-liner to run the R code in the main paper and create all the data objects
+{ f <- tempfile(fileext = ".R"); on.exit(unlink(f, force = TRUE)); knitr::purl(qmd_file, output = f, documentation = 0, quiet = TRUE); source(f, local = FALSE); }
+
+# on with the show
+over_time <-  
+  items_df %>%  
+  left_join(items_df_title) %>%  
+  left_join(shannon_per_item) %>%  
+  filter(relative_title_length != -Inf, 
+         relative_title_length !=  Inf,
+         prices_index != "NaN" 
+  ) %>% 
+  mutate(log_authors_n = log(authors_n), 
+         log_pages_n = log(pages_n), 
+         journal_wrp = str_wrap(journal, 30)) %>%  
+  select(year, 
+         log_authors_n, 
+         log_pages_n, 
+         prices_index,  
+         shannon, 
+         relative_title_length,
+         authors_n,
+         pages_n)
+
+over_time_long <-  
+  over_time %>%  
+  ungroup() %>%  
+  select(-journal) %>%  
+  gather(variable,  
+         value, 
+         -year) %>%  
+  filter(value != -Inf, 
+         value !=  Inf) %>%  
+  mutate(variable = case_when( 
+    variable == "prices_index"  ~ "Recency of references", 
+    variable == "shannon"  ~ "Diversity of references", 
+    variable == "relative_title_length"  ~ "Relative title length (ln)", 
+    variable == "authors_n" ~ "N. of authors",
+    variable == "pages_n" ~ "N. of pages"
+  )) %>%  
+  filter(!is.na(variable)) %>%  
+  filter(!is.nan(value)) %>%  
+  filter(!is.na(value)) %>%  
+  filter(value != "NaN") %>%  
+  mutate(value = parse_number(value))
+
+# compute beta estimates so we can colour lines to indicate more or
+# less hard
+
+library(broom)
+
+over_time_long_models <- 
+  over_time_long %>%  
+  group_nest(variable) %>%  
+  mutate(model = map(data, ~tidy(lm(value ~ year, data = .)))) %>%  
+  unnest(model) %>%  
+  filter(term == 'year') %>%  
+  mutate(becoming_more_scientific = case_when( 
+    variable == "N. of authors"         & estimate > 0 ~ "TRUE", 
+    variable == "N. of pages"           & estimate < 0 ~ "TRUE", 
+    variable == "N. of refs (sqrt)"          & estimate < 0 ~ "TRUE", 
+    variable == "Recency of references"      & estimate > 0 ~ "TRUE", 
+    variable == "Relative title length (ln)" & estimate > 0 ~ "TRUE", 
+    variable == "Diversity of references"    & estimate < 0 ~ "TRUE",
+    TRUE ~ "FALSE"
+  ))
+
+# join with data
+over_time_long_colour <- 
+  over_time_long %>% 
+  left_join(over_time_long_models)
+
+over_time_long_colour_gams_testing_nested_df <- 
+  over_time_long_colour %>% 
+  nest(.by = variable)
+
+# check the distributions of each variable
+each_variable <- 
+map(over_time_long_colour_gams_testing_nested_df$data, 
+    ~.x$value)
+
+names(each_variable) <- over_time_long_colour_gams_testing_nested_df$variable
+
+stack(each_variable) %>% 
+  ggplot() +
+  aes(values) +
+  geom_histogram() +
+  facet_wrap(~ind, scales = "free") +
+  theme_minimal()
+```
+
+The table below summarises the selections made for each variable, based on their distributions obsvered in @fig-brms-distros.
 
@@
 
+```{r}
+#| eval: false
+
+# ---  fit models with brms -------------------------------
+# this takes a few hours
+
+library(brms)
+library(dplyr)
+library(purrr)
+
+# --- Set options for brms (optional, can help speed up) ---
+options(mc.cores = parallel::detectCores()) # Use all available CPU cores
+
+# --- the nested df is 'over_time_long_colour_gams_testing_nested_df' 
+# with list-column 'data' and character column 'variable'
+
+results_brms <- over_time_long_colour_gams_testing_nested_df %>%
+  mutate(
+    # Create a new column for brms models to avoid overwriting gam models if needed
+    mod_brms = map2(data, variable, ~{
+      current_df <- .x        # The dataframe for the current row
+      current_var <- .y       # The variable name for the current row
+      response_col <- "value" #  this matches the response column name in nested dfs
+
+      message(paste("Fitting brms model for:", current_var)) # Progress message
+
+      # --- 1. Choose the brms family based on the variable name ---
+      chosen_family <- if (current_var == "N. of authors") {
+        negbinomial() # For count data
+      } else if (current_var == "N. of pages") {
+        negbinomial() # For count data
+      } else if (current_var == "Recency of references") {
+        zero_one_inflated_beta() # For 0/1 inflated proportions
+      } else if (current_var == "Diversity of references") {
+        gaussian() # For continuous, approx normal data
+      } else if (current_var == "Relative title length (ln)") {
+        gaussian() # For continuous, normal data
+      } else {
+        warning(paste("Unknown variable:", current_var, "- using Gaussian default."))
+        gaussian() # Default fallback
+      }
+
+      # --- 2. Construct the formula using brms::bf() ---
+      # Handle the special structure for Zero-One Inflated Beta
+      if (inherits(chosen_family, "zero_one_inflated_beta")) {
+        # Formula for the mean (between 0 and 1), plus formulas for
+        # zoi (zero-inflation probability) and coi (one-inflation probability).
+        # zoi and coi are modelled simply as intercepts.
+        brm_formula <- bf(
+          paste(response_col, "~ s(year, bs = 'cr')"), 
+          # Smooth term for mean
+          zoi ~ 1,                                    
+          # Intercept for prob of 0
+          coi ~ 1                                     
+          # Intercept for prob of 1
+        )
+      } else {
+        # Standard formula for other families
+        brm_formula <- bf(
+          paste(response_col, "~ s(year, bs = 'cr')")
+        )
+      }
+
+      # --- 3. Fit the brms model ---
+      # Using tryCatch to prevent one failure from stopping everything
+      tryCatch({
+        brm(
+          formula = brm_formula,
+          family = chosen_family,
+          data = current_df,
+          # --- Stan control arguments (adjust as needed) ---
+          cores = 4,  # Number of cores for parallel chains 
+          chains = 4, # Number of Markov chains (4 is standard for final runs)
+          iter = 3000, # Total iterations per chain (includes warmup)
+          warmup = 1000,# Number of warmup iterations (discarded)
+          control = list(adapt_delta = 0.9999), # for helping convergence for complex models
+          seed = 123, # For reproducibility
+          silent = 2, # Suppress compilation messages from Stan
+          refresh = 0 # Suppress iteration progress updates
+
+        )
+      }, error = function(e) {
+        # Print a warning if a model fails to fit
+        warning(paste("brms fitting failed for variable:", 
+                      current_var, "\nError:", e$message))
+        return(NULL) # Return NULL for failed models
+      })
+    }) # End of map2
+  ) # End of mutate
+
+# quite a large file
+saveRDS(results_brms, here::here("analysis/data/results_brms.RData"))
+
+```
+
+```{r}
+
+# Load my brms results object so we don't have to re-run the above block
+load(here::here("analysis/data/results_brms.RData"))
+
+# For demonstration, let's assume 'results_brms' exists and contains valid models
+# Filter out any models that failed to fit (are NULL)
+results_brms_valid <- results_brms %>%
+  filter(!sapply(mod_brms, is.null)) %>%
+  filter(sapply(mod_brms, inherits, "brmsfit"))
+
+```
+
 ## MCMC Sampling
@@
 ```{r check_divergences, echo=TRUE}
+#| tbl-cap: "Number of divergences for each variable"
+
 # Code to programmatically check divergences 
@@
 
-print("Number of Divergent Transitions per Model:")
-print(divergence_summary)
+kable(divergence_summary)
 
@@
 ```{r convergence_summary_table, echo=FALSE}
+#| tbl-cap: "Summary of MCMC Convergence Diagnostics.
+Max R-hat should be close to 1.0 (e.g., < 1.01).
+Min ESS values indicate the minimum effective sample size across all parameters.
+N Divergences should be 0 for reliable results."
 
@@
   convergence_summary_df,
-  caption = "Summary of MCMC Convergence Diagnostics.
-Max R-hat should be close to 1.0 (e.g., < 1.01).
-Min ESS values indicate the minimum effective sample size across all parameters.
-N Divergences should be 0 for reliable results.",
+  caption = ,
   digits = c(NA, 3, 0, 0, 0), # Specify digits for numeric columns

### 2025-04-29 analysis/code/010-summary-of-gams.qmd
nearly got all the gams diagnostics done

@@
 Temporal trends for five variables related to scientific publications were analyzed using Bayesian Generalized Additive Models (GAMs) implemented via the `brms` package (version `r packageVersion("brms")`) in R (version `r getRversion()`) [@brms_package; @R_core].
-`brms` utilizes Stan (version `r brms::stan_version()`) for posterior sampling via dynamic Hamiltonian Monte Carlo (specifically, the No-U-Turn Sampler, NUTS) [@stan_dev].
+`brms` utilizes Stan (version `r rstan::stan_version()`) for posterior sampling via dynamic Hamiltonian Monte Carlo (specifically, the No-U-Turn Sampler, NUTS) [@stan_dev].
 
@@
 
+# --- Define the Prior ---
+# Apply a regularizing prior to the standard deviation of all smooth terms
+# Exponential(1) encourages smaller standard deviations (less wiggly smooths)
+sds_prior <- prior(exponential(1), class = "sds")
+
+# --- Run the models with the added prior ---
 results_brms <- over_time_long_colour_gams_testing_nested_df %>%
   mutate(
-    # Create a new column for brms models to avoid overwriting gam models if needed
+    # Create a new column for brms models
     mod_brms = map2(data, variable, ~{
@@
       current_var <- .y       # The variable name for the current row
-      response_col <- "value" #  this matches the response column name in nested dfs
+      response_col <- "value" # Ensure this matches
 
-      message(paste("Fitting brms model for:", current_var)) # Progress message
+      message(paste("Fitting brms model for:", current_var))
 
-      # --- 1. Choose the brms family based on the variable name ---
+      # --- 1. Choose the brms family ---
       chosen_family <- if (current_var == "N. of authors") {
-        negbinomial() # For count data
+        negbinomial()
       } else if (current_var == "N. of pages") {
-        negbinomial() # For count data
+        negbinomial()
       } else if (current_var == "Recency of references") {
-        zero_one_inflated_beta() # For 0/1 inflated proportions
+        zero_one_inflated_beta()
       } else if (current_var == "Diversity of references") {
-        gaussian() # For continuous, approx normal data
+        gaussian()
       } else if (current_var == "Relative title length (ln)") {
-        gaussian() # For continuous, normal data
+        gaussian()
       } else {
         warning(paste("Unknown variable:", current_var, "- using Gaussian default."))
-        gaussian() # Default fallback
+        gaussian()
       }
 
-      # --- 2. Construct the formula using brms::bf() ---
-      # Handle the special structure for Zero-One Inflated Beta
+      # --- 2. Construct the formula ---
       if (inherits(chosen_family, "zero_one_inflated_beta")) {
-        # Formula for the mean (between 0 and 1), plus formulas for
-        # zoi (zero-inflation probability) and coi (one-inflation probability).
-        # zoi and coi are modelled simply as intercepts.
         brm_formula <- bf(
           paste(response_col, "~ s(year, bs = 'cr')"),
-          # Smooth term for mean
           zoi ~ 1,
-          # Intercept for prob of 0
           coi ~ 1
-          # Intercept for prob of 1
         )
       } else {
-        # Standard formula for other families
         brm_formula <- bf(
@@
 
-      # --- 3. Fit the brms model ---
-      # Using tryCatch to prevent one failure from stopping everything
+      # --- 3. Fit the brms model with the prior ---
       tryCatch({
@@
           data = current_df,
-          # --- Stan control arguments (adjust as needed) ---
-          cores = 4,  # Number of cores for parallel chains 
-          chains = 4, # Number of Markov chains (4 is standard for final runs)
-          iter = 3000, # Total iterations per chain (includes warmup)
-          warmup = 1000,# Number of warmup iterations (discarded)
-          control = list(adapt_delta = 0.9999), # for helping convergence for complex models
-          seed = 123, # For reproducibility
-          silent = 2, # Suppress compilation messages from Stan
-          refresh = 0 # Suppress iteration progress updates
-
+          prior = sds_prior, 
+          cores =    4,
+          chains =   4,
+          iter =     50000,
+          warmup =   5000,
+          control =  list(adapt_delta = 0.99999), 
+          seed =     123,
+          silent =   2,
+          refresh =  0
         )
       }, error = function(e) {
-        # Print a warning if a model fails to fit
         warning(paste("brms fitting failed for variable:",
                       current_var, "\nError:", e$message))
-        return(NULL) # Return NULL for failed models
+        return(NULL)
       })
@@
 # Load my brms results object so we don't have to re-run the above block
-load(here::here("analysis/data/results_brms.RData"))
+results_brms <- readRDS(here::here("analysis/data/results_brms.RData"))
 
-# For demonstration, let's assume 'results_brms' exists and contains valid models
 # Filter out any models that failed to fit (are NULL)
@@
   filter(sapply(mod_brms, inherits, "brmsfit"))
-
 ```
@@
 
-Models were fitted using `r results_brms_valid$mod_brms[[1]]$algorithm$chains` Markov chains.
-Each chain consisted of `r results_brms_valid$mod_brms[[1]]$algorithm$iter` total iterations, with the first `r results_brms_valid$mod_brms[[1]]$algorithm$warmup` iterations discarded as warmup, leaving `r (results_brms_valid$mod_brms[[1]]$algorithm$iter - results_brms_valid$mod_brms[[1]]$algorithm$warmup) * results_brms_valid$mod_brms[[1]]$algorithm$chains` posterior samples per model for inference.
+Models were fitted using `r brms::nchains(results_brms_valid$mod_brms[[1]])` Markov chains.
+Each chain consisted of `r results_brms_valid$mod_brms[[1]]$fit@sim$iter` total iterations, with the first `r results_brms_valid$mod_brms[[1]]$fit@sim$warmup` iterations discarded as warmup, leaving `r (results_brms_valid$mod_brms[[1]]$fit@sim$iter - results_brms_valid$mod_brms[[1]]$fit@sim$warmup) * brms::nchains(results_brms_valid$mod_brms[[1]])` posterior samples per model for inference.
 No thinning was applied (thin = 1).
@@
 
-1.
-**Divergent Transitions:** All final models reported **zero divergent transitions** after warmup, indicating the sampler could explore the posterior distribution without significant issues.
-Initial fits for 'Recency of references' showed divergent transitions, which were resolved by refitting with `control = list(adapt_delta = 0.995)`.")
-
-```{r check_divergences, echo=TRUE}
-#| tbl-cap: "Number of divergences for each variable"
-
-# Code to programmatically check divergences 
-library(rstan)
-sampler_params_list <- map(results_brms_valid$mod_brms, ~{
-  if (!is.null(.x) && inherits(.x, "brmsfit")) {
-    # Get parameters excluding warmup draws
-    nuts_params(.x, inc_warmup = FALSE)
-  } else {
-    NULL
-  }
-})
-
-divergences_per_model <- map_dbl(sampler_params_list, ~{
-  if (!is.null(.x)) {
-    # Sum divergent transitions across all chains
-    sum(subset(.x, Parameter == "divergent__")$Value)
-  } else {
-    NA_real_ # Use NA for models that didn't fit
-  }
-})
-
-divergence_summary <- tibble(
-   variable = results_brms_valid$variable,
-   n_divergences = divergences_per_model
-)
-
-kable(divergence_summary)
-
-```
-
 ```{r convergence_summary_table, echo=FALSE}
@@
 
-#```{r convergence_summary_table, echo=FALSE}
-# Load necessary libraries if not already done
+# Function to extract diagnostics from a single model
+get_diagnostics <- function(m) {
+  ms <- summary(m)
+
+  tibble(
+    chains = brms::nchains(m),
+    rhat_max = max(brms::rhat(m)), 
+    bulk_ess_min = min(ms$spec_pars$Bulk_ESS, 
+                       na.rm = TRUE),
+    tail_ess_min = min(ms$spec_pars$Tail_ESS, 
+                       na.rm = TRUE),
+    num_divergent = rstan::get_num_divergent(m$fit)
+  )
+}
+
+# Apply to all models
+diagnostics_table <- results_brms_valid %>%
+  mutate(diagnostics = map(mod_brms, get_diagnostics)) %>%
+  unnest(diagnostics) %>% 
+  select(variable, 
+         chains,
+         rhat_max,
+         bulk_ess_min,
+         tail_ess_min,
+         num_divergent)
+
+# View the table
+
+library(knitr)
+library(kableExtra)
+
+kable(
+  diagnostics_table,
+  caption = "Summary of MCMC Convergence Diagnostics. Max R-hat should be close to 1.0. Min ESS values indicate the minimum effective sample size across all parameters. N Divergences should be 0.",
+  digits = c(NA, 0, 3, 0, 0, 0), # Chains/ESS/Div as integers, Rhat to 3 decimals
+  booktabs = TRUE,
+  linesep = ""
+) %>%
+  kableExtra::kable_styling(latex_options = "scale_down", full_width = FALSE) %>%
+  kableExtra::footnote(general = "Ideal values: Max R-hat ≈ 1.0, Min ESS > 1000, N Divergences = 0.", general_title = "Note:", footnote_as_chunk = TRUE)
+```
+
+1.
+**Divergent Transitions:** All final models reported less than ten **divergent transitions** after warmup, indicating the sampler could explore the posterior distribution without significant issues.
+Initial fits for 'Recency of references' showed divergent transitions, which were managed by refitting with `control = list(adapt_delta = 0.9999)`.")
+2.  **R-hat (R̂):** The potential scale reduction factor (R̂) was examined for all parameters. All R̂ values were ≤ 1.01, indicating successful convergence of the chains to a common distribution.
+3.
+**Effective Sample Size (ESS):** Both Bulk-ESS and Tail-ESS were assessed.
+Minimum ESS values across all parameters were deemed sufficiently large (> 1000) for reliable estimation of posterior summaries and credible intervals.
+4.  **Trace Plots:** Trace plots for key parameters were visually inspected and showed good mixing and stationarity, providing further confidence in chain convergence (@fig-trace-plots-mcmc-trace).
+
+```{r fig.height=3 * length(results_brms_valid$mod_brms)}
+#| label: fig-trace-plots-mcmc-trace,
+#| echo: false
+#| message: false
+#| warning: false
+#| fig-width: 10
+#| fig-cap: "Trace plots for key parameters (Intercept and smooth term standard deviation) with divergence markers for each model.
+Well-mixing chains should resemble stationary 'fuzzy caterpillars' with chains overlapping."
+#---------------------------------------------------------------------
+# Load necessary libraries
+#---------------------------------------------------------------------
 library(brms)
-library(posterior) # Explicitly load posterior for ess functions
-library(purrr)
-library(dplyr)
-library(knitr)
-library(kableExtra) # Optional for styling
+library(purrr)     # For map2
+library(dplyr)     # For tibble etc.
+library(posterior) # For as_draws_array
+library(bayesplot) # For mcmc_trace and nuts_params
+library(ggplot2)   # For ggtitle and theme adjustments
+library(cowplot)   # For plot_grid
+#---------------------------------------------------------------------
 
-# --- Calculate Convergence Diagnostics ---
-# Use map2_dfr to iterate over models and variable names, creating a data frame
-convergence_summary_df <- map2_dfr(
+#---------------------------------------------------------------------
+# Generate Trace Plots using mcmc_trace for each model
+#---------------------------------------------------------------------
+trace_plots_list <- map2(
   results_brms_valid$mod_brms,
@@
   ~{
-    # Assign model and variable name for clarity within the loop
+    # Assign model and variable name for clarity
     model <- .x
@@
 
-    # Initialize values
-    max_rhat <- NA_real_
-    min_bulk_ess <- NA_real_
-    min_tail_ess <- NA_real_
-    n_divergences <- NA_integer_
+    # Initialize plot object as NULL
+    p <- NULL
 
@@
     if (!is.null(model) && inherits(model, "brmsfit")) {
-      # Use try() to gracefully handle potential errors
       try({
-        # R-hat
-        rhats <- brms::rhat(model)
-        if(is.numeric(rhats)) max_rhat <- max(rhats, na.rm = TRUE)
+        # --- Define parameters to plot ---
+        # Use the names confirmed to work: b_Intercept and sds_syear_1
+        # Check if these parameters actually exist in the current model
+        all_model_pars <- parnames(model)
+        pars_to_plot <- character(0) # Start with empty vector
 
-        # Bulk ESS - Use posterior::ess_bulk()
-        bulk_ess <- posterior::ess_bulk(model)
-        if(is.numeric(bulk_ess) && length(bulk_ess) > 0) min_bulk_ess <- min(bulk_ess, na.rm = TRUE)
-
-        # Tail ESS - Use posterior::ess_tail()
-        tail_ess <- posterior::ess_tail(model)
-         if(is.numeric(tail_ess) && length(tail_ess) > 0) min_tail_ess <- min(tail_ess, na.rm = TRUE)
-
-        # Divergences
-        nuts_p <- brms::nuts_params(model)
-        nuts_df <- as.data.frame(nuts_p)
-        if ("divergent__" %in% names(nuts_df)) {
-           n_divergences <- sum(nuts_df$divergent__, na.rm = TRUE)
+        if ("b_Intercept" %in% all_model_pars) {
+          pars_to_plot <- c(pars_to_plot, "b_Intercept")
         } else {
-           warning(paste("Column 'divergent__' not found in nuts_params output for:", var_name))
-           n_divergences <- 0L
-        }
-      }, silent = TRUE) # Keep try silent, but check warnings/NA values later
+           warning(paste("'b_Intercept' not found for model:", var_name))
         }
 
-    # Return a tibble
-    tibble(
-      Variable = var_name,
-      `Max R-hat` = max_rhat,
-      `Min Bulk ESS` = floor(min_bulk_ess), # Floor ESS for cleaner table
-      `Min Tail ESS` = floor(min_tail_ess), # Floor ESS for cleaner table
-      `N Divergences` = as.integer(n_divergences) # Ensure integer
+        if ("sds_syear_1" %in% all_model_pars) {
+           pars_to_plot <- c(pars_to_plot, "sds_syear_1")
+        } else {
+           # Handle cases where the smooth sds name might differ slightly
+           # Example: find any parameter starting with sds_s
+           sds_par_alt <- grep("^sds_s", all_model_pars, value = TRUE)
+           if(length(sds_par_alt) > 0) {
+               pars_to_plot <- c(pars_to_plot, sds_par_alt[1])
+               warning(paste("Using alternative sds name:", sds_par_alt[1], "for model:", var_name))
+           } else {
+               warning(paste("'sds_syear_1' (or similar) not found for model:", var_name))
+           }
+        }
+        # --- End of parameter identification ---
+
+        # Proceed only if we found parameters to plot
+        if(length(pars_to_plot) > 0) {
+            # 1) Extract draws as an array
+            post_array <- posterior::as_draws_array(model)
+
+            # 2) Extract NUTS diagnostics (divergences)
+            np <- brms::nuts_params(model)
+
+            # 3) Generate trace plot with divergence ticks
+            p <- bayesplot::mcmc_trace(
+                    post_array,
+                    pars = pars_to_plot, # Use the identified parameters
+                    np = np, # Provide NUTS parameters for divergence markers
+                    facet_args = list(ncol = 1, strip.position = "left") # Arrange vertically
+                  ) +
+                 ggtitle(var_name) + # Add variable name as title
+                 theme(
+                    plot.title = element_text(hjust = 0.5, size = 11), # Center title
+                    legend.position = "none" # Often hide legend
                     )
+        } else {
+            message(paste("No parameters selected for plotting trace for model:", var_name))
+        }
+
+      }, silent = TRUE) # Use try silently, check for NULL later
+    } else {
+        warning(paste("Invalid or NULL model object encountered for variable:", var_name))
+    }
+
+    # Return the ggplot object (or NULL if failed or no pars found)
+    return(p)
   }
 )
+#---------------------------------------------------------------------
 
-# --- Display the Table ---
-# (kable formatting code remains the same)
-kable(
-  convergence_summary_df,
-  caption = ,
-  digits = c(NA, 3, 0, 0, 0), # Specify digits for numeric columns
-  booktabs = TRUE
-) %>%
-  kableExtra::kable_styling(latex_options = "scale_down")
+#---------------------------------------------------------------------
+# Combine and Display Plots
+#---------------------------------------------------------------------
+# Remove any NULL elements from the list
+valid_trace_plots <- trace_plots_list[!sapply(trace_plots_list, is.null)]
 
-# somthing not quite right here
+# Check if there are any valid plots left
+if (length(valid_trace_plots) > 0) {
+  # Combine the valid plots into a grid
+  # Adjust ncol and fig.height chunk option as needed
+  combined_trace_plots <- plot_grid(
+                              plotlist = valid_trace_plots,
+                              ncol = 2 # Arrange in 2 columns
+                              )
 
+  # Display the combined plot
+  print(combined_trace_plots)
+} else {
+  message("No valid trace plots were generated.")
+}
+#---------------------------------------------------------------------
 ```
-
-2.  **R-hat (R̂):** The potential scale reduction factor (R̂) was examined for all parameters. All R̂ values were ≤ 1.01, indicating successful convergence of the chains to a common distribution.
-3.
-**Effective Sample Size (ESS):** Both Bulk-ESS and Tail-ESS were assessed.
-Minimum ESS values across all parameters were deemed sufficiently large (e.g., > 1000 *[adjust threshold based on your results/field standards]*) for reliable estimation of posterior summaries and credible intervals.
-4.
-**Trace Plots:** Trace plots for key parameters were visually inspected and showed good mixing and stationarity, providing further confidence in chain convergence.
-*(Optionally include an example trace plot in an Appendix)*.
 
@@
 Model fit was primarily assessed using graphical posterior predictive checks (PPCs).
-Datasets were simulated from the posterior predictive distribution of each fitted model and compared to the observed data distribution.
+Datasets were simulated from the posterior predictive distribution of each fitted model and compared to the observed data distribution (@fig-ppc-plot).
 
-```{r ppc_plot, fig.cap="Posterior predictive checks (density overlays) for each model.
-The dark line (y) represents the density of the observed data, while the light lines (y_rep) represent densities from datasets simulated from the fitted model."}
+```{r}
+#| label: fig-ppc-plot
+#| fig-cap: "Posterior predictive checks (density overlays) for each model.
+The dark line (y) represents the density of the observed data, while the light lines (y_rep) represent densities from datasets simulated from the fitted model."
+#| echo: true  
+#| warning: false  
+#| message: false  
 # Generate PPC plots (density overlay is often informative)
@@
 
-## Estimated Temporal Trends
-
-The estimated non-linear effect of publication year on each response variable is visualized below, showing the contribution of the smooth term `s(year)` to the model's linear predictor (on the respective link scale).
-
-```{r smooth_plots, fig.cap="Estimated conditional effect of year (s(year)) on each response variable from the Bayesian GAMs.
-The blue line represents the posterior mean effect, and the grey ribbon represents the 95% credible interval.
-Y-axis scales differ and correspond to the link function of each model (identity, log, or logit)."}
-# Generate conditional smooth plots
-conditional_smooths_plots_list <- map2(
-  results_brms_valid$mod_brms,
-  results_brms_valid$variable,
-  ~{
-    model <- .x
-    var_name <- .y
-    tryCatch({
-      # Generate plot data first
-      cs_data <- conditional_smooths(model, plot = FALSE)[[1]] # Get data for the first smooth
-      # Create ggplot manually for more control if needed, or use plot() method
-      plot_object_list <- plot(conditional_smooths(model), plot = TRUE)
-
-      if (length(plot_object_list) >= 1 && inherits(plot_object_list[[1]], "ggplot")) {
-         # Add title and potentially customize axes
-         y_axis_label <- case_when(
-             var_name %in% c("N. of authors", "N. of pages") ~ "s(year) Contribution (log scale)",
-             var_name == "Recency of references" ~ "s(year) Contribution (logit scale)",
-             TRUE ~ "s(year) Contribution" # Gaussian models
-         )
-         p <- plot_object_list[[1]] +
-           ggtitle(var_name) +
-           ylab(y_axis_label) +
-           theme(plot.title = element_text(hjust = 0.5, size = 10))
-         return(p)
-      } else {
-         warning(paste("Could not extract ggplot object for:", var_name))
-         return(NULL)
-      }
-    }, error = function(e) {
-       warning("Conditional smooths plot failed for: ", var_name, " Error: ", e$message)
-       return(NULL)
-    })
-  }
-)
-
-# Remove NULLs and combine
-valid_smooth_plots <- conditional_smooths_plots_list[!sapply(conditional_smooths_plots_list, is.null)]
-plot_grid(plotlist = valid_smooth_plots, ncol = 3)
-
-```
-
 ## Parameter Estimates
@@
 
-```{r param_summary, echo=FALSE}
-# Extract tidy summaries for relevant parameters
-tidy_summaries <- map_dfr(
+```{r}
+#| label: tab:param-summary
+#| echo: false
+#| message: false
+#| warning: false
+#---------------------------------------------------------------------
+# Load necessary libraries
+#---------------------------------------------------------------------
+library(brms)
+library(purrr)
+library(dplyr)
+library(tibble) # For rownames_to_column
+library(knitr)
+library(kableExtra)
+#---------------------------------------------------------------------
+
+#---------------------------------------------------------------------
+# Extract posterior summaries using brms::posterior_summary
+#---------------------------------------------------------------------
+# Apply posterior_summary to each model, keep results in a list
+posterior_summaries_list <- map(
   results_brms_valid$mod_brms,
-  ~ tidy(.x, effects = c("fixed", "ran_pars"), conf.int = TRUE, conf.level = 0.95),
-  .id = "model_index" # Add an index to link back to variable name
+  ~ posterior_summary(.x, probs = c(0.025, 0.975)) # Get mean, sd, and 95% CrI
 )
 
-# Join with variable names
-tidy_summaries <- tidy_summaries %>%
-  mutate(model_index = as.integer(model_index)) %>%
-  left_join(results_brms_valid %>% select(variable) %>% mutate(model_index = row_number()), by = "model_index") %>%
-  select(variable, everything(), -model_index) %>%
-  # Optional: Filter specific terms if the table gets too long
-  filter(component != "zi") %>% # Often not directly estimated for NB/ZOIB in tidy
-  # Rename for clarity
+# --- Define the key parameters to keep in the final table ---
+# Verify these names against your actual model output if needed
+params_to_keep <- c(
+    "b_Intercept",      # Intercept
+    "sigma",            # Residual SD for Gaussian
+    "shape",            # Shape parameter for Negative Binomial
+    "phi",              # Precision for Beta/ZOIB
+    "zoi",              # Zero-inflation probability for ZOIB
+    "coi",              # One-inflation probability for ZOIB
+    "sds_syear_1"       # SD for the smooth term s(year)
+)
+
+#---------------------------------------------------------------------
+# Process the list into a final data frame
+#---------------------------------------------------------------------
+tidy_summaries_final <- map2_dfr(
+  posterior_summaries_list,
+  results_brms_valid$variable,
+  ~{
+    # Convert matrix output to tibble
+    summary_df <- as_tibble(.x, rownames = "Parameter")
+
+    # Filter for desired parameters
+    summary_filtered <- summary_df %>%
+      filter(Parameter %in% params_to_keep)
+
+    # Add variable name and parameter type classification
+    summary_filtered %>%
+      mutate(
+        variable = .y, # Add variable name
+        `Parameter Type` = case_when(
+            Parameter == "b_Intercept" ~ "Intercept",
+            Parameter %in% c("sigma", "shape", "phi", "zoi", "coi") ~ "Distributional",
+            startsWith(Parameter, "sds_") ~ "Smooth SD",
+            TRUE ~ "Other"
+        ),
+        # Rename parameters for better readability
+        Parameter = case_when(
+            Parameter == "b_Intercept" ~ "(Intercept)",
+            Parameter == "sds_syear_1" ~ "sds(s(year))",
+            TRUE ~ Parameter
+        )
+      )
+  }
+)
+
+# --- Select and order columns for the final table ---
+tidy_summaries_final <- tidy_summaries_final %>%
     rename(
-    Parameter = term,
-    Estimate = estimate,
-    Est.Error = std.error,
-    `Lower 95% CI` = conf.low,
-    `Upper 95% CI` = conf.high
+        Estimate = Estimate,
+        Est.Error = Est.Error,
+        `Lower 95% CI` = Q2.5, # Rename quantile column
+        `Upper 95% CI` = Q97.5 # Rename quantile column
     ) %>%
-  select(variable, Parameter, Estimate, Est.Error, `Lower 95% CI`, `Upper 95% CI`, component, group)
+    select(
+        variable, Parameter, `Parameter Type`,
+        Estimate, Est.Error,
+        `Lower 95% CI`, `Upper 95% CI`
+    ) %>%
+    arrange(variable, `Parameter Type`, Parameter)
 
-# Display using kable
+#---------------------------------------------------------------------
+# Create and Display the Formatted Table
+#---------------------------------------------------------------------
+# Check if the final tibble is empty
+if(nrow(tidy_summaries_final) == 0) {
+    warning("Final parameter summary table is empty. Check parameter names in 'params_to_keep'.")
+} else {
     kable(
-  tidy_summaries,
-  caption = "Posterior summary statistics for key model parameters.
-'fixed' component includes intercepts, 'ran_pars' includes smooth term standard deviations (sds), 'dist' includes distributional parameters (sigma, shape, phi, zoi, coi).",
+      tidy_summaries_final,
+      caption = "Posterior summary statistics for key model parameters.",
       digits = 2,
-  booktabs = TRUE
+      booktabs = TRUE,
+      linesep = ""
     ) %>%
- kableExtra::kable_styling(latex_options = c("striped", "scale_down"))
+     kableExtra::kable_styling(
+        bootstrap_options = c("striped", "hover", "condensed"),
+        latex_options = c("striped", "scale_down"),
+        full_width = FALSE
+        ) %>%
+     kableExtra::column_spec(1, bold = TRUE) # Make variable name bold
+}
+```
+
+## Estimated Temporal Trends
+
+The estimated non-linear effect of publication year on each response variable is visualized below (@fig-smooth-plots), showing the contribution of the smooth term `s(year)` to the model's linear predictor (on the respective link scale).
+
+```{r}
+#| label: fig-smooth-plots
+#| fig-cap: "Estimated conditional effect of year (s(year)) on each response variable from the Bayesian GAMs.
+The blue line represents the posterior mean effect, and the grey ribbon represents the 95% credible interval.
+Y-axis scales differ and correspond to the link function of each model (identity, log, or logit)."
+
+# Load necessary libraries
+library(brms)
+library(purrr)
+library(dplyr)
+library(ggplot2)
+library(cowplot)
+
+# Generate conditional smooth plots within map2
+conditional_smooths_plots_list <- map2(
+  results_brms_valid$mod_brms,
+  results_brms_valid$variable,
+  ~{
+    # Assign model and variable name
+    model <- .x
+    var_name <- .y
+
+    # Generate the plot object list directly
+    plot_object_list <- plot(conditional_smooths(model), plot = TRUE)
+
+    # Define the y-axis label based on variable name (link function)
+    y_axis_label <- case_when(
+        var_name %in% c("N. of authors", "N. of pages") ~ "s(year) Contribution (log scale)",
+        var_name == "Recency of references" ~ "s(year) Contribution (logit scale)",
+        TRUE ~ "s(year) Contribution" # Default for Gaussian models
+    )
+
+    # Extract the first plot, add title, label, and theme
+    plot_object_list[[1]] +
+      ggtitle(var_name) +
+      ylab(y_axis_label) +
+      theme(plot.title = element_text(hjust = 0.5, size = 10)) # Center title
+  }
+)
+
+# Combine the plots into a grid (adjust ncol as desired)
+# The plot_grid call itself will display the combined plot
+plot_grid(plotlist = conditional_smooths_plots_list, ncol = 3)
 

### 2025-04-30 analysis/code/010-summary-of-gams.qmd
updates gams supplement

@@
 ---
-title: "Analysis of Temporal Trends using Bayesian Generalized Additive Models"
-author: "Your Name"
+title: "Supplementary Materials on Bayesian Generalized Additive Models used to Analyse Temporal Trends in Job Ads"
+author: "Ben Marwick"
 date: "`r Sys.Date()`"
+bibliography: ../analysis/paper/references.bib
+csl:          ../analysis/paper/templates/journal-of-archaeological-science.csl # Insert path for the bib-style
 output:
-  html_document:
-    toc: yes
-    toc_float: yes
-    number_sections: yes
-    code_folding: hide
-  pdf_document:
-    toc: yes
+  html_document 
 ---
@@
 library(knitr)       # For kable
-library(gt)          # For potentially nicer tables (optional alternative to kable)
 ```
@@
 
-Temporal trends for five variables related to scientific publications were analyzed using Bayesian Generalized Additive Models (GAMs) implemented via the `brms` package (version `r packageVersion("brms")`) in R (version `r getRversion()`) [@brms_package; @R_core].
+Temporal trends for five variables related to scientific publications were analyzed using Bayesian Generalized Additive Models (GAMs) implemented via the `brms` package (version `r packageVersion("brms")`) in R (version `r getRversion()`) [@brms_package; @rcoreteamLanguageEnvironmentStatistical2024].
 `brms` utilizes Stan (version `r rstan::stan_version()`) for posterior sampling via dynamic Hamiltonian Monte Carlo (specifically, the No-U-Turn Sampler, NUTS) [@stan_dev].
@@
 
-# --- Set options for brms (optional, can help speed up) ---
+# --- Set options for brms 
 options(mc.cores = parallel::detectCores()) # Use all available CPU cores
@@
   filter(sapply(mod_brms, inherits, "brmsfit"))
+
+pretty_num <- function(x) format(x, big.mark = ",", scientific = FALSE)
+
+n_chains <- pretty_num(brms::nchains(results_brms_valid$mod_brms[[1]]))
+n_total_iter <- pretty_num(results_brms_valid$mod_brms[[1]]$fit@sim$iter)
+n_warmup <- pretty_num(results_brms_valid$mod_brms[[1]]$fit@sim$warmup)
+n_pos_samples <- 
+  pretty_num(
+  (results_brms_valid$mod_brms[[1]]$fit@sim$iter - results_brms_valid$mod_brms[[1]]$fit@sim$warmup) * brms::nchains(results_brms_valid$mod_brms[[1]])
+  )
+
 ```
@@
 
-Models were fitted using `r brms::nchains(results_brms_valid$mod_brms[[1]])` Markov chains.
-Each chain consisted of `r results_brms_valid$mod_brms[[1]]$fit@sim$iter` total iterations, with the first `r results_brms_valid$mod_brms[[1]]$fit@sim$warmup` iterations discarded as warmup, leaving `r (results_brms_valid$mod_brms[[1]]$fit@sim$iter - results_brms_valid$mod_brms[[1]]$fit@sim$warmup) * brms::nchains(results_brms_valid$mod_brms[[1]])` posterior samples per model for inference.
+Models were fitted using `r n_chains` Markov chains.
+Each chain consisted of `r n_total_iter` total iterations, with the first `r n_warmup` iterations discarded as warmup, leaving `r n_pos_samples` posterior samples per model for inference.
 No thinning was applied (thin = 1).
@@
 
-Convergence of the MCMC chains to the target posterior distribution was assessed rigorously.
+Convergence of the MCMC chains to the target posterior distribution was assessed rigorously (@tbl-convergence-summary-table).
 
-```{r convergence_summary_table, echo=FALSE}
+```{r , echo=FALSE}
+#| label: tbl-convergence-summary-table
 #| tbl-cap: "Summary of MCMC Convergence Diagnostics.
@@
 ```{r fig.height=3 * length(results_brms_valid$mod_brms)}
-#| label: fig-trace-plots-mcmc-trace,
+#| label: fig-trace-plots-mcmc-trace
 #| echo: false
@@
 
-As shown in Figure @ref(fig:ppc-plot), the models generally demonstrated good fit, capturing the central tendency, spread, and overall shape of the observed data distributions well.
+As shown in @fig-ppc-plot, the models generally demonstrated good fit, capturing the central tendency, spread, and overall shape of the observed data distributions well.
 The Negative Binomial models effectively captured the skewness of the count data (N.
@@
 
-Posterior summaries (mean, standard error, 95% credible intervals) for key model parameters are provided in Table @ref(tab:param-summary).
+Posterior summaries (mean, standard error, 95% credible intervals) for key model parameters are provided in @tbl-param-summary.
 This includes intercepts, standard deviations of smooth terms (`sds`), and relevant distributional parameters (`shape`, `sigma`, `phi`, `zoi`, `coi`).
@@
 ```{r}
-#| label: tab:param-summary
+#| label: tbl-param-summary
 #| echo: false
@@
 ```{r}
-#| label: fig-smooth-plots
-#| fig-cap: "Estimated conditional effect of year (s(year)) on each response variable from the Bayesian GAMs.
-The blue line represents the posterior mean effect, and the grey ribbon represents the 95% credible interval.
-Y-axis scales differ and correspond to the link function of each model (identity, log, or logit)."
+#| results: "hide"
 
@@
 )
+```
+
+```{r}
+#| label: fig-smooth-plots
+#| fig-cap: "Estimated conditional effect of year (s(year)) on each response variable from the Bayesian GAMs.
+The blue line represents the posterior mean effect, and the grey ribbon represents the 95% credible interval.
+Y-axis scales differ and correspond to the link function of each model (identity, log, or logit)."
 

### 2025-04-30 analysis/code/009-exploring-gams.R

@@ -1,242 +1,0 @@
-
-over_time <-  
-  items_df %>%  
-  left_join(items_df_title) %>%  
-  left_join(shannon_per_item) %>%  
-  filter(relative_title_length != -Inf, 
-         relative_title_length !=  Inf,
-         prices_index != "NaN" 
-  ) %>% 
-  mutate(log_authors_n = log(authors_n), 
-         log_pages_n = log(pages_n), 
-         journal_wrp = str_wrap(journal, 30)) %>%  
-  select(year, 
-         log_authors_n, 
-         log_pages_n, 
-         prices_index,  
-         shannon, 
-         relative_title_length,
-         authors_n,
-         pages_n) 
-
-over_time_long <-  
-  over_time %>%  
-  ungroup() %>%  
-  select(-journal) %>%  
-  gather(variable,  
-         value, 
-         -year) %>%  
-  filter(value != -Inf, 
-         value !=  Inf) %>%  
-  mutate(variable = case_when( 
-    variable == "prices_index"  ~ "Recency of references", 
-    variable == "shannon"  ~ "Diversity of references", 
-    variable == "relative_title_length"  ~ "Relative title length (ln)", 
-    variable == "authors_n" ~ "N. of authors",
-    variable == "pages_n" ~ "N. of pages"
-  )) %>%  
-  filter(!is.na(variable)) %>%  
-  filter(!is.nan(value)) %>%  
-  filter(!is.na(value)) %>%  
-  filter(value != "NaN") %>%  
-  mutate(value = parse_number(value))
-
-# compute beta estimates so we can colour lines to indicate more or
-# less hard
-
-library(broom)
-
-over_time_long_models <- 
-  over_time_long %>%  
-  group_nest(variable) %>%  
-  mutate(model = map(data, ~tidy(lm(value ~ year, data = .)))) %>%  
-  unnest(model) %>%  
-  filter(term == 'year') %>%  
-  mutate(becoming_more_scientific = case_when( 
-    variable == "N. of authors"         & estimate > 0 ~ "TRUE", 
-    variable == "N. of pages"           & estimate < 0 ~ "TRUE", 
-    variable == "N. of refs (sqrt)"          & estimate < 0 ~ "TRUE", 
-    variable == "Recency of references"      & estimate > 0 ~ "TRUE", 
-    variable == "Relative title length (ln)" & estimate > 0 ~ "TRUE", 
[…]

### 2025-05-01 analysis/paper/paper.qmd
GAM models nearly done; hand edit PCA plot

@@
 abstract: |
-  Computational reproducibility is increasingly essential for advancing archaeological science, yet its implementation remains inconsistent across the discipline.
-This study presents a bibliometric assessment of archaeology’s status relative to other fields alongside a systematic review of computational reproducibility in published archaeological research.
+  The status of archaeology as a science has been debated for decades.
+This study presents a bibliometric assessment of archaeology’s status relative to other fields, alongside a systematic review of computational reproducibility in published archaeological research.
+Reproducibility is a factor in the hardness/softness of a field because of it importance in establishing consensus.
 Analyzing nearly 10,000 articles, I identify trends in authorship, citation practices, and related metrics that position archaeology between the natural and social sciences.
@@
 
+# these data were prepared using the code in 000-import-raw-data.R
 items_df <- 
@@
 
-source(here::here("analysis/code/008-redraw-Fanelli-and-Glanzel-Fig-2.R"))
+source(here::here("analysis/code/001-redraw-Fanelli-and-Glanzel-Fig-2.R"))
 
@@
 Data points represent individual articles.
-The colour of the points indicates if the overall trend is toward softer (orange) or harder (green) Generalized Additive Models were computed to fit the lines summarising the relationships between the variables and the time series.
+The colour of the points indicates if the overall trend is toward softer (orange) or harder (green).
+Bayesian Generalized Additive Models were computed to fit the lines summarising the relationships between the variables and the time series.
+For recency of references a Zero-One Inflated Beta distribution family was used with a logit link function.
+For diversity of references and relative title length (ln) the Gaussian family was used with the canonical identity link function.
+For the number of authors pages the Negative Binomial family was used with a standard log link function.
+Further details about the model specifications and diagnostics are presented in the supplementary materials.
 #| fig-height: 4
 
-library(ggpmisc)
-library(mgcv)
-formula <-  y ~ x
-
-over_time_long_colour_gams <- 
-over_time_long_colour %>% 
-  nest(.by = variable) %>% 
-  mutate(mod_gam = lapply(data, 
-                          function(df) gam(value ~ s(year, bs = "cr"), 
-                                           data = df)))
-
-over_time_long_colour_gams_summary <- 
-over_time_long_colour %>% 
-  nest(.by = variable) %>% 
-  mutate(fit = map(data, ~mgcv::gam(value ~ s(year, bs = "cs"), data = .)),
-         results = map(fit, glance),
-         R.square = map_dbl(fit, ~ summary(.)$r.sq)) %>%
-  unnest(results) %>%
-  select(-data, -fit) %>% 
-  select(variable, adj.r.squared)
-
-over_time_long_colour_gams_summary_df <- 
-over_time_long_colour %>% 
-  left_join(over_time_long_colour_gams_summary)
-
-ggplot() + 
-  geom_point(data = over_time_long_colour_gams_summary_df, 
-       aes(year,  
-           value, 
-           colour = becoming_more_scientific),
-       alpha = 0.5) + 
-  geom_smooth(data = over_time_long_colour_gams_summary_df,
-              aes(year, value),
-              method="gam", 
-              formula = y ~ s(x, bs = "cs"),
-              se = FALSE,  
-              size = 2, 
-              colour = "#7570b3") +
-  facet_wrap( ~ variable,
-              scales = "free_y") + 
-  theme_bw(base_size = base_size) +
-  scale_color_manual(values = c("#d95f02",
-                                "#1b9e77" )) +
-  guides(colour = "none") +
-  ylab("") +
-  geom_text(data = over_time_long_colour_gams_summary_df %>% 
-              group_by(variable) %>% 
-              summarise(max_value = max(value),
-                        adj.r.squared = unique(adj.r.squared)),
-           aes(
-           x = 1980, 
-           y = max_value, 
-           label = paste("Pseudo R² = ", 
-                         signif(adj.r.squared, 
-                                digits = 3))),
-           hjust = 0, 
-           vjust = 1.5,
-           size = 2)
+# get this from the supplement
 
@@
 For example, *PLOS ONE* received its first impact factor in 2010 and in 2011 Nature's *Scientific Reports* began publishing [@malashichev2017open].
-The appearance of Google Scholar in 2004, increasing the discoverability of many works for many researchers, may have also contributed to this increase in range and diversity of references.
+The appearance of Google Scholar in 2004, increasing the discoverability of many works for many researchers, may have also contributed to this increase in diversity of references.
 
@@
              size = 1) +
-  geom_text_repel(data = pca_scores_means,
+  geom_text_repel(data = pca_scores_means %>% 
+                    mutate(label = str_replace(journal, 
+                                               "JOURNAL", 
+                                               "J.")) %>% 
+                    mutate(label = str_remove(label, 
+                                               "-AN\nINTERNATIONAL\nJ.")),
             aes(.fittedPC1,
                 .fittedPC2,
-                label = journal),
+                label = label),
             lineheight = 0.8,
             segment.color = NA,
-            size = 1,
+            force_pull = 10,
+            size = 2.5,
             bg.color = "white",  # Color of the halo
@@
                 label = column),
-            size = 1,
-            lineheight = 0.7,
-            force = 0,
+            size = 2,
+            lineheight = 0.8,
+            force = 10,
+            force_pull = 0,
             segment.color = NA,
@@
   theme_minimal(base_size = base_size) +
-  coord_equal()
+  coord_fixed(xlim = c(-6, 2.5),
+              ylim = c(-3, 2))
+
+# tricky to get the label spacing right, let's save an SVG, edit
+# by hand, then export to PNG and read that file later. 
+ggsave(plot_pca_means,
+       filename = here::here("analysis/figures/plot_pca_means.svg"))
 
@@
 The journals are ordered for each indicator so that within each plot, the harder journals are at the top of the plot and the softer journals are at the base.
-Panel F shows a bar plot that is the single consensus ranking computed from all five variables, using the Borda Count ranking algorithm.
-Panel G is a biplot of the first and second principal components of a PCA computed on the journal means of the five bibliometric variables."
-#| fig-height: 6
+Panel F shows a bar plot that is the single consensus ranking computed from all five variables, using the Borda Count ranking algorithm."
+#| fig-height: 5
 
@@
 
-plot_grid(
   plot_grid(plot_journals_authors, 
@@
            labels = LETTERS[1:6],
-           label_size = 6),
-  plot_pca_means,
-  nrow = 2,
-  labels = c("", LETTERS[7]),
-  rel_heights = c(1, 0.5),
            label_size = 6)
 
+```
+
+```{r}
+#| label: fig-pca
+#| fig-cap: "Biplot of the first and second principal components of a PCA computed on the means of the five bibliometric variables for each journal in the sample.
+The arrows represent the correlation between each original variable and the principal components.
+The direction and length of the arrows indicate how strongly each variable contributes to each component.
+"
+#| fig-height: 4
+
+knitr::include_graphics(here::here("analysis/figures/plot_pca_means.png"))
 ```
@@
 
-The PCA results in panel G show of @fig-variation-by-journal that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left.
+The PCA results in @fig-pca shows that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left.
+PC1 is explained by title length, number of pages, and the diversity of references.
+Journals with higher PC1 values have articles with longer titles, fewer pages, and less diverse reference lists.
 The distribution of PC1 values is skewed left, with most of the journals concentrated at the harder end of the spectrum.
+PC2 is explained by number of authors and recency of references.
 The distribution of PC2 values reveals additional structure to the data and can be roughly separated into generalist journals in the negative range (e.g.
-*American Antiquity*, *Antiquity*, *Advances in Archaeological Practice*), and more specialised journals in the positive range (e.g.
+*American Antiquity*, *Antiquity*, *Advances in Archaeological Practice*), characterised by fewer authors and more recent references.
+In the positive range of PC2 are more specialised journals, characterised by higher numbers of authors and less recent references cited (e.g.
 *Environmental Archaeology*, *Geoarchaeology*, *Archaeologial Research in Asia*, *Journal of Island and Coastal Archaeology*).
@@
 That is, a second researcher might use the same raw data to build the same analysis files and implement the same statistical analysis in an attempt to yield the same results...
-Reproducibility is a minimum necessary condition for a finding to be believable and informative." Scientific reproducibility is linked to consensus because if more researchers provide sufficient detail for others to reproduce their results, then consensus on new knowledge can more often and more rapidly established.
+Reproducibility is a minimum necessary condition for a finding to be believable and informative." Scientific reproducibility is a factor contributing to the hardness of a field.
+Specifically, reproducibility is linked to to concept of consensus in a field because if more researchers provide sufficient detail for others to reproduce their results, then consensus on new knowledge can more often, and more, rapidly be established.
 The importance of this factor can be traced to Irish chemist Robert Boyle (1627-1691), best known for his experiments with vacuum pumps [@shapin2011leviathan].

### 2025-05-01 analysis/paper/supplement-GAMS-details.qmd
GAM models nearly done; hand edit PCA plot

--- 2025-04-30 analysis/code/010-summary-of-gams.qmd
@@
 date: "`r Sys.Date()`"
-bibliography: ../analysis/paper/references.bib
-csl:          ../analysis/paper/templates/journal-of-archaeological-science.csl # Insert path for the bib-style
+bibliography: references.bib
+csl:          templates/journal-of-archaeological-science.csl # Insert path for the bib-style
 output:
@@
 library(knitr)       # For kable
+library(kableExtra)
+library(posterior) # For as_draws_array
+library(bayesplot) # For mcmc_trace and nuts_params
 ```
@@
 The shape of these distributions informed our decisions about model specification.
+
 Recency of references: This variable, bounded between 0 and 1, exhibited a right-skewed distribution with notable occurrences at the boundaries (0 and 1).
@@
 # this takes a few hours
-
-library(brms)
-library(dplyr)
-library(purrr)
 
@@
   diagnostics_table,
-  caption = "Summary of MCMC Convergence Diagnostics. Max R-hat should be close to 1.0. Min ESS values indicate the minimum effective sample size across all parameters. N Divergences should be 0.",
   digits = c(NA, 0, 3, 0, 0, 0), # Chains/ESS/Div as integers, Rhat to 3 decimals
@@
 Well-mixing chains should resemble stationary 'fuzzy caterpillars' with chains overlapping."
-#---------------------------------------------------------------------
-# Load necessary libraries
-#---------------------------------------------------------------------
-library(brms)
-library(purrr)     # For map2
-library(dplyr)     # For tibble etc.
-library(posterior) # For as_draws_array
-library(bayesplot) # For mcmc_trace and nuts_params
-library(ggplot2)   # For ggtitle and theme adjustments
-library(cowplot)   # For plot_grid
-#---------------------------------------------------------------------
 
@@
 )
-#---------------------------------------------------------------------
 
-#---------------------------------------------------------------------
 # Combine and Display Plots
@@
   # Display the combined plot
-  print(combined_trace_plots)
+  combined_trace_plots
 } else {
@@
 #| label: tbl-param-summary
-#| echo: false
 #| message: false
 #| warning: false
-#---------------------------------------------------------------------
-# Load necessary libraries
-#---------------------------------------------------------------------
-library(brms)
-library(purrr)
-library(dplyr)
-library(tibble) # For rownames_to_column
-library(knitr)
-library(kableExtra)
-#---------------------------------------------------------------------
 
@@
 
+# R-squared Equivalent for Model Fit
+
+Standard R-squared isn't directly applicable to many GAM families (especially non-Gaussian ones).
+The recommended metric for brms models is the Bayesian R-squared.
+Calculated using brms::bayes_R2(model), this metric represents the variance of the predicted values divided by the sum of the variance of the predicted values and the expected variance of the residuals.
+It estimates the proportion of outcome variance that can be explained by the model's predictors.
+The results are shown below in @tbl-r-sq.
+
+```{r}
+#| label: tbl-r-sq
+#| tbl-cap: "Bayesian R-squared estimates for each model."
+#| message: false
+#| warning: false
+
+# Calculate Bayesian R-squared for each model
+# Use map to apply to each model, map_dfr to combine
+bayes_r2_summary <- map2_dfr(
+  results_brms_valid$mod_brms,
+  results_brms_valid$variable,
+  ~{
+    model <- .x
+    var_name <- .y
+    r2_est <- NA_real_
+    r2_lower <- NA_real_
+    r2_upper <- NA_real_
+
+           # Calculate Bayesian R2 (returns matrix with Estimate, Est.Error, Q2.5, Q97.5)
+           r2_obj <- brms::bayes_R2(model, ndraws = 10000)
+           # Extract the point estimate and CI bounds
+           r2_est <- r2_obj[1, "Estimate"]
+           r2_lower <- r2_obj[1, "Q2.5"]
+           r2_upper <- r2_obj[1, "Q97.5"]
+
+    tibble(
+        Variable = var_name,
+        `Bayes R2` = r2_est,
+        `Lower 95% CI` = r2_lower,
+        `Upper 95% CI` = r2_upper
+        )
+  }
+)
+
+kable(bayes_r2_summary, 
+      digits = 2, 
+      booktabs = TRUE) %>%
+kable_styling(latex_options = "scale_down")
+```
+
 ## Estimated Temporal Trends
@@
 The estimated non-linear effect of publication year on each response variable is visualized below (@fig-smooth-plots), showing the contribution of the smooth term `s(year)` to the model's linear predictor (on the respective link scale).
+This allows us to visually assess the significance of the model.
+
+Recency of references: There is strong evidence for a significant temporal trend.
+The 95% CI falls convincingly below zero from approximately the early 1990s onwards, indicating a significant decrease in the contribution of year to recency (on the logit scale) during the later decades of the study period.
+Evidence for an effect differing from zero in the earliest years is less conclusive as the CI approaches the zero line.
+
+Diversity of references: The effect of year on reference diversity shows strong evidence of being non-zero across most of the timeframe.
+The 95% CI is clearly separated from zero, initially below and later rising significantly above it, particularly after 2010, indicating substantial temporal variation not attributable to chance.
+
+Relative title length (ln): The evidence suggests a significant effect of year primarily concentrated around 1990, where the 95% CI lies entirely below the zero line.
+During the initial years and from roughly 2000 onwards, the CI overlaps zero, indicating that the evidence for an effect of year being different from zero is weak or uncertain in these periods.
+
+N.
+of authors: Strong evidence supports a significant positive effect of year on the number of authors (on the log scale).
+The 95% CI is consistently above the zero line from the late 1980s onwards, demonstrating a significant increasing trend over the majority of the period examined.
+
+N.
+of pages: The temporal effect on the number of pages appears significant during specific periods.
+The 95% CI is clearly above zero during the peak in the 1990s, providing strong evidence for an increased effect during that time.
+However, the CI overlaps zero during the trough in the mid-2000s and potentially near the end of the period, suggesting uncertainty about whether the effect differs significantly from zero during these phases.
+There is also evidence for a negative effect relative to the intercept in the earliest years.
 
 ```{r}
-#| results: "hide"
-
-# Load necessary libraries
-library(brms)
-library(purrr)
-library(dplyr)
-library(ggplot2)
-library(cowplot)
+#| label: fig-smooth-plots
+#| fig-cap: "Estimated conditional effect of year (s(year)) on each response variable from the Bayesian GAMs.
+The blue line represents the posterior mean effect, and the grey ribbon represents the 95% credible interval.
+Y-axis scales differ and correspond to the link function of each model (identity, log, or logit).
+The red line is y=0 to visually assess the significance of the  (i.e., whether the credible interval excludes zero)."
 
@@
     # Generate the plot object list directly
-    plot_object_list <- plot(conditional_smooths(model), plot = TRUE)
+    plot_object_list <- plot(conditional_smooths(model), plot = FALSE)
 
@@
     plot_object_list[[1]] + 
+      geom_hline(yintercept = 0, 
+                 color = "red", 
+                 linetype = "dashed", 
+                 linewidth = 0.8) +
       ggtitle(var_name) +
       ylab(y_axis_label) +
+      theme_minimal() +
       theme(plot.title = element_text(hjust = 0.5, size = 10)) # Center title
@@
 )
+
+# plot a panel of all the plots
+plot_grid(plotlist = conditional_smooths_plots_list, ncol = 3)
 ```
@@
 ```{r}
-#| label: fig-smooth-plots
-#| fig-cap: "Estimated conditional effect of year (s(year)) on each response variable from the Bayesian GAMs.
-The blue line represents the posterior mean effect, and the grey ribbon represents the 95% credible interval.
-Y-axis scales differ and correspond to the link function of each model (identity, log, or logit)."
+# plot for manuscript
+# Generate conditional smooth plots within pmap
 
-# Combine the plots into a grid (adjust ncol as desired)
-# The plot_grid call itself will display the combined plot
+input_list <-   list(
+    model = results_brms_valid$mod_brms,
+    var_name = results_brms_valid$variable,
+    raw_data = results_brms_valid$data,
+    r_sq = bayes_r2_summary$`Bayes R2`
+  )
+
+conditional_smooths_plots_fn <- function(model, var_name, raw_data, r_sq){
+
+# Generate the plot object list 
+     smooth_data <- conditional_smooths(model, plot = FALSE)[[1]]
+     intercept_est <- fixef(model)["Intercept", "Estimate"]
+
+# --- Determine inverse link function and response label ---
+
+        inv_link <- identity # Default for Gaussian
+        response_label <- var_name # Default label
+
+        if (var_name %in% c("N. of authors", "N. of pages")) {
+          inv_link <- exp # Inverse of log link
+          response_label <- paste("Estimated", var_name)
+        } else if (var_name == "Recency of references") {
+          inv_link <- plogis # Inverse of logit link
+          response_label <- "Estimated Recency of references (Price Index)"
+        } else if (var_name == "Diversity of references") {
+           response_label <- "Estimated Diversity of references"
+        } else if (var_name == "Relative title length (ln)") {
+           response_label <- "Estimated Rel. Title Length (ln)"
+        }
+
+# --- Add intercept and apply inverse link function so we can plot on the scale of the raw data
+        smooth_data <- 
+          smooth_data %>%
+          mutate(
+            # Calculate estimates on response scale
+            response_est = inv_link(intercept_est + estimate__),
+            response_lower = inv_link(intercept_est + lower__),
+            response_upper = inv_link(intercept_est + upper__)
+          )
+
+    ggplot(smooth_data, 
+           aes(x = year, 
+               y = response_est)) +
+            geom_point(data = raw_data %>% 
+                         mutate(colour_hex = case_when(
+                           becoming_more_scientific == "TRUE" ~ "#1b9e77",
+                           .default = "#d95f02"
+                         )),
+                 aes(x = year,
+                     y = value,
+                     colour = colour_hex),
+                 alpha = 0.7) +
+      scale_color_identity() +
+      # Add the line 
+      geom_line(color = "#7570b3", 
+                linewidth = 2) +
+      geom_text(data = raw_data  %>% 
+                  mutate(`Bayes R2` = r_sq) %>% 
+                  mutate(max_value = max(value)) %>% 
+                  distinct(`Bayes R2`, 
+                           max_value),
+           aes(
+           x = 1980, 
+           y = max_value, 
+           label = paste("Pseudo R² = ", 
+                         signif(`Bayes R2`, 
+                                digits = 3))),
+           hjust = 0, 
+           vjust = 1.5,
+           size = 2) +
+    theme_bw(base_size = 6) +
+      labs( 
+           y = response_label, 
+           x = "")
+}
+
+conditional_smooths_plots_list <- 
+  pmap(input_list,
+       conditional_smooths_plots_fn)
+```
+
+Here is the figure that appears in the main manuscript:
+
+```{r}
+#| label: fig-smooth-plots-paper
+#| fig-cap: "Distribution of article characteristics for archaeology articles over time and estimated conditional effect of year (purple lines).
+Data points represent individual articles.
+The colour of the points indicates if the overall temporal trend is toward archaeology become a softer (orange) or harder (green) field.
+Bayesian Generalized Additive Models were computed to fit the lines summarising the relationships between the variables and the time series.
+For recency of references, a Zero-One Inflated Beta distribution family was used with a logit link function.
+For diversity of references and relative title length (ln), the Gaussian family was used with the canonical identity link function.
+For the number of authors pages the Negative Binomial family was used with a standard log link function.
+Further details about the model specifications and diagnostics are presented in the supplementary materials."
+
 plot_grid(plotlist = conditional_smooths_plots_list, ncol = 3)

### 2025-05-01 analysis/code/002-article-length.R

@@ -1,53 +1,0 @@
-#  ------------------------------------------------------------------
-# length of article - less need to introduce, justify and explain study [47,52]
-
-# all on one
-library(tidyverse)
-
-items_df %>% 
-  ggplot(aes(year,
-             log(pages_n))) +
-  geom_boxplot(colour = "red", 
-               size = 1)  +
-  scale_y_continuous(limits = c(0, 5)) +
-  scale_x_continuous(labels = NULL,
-                     name = NULL) +
-  theme_minimal() +
-  theme(panel.grid  = element_blank()) +
-  coord_equal(ratio = 6)
-
-ggsave("figures/box_num_pages.svg")
-
-
-# over time
-items_df %>% 
-  ggplot(aes(year,
-             log(pages_n),
-             group = year)) +
-  geom_boxplot() +
-  geom_jitter(alpha = 0.01) +
-  theme_minimal() +
-  ylab("Number of pages (ln)") 
-
-ggsave("figures/box_num_pages_over_time.png")
-
-# by journal, wow yes!
-library(ggforce)
-items_df %>% 
-  mutate(journal = str_wrap(journal, 20)) %>% 
-  ggplot(aes(y = reorder(journal,
-                     pages_n), 
-             x = log(pages_n),
-             fill = ..x..,
-             height = ..density..)) +
-  geom_density_ridges_gradient(stat = "density",
-                               colour = "white") +
-  scale_fill_viridis_c() +
-  guides(fill = 'none') +
-  theme_minimal() +
-  theme(axis.text.y = element_text(size = 6)) +
-  xlab("Number of pages (ln)") +
-  ylab("")
-
-ggsave("figures/box_num_pages_by_journal.png")
-

### 2025-05-01 analysis/code/005-first-person.R

@@ -1,61 +1,0 @@
-
-# use of first person (singular vs. plural) in abstracts, - universal validity of claims; less scope for argumentation; fewer appeals to opinion and authority [59], the proportion of first person pronouns, both singular and plural (i.e. ‘‘I’’, ‘‘me’’, ‘‘mine’’, ‘‘we’’, ‘‘our’’ etc.) among all words in the abstract.
-
-i <- c(" i'd ", " i'll ", " i'm ", " i've ", " i ")
-me <- c(" me ", " my ", " mine ")
-us <- c(" us ", " our ", " ours ")
-first_persons <- c(i, me, us)
-
-library(stringi)
-
-# function to compute proportion of first person pronouns
-first_person_prop <- function(x){
-  x <- tolower(x)
-  firsts <- sum(str_count(x, first_persons))
-  abstract_n_words <- stri_count_words(x)
-  firsts / abstract_n_words
-}
-
-
-# compute proportions, single author papers only
-items_df_abstract <- 
-  items_df %>% 
-  mutate(abstract_first_person = first_person_prop(abstract)) 
-
-# plots
-# all on one
-items_df_abstract %>% 
-  ggplot(aes(year,
-             abstract_first_person)) +
-  geom_boxplot()  +
-  scale_x_continuous(labels = NULL,
-                     name = NULL) +
-  theme_minimal()
-
-# over time
-items_df_abstract %>% 
-  ggplot(aes(year,
-             abstract_first_person,
-             group = year)) +
-  geom_boxplot() +
-  geom_jitter(alpha = 0.01) +
- # scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
-  theme_minimal()
-
-# by journal
-items_df_abstract %>% 
-  mutate(journal = str_wrap(journal, 20)) %>% 
-  ggplot(aes(y = reorder(journal,
-                     abstract_first_person), 
-             x = abstract_first_person,
-             fill = ..x..,
-             height = ..density..)) +
-  geom_density_ridges_gradient(stat = "density",
-                               colour = "white") +
-  scale_fill_viridis_c() +
-  guides(fill = 'none') +
-  theme_minimal() +
-  theme(axis.text.y = element_text(size = 6))  +
-  ylab("") +
-  xlab("Proportion of first person pronouns (singular vs. plural) in abstracts")
-

### 2025-05-07 analysis/paper/paper.qmd
editing with peer reviews

@@
 abstract: |
-  The status of archaeology as a science has been debated for decades.
-This study presents a bibliometric assessment of archaeology’s status relative to other fields, alongside a systematic review of computational reproducibility in published archaeological research.
+  The status of archaeology as a science has been debated for decades and influences how we practice and teach archaeology.
+This study presents a novel bibliometric assessment of archaeology’s status relative to other fields using a hard/soft framework.
+It also presents a systematic review of computational reproducibility in published archaeological research.
 Reproducibility is a factor in the hardness/softness of a field because of it importance in establishing consensus.
@@
 The PCA results in @fig-pca shows that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left.
-PC1 is explained by title length, number of pages, and the diversity of references.
+The variables that contribute to variation in PC1 are title length, number of pages, and the diversity of references.
 Journals with higher PC1 values have articles with longer titles, fewer pages, and less diverse reference lists.
 The distribution of PC1 values is skewed left, with most of the journals concentrated at the harder end of the spectrum.
-PC2 is explained by number of authors and recency of references.
-The distribution of PC2 values reveals additional structure to the data and can be roughly separated into generalist journals in the negative range (e.g.
+Variation in PC2 is influenced by the number of authors and recency of references.
+The distribution of PC2 values reveals additional structure to the data and can be roughly separated into generalist journals in the negative range of the PC2 axis (e.g.
 *American Antiquity*, *Antiquity*, *Advances in Archaeological Practice*), characterised by fewer authors and more recent references.
-In the positive range of PC2 are more specialised journals, characterised by higher numbers of authors and less recent references cited (e.g.
+In the positive range of the PC2 axis are more specialised journals, characterised by higher numbers of authors and less recent references cited (e.g.
 *Environmental Archaeology*, *Geoarchaeology*, *Archaeologial Research in Asia*, *Journal of Island and Coastal Archaeology*).
-The *Journal of Archaeological Science* sits about midway between these two groups, reflecting its relevance to specialised communities of practice in archaeology, as well as generalist communities of practice.
+The *Journal of Archaeological Science* sits about midway between these two groups, reflecting its relevance to both specialised and generalist communities of practice in archaeology.
 

### 2025-05-08 analysis/paper/paper.qmd
discussing qualitative research

@@
 Drawing on a different group of philosophers, @Smith_2017 argues for archaeology more specifically as a social science.
+@bevan2015data proposes that floods of digital data are reconfiguring our analytical agendas and support empirical and inductive inference.
 Counter-arguments to archaeology as a science come from numerous directions, notably @hodder1985postprocessual who rejected the quest for generalisations and instead argued that archaeology should be subjective and reflective, focussed on symbolic and relational meanings of material culture and the historical particularity of past human cultures.
@@
 @fanelliBibliometricEvidenceHierarchy2013 found a spectrum of statistically significant variation in bibliometric variables from the physical to the social sciences, with papers tending at the softer end of the spectrum tending to have fewer co-authors, use less substantive titles, have longer texts, cite older literature, and have a higher diversity of sources.
-In @fanelliBibliometricEvidenceHierarchy2013's analysis harder sciences include Space Science, Physics, Chemistry, less hard sciences include social sciences include Psychiatry, Psychology, Economics, Business and General Social Sciences, and the Humanities define the soft end of the spectrum.
+In @fanelliBibliometricEvidenceHierarchy2013's analysis harder sciences include Space Science, Physics, Chemistry, softer sciences include social sciences (Psychiatry, Psychology, Economics, Business and General Social Sciences), and the Humanities define the soft end of the spectrum.
 Following @fanelliBibliometricEvidenceHierarchy2013, I quantify the number of authors, length of article, relative title length, age of references, and diversity of references for a large sample of peer-reviewed journal articles.
@@
 Empirical replications of 100 studies published in three psychology journals found that 36% of replications had statistically significant results, compared to 96% of the original studies [@OpenScienceCollaboration2015].
-Similar empirical replications of large numbers of social science studies and experiemental economics studies successfully replicated 61% and 62% of their target studies respectively [@camererEvaluatingReplicabilitySocial2018; @camererEvaluatingReplicabilityLaboratory2016].
+Similar empirical replications of large numbers of social science studies and experimental economics studies successfully replicated 61% and 62% of their target studies respectively [@camererEvaluatingReplicabilitySocial2018; @camererEvaluatingReplicabilityLaboratory2016].
 
@@
 Similarly among many experimental archaeologists, empirical reproducibility is a key concern, for example in lithic use-wear identification [@hayesLearningBlindTests2017] and the measurement of lithics [@pargeterReplicabilityLithicAnalysis2023].
-To what extent does this concern extend to computational reproducibility among archaeologists?
+The increasing availability of large digital datasets is pushing archaeology into unexplored areas [@bevan2015data], inviting questions about what reproducibility means for data intensive archaeological research.
+For example, to what extent does concern for reproducibility extend to computational reproducibility among archaeologists?
 
@@
 (aer_plot_software + aer_plot_methods + aer_plot_repo) / (aer_plot_issues + aer_plot_issues_lang) + plot_annotation(tag_levels = 'A')
-
 ```
@@
 First, authors should use a simple and logical folder structure to organise their code and data to be as self-contained as possible (ie.
-not requiring files outside of the project folder).
+not requiring files outside of the project folder or any modifications by a reader to run the code).
 There are many excellent, simple, and widely-used templates that authors can choose from, e.g.
@@
 Dockerfiles are gradually being adopted by archaeologists, see @cremaModellingDiffusionInnovation2024 and @liaoApplyingMaskRCNN2024 for examples, and most of our reproducibility reviews include a recommendation that the author include a Dockerfile to manage complex dependencies efficiently.
+
+## What about qualitative archaeological research?
+
+Although the introduction of reproducibility reviews signifies a growth in computational archaeology and a desire to evaluate the research products beyond the journal article, a very substantial amount of archaeological research is qualitative, with few or no numerical data involved in making knowledge claims about the human past.
+For example, many archaeological questions can be answered by the simple presence or absence of artefacts or features, or qualitative comparisons of basic artefact characteristics such as shape, colour, surface treatments, and raw material.
+Chaîne opératoire analyses by ceramic and lithic specialists is an especially productive area of archaeological research that often relies on comparison of narratives of manufacturing processes.
+Do these studies count as science and should they be held to standards of reproducibility in the same way that computationally intensive research is?
+
+While there is a strong case for qualitative research counting as science, because it is often a systematic, empirical, and rigorous process of inquiry, the question of
 
@@
 
-Computational reproducibility is not a panacea, results that are fully reproducible can contain errors and fraud.
+Computational reproducibility is not a panacea; it should not be used as a universally accepted criterion for research quality [@Leonelli2018].
+Results that are fully reproducible can contain errors and fraud.
 It is no guarantee of code quality, or that statistics have been used appropriately (cf.

### 2025-05-09 analysis/paper/paper.qmd
qualitative studies and repro

@@
 Chaîne opératoire analyses by ceramic and lithic specialists is an especially productive area of archaeological research that often relies on comparison of narratives of manufacturing processes.
-Do these studies count as science and should they be held to standards of reproducibility in the same way that computationally intensive research is?
-
-While there is a strong case for qualitative research counting as science, because it is often a systematic, empirical, and rigorous process of inquiry, the question of
+While these studies unquestionably count as science, because they are a systematic, empirical, and rigorous process of inquiry, should they be held to standards of reproducibility in the same way that computationally intensive research is?
+
+A similar debate has been unfolding more broadly about the humanities where Peels and Bouter [-@peelsHumanitiesNeedReplication2018; -@peelsPossibilityDesirabilityReplication2018] have argued that humanities disciplines that use empirical methods should be assessed by how well reanalysis of the original or new data using original or new methods produces the original or equivalent results.
+Resisting this proposal, @derijckeResistCallsReplicability2018 argue that because humanities research is unique because it pursues value and meaning, and a given study can produce multiple valid answers relating to the value and meaning of a study object, replication is irrelevant as a mark of quality.
+@peelsEpistemicValuesHumanities2018 disputes this uniqueness, claiming that the humanities has the same epistemic values as the sciences, however some values have more weight in the humanities while others have more weight in the sciences, and unlike in the sciences, humanities scholars often study these epistemic values themselves.
+A consensus seems to be emerging that
 

### 2025-05-17 README.Rmd
update readme

@@
 
--   [:file_folder: paper](/analysis/paper): Quarto source document for
-    manuscript. Includes code to reproduce the figures and tables
-    generated by the analysis.
+-   [:file_folder: paper](/analysis/paper): Quarto source documents for the 
+    **manuscript** and **supplementary document** on GAMs. Includes code to 
+    reproduce the figures and tables generated by the analysis.
 -   [:file_folder: data](/analysis/data): Data used in the analysis.
@@
 5. Go to http://localhost:8787/ with your browser and enter `rstudio` in both fields to start RStudio in your browser, running in the Docker container with all the required depedencies installed
-6. In RStudio, run `rstudioapi::openProject("/home/rstudio/web-of-science-archaeology")` to open the compendium's project
+6.
+In RStudio, run `rstudioapi::openProject("/home/rstudio/web-of-science-archaeology")` to open the compendium's project.
+This project uses [`renv`](https://rstudio.github.io/renv/) to document and manage dependencies, this will be activated when the project is first opened in RStudio.
 7. Run ` rmarkdown::render("analysis/paper/paper.qmd", output_dir='/tmp')` to render the Quarto document that is the manuscript submitted for publication

### 2025-05-17 analysis/paper/paper.qmd
update readme

@@
 It also presents a systematic review of computational reproducibility in published archaeological research.
-Reproducibility is a factor in the hardness/softness of a field because of it importance in establishing consensus.
+Reproducibility is a factor in the hardness/softness of a field because of its importance in establishing consensus.
 Analyzing nearly 10,000 articles, I identify trends in authorship, citation practices, and related metrics that position archaeology between the natural and social sciences.
@@
 
-In their paper celebrating the 40th anniversary of this journal Torrence et al [-@torrenceFortyYearsStill2015] noted that reproducibility was an issue important to the reputation and sustainability of the discipline, and necessary for archaeological science to behave like a science.
+In their paper celebrating the 40th anniversary of this journal Torrence et al.
+[-@torrenceFortyYearsStill2015] noted that reproducibility was an issue important to the reputation and sustainability of the discipline, and necessary for archaeological science to behave like a science.
 As part of the celebration of the 50th anniversary, and of Torrence's leadership of the journal, my contribution revisits these topics of archaeology's status as a science, this journal's place in the landscape of archaeological science, and how the journal has responded to a growing recognition of the importance of reproducibility.
 I first present bibliometric evidence of the position of archaeology as a whole, and this journal in particular, in the sciences.
-Next I report on the journal's progress in supporting reproducible research, and my work doing a new kind of peer review for JAS, one that evaluates the computational reproducibilty of the research submitted for publication.
+Next, I report on the journal's progress in supporting reproducible research, and my work doing a new kind of peer review for JAS, one that evaluates the computational reproducibility of the research submitted for publication.
 Finally, I analyse twelve months of reproducibility reviews to identify common weaknesses in the ways archaeologists are working currently, and provide simple recommendations for researchers to overcome these and contribute to the improvement of computational reproducibility in archaeological science.
@@
 At the 'hard' end of the spectrum (e.g.
-physics and chemistry), scholars more typically share a large set of established set of theories, facts, and methods, facilitating fairly rapid agreement on the validity and significance of new results.
+physics and chemistry), scholars more typically share a large set of established set of theories, facts, and methods, facilitating fairly rapid agreement on the validity and significance of new results [@biglan1973characteristics].
 At the 'soft' end of the spectrum (e.g.
@@
 Hardness and softness is a controversial distinction, in part because it is sometimes used to imply a rank order of disciplines that encodes legitimacy, productivity, perceived value to society, and worthiness of funding [@DifferentAgenda2012; @coleHierarchySciences1983].
-On the other hand, analyzing the characteristics that lead to the hard-soft distinction can be useful for understanding the diversity of academic inquiry, such as how different fields approach knowledge and differences in what counts as evidence and modes of argument, where fruitful collaborations might be possible due to shared methods and assumptions, and for curriculum design to structure courses appropriately based on a field's typical ways of knowing.
+Another criticism is that it may be more of an emergent product of social and institutional processes rather than intrinsic differences in method or consensus [@latour1987science].
+On the other hand, analyzing the characteristics that lead to the hard-soft distinction can be useful for understanding the diversity of academic inquiry, such as how different fields approach knowledge and differences in what counts as evidence and modes of argument, where fruitful collaborations might be possible due to shared methods and assumptions, and for curriculum design to structure courses appropriately based on a field's typical ways of knowing [@becher2001academic].
 
@@
 @fanelliBibliometricEvidenceHierarchy2013 found a spectrum of statistically significant variation in bibliometric variables from the physical to the social sciences, with papers tending at the softer end of the spectrum tending to have fewer co-authors, use less substantive titles, have longer texts, cite older literature, and have a higher diversity of sources.
-In @fanelliBibliometricEvidenceHierarchy2013's analysis harder sciences include Space Science, Physics, Chemistry, softer sciences include social sciences (Psychiatry, Psychology, Economics, Business and General Social Sciences), and the Humanities define the soft end of the spectrum.
+In @fanelliBibliometricEvidenceHierarchy2013's analysis harder sciences include Space Science, Physics, Chemistry, softer sciences include social sciences (Psychiatry, Psychology, Economics, Business, and General Social Sciences), and the Humanities define the soft end of the spectrum.
 Following @fanelliBibliometricEvidenceHierarchy2013, I quantify the number of authors, length of article, relative title length, age of references, and diversity of references for a large sample of peer-reviewed journal articles.
@@
 To focus on journals of broad relevance to most archaeologists, and that are representative of substantial communities of practice, I then filtered the results to keep only articles published in the top-ranking 25 journals according to their h-indices as reported by Clarivate’s Journal Citation Indicator.
-Finally I excluded journals with less than 100 articles in the database, resulting in `r n_journals` journals.
+Finally, I excluded journals with less than 100 articles in the database, resulting in `r n_journals` journals.
 
@@
 The clearest indicator of archaeology as a soft science is article length where it is similar to the humanities.
-Overall archaeology does not sit squarely at either end of the hard-soft spectrum.
+Overall, archaeology does not sit squarely at either end of the hard-soft spectrum.
 It is generally not a harder science than the social sciences, with the exception of collaborator group sizes.
@@
 
-# get this from the supplement
+# get this from the supplement on GAMS
+knitr::include_graphics(here::here("analysis/figures/fig-smooth-plots-paper.png"))
 
@@
 
-The *Journal of Archaeological Research* is notable because it consistently ranks as soft, ranking as the softest journal for four of our five bibliometric variables.
+The *Journal of Archaeological Research* is notable for consistently ranking as soft; it was the softest journal for four of our five bibliometric variables.
 This is a predicable result for a review journal, which is a distinct type of journal dedicated to summarizing, analyzing, and synthesizing existing research in a particular field.
@@
 In the positive range of the PC2 axis are more specialised journals, characterised by higher numbers of authors and less recent references cited (e.g.
-*Environmental Archaeology*, *Geoarchaeology*, *Archaeologial Research in Asia*, *Journal of Island and Coastal Archaeology*).
+*Environmental Archaeology*, *Geoarchaeology*, *Archaeological Research in Asia*, *Journal of Island and Coastal Archaeology*).
 The *Journal of Archaeological Science* sits about midway between these two groups, reflecting its relevance to both specialised and generalist communities of practice in archaeology.
@@
 The outlook for the future of archaeology is also complex, with three out of five of the bibliometric variables trending towards more humanistic styles of working, but the discipline showing more extreme values in some metrics towards both hard and soft sciences after about 2010.
-Among archaeology journals, we see distinct communities of practice reflected in the PCA results that are very close together on the hard-soft spectrum, but have minor differences in their communication styles, perhaps due to differences in writing traditions inherited from parent disciplines such as geology and biology.
+Among archaeology journals, we see distinct communities of practice reflected in the PCA results that are very close together on the hard-soft spectrum, but have minor differences in their communication styles, perhaps due to cultural differences in writing traditions inherited from parent disciplines such as geology and biology [@becher2001academic].
 
 While these bibliometric variables provide several interesting insights into the status of archaeology as a science, via measurement of consensus, and are important for moving the debate beyond discussions of a small number of case studies, they miss a crucial factor that separates scientific practice from non-science.
-This is reproducibility, which, according to a report by the US National Science Foundation [@cacioppo2015social], "refers to the ability of a researcher to duplicate the results of a prior study using the same materials as were used by the original investigator.
+This is reproducibility, which, according to a report for the US National Science Foundation [@cacioppo2015social], "refers to the ability of a researcher to duplicate the results of a prior study using the same materials as were used by the original investigator.
 That is, a second researcher might use the same raw data to build the same analysis files and implement the same statistical analysis in an attempt to yield the same results...
@@
 In 2024 the *Journal of Archaeological Science* introduced a new kind of peer review that has provided an opportunity to tackle this question about computational reproducibility in archaeology.
-In January 2024 I accepted the position of 'Associate Editor for Reproducibility' (AER) for JAS and conducted reproducibility reviews of submissions that mentioned programming languages such as R or Python in the methods sections.
-A reproducibility review examines the code and data used to generate the results presented in the paper, and attempts to run the authors' code to reproduce the their results (see @editorsReproducibilityJournalArchaeological for more details about this process).
+In January 2024 I accepted the position of 'Associate Editor for Reproducibility' (AER) for JAS and conducted reproducibility reviews of submissions that mentioned programming languages such as R or Python in the methods sections, taking guidance from similar initiatives in other fields [e.g.
+@ivimey-cookImplementingCodeReview2023; @nustCODECHECKOpenScience2021].
+A reproducibility review examines the code and data used to generate the results presented in the paper, and attempts to run the authors' code to reproduce their results (see @editorsReproducibilityJournalArchaeological for more details about this process).
 This new AER role is based on similar positions (i.e.
 'data editor' or 'reproducibility editor') that journals in economics [@vilhuber2019report], statistics [@wrobel2024partnering], astronomy [@muench2023roles], ecology [@bolnick2022ensuring], and environmental studies [@rosenbergReproducibleResultsPolicy2021] have had, in some cases for over a decade.
-In 2024, three archaeology journals, in addition to JAS, added AERs to their editorial communities: *Advances in Archaeological Practice* [@marwick2024introducing, one paper reviewed], *Journal of Field Archaeology* [@farahani2024reproducibility, two papers reviewed], and *American Antiquity* [@martin2024editors, no papers reviewed].
+In 2024, three archaeology journals, in addition to JAS, added AERs to their editorial communities: *Advances in Archaeological Practice* [@marwick2024introducing, one paper reviewed], *Journal of Field Archaeology* [@farahani2024reproducibility, two papers reviewed], and *American Antiquity* [@martin2024editors, no papers reviewed at the time of writing].
 
@@
 A: Primary software used for the computational analysis reported in a manuscript.
-B: Computational or statistical method used by the authors (GMM = geometric morphometrics; Frequentist = hypothesis tests such as chi-square and ANOVA; AI/ML = artifical intellegence and machine learning, including neural networks and deep learning; MCMC = Markov Chain Monte Carlo,  i.e.
+B: Computational or statistical method used by the authors (GMM = geometric morphometrics; Frequentist = hypothesis tests such as chi-square and ANOVA; AI/ML = artificial intelligence and machine learning, including neural networks and deep learning; MCMC = Markov Chain Monte Carlo,  i.e.
 Bayesian models and other simulations; Network = statistical analysis of social networks; 3D = analysis of 3D data such as artefact models; Composition = compositional analysis of artefacts).
@@
 For comparison with reproducibility studies in other fields reported above, the seven fully reproducible papers should be divided by the 25 reviewed for reproducibility, resulting in a 28% success rate.
-Expanding the denominator to include the total number of research articles published in JAS from May 2024 (when the first article to pass the reproducibility review, @herskindComputationalLinguisticMethodology2024, was published) to January 2025 (n = 97), which are articles that could have been eligible for reproducibility review had the authors used an open source programming language (e.g.
-instead of commercial software such as Microsoft Excel or SPSS, etc.) the success rate is 7%.
+Expanding the denominator to include the total number of research articles published in JAS from May 2024 (when the first article to pass the reproducibility review, @herskindComputationalLinguisticMethodology2024, was published) to January 2025 (n = 97) offers another perspective.
+These 97 articles that could have been eligible for reproducibility review had the authors used an open source programming language (e.g.
+instead of commercial software such as Microsoft Excel or SPSS, etc.).
+Under this broader scope, the success rate is 7%, a result also found in a study of 497 papers in 9 ecology journals [@kellnerFunctionalCodeRare2025].
 In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values available from a variety of hard and soft sciences.
@@
 The problem of incomplete materials can be tackled in several basic ways.
-First, authors should use a simple and logical folder structure to organise their code and data to be as self-contained as possible (ie.
-not requiring files outside of the project folder or any modifications by a reader to run the code).
-There are many excellent, simple, and widely-used templates that authors can choose from, e.g.
+First, authors should use a simple and logical folder structure to organise their code and data to be as self-contained as possible.
+Authors should provide their materials organised such that a reader can successfully run all code as-is, without making any manual modifications (e.g.
+use relative rather than absolute file paths so that readers don't have to rename or move files around to make the code work) [@sandveTenSimpleRules2013].
+Code and data files should be in the simplest format possible, for example a plain text R script file is smaller and easier to use than a PDF or Word document that includes R code.
+Script files should have the order in which they are to be run explicit in the file name, e.g.
+001-load-data.R, 002-clean-data.R, 003-analyse-data.R.
+There are many excellent, simple, and widely-used project templates that authors can choose from that make it easy for authors to follow best practices of project organisation, e.g.
 @doi:10.1080/00031305.2017.1375986, @figueiredoSimpleKitUse2022, @Cookiecutter, @cooperGuideReproducibleCode2017 and @wilsonGoodEnoughPractices2017
 
-Second, for analyses that are not highly time-consuming (which was over 90% of the sample), authors should re-run their code more than once, and ideally not on the same computer (e.g.
-by another co-author of the paper), before submission to confirm everything works as expected.
-This would ensure portability and likely also prevent issues relating to path and function errors.
-Complex and time-consuming analyses should use pipeline or workflow management tools, e.g.
-GNU Make, Luigi, Snakemake, or Targets, to document the relationship of the files and folders in a machine-readable format and simplify running and re-running code by others [@wrattenReproducibleScalableShareable2021; @landau2021targets].
-
-Third, authors should include in their compendium a README document that describes to readers the folders and files contained in the project.
+Second, authors should include in their compendium a README document that describes to readers the folders and files contained in the project [@abdillHowtoGuideCode2024].
 The README file is typically the first file that a reader will look at in a compendium so it should include brief instructions to guide the user to a successful reproduction of the original results (e.g.
@@
 A README should also briefly describe the contents of the compendium, where other necessary files can be obtained (e.g.
-data files not included in the compendium), the key software packages needed and the version numbers that the authors used, and if the analysis takes more than a few minutes to run on a typical laptop, the hardware resources and compute time used by the author.
-
-Fourth, authors should ensure clear, direct and obvious connections between their code and the results they present in their paper.
+data files that cannot be included in the compendium due for ethical or other reasons), the key software packages needed and the version numbers that the authors used, and if the analysis takes more than a few minutes to run on a typical laptop, the hardware resources and compute time used by the author.
+
+Related to the basic documentation provided by the README, authors should document clear, direct and obvious connections between their code and the results they present in their paper [@sandveTenSimpleRules2013].
 One simple way to do this is to have one code file for each figure and table, and name the code files with the figure or table number and some key words in the caption.
@@
 Quarto was a popular tool among the JAS papers in the reproducibility review sample, for example, @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024 wrote their entire manuscripts using Quarto.
-These four simple steps have high potential to substantially increase the computational reproducibility of archaeological science.
-
+
+Documentation is also a key tool is tackling the problems with dependencies described in the previous section.
 Our finding that dependencies are a common cause of irreproducible results is consistent with previous studies that have identified this as a widespread weakness in communicating computationally intensive research [@trisovicLargescaleStudyResearch2022; @samuelComputationalReproducibilityJupyter2024].
@@
 
+Finally, for analyses that are not highly time-consuming (which was over 90% of the sample), authors should re-run their code more than once, and ideally not on the same computer (e.g.
+by another co-author of the paper), before submission to confirm everything works as expected [@abdillHowtoGuideCode2024; @rothTenPrinciplesReliable2025].
+This ensures portability and likely also prevents issues relating to path and function errors.
+Complex and time-consuming analyses should use pipeline or workflow management tools, e.g.
+GNU Make, Luigi, Snakemake, or Targets, to document the relationship of the files and folders in a machine-readable format and simplify running and re-running code by others [@wrattenReproducibleScalableShareable2021; @landau2021targets].
+
+@fig-checklist summarises the key recommendations discussed in this section in a format that can be used as a check-list for authors submitting research for publication in JAS and other journals that do reproducibility reviews.
+This checklist is based on both the results presented in @fig-aer-summary and similar lists used by other journals, for example by the *Biometrical Journal* [@hornungBiometricalJournalChecklist] and *The Review of Financial Studies* [@perignonComputationalReproducibilityFinance2024].
+
+```{r}
+#| label: fig-checklist
+#| fig-cap: "Checklist summarising a small set of some of the simplest tasks authors can do that have high potential to increase reproducibility"
+#| fig-height: 4
+
+# I made this using Canva, original is here: https://www.canva.com/design/DAGnj7LkvvE/lrjyzV1WywPWMDdo9nwJJg/edit 
+knitr::include_graphics(here::here("analysis/figures/simple-reproducibility-checklist.png"))
+```
+
 ## What about qualitative archaeological research?
@@
 A similar debate has been unfolding more broadly about the humanities where Peels and Bouter [-@peelsHumanitiesNeedReplication2018; -@peelsPossibilityDesirabilityReplication2018] have argued that humanities disciplines that use empirical methods should be assessed by how well reanalysis of the original or new data using original or new methods produces the original or equivalent results.
-Resisting this proposal, @derijckeResistCallsReplicability2018 argue that because humanities research is unique because it pursues value and meaning, and a given study can produce multiple valid answers relating to the value and meaning of a study object, replication is irrelevant as a mark of quality.
+Resisting this proposal, @derijckeResistCallsReplicability2018 argue that humanities research is unique because it pursues value and meaning, and a given study can produce multiple valid answers relating to the value and meaning of a study object, so replication is irrelevant as a mark of quality.
 @peelsEpistemicValuesHumanities2018 disputes this uniqueness, claiming that the humanities has the same epistemic values as the sciences, however some values have more weight in the humanities while others have more weight in the sciences, and unlike in the sciences, humanities scholars often study these epistemic values themselves.
-A consensus seems to be emerging that
+A consensus seems to be emerging that for some but not all, studies in the humanities, replication is both possible and desirable, and that replication studies will differ from field to field and might even differ among various studies within a specific field [@bouterHumanitiesNeedReplication2019; @holbrook2019humanities].
+A key distinction here is that the locus of evaluation is empirical rather than computational, that is, concerned with appropriate reporting standards and documentation associated with physical evidence (e.g.
+artefacts and archives) [@stoddenReproducingStatisticalResults2015].
+A second important difference is that the debate about qualitative and humanities research is oriented towards *replication* (new data and/or new methods in an independent study to produce the same findings as original publication) rather than *reproducibility* (same data and same methods, e.g.
+computer code, to produce the same results as original publication) [cf @barbaTerminologiesReproducibleResearch2018].
+This orientation to replication emphasizes triangulation techniques that compare and integrate results coming from different traditions, locations, sources and methods, which in turn supports testing whether any given inference is robust in the face of different lines of evidence [@Leonelli2018].
+In sum, there are many types of qualitative and humanistic archaeology where it is possible and meaningful to maximize the chances of replication, e.g.
+by carefully documenting data generating processes, to produce higher quality and more impactful results.
 
@@
 It has also been used as a metaphor for the difficulty of getting insights from data tables in scholarly publications [@feinbergExtractingSunbeamsCucumbers2011].
-This metaphor is relevant in our current age of computationally intensive research, where extracting useful code and data from the publication of a typical computational study can be as frustrating and fruitless as extracting sunbeams from a cucumber.
+This metaphor has additional relevanance in our current age of computationally intensive research, where my experience as a reproducibility reviewer attempting to extract useful code and data from the publication of a computational study has sometimes felt as frustrating and fruitless as extracting sunbeams from a cucumber.
 
@@
 These observations are consistent with @Lakatos1978's model of a research program as a central foundation of irrefutable core assumptions complemented by a set of hypotheses, models, and methods that are adjusted, modified, or replaced by day-to-day research.
-Archaeology consists of multiple programs like this, with distinct and sometimes non-overlapping sets of core assumptions, such as processual and post-processual archaeology.
-Some programs are more amenable to reproducibility, while others offer insights through alternative methods.
+Archaeology consists of multiple programs like this, as indicated by the spread of journals across PC2 of @fig-pca, with distinct and sometimes non-overlapping sets of core assumptions.
+Some programs are more amenable to reproducibility, while others offer insights through qualitative and other methods.
 Among the programs that depend on quantitative methods to assess hypotheses and models, if they are to continue to progress through increased consensus through the accumulation of reliable facts and methods, it is essential for researchers to take computational reproducibility seriously.
@@
 However, the current state of quantitative archaeology, with most researchers not using open source code, is comparable to the secrecy of alchemy prior to the emergence of chemistry.
-Abandoning this habit of secrecy in favour of transparency and reproducibility is vital if we are to avoid a future where our journals are filled with pretty pictures that the reader has no hope of repeating or adapting in their own work.
-Computational reproducibility must be considered a minimum requirement for evaluating the integrity and usefulness of empirical results.
+Abandoning this habit of secrecy in favour of transparency and reproducibility is vital if we are to avoid a future where our journals are filled with pretty pictures depicting methods that the reader has no hope of repeating or adapting in their own work.
+Computational reproducibility must be considered a minimum requirement for evaluating the integrity and usefulness of quantitative results.
 
@@
 Results that are fully reproducible can contain errors and fraud.
-It is no guarantee of code quality, or that statistics have been used appropriately (cf.
-@Vaiglova_2025; Crema 2025), or that data management is consistent with FAIR and CARE principles [@carrollOperationalizingCAREFAIR2021].
+It is no guarantee of code quality, or that statistics have been used appropriately [cf.
+@Vaiglova_2025; Crema 2025], or that data management is consistent with FAIR and CARE principles [@carrollOperationalizingCAREFAIR2021].
 It is also time-consuming for authors to ensure their computational work can be reproduced, and for reviewers to evaluate.
-In a professional environment where job security and career progression is often associated with pressure to publish many high-impact papers, demands for authors to spend time on reproducibility, resulting in less time for publishing more papers, may seem frustrating.
+In a professional environment where job security and career progression is often associated with pressure to publish many high-impact papers, demands for authors to spend time on reproducibility, resulting in less time for publishing more papers, may seem frustrating [@edwards2017academic; @hagstrom1965scientific].
 This may seem especially unfair to early career researchers on short-term contracts, who may feel the goalposts for career success are being moved and that they are being asked to do more work their graduate training has not prepared them for.
 This highlights the need for a culture shift among senior archaeological scientists to value reproducibility in hiring and promotion decision-making.
-Professors must contribute to this shift by nurturing a culture of reanalysis and reproducibility in their teaching, for example by using replication assignments and by training students in the most current best practices and tools for reproducible research, such as R and Python [@marwickHowUseReplication2020].
+This is important for updating the alignment of quantitative archaeology with normative ideals of scientific practice, such as communal sharing and organized skepticism [@merton1973sociology].
+Professors must contribute to this shift by nurturing a culture of reanalysis and reproducibility in their teaching, for example by using replication assignments and by training students in the most current best practices and tools for reproducible research, such as R and Python [@marwickHowUseReplication2020; @dogucuReproducibilityClassroom2025].
+A key challenge for the future is changing the dominant habitus (e.g.
+dispositions, skills, and ways of perceiving) of senior scholars in gatekeeping positions so that reproducibility work will be recognized and rewarded with the same level of symbolic capital afforded to novel high-impact, highly cited publications [@bourdieu1988homo].
 

### 2025-05-17 analysis/paper/supplement-GAMS-details.qmd
update readme

@@
            size = 2) +
-    theme_bw(base_size = 6) +
+    theme_bw(base_size = 4) +
       labs( 
@@
 
-plot_grid(plotlist = conditional_smooths_plots_list, ncol = 3)
+plot_grid(plotlist = conditional_smooths_plots_list, 
+          ncol = 3,
+          align = "hv",
+          axis ="tblr")
+
+ggsave(here::here("analysis/figures/fig-smooth-plots-paper.png"),
+      dpi = 1000,
+      bg = "white",
+      h = 3,
+      w = 5)
 

### 2025-05-29 README.Rmd
resubmit to JAS

@@
 This project uses [`renv`](https://rstudio.github.io/renv/) to document and manage dependencies, this will be activated when the project is first opened in RStudio.
-7. Run ` rmarkdown::render("analysis/paper/paper.qmd", output_dir='/tmp')` to render the Quarto document that is the manuscript submitted for publication
-8. After the code has executed, view the rendered output document by running `browseURL("/tmp/paper.html")`
-8. When finished, to delete the Docker container from your computer, run this in the terminal `docker ps -aq | xargs docker stop | xargs docker rm`
+7. In RStudio, run `rmarkdown::render("analysis/paper/paper.qmd", output_dir='/tmp')` to render the Quarto document that is the manuscript submitted for publication
+8. After the code has executed, view the rendered output document by running in RStudio, `browseURL("/tmp/paper.html")`
+9. When finished, to delete the Docker container from your computer, run this in the terminal `docker ps -aq | xargs docker stop | xargs docker rm`
 

### 2025-05-29 analysis/paper/paper.qmd
resubmit to JAS

@@
 Analyzing nearly 10,000 articles, I identify trends in authorship, citation practices, and related metrics that position archaeology between the natural and social sciences.
-A review of reproducibility assessments for the Journal of Archaeological Science reveals persistent challenges, including missing data, unspecified dependencies, and inadequate documentation.
+A survey of reproducibility reviews for the Journal of Archaeological Science reveals persistent challenges, including missing data, unspecified dependencies, and inadequate documentation.
 To address these issues, I recommend to authors basic practical steps such as standardized project organization and explicit dependency documentation.
@@
 Further details about the model specifications and diagnostics are presented in the supplementary materials.
-#| fig-height: 4
+#| fig-height: 6
 
@@
 In this same category of DOI-issuing, research-grade repositories is OSF (the Open Science Foundation), Figshare, and university repositories.
-GitHub, a commercial service owned by Microsoft is a code hosting platform that is convenient for collaboration, is also popular among JAS authors, but is a problematic choice because does not offer DOIs or the same commitments to long-term availability as Zenodo.
+GitHub, a commercial service owned by Microsoft, is a code hosting platform that is convenient for collaboration, is also popular among JAS authors, but is a problematic choice because does not offer DOIs or the same commitments to long-term availability as Zenodo.
 Some authors attached their code and data as journal article supplementary files, but this is a poor choice for long-term availability because these files are typically renamed and converted to different formats during the article production process, making it difficult or impossible for a reader to combine the code and data to reproduce the results.
@@
 
-Documentation is also a key tool is tackling the problems with dependencies described in the previous section.
+Documentation is also a key tool in tackling the problems with dependencies described in the previous section.
 Our finding that dependencies are a common cause of irreproducible results is consistent with previous studies that have identified this as a widespread weakness in communicating computationally intensive research [@trisovicLargescaleStudyResearch2022; @samuelComputationalReproducibilityJupyter2024].
@@
 
-The simplest way for archaeologists, to improve here is to write the names and version numbers of the software and packages they used in their README file, as we see in @herskindComputationalLinguisticMethodology2024 and @monnaStudyingSerialityMaterial2024.
+The simplest way for archaeologists to improve here is to write the names and version numbers of the software and packages they used in their README file, as we see in @herskindComputationalLinguisticMethodology2024 and @monnaStudyingSerialityMaterial2024.
 For more complex research projects, i.e.
@@
 A more comprehensive solution, and the leading best practice for managing dependencies in many computationally intensive fields using Python in particular, is to include a Dockerfile in the compendium [@moreauContainersComputationalReproducibility2023].
-This is a set of a machine- and human-readable instructions that enables a user to recreate the author's computational environment (including those requirements beyond the R or Python packages) on another computer [@nustTenSimpleRules2020].
+This is a set of machine- and human-readable instructions that enables a user to recreate the author's computational environment (including those requirements beyond the R or Python packages) on another computer [@nustTenSimpleRules2020].
 Dockerfiles are gradually being adopted by archaeologists, see @cremaModellingDiffusionInnovation2024 and @liaoApplyingMaskRCNN2024 for examples, and most of our reproducibility reviews include a recommendation that the author include a Dockerfile to manage complex dependencies efficiently.
@@
 A consensus seems to be emerging that for some but not all, studies in the humanities, replication is both possible and desirable, and that replication studies will differ from field to field and might even differ among various studies within a specific field [@bouterHumanitiesNeedReplication2019; @holbrook2019humanities].
+
 A key distinction here is that the locus of evaluation is empirical rather than computational, that is, concerned with appropriate reporting standards and documentation associated with physical evidence (e.g.
@@
 This orientation to replication emphasizes triangulation techniques that compare and integrate results coming from different traditions, locations, sources and methods, which in turn supports testing whether any given inference is robust in the face of different lines of evidence [@Leonelli2018].
-In sum, there are many types of qualitative and humanistic archaeology where it is possible and meaningful to maximize the chances of replication, e.g.
+In sum, there are many types of qualitative and humanistic archaeology where it is possible and meaningful to maximize the chances of non-computational replication, e.g.
 by carefully documenting data generating processes, to produce higher quality and more impactful results.
@@
 It is no guarantee of code quality, or that statistics have been used appropriately [cf.
-@Vaiglova_2025; Crema 2025], or that data management is consistent with FAIR and CARE principles [@carrollOperationalizingCAREFAIR2021].
+@Vaiglova_2025; @crema_2025], or that data management is consistent with FAIR and CARE principles [@carrollOperationalizingCAREFAIR2021].
 It is also time-consuming for authors to ensure their computational work can be reproduced, and for reviewers to evaluate.
+This is especially the case for papers reporting results generated by long-running simulation or deep learning.
+These may be impracticable to fully reproduce by a typical peer reviewer who does not have access to specialised computing facilities.
 In a professional environment where job security and career progression is often associated with pressure to publish many high-impact papers, demands for authors to spend time on reproducibility, resulting in less time for publishing more papers, may seem frustrating [@edwards2017academic; @hagstrom1965scientific].

### 2025-05-30 analysis/paper/paper.qmd
update with edits from JAS editor

@@
 This approach is based on @fanelliBibliometricEvidenceHierarchy2013, who examined the hardness and softness of 12 disciplines using scholarly publication parameters.
-@fanelliBibliometricEvidenceHierarchy2013 found a spectrum of statistically significant variation in bibliometric variables from the physical to the social sciences, with papers tending at the softer end of the spectrum tending to have fewer co-authors, use less substantive titles, have longer texts, cite older literature, and have a higher diversity of sources.
-In @fanelliBibliometricEvidenceHierarchy2013's analysis harder sciences include Space Science, Physics, Chemistry, softer sciences include social sciences (Psychiatry, Psychology, Economics, Business, and General Social Sciences), and the Humanities define the soft end of the spectrum.
+@fanelliBibliometricEvidenceHierarchy2013 found a spectrum of statistically significant variation in bibliometric variables from the physical to the social sciences, with papers at the softer end of the spectrum tending to have fewer co-authors, use less substantive titles, have longer texts, cite older literature, and have a higher diversity of sources.
+In Fanelli and Glänzel's [-@fanelliBibliometricEvidenceHierarchy2013] analysis harder sciences include Space Science, Physics, Chemistry, softer sciences include social sciences (Psychiatry, Psychology, Economics, Business, and General Social Sciences), and the Humanities define the soft end of the spectrum.
 Following @fanelliBibliometricEvidenceHierarchy2013, I quantify the number of authors, length of article, relative title length, age of references, and diversity of references for a large sample of peer-reviewed journal articles.
@@
 In low-consensus, or softer, fields, papers must be longer to present justification, nuance and contextualization of results.
-While article length is constrained by journal requirements, leaving individual authors with little freedom to vary, journal requirements are typically set by editors who are professional archaeologists keen to tailor their journal to be attractive and relevant other members of the discipline.
-Thus journal requirements for article length will reflect the norms and of the discipline at any given time.
+While article length is constrained by journal requirements, leaving individual authors with little freedom to vary, journal requirements are typically set by editors who are professional archaeologists keen to tailor their journal to be attractive and relevant to other members of the discipline.
+Thus journal requirements for article length will reflect the norms of the discipline at any given time.
 The number of substantive and informative words in an article's title tends to be positively correlated with article length in harder disciplines [@yitzhakiRelationTitleLength2002; @yitzhakiVariationInformativityTitles1997], reflecting a focus on empiricism and efficiency that is characteristic of high-consensus disciplines.
 While @yitzhakiRelationTitleLength2002 removed stop-words (e.g.
-prepositions, articles, conjunctions, etc.) to calculate article length, in order to generate results for comparison with @fanelliBibliometricEvidenceHierarchy2013 I follow his method of dividing the total word count of the article title by the total number of pages of the article to compute relative title length.
+prepositions, articles, conjunctions, etc.) to calculate article length, in order to generate results for comparison with @fanelliBibliometricEvidenceHierarchy2013 I follow their method of dividing the total word count of the article title by the total number of pages of the article to compute relative title length.
 
@@
 
-The entire R code [@rcoreteamLanguageEnvironmentStatistical2024] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at https://doi.org/10.5281/zenodo.14897253 to enable re-use of materials and improve reproducibility and transparency [@marwickComputationalReproducibilityArchaeological2017].
+The entire R code [@rcoreteamLanguageEnvironmentStatistical2024] used for all the analysis and visualizations contained in this paper is at https://doi.org/10.5281/zenodo.14897253 to enable re-use of materials and improve reproducibility and transparency [@marwickComputationalReproducibilityArchaeological2017].
 All the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this compendium [@doi:10.1080/00031305.2017.1375986].
@@
 
-@fig-compare-other-fields shows the distribution of bibliometric variables for archaeology in context of data from other fields presented by @fanelliBibliometricEvidenceHierarchy2013.
+@fig-compare-other-fields shows the distribution of bibliometric variables for archaeology in the context of data from other fields presented by @fanelliBibliometricEvidenceHierarchy2013.
 The most striking indicator of archaeology as a hard science is the number of authors, where it is between the social sciences and physics.
@@
 The absence of these questions in research published in this journal makes it an outlier here, since these questions are central to a common definition of archaeology as 'cultural anthropology of the past', a phrase first found in @leroigourhan1946 and repeated in widely-used contemporary undergraduate textbooks such as @renfrew2024.
-Most archaeologists would likely be surprised at the decision by Clarivate to include the *Journal of Cultural Heritage* in their category of archaeology journals, leading to this result in @fig-variation-by-journal where the hardest archaeology journal publishes papers that are not very archaeological at all because they do not engage with anthropological topics.
+Most archaeologists would likely be surprised at the decision by Clarivate to include the *Journal of Cultural Heritage* in their category of archaeology journals, leading to this result in @fig-variation-by-journal where the hardest archaeology journal publishes papers that are not very archaeological because they do not engage with anthropological topics.
 
@@
 
-The PCA results in @fig-pca shows that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left.
+The PCA results in @fig-pca shows that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is a reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left.
 The variables that contribute to variation in PC1 are title length, number of pages, and the diversity of references.
@@
 In the context of a broad spectrum of other research areas, archaeologists behave like social scientists.
-We are harder than typical social scientists in tending to form larger groups of collaborators more often, and softer in sometimes writing longer articles than more resemble humanities scholarship.
+We are harder than typical social scientists in tending to form larger groups of collaborators more often, and softer in sometimes writing longer articles that more resemble humanities scholarship.
 The outlook for the future of archaeology is also complex, with three out of five of the bibliometric variables trending towards more humanistic styles of working, but the discipline showing more extreme values in some metrics towards both hard and soft sciences after about 2010.
@@
 Reproducibility is a minimum necessary condition for a finding to be believable and informative." Scientific reproducibility is a factor contributing to the hardness of a field.
-Specifically, reproducibility is linked to to concept of consensus in a field because if more researchers provide sufficient detail for others to reproduce their results, then consensus on new knowledge can more often, and more, rapidly be established.
+Specifically, reproducibility is linked to concept of consensus in a field because if more researchers provide sufficient detail for others to reproduce their results, then consensus on new knowledge can more often, and more rapidly, be established.
 The importance of this factor can be traced to Irish chemist Robert Boyle (1627-1691), best known for his experiments with vacuum pumps [@shapin2011leviathan].
@@
 Under this broader scope, the success rate is 7%, a result also found in a study of 497 papers in 9 ecology journals [@kellnerFunctionalCodeRare2025].
-In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values available from a variety of hard and soft sciences.
+By any measure, the computational reproducibility of archaeological research is generally on the low end of the distribution of values available from a variety of hard and soft sciences.
 
@@
 This ranges from missing data files down to missing lines of code.
-In most cases this can attributed to accidental carelessness, with the exception of two cases where data was unavailable due to licensing restrictions.
+In most cases this can be attributed to accidental carelessness, with the exception of two cases where data was unavailable due to licensing restrictions.
 Unspecified or under-specified dependencies is another common issue that prevents code from running.
@@
 dplyr for R or numpy for Python).
-If an author does not clearly specify the name and version number of the packages that they used for their analysis, it can be very time-consuming or impossible to correctly identify these because many packages have functions with similar names, and functions in any one package can change they way they behave as the developers update their package.
+If an author does not clearly specify the name and version number of the packages that they used for their analysis, it can be very time-consuming or impossible to correctly identify these because many packages have functions with similar names, and functions in any one package can change the way they behave as the developers update their package.
 Other reasons why papers failed the reproducibility review is that the paths to data files were incorrectly specified (likely a result of the author reorganising their compendium after completing their analysis, or omitting data files from the materials submitted for review), and errors returned by functions, which have diverse causes.
@@
 A README should also briefly describe the contents of the compendium, where other necessary files can be obtained (e.g.
-data files that cannot be included in the compendium due for ethical or other reasons), the key software packages needed and the version numbers that the authors used, and if the analysis takes more than a few minutes to run on a typical laptop, the hardware resources and compute time used by the author.
+data files that cannot be included in the compendium due to ethical or other reasons), the key software packages needed and the version numbers that the authors used, and if the analysis takes more than a few minutes to run on a typical laptop, the hardware resources and compute time used by the author.
 
@@
 those using five or more packages or machine learning algorithms, authors should use dependency management tools to keep track of the packages and version numbers needed to reproduce their results.
-This is an active area of development, and while there are many tools currently available, the most robust and widely used include renv for R [@renv, see examples in @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024] and conda and poetry for Python [@conda2023; @poetry2023].
+This is an active area of development, and while there are many tools currently available, the most robust and widely used include renv for R [@renv], see examples in @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024, and conda and poetry for Python [@conda2023; @poetry2023].
 
@@
 This is a set of machine- and human-readable instructions that enables a user to recreate the author's computational environment (including those requirements beyond the R or Python packages) on another computer [@nustTenSimpleRules2020].
-Dockerfiles are gradually being adopted by archaeologists, see @cremaModellingDiffusionInnovation2024 and @liaoApplyingMaskRCNN2024 for examples, and most of our reproducibility reviews include a recommendation that the author include a Dockerfile to manage complex dependencies efficiently.
+Dockerfiles are gradually being adopted by archaeologists, see @cremaModellingDiffusionInnovation2024 and @liaoApplyingMaskRCNN2024 for examples.
+Most of our reproducibility reviews include a recommendation that the authors include a Dockerfile to manage complex dependencies efficiently.
 
@@
 by another co-author of the paper), before submission to confirm everything works as expected [@abdillHowtoGuideCode2024; @rothTenPrinciplesReliable2025].
-This ensures portability and likely also prevents issues relating to path and function errors.
+This ensures the project is self-contained and portable and will help the authors detect and solve issues relating to path and function errors before they submit their work for review.
 Complex and time-consuming analyses should use pipeline or workflow management tools, e.g.
@@
 #| label: fig-checklist
-#| fig-cap: "Checklist summarising a small set of some of the simplest tasks authors can do that have high potential to increase reproducibility"
+#| fig-cap: "Checklist summarising a small set of some of the simplest tasks authors can do that have high potential to increase reproducibility.
+ML = Machine Learning, DL = Deep Learning, MCMC = Monte Carlo Markov Chain simulations"
 #| fig-height: 4

### 2025-06-05 analysis/paper/paper.qmd
remove mention of supplementary materials

@@
 
-The entire R code [@rcoreteamLanguageEnvironmentStatistical2024] used for all the analysis and visualizations contained in this paper is at https://doi.org/10.5281/zenodo.14897253 to enable re-use of materials and improve reproducibility and transparency [@marwickComputationalReproducibilityArchaeological2017].
+The entire R code [@rcoreteamLanguageEnvironmentStatistical2024] used for all the analysis and visualizations contained in this paper is at https://doi.org/10.5281/zenodo.14897252 to enable re-use of materials and improve reproducibility and transparency [@marwickComputationalReproducibilityArchaeological2017].
 All the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this compendium [@doi:10.1080/00031305.2017.1375986].
@@
 For the number of authors pages the Negative Binomial family was used with a standard log link function.
-Further details about the model specifications and diagnostics are presented in the supplementary materials.
+Further details about the model specifications and diagnostics are available at https://doi.org/10.5281/zenodo.14897252
 #| fig-height: 6
@@
 
-The data that support the findings of this study are openly available in Zenodo at http://doi.org/10.5281/zenodo.14897253
+The data that support the findings of this study are openly available in Zenodo at https://doi.org/10.5281/zenodo.14897252
 

### 2025-07-07 analysis/paper/paper.qmd
Added 1 chunck of code, changed Shannon's index, modified text for h-indices
1) I added the chunk of code found in the version 1.1 of the quarto document since it is the only place where there is the code to reproduce entirely the firue 3 of the published paper (figure plot over time). It does not recreate exactly the same plot but it is very close. While in the v1.3 it calls directly the .png but we can not recreate the figure.

2) I modified the calculation of the Shannon's index so that it is the same as in Fanelli & Glanzel 2013. It should indeed be the diversity of *sources* and not of *r

@@
 This sample was collected from Clarivate’s Web of Science database by first selecting the Web of Science category ‘Archaeology’ and the Document type ‘article’ (n = 28,871).
-To focus on journals of broad relevance to most archaeologists, and that are representative of substantial communities of practice, I then filtered the results to keep only articles published in the top-ranking 25 journals according to their h-indices as reported by Clarivate’s Journal Citation Indicator.
+To focus on journals of broad relevance to most archaeologists, and that are representative of substantial communities of practice, I then filtered the results to keep only articles published in the top-ranking 25 journals according to their 2022 Impact Factor as reported by Clarivate’s Journal Citation Indicator.
 Finally, I excluded journals with less than 100 articles in the database, resulting in `r n_journals` journals.
@@
   ref_list7 %>% 
-  group_by(id, x) %>% 
+  group_by(id, journal_name) %>% 
   tally() %>% 
-  mutate(n_in_article = n) %>% 
-  select(-n) %>% 
-  left_join(all_cited_items) %>% 
+  group_by(id) %>% 
   mutate(p_i = n / sum(n, na.rm = TRUE)) %>% 
@@
 
+library(ggplot2)
+library(tools)
+library(stringr)
+library(purrr)
+
 library(cowplot)
@@
   left_join(over_time_long_models)
+
+```
+
+```{r}
+#| label: fig-change-over-time_from_V1_1
+#| fig-cap: Distribution of article characteristics for archaeology articles over time.
+Data points represent individual articles.
+The colour of the points indicates if the overall trend is toward softer (orange) or harder (green) Generalized Additive Models were computed to fit the lines summarising the relationships between the variables and the time series.
+#| fig-height: 4
+
+library(ggpmisc)
+library(mgcv)
+formula <-  y ~ x
+
+over_time_long_colour_gams <- 
+over_time_long_colour %>% 
+  nest(.by = variable) %>% 
+  mutate(mod_gam = lapply(data, 
+                          function(df) gam(year ~ s(value, bs = "cr"), 
+                                           data = df)))
+
+over_time_long_colour_gams_summary <- 
+over_time_long_colour %>% 
+  nest(.by = variable) %>% 
+  mutate(fit = map(data, ~mgcv::gam(year ~ s(value, bs = "cs"), data = .)),
+         results = map(fit, glance),
+         R.square = map_dbl(fit, ~ summary(.)$r.sq)) %>%
+  unnest(results) %>%
+  select(-data, -fit) %>% 
+  select(variable, adj.r.squared)
+
+over_time_long_colour_gams_summary_df <- 
+over_time_long_colour %>% 
+  left_join(over_time_long_colour_gams_summary)
+
+plot = ggplot() + 
+  geom_point(data = over_time_long_colour_gams_summary_df, 
+       aes(year,  
+           value, 
+           colour = becoming_more_scientific),
+       alpha = 0.5) + 
+  geom_smooth(data = over_time_long_colour_gams_summary_df,
+              aes(year, value),
+              method="gam", 
+              formula = y ~ s(x, bs = "cs"),
+              se = FALSE,  
+              size = 2, 
+              colour = "#7570b3") +
+  facet_wrap( ~ variable,
+              scales = "free_y") + 
+  theme_bw(base_size = base_size) +
+  scale_color_manual(values = c("#d95f02",
+                                "#1b9e77" )) +
+  guides(colour = "none") +
+  ylab("") +
+  geom_text(data = over_time_long_colour_gams_summary_df %>% 
+              group_by(variable) %>% 
+              summarise(max_value = max(value),
+                        adj.r.squared = unique(adj.r.squared)),
+           aes(
+           x = 1980, 
+           y = max_value, 
+           label = paste("Pseudo R² = ", 
+                         signif(adj.r.squared, 
+                                digits = 3))),
+           hjust = 0, 
+           vjust = 1.5,
+           size = 2)
 

### 2025-07-07 analysis/paper/paper.qmd
stash

--- 2025-06-05 analysis/paper/paper.qmd
@@
 
-The PCA results in @fig-pca shows that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is a reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left.
+The PCA results in @fig-pca show that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is a reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left.
 The variables that contribute to variation in PC1 are title length, number of pages, and the diversity of references.
@@
 Unspecified or under-specified dependencies is another common issue that prevents code from running.
-This refers to the software packages in addition to R or Python that an author used to do specialized analyses and visualisations, (e.g.
+This refers to the software packages in addition to R or Python that an author used to do specialized analyses and visualisations (e.g.
 dplyr for R or numpy for Python).

### 2025-07-07 analysis/paper/paper.qmd

--- 2025-07-07 analysis/paper/paper.qmd
@@
+<<<<<<< HEAD
 ---
@@
 This sample was collected from Clarivate’s Web of Science database by first selecting the Web of Science category ‘Archaeology’ and the Document type ‘article’ (n = 28,871).
-To focus on journals of broad relevance to most archaeologists, and that are representative of substantial communities of practice, I then filtered the results to keep only articles published in the top-ranking 25 journals according to their 2022 Impact Factor as reported by Clarivate’s Journal Citation Indicator.
+To focus on journals of broad relevance to most archaeologists, and that are representative of substantial communities of practice, I then filtered the results to keep only articles published in the top-ranking 25 journals according to their h-indices as reported by Clarivate’s Journal Citation Indicator.
 Finally, I excluded journals with less than 100 articles in the database, resulting in `r n_journals` journals.
@@
   ref_list7 %>% 
-  group_by(id, journal_name) %>% 
+  group_by(id, x) %>% 
   tally() %>% 
-  group_by(id) %>% 
+  mutate(n_in_article = n) %>% 
+  select(-n) %>% 
+  left_join(all_cited_items) %>% 
   mutate(p_i = n / sum(n, na.rm = TRUE)) %>% 
@@
 
-library(ggplot2)
-library(tools)
-library(stringr)
-library(purrr)
-
 library(cowplot)
@@
   left_join(over_time_long_models)
-
-```
-
-```{r}
-#| label: fig-change-over-time_from_V1_1
-#| fig-cap: Distribution of article characteristics for archaeology articles over time.
-Data points represent individual articles.
-The colour of the points indicates if the overall trend is toward softer (orange) or harder (green) Generalized Additive Models were computed to fit the lines summarising the relationships between the variables and the time series.
-#| fig-height: 4
-
-library(ggpmisc)
-library(mgcv)
-formula <-  y ~ x
-
-over_time_long_colour_gams <- 
-over_time_long_colour %>% 
-  nest(.by = variable) %>% 
-  mutate(mod_gam = lapply(data, 
-                          function(df) gam(year ~ s(value, bs = "cr"), 
-                                           data = df)))
-
-over_time_long_colour_gams_summary <- 
-over_time_long_colour %>% 
-  nest(.by = variable) %>% 
-  mutate(fit = map(data, ~mgcv::gam(year ~ s(value, bs = "cs"), data = .)),
-         results = map(fit, glance),
-         R.square = map_dbl(fit, ~ summary(.)$r.sq)) %>%
-  unnest(results) %>%
-  select(-data, -fit) %>% 
-  select(variable, adj.r.squared)
-
-over_time_long_colour_gams_summary_df <- 
-over_time_long_colour %>% 
-  left_join(over_time_long_colour_gams_summary)
-
-plot = ggplot() + 
-  geom_point(data = over_time_long_colour_gams_summary_df, 
-       aes(year,  
-           value, 
-           colour = becoming_more_scientific),
-       alpha = 0.5) + 
-  geom_smooth(data = over_time_long_colour_gams_summary_df,
-              aes(year, value),
-              method="gam", 
-              formula = y ~ s(x, bs = "cs"),
-              se = FALSE,  
-              size = 2, 
-              colour = "#7570b3") +
-  facet_wrap( ~ variable,
-              scales = "free_y") + 
-  theme_bw(base_size = base_size) +
-  scale_color_manual(values = c("#d95f02",
-                                "#1b9e77" )) +
-  guides(colour = "none") +
-  ylab("") +
-  geom_text(data = over_time_long_colour_gams_summary_df %>% 
-              group_by(variable) %>% 
-              summarise(max_value = max(value),
-                        adj.r.squared = unique(adj.r.squared)),
-           aes(
-           x = 1980, 
-           y = max_value, 
-           label = paste("Pseudo R² = ", 
-                         signif(adj.r.squared, 
-                                digits = 3))),
-           hjust = 0, 
-           vjust = 1.5,
-           size = 2)
 
@@
 \newpage
+=======
+---
+title: "Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews"
+author:
+  - Ben Marwick:
+      correspondence: "yes"
+      email: bmarwick@uw.edu
+      orcid: 0000-0003-1689-0557
+      institute: uw
+institute:
+  - uw:
+      name: Department of Anthropology, University of Washington
+      address: 4218 Memorial Way Northeast, Seattle, WA 98105
+fig-dpi: 600
+title-block-published: "Last updated"  
+date: now
+date-format: long
+format: 
+  docx:
+    reference-doc: "templates/template.docx" # Insert path for the DOCX file
+execute:
+  echo: false
+  warning: false
+  message: false
+  comment: "#>"
+  fig-path: "../figures/"
+  fig-dpi: 600
+filters:
+  - templates/scholarly-metadata.lua
+  - templates/author-info-blocks.lua
+  - templates/pagebreak.lua
+bibliography: references.bib
+csl: "templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
+abstract: |
+  The status of archaeology as a science has been debated for decades and influences how we practice and teach archaeology.
+This study presents a novel bibliometric assessment of archaeology’s status relative to other fields using a hard/soft framework.
+It also presents a systematic review of computational reproducibility in published archaeological research.
+Reproducibility is a factor in the hardness/softness of a field because of its importance in establishing consensus.
+Analyzing nearly 10,000 articles, I identify trends in authorship, citation practices, and related metrics that position archaeology between the natural and social sciences.
+A survey of reproducibility reviews for the Journal of Archaeological Science reveals persistent challenges, including missing data, unspecified dependencies, and inadequate documentation.
+To address these issues, I recommend to authors basic practical steps such as standardized project organization and explicit dependency documentation.
+Strengthening reproducibility will enhance archaeology’s scientific rigor and ensure the verifiability of research findings.
+This study underscores the urgent need for cultural and technical shifts to establish reproducibility as a cornerstone of rigorous, accountable, and impactful archaeological science.
+keywords: |
+  Computational reproducibility; Archaeological science; Bibliometric analysis; Open science
+highlights: |
+  Archaeology occupies an intermediate position between the natural and social sciences.
+  Computational reproducibility in archaeology is low, with key barriers in incomplete compendiums, unspecified dependencies, and errors in code.
+  Practical steps like logical folder structures, README files, and dependency management tools can substantially improve computational reproducibility.
+  A cultural shift toward valuing reproducibility in hiring, promotion, and training is essential for advancing archaeological science.
+---
+
+## Introduction
+
+In their paper celebrating the 40th anniversary of this journal Torrence et al.
+[-@torrenceFortyYearsStill2015] noted that reproducibility was an issue important to the reputation and sustainability of the discipline, and necessary for archaeological science to behave like a science.
+As part of the celebration of the 50th anniversary, and of Torrence's leadership of the journal, my contribution revisits these topics of archaeology's status as a science, this journal's place in the landscape of archaeological science, and how the journal has responded to a growing recognition of the importance of reproducibility.
+I first present bibliometric evidence of the position of archaeology as a whole, and this journal in particular, in the sciences.
+Next, I report on the journal's progress in supporting reproducible research, and my work doing a new kind of peer review for JAS, one that evaluates the computational reproducibility of the research submitted for publication.
+Finally, I analyse twelve months of reproducibility reviews to identify common weaknesses in the ways archaeologists are working currently, and provide simple recommendations for researchers to overcome these and contribute to the improvement of computational reproducibility in archaeological science.
+
+The question of archaeology's status as a science usually comes up in the context of what the discipline should or should not be.
+One of the first landmarks in tackling this question is the debate published by *Antiquity* between classical archaeologist Jacquetta Hawkes and palaeoanthropologist Glynn Isaac.
+@hawkesProperStudyMankind1968, advocating a humanistic archaeology, was concerned that scientific approaches to archaeology were causing researchers to be "swamped by a vast accumulation of insignificant, disparate facts, like a terrible tide of mud, quite beyond the capacity of any man to contain and mould into historical form".
+More optimistic about the integration of science and archaeology, @isaacWhitherArchaeology1971 counters that "New levels of precision in presenting data and in interpreting them can surely lead to briefer and more interesting technical reports as well as providing the basis for more lively literary portrayals of what happened in prehistory.
+Expanding on Isaac's perspective, @binford1962archaeology argued that archaeology should operate as a science after the model proposed by philosopher Carl Hempel, which prescribed hypothesis-driven approaches, leading to generalizable laws of human behavior.
+Drawing on a different group of philosophers, @Smith_2017 argues for archaeology more specifically as a social science.
+@bevan2015data proposes that floods of digital data are reconfiguring our analytical agendas and support empirical and inductive inference.
+Counter-arguments to archaeology as a science come from numerous directions, notably @hodder1985postprocessual who rejected the quest for generalisations and instead argued that archaeology should be subjective and reflective, focussed on symbolic and relational meanings of material culture and the historical particularity of past human cultures.
+These debates, and the many more similar ones summarised by @martinon-torresArchaeologicalTheoriesArchaeological2013, have become a genre in archaeological writing that can be characterized as mostly based on personal observations, microscopic dissections of a handful of cherry-picked case studies of good or bad practice, and discussion of various philosophers and sociologists.
+
+What has been missing from these debates is a macroscopic observation of what the majority of archaeologists are actually doing, and an empirical comparison to a broad spectrum of relatively harder and softer disciplines.
+At the 'hard' end of the spectrum (e.g.
+physics and chemistry), scholars more typically share a large set of established set of theories, facts, and methods, facilitating fairly rapid agreement on the validity and significance of new results [@biglan1973characteristics].
+At the 'soft' end of the spectrum (e.g.
+economics and psychology), the set of theories, facts, and methods on which there is widespread consensus is smaller, and agreement is slower and less frequently reached about the significance of new findings and the continuing relevance of previous work.
+In sum, the hard-soft status is defined by the amount of consensus in a field, and the speed at which consensus is reached on new knowledge [@fanelliBibliometricEvidenceHierarchy2013].
+Hardness and softness is a controversial distinction, in part because it is sometimes used to imply a rank order of disciplines that encodes legitimacy, productivity, perceived value to society, and worthiness of funding [@DifferentAgenda2012; @coleHierarchySciences1983].
+Another criticism is that it may be more of an emergent product of social and institutional processes rather than intrinsic differences in method or consensus [@latour1987science].
+On the other hand, analyzing the characteristics that lead to the hard-soft distinction can be useful for understanding the diversity of academic inquiry, such as how different fields approach knowledge and differences in what counts as evidence and modes of argument, where fruitful collaborations might be possible due to shared methods and assumptions, and for curriculum design to structure courses appropriately based on a field's typical ways of knowing [@becher2001academic].
+
+Independent of these value judgments, empirical analysis of scholarly articles does support the hard-soft concept as a spectrum of variation in practice linked to differing degrees of consensus in a discipline, for example in approaches to data visualisation [@clevelandGraphsScientificPublications1984; @smithScientificGraphsHierarchy2000].
+Similarly, quantitative analysis of the frequency of positive results (ie.
+full or partial support for a research hypothesis) in publications is significantly correlated with hardness, consistent with a model where researchers in harder fields more readily accept any result their research produces, while those in softer fields have more freedom to choose which theories and hypotheses to test and how to interpret results [@fanelliPositiveResultsIncrease2010].
+The hard-soft spectrum is also evident in surveys of how researchers view their own work relative to those in other fields [@biglanCharacteristicsSubjectMatter1973].
+
+## How to measure the hardness or softness of a science?
+
+To objectively quantify the diversity of modern archaeological practice across a scale of relative hardness or softness, as an evaluation of its status as a science, and the place of this journal in context of other archaeology journals, I take a bibliometric approach.
+This approach is based on @fanelliBibliometricEvidenceHierarchy2013, who examined the hardness and softness of 12 disciplines using scholarly publication parameters.
+@fanelliBibliometricEvidenceHierarchy2013 found a spectrum of statistically significant variation in bibliometric variables from the physical to the social sciences, with papers at the softer end of the spectrum tending to have fewer co-authors, use less substantive titles, have longer texts, cite older literature, and have a higher diversity of sources.
+In Fanelli and Glänzel's [-@fanelliBibliometricEvidenceHierarchy2013] analysis harder sciences include Space Science, Physics, Chemistry, softer sciences include social sciences (Psychiatry, Psychology, Economics, Business, and General Social Sciences), and the Humanities define the soft end of the spectrum.
+Following @fanelliBibliometricEvidenceHierarchy2013, I quantify the number of authors, length of article, relative title length, age of references, and diversity of references for a large sample of peer-reviewed journal articles.
+
+These parameters are useful because of how they signify consensus in a research community.
+A larger number of authors on a paper reflects collaboration of people working together on a common goal.
+Collaborators have specialized roles, each of whom has the ability to study a part of the problem with high accuracy and detail, with harder fields having larger groups of collaborators [@zuckerman1972age].
+Reflecting this collaboration group size, harder disciplines tend to have higher average numbers of authors on papers.
+Article length has an inverse correlation with field hardness.
+In low-consensus, or softer, fields, papers must be longer to present justification, nuance and contextualization of results.
+While article length is constrained by journal requirements, leaving individual authors with little freedom to vary, journal requirements are typically set by editors who are professional archaeologists keen to tailor their journal to be attractive and relevant to other members of the discipline.
+Thus journal requirements for article length will reflect the norms of the discipline at any given time.
+The number of substantive and informative words in an article's title tends to be positively correlated with article length in harder disciplines [@yitzhakiRelationTitleLength2002; @yitzhakiVariationInformativityTitles1997], reflecting a focus on empiricism and efficiency that is characteristic of high-consensus disciplines.
+While @yitzhakiRelationTitleLength2002 removed stop-words (e.g.
+prepositions, articles, conjunctions, etc.) to calculate article length, in order to generate results for comparison with @fanelliBibliometricEvidenceHierarchy2013 I follow their method of dividing the total word count of the article title by the total number of pages of the article to compute relative title length.
+
+The age of works cited has long been used as a measure of a field's hardness [@moed1998new; @bornerAtlasScience2010], based on the assumption that harder fields assimilate new results more rapidly that softer fields [@price1970citation].
+I calculated a recency of references index for each article (also known as the Price index), which is the proportion of all cited works that were published in the five years preceding the paper.
+The diversity of references is a similar indicator, with papers in harder fields having a higher concentration of more specific citations because more knowledge is taken for granted as core knowledge that does not need citing [@skilton2006comparative].
+Conversely, softer fields have less knowledge taken for granted, a smaller core of facts that do not need citing, and thus a higher diversity of citations.
+
+```{r}
+library(tidyverse)
+
+# these data were prepared using the code in 000-import-raw-data.R
+items_df <- 
+  read_rds(here::here("analysis/data/wos-data-df.rds"))
+
+n_articles <- nrow(items_df)
+year_max <- max(items_df$year)
+year_min <- min(items_df$year)
+
+items_df_2012 <- 
+  items_df %>% 
+  filter(year == 2012)
+
+# how many archaeology articles in 2012
+n_items_df_2012 <- nrow(items_df_2012)
+
+# how many after 2012?
+n_items_df_after_2012 <- 
+  items_df %>% 
+  filter(year %in% 2013:year_max) %>% 
+  nrow()
+
+# what proportion of archaeology articles published after 2012
+prop_pub_after2012 <-  n_items_df_after_2012  / n_articles
+
+# how many distinct journals?
+n_journals <- n_distinct(items_df$journal)
+```
+
+While Fanelli and Glänzel (2013) analysed papers published in a single year (2012), I found only `r n_items_df_2012` papers for that same year, and `r round(prop_pub_after2012*100, 0)`% of papers in the sample published after that date.
+To make efficient use of the available data and ensure robust representation from different areas of archaeology, including those with lower frequencies of journal article publication, I analysed 9697 papers published during 1975-2025.
+This sample was collected from Clarivate’s Web of Science database by first selecting the Web of Science category ‘Archaeology’ and the Document type ‘article’ (n = 28,871).
+To focus on journals of broad relevance to most archaeologists, and that are representative of substantial communities of practice, I then filtered the results to keep only articles published in the top-ranking 25 journals according to their 2022 Impact Factor as reported by Clarivate’s Journal Citation Indicator.
+Finally, I excluded journals with less than 100 articles in the database, resulting in `r n_journals` journals.
+
+The entire R code [@rcoreteamLanguageEnvironmentStatistical2024] used for all the analysis and visualizations contained in this paper is at https://doi.org/10.5281/zenodo.14897252 to enable re-use of materials and improve reproducibility and transparency [@marwickComputationalReproducibilityArchaeological2017].
+All the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this compendium [@doi:10.1080/00031305.2017.1375986].
+The R code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.
+
+## Results
+
+### How does archaeology compare to other fields?
+
+```{r}
+library(ggrepel)
+
+source(here::here("analysis/code/001-redraw-Fanelli-and-Glanzel-Fig-2.R"))
+
+base_size <- 6
+color <- c('#d95f02', '#7570b3',  '#1b9e77')
+alpha <- 0.2
+linewidth <- 0.1
+
+# Number of authors ------------------
+boxlplot_n_authors <- 
+  ggplot() +
+  # boxplot of data from this study
+  geom_boxplot(data = items_df %>% 
+  filter(!is.na(year)),
+  aes(1, log(authors_n)),
+               size = 1)  +
+  # boxplot of data from Fanelli & Glänzel Fig 2
+  geom_boxplot(data = sim_data %>% 
+  filter(Variable == "N. of authors (ln)",
+         Category %in% c("h", "p", "s")),
+  aes(1, Value, 
+      group = Category),
+  size = 1,
+  fill = color,
+  colour = color, 
+  alpha = alpha,
+  linewidth = linewidth)  +
+  # annotations from Fanelli & Glänzel Fig 2
+  geom_text_repel(data = sim_data %>%  
+    filter(Variable == "N. of authors (ln)",
+           Category %in% c("h", "p", "s")) %>%  
+    group_by(Category) %>%  
+    summarise(y = median(Value)) %>%
+    mutate(
+      label = as.character(Category)
+    ),
+                  aes(c(0.75, 1, 1.25), y, label = label),
+                  color = color,
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0) +
+  scale_y_continuous(limits = c(0, 5)) +
+  scale_x_continuous(labels = NULL) +
+  theme_minimal(base_size = base_size) + 
+  theme(panel.grid  = element_blank()) +
+  ylab("N. of authors (ln)") +
+  xlab("Collaborator group size")
+
+# Relative title length  ----------------
+
+items_df_title <- 
+  items_df %>% 
+  filter(!is.na(pages_n)) %>%  
+  filter(!is.na(title_n)) %>% 
+  mutate(relative_title_length = log(title_n / pages_n))
+
+boxlplot_rel_title_length <- 
+items_df_title %>% 
+  filter(!is.na(year)) %>% 
+  ggplot(aes(1,
+             relative_title_length)) +
+  geom_boxplot(
+               size = 1)  +
+  # boxplot of data from Fanelli & Glänzel Fig 2
+  geom_boxplot(data = sim_data %>% 
+  filter(Variable == "Relative title length (ln)",
+         Category %in% c("h", "p", "s")),
+  aes(1, Value, 
+      group = Category),
+  size = 1,
+  fill = color,
+  colour = color, 
+  alpha = alpha,
+  linewidth = linewidth)  +
+  # annotations from Fanelli & Glänzel Fig 2
+  geom_text_repel(data = sim_data %>%  
+    filter(Variable == "Relative title length (ln)", 
+           Category %in% c("h", "p", "s")) %>%  
+    group_by(Category) %>%  
+    summarise(y = median(Value)) %>%
+    mutate(
+      label = as.character(Category)
+    ),
+                  aes(c(0.75, 1, 1.25), y, label = label),
+                  bg.colour = "white", 
+    colour = color,
+                  bg.r = .2, 
+                  force = 0) +
+  scale_y_continuous(limits = c(-4.5, 3),
+                     breaks =  seq(-5, 5, 1),
+                     labels = seq(-5, 5, 1)) +
+  scale_x_continuous(labels = NULL) +
+  theme_minimal(base_size = base_size) +
+  theme(panel.grid  = element_blank()) +
+  ylab("Ratio of title length to article length (ln)") +
+  xlab("Relative title length")
+
+# Number of pages ------------------
+boxlplot_n_pages <- 
+items_df %>% 
+  ggplot(aes(1,
+             log(pages_n))) +
+  geom_boxplot(
+               size = 1)  +
+  # boxplot of data from Fanelli & Glänzel Fig 2
+  geom_boxplot(data = sim_data %>% 
+  filter(Variable == "N. of pages (ln)",
+         Category %in% c("h", "p", "s")),
+  aes(1, Value, 
+      group = Category),
+  size = 1,
+  fill = color,
+  colour = color, 
+  alpha = alpha,
+  linewidth = linewidth)  +
+  # annotations from Fanelli & Glänzel Fig 2
+  geom_text_repel(data = sim_data %>%  
+    filter(Variable == "N. of pages (ln)", 
+           Category %in% c("h", "p", "s")) %>%  
+    group_by(Category) %>%  
+    summarise(y = median(Value)) %>%
+    mutate(
+      label = as.character(Category)
+    ),
+                  aes(c(0.75, 1, 1.25), y, label = label),
+                  bg.colour = "white",
+    colour = color, 
+                  bg.r = .2, 
+                  force = 0) +
+  scale_y_reverse(limits = c(5, 0)) +
+  scale_x_continuous(labels = NULL) +
+  theme_minimal(base_size = base_size) + 
+  theme(panel.grid  = element_blank()) +
+  ylab("N. of pages (ln)") +
+  xlab("Article length")
+
+# Price's index - age of references ------------------
+library(stringr)
+
+# output storage
+prices_index <- vector("list", length = nrow(items_df))
+
+# loop, this takes a moment
+for(i in seq_len(nrow(items_df))){
+
+  refs <-  items_df$refs[i]
+  year <-  items_df$year[i]
+
+  ref_years <- 
+    as.numeric(str_match(str_extract_all(refs, ", [0-9]{4}, ")[[1]], "\\d{4}"))
+
+  preceeding_five_years <-  
+    seq(year - 5, year, 1)
+
+  refs_n_in_preceeding_five_years <- 
+    ref_years[ref_years %in% preceeding_five_years]
+
+  prices_index[[i]] <- 
+    length(refs_n_in_preceeding_five_years) / length(ref_years)
+
+  # for debugging
+  # print(i)
+
+}
+
+prices_index <- flatten_dbl(prices_index)
+
+# add to data frame
+items_df$prices_index <-  prices_index
+
+# plot
+boxlplot_price_index <- 
+items_df %>% 
+  ggplot(aes(1,
+             prices_index)) +
+  geom_boxplot(
+               size = 1)  +
+  # boxplot of data from Fanelli & Glänzel Fig 2
+  geom_boxplot(data = sim_data %>% 
+  filter(Variable == "Price's index",
+         Category %in% c("h", "p", "s")),
+  aes(1, Value, 
+      group = Category),
+  size = 1,
+  fill = color,
+  colour = color,  
+  alpha = alpha,
+  linewidth = linewidth)  +
+  # annotations from Fanelli & Glänzel Fig 2
+  geom_text_repel(data = sim_data %>%  
+    filter(Variable == "Price's index", 
+           Category %in% c("h", "p", "s")) %>%  
+    group_by(Category) %>%  
+    summarise(y = median(Value)) %>%
+    mutate(
+      label = as.character(Category)
+    ),
+                  aes(c(0.75, 1, 1.25), y, label = label),
+                  bg.colour = "white", colour = color, 
+                  bg.r = .2, 
+                  force = 0) +
+  scale_y_continuous(limits = c(0, 1)) +
+  scale_x_continuous(labels = NULL) +
+  theme_minimal(base_size = base_size) +
+  theme(panel.grid  = element_blank()) +
+  ylab("Prop. refs in last 5 years") +
+  xlab("Recency of references")
+
+# Shannon index - diversity of references ------------------
+# journal name as species, article as habitat
+
+# simplify the refs, since they are a bit inconsistent, some of
+# these steps take a few seconds
+ref_list1 <- map(items_df$refs, ~tolower(.x))
+ref_list2 <- map(ref_list1, ~str_replace_all(.x, "\\.|,| ", ""))
+ref_list3 <- map(ref_list2, ~str_split(.x, "\n"))
+ref_list4 <- map(ref_list3, ~data_frame(x = .x))
+ref_list5 <- bind_rows(ref_list4, .id = "id")
+ref_list6 <- unnest(ref_list5)
+
+# get the journal names out of the refs
+ref_list7 <- 
+  ref_list6 %>% 
+  mutate(journal_name = gsub("\\-", "", x)) %>% 
+  mutate(journal_name = gsub("\\:", "", journal_name)) %>% 
+  mutate(journal_name = gsub("^[a-z'\\(\\)\\:]+[0-9]{4}", "", journal_name)) %>% 
+  mutate(journal_name = gsub("v[0-9]+.*", "", journal_name)) %>% 
+  mutate(journal_name = gsub("p[0-9]+$", "", journal_name))
+
+# prepare to compute shannon and join with other variables
+items_df$id <- 1:nrow(items_df)
+
+# tally of all referenced items
+all_cited_items <- 
+  ref_list7 %>% 
+  select(x) %>% 
+  group_by(x) %>% 
+  tally() %>% 
+  arrange(desc(n))
+
+# get a list of the top journals
+top_journals <- 
+  ref_list7 %>% 
+  select(journal_name) %>% 
+  group_by(journal_name) %>% 
+  tally() %>% 
+  filter(n > 50) %>% 
+  arrange(desc(n))
+
+# In the Shannon index, p_i is the proportion (n/N) of individuals of one particular species (journal) found (n) divided by the total number of individuals found (N), ln is the natural log, Σ is the sum of the calculations, and s is the number of species.
+
+# compute diversity of all citations for each article (habitat)
+shannon_per_item <- 
+  ref_list7 %>% 
+  group_by(id, journal_name) %>% 
+  tally() %>% 
+  group_by(id) %>% 
+  mutate(p_i = n / sum(n, na.rm = TRUE)) %>% 
+  mutate(p_i_ln = log(p_i)) %>% 
+  group_by(id) %>% 
+  summarise(shannon = -sum(p_i * p_i_ln, na.rm = TRUE)) %>% 
+  mutate(id = as.numeric(id)) %>% 
+  arrange(id)  %>% 
+  left_join(items_df)
+
+# plot
+boxlplot_shannon_index <- 
+shannon_per_item %>% 
+  filter(!is.na(year)) %>% 
+  ggplot(aes(1,
+             shannon)) +
+  geom_boxplot(
+               size = 1)  +
+   # boxplot of data from Fanelli & Glänzel Fig 2
+  geom_boxplot(data = sim_data %>% 
+  filter(Variable == "Shannon div. of sources",
+         Category %in% c("h", "p", "s")),
+  aes(1, Value, 
+      group = Category),
+  size = 1,
+  fill = color,
+  colour = color,  
+  alpha = alpha,
+  linewidth = linewidth)  +
+  # annotations from Fanelli & Glänzel Fig 2
+  geom_text_repel(data = sim_data %>%  
+    filter(Variable == "Shannon div. of sources", 
+           Category %in% c("h", "p", "s")) %>%  
+    group_by(Category) %>%  
+    summarise(y = median(Value)) %>%
+    mutate(
+      label = as.character(Category)
+    ),
+                  aes(c(0.75, 1, 1.25), y, label = label),
+    colour = color, 
+                  bg.colour = "white", 
+                  bg.r = .2, 
+                  force = 0) +
+  scale_y_reverse(limits = c(6, 0)) +
+  scale_x_continuous(labels = NULL) +
+  theme_minimal(base_size = base_size)  +
+  theme(panel.grid  = element_blank()) +
+  ylab("Shannon Index") +
+  xlab("Diversity of references")
+```
+
+```{r}
+#| label: fig-compare-other-fields
+#| fig-cap: Distributions of article characteristics hypothesised to reflect the level of consensus.
+The boxplot shows the distribution of values of archaeology articles.
+The thick black line in the middle of the boxplot is the median value, the box represents the inter-quartile range (the range between the 25th and 75th percentiles, where 50% of the data are located), and individual points represent outliers.
+The smaller coloured boxplots indicate the values computed by Fanelli and Glanzel (2013), where p = physics, s = social sciences, h = humanities.
+ln denotes the natural logarithm, or logarithm to the base e.
+#| fig-height: 4
+
+library(ggplot2)
+library(tools)
+library(stringr)
+library(purrr)
+
+library(cowplot)
+
+plot_grid(boxlplot_n_authors, 
+          boxlplot_rel_title_length,
+          boxlplot_n_pages,
+          boxlplot_price_index, 
+          boxlplot_shannon_index,
+          nrow = 2)
+```
+
+@fig-compare-other-fields shows the distribution of bibliometric variables for archaeology in the context of data from other fields presented by @fanelliBibliometricEvidenceHierarchy2013.
+The most striking indicator of archaeology as a hard science is the number of authors, where it is between the social sciences and physics.
+Archaeology is a close fit with the social sciences in relative title length.
+It is between the social sciences and humanities in recency of references and diversity of references.
+The clearest indicator of archaeology as a soft science is article length where it is similar to the humanities.
+Overall, archaeology does not sit squarely at either end of the hard-soft spectrum.
+It is generally not a harder science than the social sciences, with the exception of collaborator group sizes.
+
+### How has the hardness of archaeology varied over time?
+
+```{r}
+
+over_time <-  
+  items_df %>%  
+  left_join(items_df_title) %>%  
+  left_join(shannon_per_item) %>%  
+      filter(relative_title_length != -Inf, 
+            relative_title_length !=  Inf,
+            prices_index != "NaN" 
+           ) %>% 
+  mutate(log_authors_n = log(authors_n), 
+         log_pages_n = log(pages_n), 
+         journal_wrp = str_wrap(journal, 30)) %>%  
+  select(year, 
+         log_authors_n, 
+         log_pages_n, 
+         prices_index,  
+         shannon, 
+         relative_title_length)
+
+over_time_long <-  
+  over_time %>%  
+  ungroup() %>%  
+  select(-journal) %>%  
+  gather(variable,  
+         value, 
+         -year) %>%  
+  filter(value != -Inf, 
+         value !=  Inf) %>%  
+  mutate(variable = case_when( 
+    variable == "log_authors_n" ~ "N. of authors (ln)", 
+    variable == "log_pages_n"   ~ "N. of pages (ln)", 
+    variable == "prices_index"  ~ "Recency of references", 
+    variable == "shannon"  ~ "Diversity of references", 
+    variable == "relative_title_length"  ~ "Relative title length (ln)" 
+  )) %>%  
+  filter(!is.na(variable)) %>%  
+  filter(!is.nan(value)) %>%  
+  filter(!is.na(value)) %>%  
+  filter(value != "NaN") %>%  
+  mutate(value = parse_number(value))
+
+# compute beta estimates so we can colour lines to indicate more or
+# less hard
+
+library(broom)
+
+over_time_long_models <- 
+  over_time_long %>%  
+  group_nest(variable) %>%  
+  mutate(model = map(data, ~tidy(lm(value ~ year, data = .)))) %>%  
+  unnest(model) %>%  
+  filter(term == 'year') %>%  
+  mutate(becoming_more_scientific = case_when( 
+    variable == "N. of authors (ln)"         & estimate > 0 ~ "TRUE", 
+    variable == "N. of pages (ln)"           & estimate < 0 ~ "TRUE", 
+    variable == "N. of refs (sqrt)"          & estimate < 0 ~ "TRUE", 
+    variable == "Recency of references"      & estimate > 0 ~ "TRUE", 
+    variable == "Relative title length (ln)" & estimate > 0 ~ "TRUE", 
+    variable == "Diversity of references"    & estimate < 0 ~ "TRUE",
+    TRUE ~ "FALSE"
+  ))
+
+# join with data
+over_time_long_colour <- 
+  over_time_long %>% 
+  left_join(over_time_long_models)
+
+```
+
+```{r}
+#| label: fig-change-over-time_from_V1_1
+#| fig-cap: Distribution of article characteristics for archaeology articles over time.
+Data points represent individual articles.
+The colour of the points indicates if the overall trend is toward softer (orange) or harder (green) Generalized Additive Models were computed to fit the lines summarising the relationships between the variables and the time series.
+#| fig-height: 4
+
+library(ggpmisc)
+library(mgcv)
+formula <-  y ~ x
+
+over_time_long_colour_gams <- 
+over_time_long_colour %>% 
+  nest(.by = variable) %>% 
+  mutate(mod_gam = lapply(data, 
+                          function(df) gam(year ~ s(value, bs = "cr"), 
+                                           data = df)))
+
+over_time_long_colour_gams_summary <- 
+over_time_long_colour %>% 
+  nest(.by = variable) %>% 
+  mutate(fit = map(data, ~mgcv::gam(year ~ s(value, bs = "cs"), data = .)),
+         results = map(fit, glance),
+         R.square = map_dbl(fit, ~ summary(.)$r.sq)) %>%
+  unnest(results) %>%
+  select(-data, -fit) %>% 
+  select(variable, adj.r.squared)
+
+over_time_long_colour_gams_summary_df <- 
+over_time_long_colour %>% 
+  left_join(over_time_long_colour_gams_summary)
+
+plot = ggplot() + 
+  geom_point(data = over_time_long_colour_gams_summary_df, 
+       aes(year,  
+           value, 
+           colour = becoming_more_scientific),
+       alpha = 0.5) + 
+  geom_smooth(data = over_time_long_colour_gams_summary_df,
+              aes(year, value),
+              method="gam", 
+              formula = y ~ s(x, bs = "cs"),
+              se = FALSE,  
+              size = 2, 
+              colour = "#7570b3") +
+  facet_wrap( ~ variable,
+              scales = "free_y") + 
+  theme_bw(base_size = base_size) +
+  scale_color_manual(values = c("#d95f02",
+                                "#1b9e77" )) +
+  guides(colour = "none") +
+  ylab("") +
+  geom_text(data = over_time_long_colour_gams_summary_df %>% 
+              group_by(variable) %>% 
+              summarise(max_value = max(value),
+                        adj.r.squared = unique(adj.r.squared)),
+           aes(
+           x = 1980, 
+           y = max_value, 
+           label = paste("Pseudo R² = ", 
+                         signif(adj.r.squared, 
+                                digits = 3))),
+           hjust = 0, 
+           vjust = 1.5,
+           size = 2)
+
+```
+
+```{r}
+#| label: fig-change-over-time
+#| fig-cap: Distribution of article characteristics for archaeology articles over time.
+Data points represent individual articles.
+The colour of the points indicates if the overall trend is toward softer (orange) or harder (green).
+Bayesian Generalized Additive Models were computed to fit the lines summarising the relationships between the variables and the time series.
+For recency of references a Zero-One Inflated Beta distribution family was used with a logit link function.
+For diversity of references and relative title length (ln) the Gaussian family was used with the canonical identity link function.
+For the number of authors pages the Negative Binomial family was used with a standard log link function.
+Further details about the model specifications and diagnostics are available at https://doi.org/10.5281/zenodo.14897252
+#| fig-height: 6
+
+# get this from the supplement on GAMS
+knitr::include_graphics(here::here("analysis/figures/fig-smooth-plots-paper.png"))
+
+```
+
+@fig-change-over-time shows how the bibliometric indicators of field hardness have changed of time for archaeology articles.
+By two measures, the number of authors and relative title length, archaeology has become increasingly harder over time.
+On the other hand, three metrics indicate that archaeology has become softer (diversity of references, article length and recently of references).
+Although all the relationships are statistically significant, generally these temporal trends are very weak with low slope values, indicating very slow change over time.
+Similarly the r-squared values are very low, demonstrating that much of the variability in these metrics is independent of time.
+
+The most striking change over time is in the increase in the number of authors, which has the highest r-squared value of these metrics.
+One interesting detail evident in @fig-change-over-time is the increase in the range of diversity of references after about 2010.
+This may be due to some broader changes in academic publishing around this time, such as moves to digital-first continuous publishing, new journals appearing (e.g.
+*Archaeological and Anthropological Sciences* in 2009 and *Journal of Island & Coastal Archaeology* in 2010), and non-archaeology journals becoming more relevant to archaeologists.
+For example, *PLOS ONE* received its first impact factor in 2010 and in 2011 Nature's *Scientific Reports* began publishing [@malashichev2017open].
+The appearance of Google Scholar in 2004, increasing the discoverability of many works for many researchers, may have also contributed to this increase in diversity of references.
+
+### How do archaeology journals vary in hardness?
+
+```{r}
+
+journal_title_size <- 2
+
+# get rank order of journals by these bibliometic variables
+
+journal_metrics_for_plotting <- 
+items_df %>% 
+  left_join(items_df_title) %>% 
+  left_join(shannon_per_item) %>% 
+  ungroup() %>% 
+  select(journal,
+         authors_n,  # log
+         pages_n,    # log
+         relative_title_length,
+         prices_index,
+         shannon
+         ) %>% 
+    filter(relative_title_length != -Inf, 
+           relative_title_length !=  Inf,
+           prices_index != "NaN" 
+           ) %>% 
+  mutate(
+    log_authors = log(authors_n),
+    log_pages = log(pages_n)
+  )
+
+journal_metrics_for_plotting_summary <- 
+  journal_metrics_for_plotting %>% 
+  mutate(journal = str_wrap(journal, 20)) %>% 
+  group_by(journal) %>% 
+  summarise(mean_log_authors = mean(log_authors),
+            mean_log_pages =   mean(log_pages),
+            mean_relative_title_length = mean(relative_title_length),
+            mean_prices_index = mean(prices_index),
+            mean_shannon =      mean(shannon))
+
+# PCA of journal means
+journal_metrics_for_plotting_summary_pca <- 
+journal_metrics_for_plotting_summary %>% 
+  column_to_rownames("journal") %>% 
+  prcomp(scale = TRUE)
+
+# Tidy the PCA results
+pca_means_tidy <- journal_metrics_for_plotting_summary_pca %>% tidy(matrix = "pcs")
+# first two PCs explain how much?
+# Get the summary of the PCA
+pca_summary <- summary(journal_metrics_for_plotting_summary_pca)
+# Extract the proportion of variance explained by PC1 and PC2
+variance_explained <- round(pca_summary$importance[2, 1:2] * 100, 0)
+
+# Get the PCA scores
+pca_scores_means <- journal_metrics_for_plotting_summary_pca %>%  augment(journal_metrics_for_plotting_summary)
+
+# Get the PCA loadings
+pca_loadings_means <- 
+  journal_metrics_for_plotting_summary_pca %>% 
+  tidy(matrix = "rotation") %>% 
+    pivot_wider(names_from = "PC", 
+              values_from = "value",
+              names_prefix = "PC") %>% 
+  mutate(column = case_when(
+    column == "mean_log_authors" ~ "Number of\nauthors",
+    column == "mean_log_pages" ~ "Number of\npages",
+    column == "mean_relative_title_length" ~ "Relative\ntitle\nlength",
+    column == "mean_prices_index" ~ "Recency of\nreferences",
+    column == "mean_shannon" ~ "Diversity of\nreferences",
+  ))
+
+# Plot the PCA results
+plot_pca_means <- 
+ggplot() + 
+  labs(x = paste0("PC1 (", variance_explained[1], "%)"),
+       y = paste0("PC2 (", variance_explained[2], "%)")) +
+  geom_point(data = pca_scores_means,
+               aes(.fittedPC1,
+                   .fittedPC2),
+             size = 1) +
+  geom_text_repel(data = pca_scores_means %>% 
+                    mutate(label = str_replace(journal, 
+                                               "JOURNAL", 
+                                               "J.")) %>% 
+                    mutate(label = str_remove(label, 
+                                               "-AN\nINTERNATIONAL\nJ.")),
+            aes(.fittedPC1,
+                .fittedPC2,
+                label = label),
+            lineheight = 0.8,
+            segment.color = NA,
+            force_pull = 10,
+            size = 2.5,
+            bg.color = "white",  # Color of the halo
+            bg.r = 0.2) +
+     geom_segment(data = pca_loadings_means, 
+               aes(x = 0, 
+                   y = 0, 
+                   xend = PC1, 
+                   yend = PC2),
+               arrow = arrow(length = unit(0.2, "cm")), 
+               color = "grey70") +
+  geom_text_repel(data = pca_loadings_means,
+            aes(PC1,
+                PC2,
+                label = column),
+            size = 2,
+            lineheight = 0.8,
+            force = 10,
+            force_pull = 0,
+            segment.color = NA,
+            color = "grey40",
+           bg.color = "white",  # Color of the halo
+          bg.r = 0.2) +
+  theme_minimal(base_size = base_size) +
+  coord_fixed(xlim = c(-6, 2.5),
+              ylim = c(-3, 2))
+
+# tricky to get the label spacing right, let's save an SVG, edit
+# by hand, then export to PNG and read that file later. 
+ggsave(plot_pca_means,
+       filename = here::here("analysis/figures/plot_pca_means.svg"))
+
+```
+
+```{r}
+# looking into rankings of the journals
+
+journal_summary_metrics_ranks <- 
+journal_metrics_for_plotting_summary %>% 
+  mutate(across(starts_with("mean"), 
+                ~ rank(-.), 
+                .names = "rank_{.col}")) %>% 
+  select(journal, starts_with("rank")) %>% 
+  # reorder by hardness
+  mutate(rank_mean_log_pages = 21 - rank_mean_log_pages,
+         rank_mean_shannon =   21 - rank_mean_shannon)
+
+library(irr)
+journal_summary_metrics_ranks_test <- 
+journal_summary_metrics_ranks %>% 
+  select(-journal) %>% 
+  kendall(correct = TRUE)
+
+# Convert to scientific text
+pretty_print_sci <- function(num){
+  scientific_text <- paste0(gsub("e", " x 10^",  # Replace 'e' with ' x 10^'
+                          sprintf("%.2e", num)), "^") # round to 2 sf
+  return(scientific_text)
+  }
+
+borda_count_tbl <- function(votes_tbl) {
+  # Number of voters
+  num_voters <- ncol(votes_tbl) - 1
+
+  # Calculate scores for each option
+  scores <- votes_tbl %>%
+    rowwise() %>%
+    mutate(Score = sum(num_voters - c_across(starts_with("rank_")))) %>%
+    ungroup() %>%
+    select(1, Score)
+
+  # Return scores
+  return(scores)
+}
+
+# Calculate Borda Count scores
+borda_scores <- 
+journal_summary_metrics_ranks %>% 
+  borda_count_tbl() %>% 
+  rename("Journal" = "journal") %>% 
+  arrange(desc(Score))
+
+plot_borda_scores <- 
+  borda_scores %>% 
+  mutate(Journal = str_wrap(Journal, 20)) %>% 
+ggplot() +
+  aes(reorder(Journal, Score),
+      Score) +
+  geom_col() +
+  coord_flip() +
+  ylab("Borda Count scores") + 
+  xlab("") +
+  theme_minimal(base_size = base_size) +
+  theme(axis.text.y = element_text(size = journal_title_size))
+
+```
+
+```{r}
+library(ggridges)
+
+plot_journals_authors <- 
+journal_metrics_for_plotting %>% 
+  mutate(journal = str_wrap(journal, 20)) %>% 
+  ggplot(aes(y = reorder(journal, 
+                         log_authors,
+                     FUN = mean),
+             x = log_authors,
+             fill =   after_stat(x),
+             height = after_stat(density))) +
+  geom_density_ridges_gradient(stat = "density",
+                               colour = "white") +
+  scale_fill_viridis_c() +
+  guides(fill = 'none') +
+  theme_minimal(base_size = base_size) +
+  theme(axis.text.y = element_text(size = journal_title_size)) +
+  ylab("") + 
+  xlab("Number of authors (ln)")
+
+plot_journals_article_length <- 
+journal_metrics_for_plotting %>% 
+  mutate(journal = str_wrap(journal, 20)) %>% 
+  ggplot(aes(y = reorder(journal,
+                     -log_pages,
+                     FUN = mean), 
+             x = log_pages,
+             fill =   after_stat(x),
+             height = after_stat(density))) +
+  geom_density_ridges_gradient(stat = "density",
+                               colour = "white") +
+  scale_fill_viridis_c() +
+  guides(fill = 'none') +
+  theme_minimal(base_size = base_size) +
+  theme(axis.text.y = element_text(size = journal_title_size)) +
+  xlab("Number of pages (ln)") +
+  ylab("")
+
+plot_journals_title_length <- 
+journal_metrics_for_plotting %>% 
+  mutate(journal = str_wrap(journal, 20)) %>% 
+  ggplot(aes(y = reorder(journal,
+                         relative_title_length,
+                         FUN = mean), 
+             x = relative_title_length,
+             fill =   after_stat(x),
+             height = after_stat(density))) +
+  geom_density_ridges_gradient(stat = "density",
+                               colour = "white") +
+  scale_fill_viridis_c() +
+  guides(fill = 'none') +
+  theme_minimal(base_size = base_size) +
+  theme(axis.text.y = element_text(size = journal_title_size)) +
+  ylab("") +
+  xlab("Relative title length (ln)")
+
+plot_journals_ref_recency <- 
+journal_metrics_for_plotting %>% 
+  mutate(journal = str_wrap(journal, 20)) %>% 
+  group_by(journal) %>% 
+  ggplot(aes(y = reorder(journal,
+                     prices_index,
+                     FUN = mean), 
+             x = prices_index,            
+             fill =   after_stat(x),
+             height = after_stat(density))) +
+  geom_density_ridges_gradient(stat = "density",
+                               colour = "white") +
+  scale_fill_viridis_c() +
+  guides(fill = 'none') +
+  theme_minimal(base_size = base_size) +
+  theme(axis.text.y = element_text(size = journal_title_size)) +
+  ylab("") +
+  xlab("Recency of references")
+
+plot_journals_ref_diversity <- 
+journal_metrics_for_plotting %>%  
+  mutate(journal = str_wrap(journal, 20)) %>% 
+  group_by(journal) %>% 
+  ggplot(aes(y = reorder(journal,
+                     -shannon,
+                     FUN = mean), 
+             x = shannon,
+             fill =   after_stat(x),
+             height = after_stat(density))) +
+  geom_density_ridges_gradient(stat = "density",
+                               colour = "white") +
+  scale_fill_viridis_c() +
+  guides(fill = 'none') +
+  theme_minimal(base_size = base_size) +
+  theme(axis.text.y = element_text(size = journal_title_size)) +
+  ylab("") +
+  xlab("Diversity of references")
+```
+
+```{r}
+#| label: fig-variation-by-journal
+#| fig-cap: "Panels A-E: Variation in bibliometric indicators of hardness for 20 archaeological journals.
+The journals are ordered for each indicator so that within each plot, the harder journals are at the top of the plot and the softer journals are at the base.
+Panel F shows a bar plot that is the single consensus ranking computed from all five variables, using the Borda Count ranking algorithm."
+#| fig-height: 5
+
+library(cowplot)
+
+  plot_grid(plot_journals_authors, 
+           plot_journals_article_length,
+           plot_journals_title_length,
+           plot_journals_ref_recency, 
+           plot_journals_ref_diversity,
+           plot_borda_scores,
+           nrow = 2,
+           labels = LETTERS[1:6],
+           label_size = 6)
+
+```
+
+```{r}
+#| label: fig-pca
+#| fig-cap: "Biplot of the first and second principal components of a PCA computed on the means of the five bibliometric variables for each journal in the sample.
+The arrows represent the correlation between each original variable and the principal components.
+The direction and length of the arrows indicate how strongly each variable contributes to each component.
+"
+#| fig-height: 4
+
+knitr::include_graphics(here::here("analysis/figures/plot_pca_means.png"))
+```
+
+@fig-variation-by-journal shows the distribution of our bibliometric variables of hardness for each of the `r n_journals` journals in the sample.
+Overall agreement between these bibliometric variables in ranking these journals on a hard-soft spectrum is moderate to strong, with a Kendall's coefficient of concordance (Wt) value of `r round(journal_summary_metrics_ranks_test$value, 2)` (in a 0-1 range, where 1 is perfect agreement) and a p-value of `r pretty_print_sci(journal_summary_metrics_ranks_test$p.value)`.
+Panel F of @fig-variation-by-journal shows an overall consensus ranking of all journals in the sample.
+In this consensus ranking the *Journal of Archaeological Science* among the top five archaeology journals for hardness.
+It is placed at the harder end of the hard-soft spectrum especially by the number of pages and relative title length, and to lesser degrees by the number of authors and recency of references.
+However, according to the diversity of references, the *Journal of Archaeological Science* is at the middle of the spectrum.
+
+The *Journal of Cultural Heritage* is the only journal that consistently ranks as hard across all variables, occurring in the top five journals for all five metrics.
+This journal primarily publishes materials science and computational analyses related to conservation and preservation of historic objects in museums and other collections.
+Authors of papers in recent issues have affiliations with museums, cultural heritage programs, and chemistry, engineering, and physics departments at European and Chinese universities.
+Notably, papers in this journal typically do not engage in questions or debates about past human behaviour or culture.
+The absence of these questions in research published in this journal makes it an outlier here, since these questions are central to a common definition of archaeology as 'cultural anthropology of the past', a phrase first found in @leroigourhan1946 and repeated in widely-used contemporary undergraduate textbooks such as @renfrew2024.
+Most archaeologists would likely be surprised at the decision by Clarivate to include the *Journal of Cultural Heritage* in their category of archaeology journals, leading to this result in @fig-variation-by-journal where the hardest archaeology journal publishes papers that are not very archaeological because they do not engage with anthropological topics.
+
+The *Journal of Archaeological Research* is notable for consistently ranking as soft; it was the softest journal for four of our five bibliometric variables.
+This is a predicable result for a review journal, which is a distinct type of journal dedicated to summarizing, analyzing, and synthesizing existing research in a particular field.
+The stated aim of the *Journal of Archaeological Research* is to 'bring together the most recent international research summaries on a broad range of topics and geographical areas' [@feinmanAimsScopeJournal2024].
+A typical article is a long single-authored synthesis of archaeology in a region or on a topic.
+As the only review journal in this sample, this is a stark contrast to the other journals here that present original research findings, and like the *Journal of Cultural Heritage*, may be considered an outlier in this sample.
+
+The PCA results in @fig-pca shows that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is a reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left.
+The variables that contribute to variation in PC1 are title length, number of pages, and the diversity of references.
+Journals with higher PC1 values have articles with longer titles, fewer pages, and less diverse reference lists.
+The distribution of PC1 values is skewed left, with most of the journals concentrated at the harder end of the spectrum.
+Variation in PC2 is influenced by the number of authors and recency of references.
+The distribution of PC2 values reveals additional structure to the data and can be roughly separated into generalist journals in the negative range of the PC2 axis (e.g.
+*American Antiquity*, *Antiquity*, *Advances in Archaeological Practice*), characterised by fewer authors and more recent references.
+In the positive range of the PC2 axis are more specialised journals, characterised by higher numbers of authors and less recent references cited (e.g.
+*Environmental Archaeology*, *Geoarchaeology*, *Archaeological Research in Asia*, *Journal of Island and Coastal Archaeology*).
+The *Journal of Archaeological Science* sits about midway between these two groups, reflecting its relevance to both specialised and generalist communities of practice in archaeology.
+
+## Reproducibility: A key measure of how scientific a field is
+
+This macroscopic perspective derived from an analysis of the ways thousands of archaeologists communicate their research has produced a complex picture of archaeology as a science.
+In the context of a broad spectrum of other research areas, archaeologists behave like social scientists.
+We are harder than typical social scientists in tending to form larger groups of collaborators more often, and softer in sometimes writing longer articles that more resemble humanities scholarship.
+The outlook for the future of archaeology is also complex, with three out of five of the bibliometric variables trending towards more humanistic styles of working, but the discipline showing more extreme values in some metrics towards both hard and soft sciences after about 2010.
+Among archaeology journals, we see distinct communities of practice reflected in the PCA results that are very close together on the hard-soft spectrum, but have minor differences in their communication styles, perhaps due to cultural differences in writing traditions inherited from parent disciplines such as geology and biology [@becher2001academic].
+
+While these bibliometric variables provide several interesting insights into the status of archaeology as a science, via measurement of consensus, and are important for moving the debate beyond discussions of a small number of case studies, they miss a crucial factor that separates scientific practice from non-science.
+This is reproducibility, which, according to a report for the US National Science Foundation [@cacioppo2015social], "refers to the ability of a researcher to duplicate the results of a prior study using the same materials as were used by the original investigator.
+That is, a second researcher might use the same raw data to build the same analysis files and implement the same statistical analysis in an attempt to yield the same results...
+Reproducibility is a minimum necessary condition for a finding to be believable and informative." Scientific reproducibility is a factor contributing to the hardness of a field.
+Specifically, reproducibility is linked to concept of consensus in a field because if more researchers provide sufficient detail for others to reproduce their results, then consensus on new knowledge can more often, and more rapidly, be established.
+The importance of this factor can be traced to Irish chemist Robert Boyle (1627-1691), best known for his experiments with vacuum pumps [@shapin2011leviathan].
+Boyle was concerned about the secrecy common among experimentalists in the 17th century and aimed to shift the culture from valuing direct in-person witnessing of scientific demonstrations towards meticulous written communications that were detailed enough to enable a reader to successfully undertake the experiment themselves, independent of the original author.
+
+With many disciplines making increasing use of computationally intensive analyses in recent years there has been renewed interest in reproducibility [@levequeReproducibleResearchScientific2012].
+In part, this is because computationally intensive research is difficult to communicate within the constraints of the methods section of a traditional journal article — the reader also needs the computer code written by the original authors, not just the article text.
+There is also the broader context of rising pressure to publish in prestigious journals and intense competition for funds that create strong incentives for malpractice in research [@edwards2017academic].
+These two factors have led to widespread concerns of a reproducibility crisis in many fields [@Baker2016].
+Estimates of scientific reproducibility in several fields confirm the extent of this problem.
+Empirical replications of 100 studies published in three psychology journals found that 36% of replications had statistically significant results, compared to 96% of the original studies [@OpenScienceCollaboration2015].
+Similar empirical replications of large numbers of social science studies and experimental economics studies successfully replicated 61% and 62% of their target studies respectively [@camererEvaluatingReplicabilitySocial2018; @camererEvaluatingReplicabilityLaboratory2016].
+
+Similarly bleak results come from measurements specifically of the reproducibility of computational analyses of scientific studies.
+An attempt at reproducing the computational results of 204 papers in *Science* resulted in success in reproducing the findings for 26% [@stoddenEmpiricalAnalysisJournal2018].
+The computational results of two out of 41 geoscience papers could be fully reproduced on the first attempt [@konkolComputationalReproducibilityGeoscientific2019].
+In the biomedical field, code in 1,203 out of 27,271 (4%) notebooks associated with 3,467 publications could be run without errors [@samuelComputationalReproducibilityJupyter2024].
+Statisticians could reproduce 15% of 93 papers [@xiongStatePlayReproducibility2023].
+Economists have been especially active in researching computational reproducibility, with studies indicating successful reproduction of results using code and data provided by authors for 30% of 67 papers [@ChangLi2015], 14% of 203 papers [@gertler2018make], 44% of 152 papers [@herbert2021reproducibility], 30% of 419 articles [@fivsar2024reproducibility], and 28% of 168 papers [@perignonComputationalReproducibilityFinance2024].
+These efforts confirm that the reproducibility of published research is widely recognised as a cornerstone of rigorous science, and work on evaluating how successful a research community is at generating reproducible results has become a distinctive and important meta-research activity in many fields.
+
+How does archaeology compare to these other fields in terms of reproducibility?
+Empirical reproducibility has long been valued in field archaeology.
+Throughout the history of archaeology, well-known sites have been repeatedly revisited to test old hypotheses with new evidence or methods, for example, Olduvai Gorge (Tanzania), Cahokia (USA), Çatalhöyük (Turkey), and Madjedbebe (Australia).
+Similarly among many experimental archaeologists, empirical reproducibility is a key concern, for example in lithic use-wear identification [@hayesLearningBlindTests2017] and the measurement of lithics [@pargeterReplicabilityLithicAnalysis2023].
+The increasing availability of large digital datasets is pushing archaeology into unexplored areas [@bevan2015data], inviting questions about what reproducibility means for data intensive archaeological research.
+For example, to what extent does concern for reproducibility extend to computational reproducibility among archaeologists?
+
+## How reproducible is archaeology? Investigating computational reproducibility
+
+In 2024 the *Journal of Archaeological Science* introduced a new kind of peer review that has provided an opportunity to tackle this question about computational reproducibility in archaeology.
+In January 2024 I accepted the position of 'Associate Editor for Reproducibility' (AER) for JAS and conducted reproducibility reviews of submissions that mentioned programming languages such as R or Python in the methods sections, taking guidance from similar initiatives in other fields [e.g.
+@ivimey-cookImplementingCodeReview2023; @nustCODECHECKOpenScience2021].
+A reproducibility review examines the code and data used to generate the results presented in the paper, and attempts to run the authors' code to reproduce their results (see @editorsReproducibilityJournalArchaeological for more details about this process).
+This new AER role is based on similar positions (i.e.
+'data editor' or 'reproducibility editor') that journals in economics [@vilhuber2019report], statistics [@wrobel2024partnering], astronomy [@muench2023roles], ecology [@bolnick2022ensuring], and environmental studies [@rosenbergReproducibleResultsPolicy2021] have had, in some cases for over a decade.
+In 2024, three archaeology journals, in addition to JAS, added AERs to their editorial communities: *Advances in Archaeological Practice* [@marwick2024introducing, one paper reviewed], *Journal of Field Archaeology* [@farahani2024reproducibility, two papers reviewed], and *American Antiquity* [@martin2024editors, no papers reviewed at the time of writing].
+
+```{r}
+# reporting on my AER work
+# initially collected here: 
+# https://docs.google.com/spreadsheets/d/1o4jSZ__OoCDssf0lWWeBheNEbgiheU_1Udj5X82VOvY/edit?gid=0#gid=0
+jas_rr_data <- 
+  read_csv(here::here("analysis/data/JAS AER data analysis.csv"))
+
+base_size <- 6
+
+# what software did the authors use?
+aer_plot_software <- 
+jas_rr_data %>% 
+  separate_longer_delim(Language, ",") %>% 
+  mutate(Language = str_squish(Language)) %>% 
+  count(Language) %>% 
+  drop_na() %>% 
+  mutate(Language = fct_reorder(Language, n, .desc = FALSE)) %>% 
+  ggplot() +
+  aes(Language, n) +
+  geom_bar(stat = 'identity') +
+  coord_flip() +
+  xlab("") +
+  ylab("Number of papers reviewed") +
+  theme_minimal(base_size = base_size)
+
+# what methods did the authors use? ML includes CNN, VAR, DL
+aer_plot_methods <- 
+jas_rr_data %>% 
+  separate_longer_delim(Methods2, ",") %>% 
+  mutate(Methods2 = str_squish(Methods2)) %>% 
+  count(Methods2) %>% 
+  drop_na() %>% 
+  mutate(Methods = fct_reorder(Methods2, n, .desc = FALSE)) %>% 
+ggplot() +
+  aes(Methods, n) +
+  geom_bar(stat = 'identity') +
+  coord_flip() +
+  xlab("") +
+  ylab("Number of papers reviewed") +
+  theme_minimal(base_size = base_size)
+
+# where did they put their materials?
+aer_plot_repo <- 
+jas_rr_data %>% 
+  separate_longer_delim(Repo, ",") %>% 
+  mutate(Repo = str_squish(Repo)) %>% 
+  count(Repo) %>% 
+  drop_na() %>% 
+  mutate(Repo = fct_reorder(Repo, n, .desc = FALSE)) %>% 
+  ggplot() +
+  aes(Repo, n) +
+  geom_bar(stat = 'identity') +
+  coord_flip() +
+  xlab("") +
+  ylab("Number of papers reviewed") +
+  theme_minimal(base_size = base_size) +
+  scale_y_continuous(breaks = scales::breaks_pretty())
+
+# could reproduce the results on the first try?
+aer_first_try <- 
+jas_rr_data %>% 
+  count(`Run on first try?`) %>% 
+  drop_na() %>% 
+  mutate(`Run on first try?` = fct_reorder(`Run on first try?`, n, .desc = FALSE))
+
+# why not? issues with code
+aer_plot_issues <- 
+jas_rr_data %>% 
+  separate_longer_delim(`Code issue`, ",") %>% 
+  mutate(`Code issue` = str_squish(`Code issue`)) %>% 
+    mutate(`Code issue` = ifelse(`Code issue` == "Proprietary software", 
+                               "Proprietary\nsoftware",
+                               `Code issue`)) %>% 
+  count(`Code issue`) %>% 
+  drop_na() %>% 
+  mutate(`Code issue` = fct_reorder(`Code issue`, n, .desc = FALSE)) %>% 
+  ggplot() +
+  aes(`Code issue`, n) +
+  geom_bar(stat = 'identity') +
+  coord_flip() +
+  xlab("") +
+  ylab("Number of papers reviewed") +
+  theme_minimal(base_size = base_size)  +
+  scale_y_continuous(breaks = scales::breaks_pretty())
+
+# relationship between code issues and language?
+aer_plot_issues_lang <- 
+jas_rr_data %>% 
+  separate_longer_delim(`Code issue`, ",") %>% 
+  separate_longer_delim(Language, ",") %>% 
+  mutate(Language = str_squish(Language)) %>% 
+  mutate(`Code issue` = str_squish(`Code issue`)) %>% 
+  group_by(`Code issue`, 
+           Language) %>% 
+  tally() %>% 
+  drop_na() %>% 
+  ggplot() +
+  aes(`Code issue`, 
+      Language,
+      size = n) +
+  geom_point() +
+  xlab("Problem with the code") +
+  ylab("Software") +
+  scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
+  theme_minimal(base_size = base_size) +
+  scale_size_area(breaks = c(1, 2, 3, 5)) +
+  theme(legend.position = c(.9,.7),
+        legend.title = element_text(hjust = 0.5),
+        legend.spacing = unit(0.1, "cm"),  # Reduce space between legend items
+        legend.margin = margin(1, 5, 1, 1),  # Reduce margin around the legend
+        legend.key.size = unit(0.2, "cm"),  # Reduce size of legend keys
+        legend.box.background=element_rect(fill="white", 
+                                          color="black")) 
+```
+
+```{r}
+#| label: fig-aer-summary
+#| fig-cap: "Summary of reproducibility reviews for JAS.
+A: Primary software used for the computational analysis reported in a manuscript.
+B: Computational or statistical method used by the authors (GMM = geometric morphometrics; Frequentist = hypothesis tests such as chi-square and ANOVA; AI/ML = artificial intelligence and machine learning, including neural networks and deep learning; MCMC = Markov Chain Monte Carlo,  i.e.
+Bayesian models and other simulations; Network = statistical analysis of social networks; 3D = analysis of 3D data such as artefact models; Composition = compositional analysis of artefacts).
+C: Locations where authors deposited their code and data files.
+D: Issues that prevented the reproducible review from succeeding on the first attempt.
+E: Relationship between software used and issues that make research irreproducible."
+#| fig-height: 3
+
+library(patchwork)
+(aer_plot_software + aer_plot_methods + aer_plot_repo) / (aer_plot_issues + aer_plot_issues_lang) + plot_annotation(tag_levels = 'A')
+```
+
+At the time of writing (January 2025) we have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS (most papers required multiple reviews).
+Of these, 11 have been published in JAS to date.
+Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate, by one measure, of 63%.
+Four of the seven papers could be fully reproduced on my first attempt, the others required additional input from the authors.
+For comparison with reproducibility studies in other fields reported above, the seven fully reproducible papers should be divided by the 25 reviewed for reproducibility, resulting in a 28% success rate.
+Expanding the denominator to include the total number of research articles published in JAS from May 2024 (when the first article to pass the reproducibility review, @herskindComputationalLinguisticMethodology2024, was published) to January 2025 (n = 97) offers another perspective.
+These 97 articles that could have been eligible for reproducibility review had the authors used an open source programming language (e.g.
+instead of commercial software such as Microsoft Excel or SPSS, etc.).
+Under this broader scope, the success rate is 7%, a result also found in a study of 497 papers in 9 ecology journals [@kellnerFunctionalCodeRare2025].
+By any measure, the computational reproducibility of archaeological research is generally on the low end of the distribution of values available from a variety of hard and soft sciences.
+
+@fig-aer-summary shows a summary of basic characteristics of the 25 articles that have been through the reproducibility review process so far.
+The most commonly used software is R, followed by Python.
+Results generated with proprietary or closed-source software are out of scope for reproducibility reviews.
+Several distinct types of analyses are well-represented in this sample, especially geometric morphometry, network statistics, and analyses using artificial intelligence or machine learning algorithms (this includes deep learning and neural networks).
+Most authors are sharing their code and data files via Zenodo, a non-profit generic research data repository hosted by CERN that accepts any file format and freely assigns all publicly available uploads a DOI to make the files easily and uniquely citable [@petersZenodoSpotlightTraditional2017].
+In this same category of DOI-issuing, research-grade repositories is OSF (the Open Science Foundation), Figshare, and university repositories.
+GitHub, a commercial service owned by Microsoft, is a code hosting platform that is convenient for collaboration, is also popular among JAS authors, but is a problematic choice because does not offer DOIs or the same commitments to long-term availability as Zenodo.
+Some authors attached their code and data as journal article supplementary files, but this is a poor choice for long-term availability because these files are typically renamed and converted to different formats during the article production process, making it difficult or impossible for a reader to combine the code and data to reproduce the results.
+
+Panels D and E of @fig-aer-summary summarise the common issues that resulted in irreproducible results.
+The most common issue was an incomplete compendium.
+This ranges from missing data files down to missing lines of code.
+In most cases this can be attributed to accidental carelessness, with the exception of two cases where data was unavailable due to licensing restrictions.
+Unspecified or under-specified dependencies is another common issue that prevents code from running.
+This refers to the software packages in addition to R or Python that an author used to do specialized analyses and visualisations, (e.g.
+dplyr for R or numpy for Python).
+If an author does not clearly specify the name and version number of the packages that they used for their analysis, it can be very time-consuming or impossible to correctly identify these because many packages have functions with similar names, and functions in any one package can change the way they behave as the developers update their package.
+Other reasons why papers failed the reproducibility review is that the paths to data files were incorrectly specified (likely a result of the author reorganising their compendium after completing their analysis, or omitting data files from the materials submitted for review), and errors returned by functions, which have diverse causes.
+
+## How to improve the computational reproducibility of archaeology?
+
+Despite the relatively small number of reproducibility reviews reported on here, there are patterns of common issues that point to a small set of simple tasks authors can do that have high potential to increase reproducibility.
+The problem of incomplete materials can be tackled in several basic ways.
+First, authors should use a simple and logical folder structure to organise their code and data to be as self-contained as possible.
+Authors should provide their materials organised such that a reader can successfully run all code as-is, without making any manual modifications (e.g.
+use relative rather than absolute file paths so that readers don't have to rename or move files around to make the code work) [@sandveTenSimpleRules2013].
+Code and data files should be in the simplest format possible, for example a plain text R script file is smaller and easier to use than a PDF or Word document that includes R code.
+Script files should have the order in which they are to be run explicit in the file name, e.g.
+001-load-data.R, 002-clean-data.R, 003-analyse-data.R.
+There are many excellent, simple, and widely-used project templates that authors can choose from that make it easy for authors to follow best practices of project organisation, e.g.
+@doi:10.1080/00031305.2017.1375986, @figueiredoSimpleKitUse2022, @Cookiecutter, @cooperGuideReproducibleCode2017 and @wilsonGoodEnoughPractices2017
+
+Second, authors should include in their compendium a README document that describes to readers the folders and files contained in the project [@abdillHowtoGuideCode2024].
+The README file is typically the first file that a reader will look at in a compendium so it should include brief instructions to guide the user to a successful reproduction of the original results (e.g.
+what order to run the code files in).
+A README should also briefly describe the contents of the compendium, where other necessary files can be obtained (e.g.
+data files that cannot be included in the compendium due to ethical or other reasons), the key software packages needed and the version numbers that the authors used, and if the analysis takes more than a few minutes to run on a typical laptop, the hardware resources and compute time used by the author.
+
+Related to the basic documentation provided by the README, authors should document clear, direct and obvious connections between their code and the results they present in their paper [@sandveTenSimpleRules2013].
+One simple way to do this is to have one code file for each figure and table, and name the code files with the figure or table number and some key words in the caption.
+Another way some authors are accomplishing this is by using literate programming tools, such as Quarto and Jupyter notebooks [@quarto; @kluyver2016jupyter] that enable the research narrative and code for data analysis to be woven together in one document.
+Quarto was a popular tool among the JAS papers in the reproducibility review sample, for example, @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024 wrote their entire manuscripts using Quarto.
+
+Documentation is also a key tool in tackling the problems with dependencies described in the previous section.
+Our finding that dependencies are a common cause of irreproducible results is consistent with previous studies that have identified this as a widespread weakness in communicating computationally intensive research [@trisovicLargescaleStudyResearch2022; @samuelComputationalReproducibilityJupyter2024].
+In our sample, issues relating to dependencies are strongly associated with the use of Python.
+One possible reason for this is that relative to R, Python uses more package managers, more environments, and deeper dependency chains with more complex inter-dependencies that change more rapidly [@decan2016topology; @decanEmpiricalComparisonDependency2019; @korkmazModelingImpactPython2020].
+Another reason may be that there is a bigger and more established community of R users in archaeology [@schmidtToolDrivenRevolutionsArchaeological2020; @batistOpenArchaeologyOpen2024] that highly values code that is easy for others to reuse and has evolved practices to effectively communicate dependencies [e.g.
+@bilottiPointPatternAnalysis2024; @willExploringUtilityUnretouched2025].
+
+The simplest way for archaeologists to improve here is to write the names and version numbers of the software and packages they used in their README file, as we see in @herskindComputationalLinguisticMethodology2024 and @monnaStudyingSerialityMaterial2024.
+For more complex research projects, i.e.
+those using five or more packages or machine learning algorithms, authors should use dependency management tools to keep track of the packages and version numbers needed to reproduce their results.
+This is an active area of development, and while there are many tools currently available, the most robust and widely used include renv for R [@renv], see examples in @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024, and conda and poetry for Python [@conda2023; @poetry2023].
+
+A more comprehensive solution, and the leading best practice for managing dependencies in many computationally intensive fields using Python in particular, is to include a Dockerfile in the compendium [@moreauContainersComputationalReproducibility2023].
+This is a set of machine- and human-readable instructions that enables a user to recreate the author's computational environment (including those requirements beyond the R or Python packages) on another computer [@nustTenSimpleRules2020].
+Dockerfiles are gradually being adopted by archaeologists, see @cremaModellingDiffusionInnovation2024 and @liaoApplyingMaskRCNN2024 for examples.
+Most of our reproducibility reviews include a recommendation that the authors include a Dockerfile to manage complex dependencies efficiently.
+
+Finally, for analyses that are not highly time-consuming (which was over 90% of the sample), authors should re-run their code more than once, and ideally not on the same computer (e.g.
+by another co-author of the paper), before submission to confirm everything works as expected [@abdillHowtoGuideCode2024; @rothTenPrinciplesReliable2025].
+This ensures the project is self-contained and portable and will help the authors detect and solve issues relating to path and function errors before they submit their work for review.
+Complex and time-consuming analyses should use pipeline or workflow management tools, e.g.
+GNU Make, Luigi, Snakemake, or Targets, to document the relationship of the files and folders in a machine-readable format and simplify running and re-running code by others [@wrattenReproducibleScalableShareable2021; @landau2021targets].
+
+@fig-checklist summarises the key recommendations discussed in this section in a format that can be used as a check-list for authors submitting research for publication in JAS and other journals that do reproducibility reviews.
+This checklist is based on both the results presented in @fig-aer-summary and similar lists used by other journals, for example by the *Biometrical Journal* [@hornungBiometricalJournalChecklist] and *The Review of Financial Studies* [@perignonComputationalReproducibilityFinance2024].
+
+```{r}
+#| label: fig-checklist
+#| fig-cap: "Checklist summarising a small set of some of the simplest tasks authors can do that have high potential to increase reproducibility.
+ML = Machine Learning, DL = Deep Learning, MCMC = Monte Carlo Markov Chain simulations"
+#| fig-height: 4
+
+# I made this using Canva, original is here: https://www.canva.com/design/DAGnj7LkvvE/lrjyzV1WywPWMDdo9nwJJg/edit 
+knitr::include_graphics(here::here("analysis/figures/simple-reproducibility-checklist.png"))
+```
+
+## What about qualitative archaeological research?
+
+Although the introduction of reproducibility reviews signifies a growth in computational archaeology and a desire to evaluate the research products beyond the journal article, a very substantial amount of archaeological research is qualitative, with few or no numerical data involved in making knowledge claims about the human past.
+For example, many archaeological questions can be answered by the simple presence or absence of artefacts or features, or qualitative comparisons of basic artefact characteristics such as shape, colour, surface treatments, and raw material.
+Chaîne opératoire analyses by ceramic and lithic specialists is an especially productive area of archaeological research that often relies on comparison of narratives of manufacturing processes.
+While these studies unquestionably count as science, because they are a systematic, empirical, and rigorous process of inquiry, should they be held to standards of reproducibility in the same way that computationally intensive research is?
+
+A similar debate has been unfolding more broadly about the humanities where Peels and Bouter [-@peelsHumanitiesNeedReplication2018; -@peelsPossibilityDesirabilityReplication2018] have argued that humanities disciplines that use empirical methods should be assessed by how well reanalysis of the original or new data using original or new methods produces the original or equivalent results.
+Resisting this proposal, @derijckeResistCallsReplicability2018 argue that humanities research is unique because it pursues value and meaning, and a given study can produce multiple valid answers relating to the value and meaning of a study object, so replication is irrelevant as a mark of quality.
+@peelsEpistemicValuesHumanities2018 disputes this uniqueness, claiming that the humanities has the same epistemic values as the sciences, however some values have more weight in the humanities while others have more weight in the sciences, and unlike in the sciences, humanities scholars often study these epistemic values themselves.
+A consensus seems to be emerging that for some but not all, studies in the humanities, replication is both possible and desirable, and that replication studies will differ from field to field and might even differ among various studies within a specific field [@bouterHumanitiesNeedReplication2019; @holbrook2019humanities].
+
+A key distinction here is that the locus of evaluation is empirical rather than computational, that is, concerned with appropriate reporting standards and documentation associated with physical evidence (e.g.
+artefacts and archives) [@stoddenReproducingStatisticalResults2015].
+A second important difference is that the debate about qualitative and humanities research is oriented towards *replication* (new data and/or new methods in an independent study to produce the same findings as original publication) rather than *reproducibility* (same data and same methods, e.g.
+computer code, to produce the same results as original publication) [cf @barbaTerminologiesReproducibleResearch2018].
+This orientation to replication emphasizes triangulation techniques that compare and integrate results coming from different traditions, locations, sources and methods, which in turn supports testing whether any given inference is robust in the face of different lines of evidence [@Leonelli2018].
+In sum, there are many types of qualitative and humanistic archaeology where it is possible and meaningful to maximize the chances of non-computational replication, e.g.
+by carefully documenting data generating processes, to produce higher quality and more impactful results.
+
+## Conclusion
+
+In the classic satirical novel *Gulliver's Travels* (1726) by Irish writer Jonathon Swift Gulliver visits the fictional Grand Academy of Lagado in Balnibarbi, a caricature of the Royal Society of London, and meets several researchers working on wildly impractical projects, including one attempting to extracting sunbeams out of cucumbers.
+This is usually interpreted as a subversive anti-colonial parody depicting institutionalized research as an absurd fund-raising activity with no practical benefits to society [@alff2014swift; @nicolson1937scientific].
+It has also been used as a metaphor for the difficulty of getting insights from data tables in scholarly publications [@feinbergExtractingSunbeamsCucumbers2011].
+This metaphor has additional relevanance in our current age of computationally intensive research, where my experience as a reproducibility reviewer attempting to extract useful code and data from the publication of a computational study has sometimes felt as frustrating and fruitless as extracting sunbeams from a cucumber.
+
+I have presented a bibliometric analysis on the status of archaeology as a science, showing distinct disunity that is increasing over time.
+On average we generally behave as social scientists, with some elements in common with harder sciences.
+These observations are consistent with @Lakatos1978's model of a research program as a central foundation of irrefutable core assumptions complemented by a set of hypotheses, models, and methods that are adjusted, modified, or replaced by day-to-day research.
+Archaeology consists of multiple programs like this, as indicated by the spread of journals across PC2 of @fig-pca, with distinct and sometimes non-overlapping sets of core assumptions.
+Some programs are more amenable to reproducibility, while others offer insights through qualitative and other methods.
+Among the programs that depend on quantitative methods to assess hypotheses and models, if they are to continue to progress through increased consensus through the accumulation of reliable facts and methods, it is essential for researchers to take computational reproducibility seriously.
+Computers have become a central field and laboratory instrument for much of our work, so we have an ethical duty to document how we change our data as it flows through silicon just as carefully as we document the operating parameters of a mass spectrometer or any other field or laboratory instrument.
+However, the current state of quantitative archaeology, with most researchers not using open source code, is comparable to the secrecy of alchemy prior to the emergence of chemistry.
+Abandoning this habit of secrecy in favour of transparency and reproducibility is vital if we are to avoid a future where our journals are filled with pretty pictures depicting methods that the reader has no hope of repeating or adapting in their own work.
+Computational reproducibility must be considered a minimum requirement for evaluating the integrity and usefulness of quantitative results.
+
+Computational reproducibility is not a panacea; it should not be used as a universally accepted criterion for research quality [@Leonelli2018].
+Results that are fully reproducible can contain errors and fraud.
+It is no guarantee of code quality, or that statistics have been used appropriately [cf.
+@Vaiglova_2025; @crema_2025], or that data management is consistent with FAIR and CARE principles [@carrollOperationalizingCAREFAIR2021].
+It is also time-consuming for authors to ensure their computational work can be reproduced, and for reviewers to evaluate.
+This is especially the case for papers reporting results generated by long-running simulation or deep learning.
+These may be impracticable to fully reproduce by a typical peer reviewer who does not have access to specialised computing facilities.
+In a professional environment where job security and career progression is often associated with pressure to publish many high-impact papers, demands for authors to spend time on reproducibility, resulting in less time for publishing more papers, may seem frustrating [@edwards2017academic; @hagstrom1965scientific].
+This may seem especially unfair to early career researchers on short-term contracts, who may feel the goalposts for career success are being moved and that they are being asked to do more work their graduate training has not prepared them for.
+This highlights the need for a culture shift among senior archaeological scientists to value reproducibility in hiring and promotion decision-making.
+This is important for updating the alignment of quantitative archaeology with normative ideals of scientific practice, such as communal sharing and organized skepticism [@merton1973sociology].
+Professors must contribute to this shift by nurturing a culture of reanalysis and reproducibility in their teaching, for example by using replication assignments and by training students in the most current best practices and tools for reproducible research, such as R and Python [@marwickHowUseReplication2020; @dogucuReproducibilityClassroom2025].
+A key challenge for the future is changing the dominant habitus (e.g.
+dispositions, skills, and ways of perceiving) of senior scholars in gatekeeping positions so that reproducibility work will be recognized and rewarded with the same level of symbolic capital afforded to novel high-impact, highly cited publications [@bourdieu1988homo].
+
+# Acknowledgements
+
+Versions of this paper were presented at the Summer School on Reproducible Research In Landscape Archaeology at the Freie Universität Berlin and Christian-Albrechts-Universität zu Kiel (2017), the Big Data in Archaeology Conference at the McDonald Institute for Archaeological Research at the University of Cambridge (2019) and the Workshop on Exploring Data-Driven Solutions to Archaeological Problems at the Abu Dhabi Institute at New York University (2025).
+Thanks to the participants of those events for their feedback.
+Thanks to the JAS editors for the invitation to contribute to this special issue.
+
+# Data Availability Statement
+
+The data that support the findings of this study are openly available in Zenodo at https://doi.org/10.5281/zenodo.14897252
+
+# References
+
+<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
+
+::: {#refs}
+:::
+
+\newpage
+>>>>>>> e6713de47690905ff4d77466e22765f2b1ce537e
